---
title: "QC correlation of classifier score with RNA-Seq expression"
author: "K S Gaonkar (D3B)"
output: html_notebook

---
    
In this notebook we will find correlation between RNA-Seq and classifier score
We will check if the values are inversely correlated as we would expect?

### Setup
```{r}
library("ggpubr")
library("ggthemes")
library("tidyverse")
library("broom")

# rootdir
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
results_dir <- "results"
```

### Classifier scores

 1) Read in scores from tp53-nf1-classifier for stranded
```{r}

score_stranded_df <- read_tsv(file.path(results_dir, "pbta-gene-expression-rsem-fpkm-collapsed.stranded_classifier_scores.tsv")) %>%
  dplyr::select(
    "tp53_score",
    "sample_id"
  ) %>%
  unique() %>%
  as.data.frame()

```

 2) Read in scores from tp53-nf1-classifier for polya
```{r}
score_polya_df <- read_tsv(file.path(results_dir, "pbta-gene-expression-rsem-fpkm-collapsed.polya_classifier_scores.tsv")) %>%
  dplyr::select(
    "tp53_score",
    "sample_id"
  ) %>%
  unique() %>%
  as.data.frame()

```

### Expression matrices

```{r}

stranded_exp_file <- file.path(data_dir,"pbta-gene-expression-rsem-fpkm-collapsed.stranded.rds")
polya_exp_file <- file.path(data_dir,"pbta-gene-expression-rsem-fpkm-collapsed.polya.rds")

```

Format expression matrices to long format

```{r}

expression_mat_to_long <- function(expressionFile,score_df,gene){
# load expressionMatrix RDS for expression based filtering for less than given threshold
expressionMatrix<-readRDS(expressionFile) %>% 
  rownames_to_column()

# Taking snippet from @jaclyn-taroni idea used in expression based fusion call filtering

expression_long_df <- expressionMatrix %>%
  # rename rowname to GeneSymbol
  dplyr::rename(GeneSymbol=rowname) %>%
  # Get the data into long format
  reshape2::melt(variable.name = "Kids_First_Biospecimen_ID",
                 value.name = "expression_value") %>%
  dplyr::mutate(Kids_First_Biospecimen_ID= as.character(Kids_First_Biospecimen_ID)) %>%
  dplyr::filter(GeneSymbol %in% gene) 

# merge score and expression file
merge_score_exp <- expression_long_df %>%
  left_join(score_df, by=c("Kids_First_Biospecimen_ID"="sample_id"))

return(merge_score_exp)
}

stranded_exp_score <- expression_mat_to_long(stranded_exp_file,score_stranded_df,"TP53")
polya_exp_score <- expression_mat_to_long(polya_exp_file,score_polya_df,"TP53")

```


### Correlation between RNA-expression and classifier sccore
```{r}
tidy(cor.test(stranded_exp_score$expression_value,stranded_exp_score$tp53_score))
tidy(cor.test(polya_exp_score$expression_value,polya_exp_score$tp53_score))

```


stranded TP53 expression to classifier_score has a negligible 
negative correlation estimate -0.05 at p value 0.067

polya TP53 expression to classifier_score has a negligible 
negative correlation estimate -0.139 at p value 0.297

