---
title: "Find CNV losses that overlap with TP53 domains"
author: "K S Gaonkar (D3B)"
output: html_notebook
params:
  base_run:
    label: "1/0 to run with base histology"
    value: 0
    input: integer
---
    
In this script we will find CNV losses that overlap with TP53 domains:
 
 - TAD = trans-activating domain (essential for function)
 - DBD = DNA-binding domain (residues 102–292)
 - TD = tetramerization domain (residues 326–356)

We want to subset CNV calls where the domain are lost which will possibly	lead to loss of function to use for evaluation of TP53 inactivation score at a later step.

### Setup
```{r}
library("ggpubr")
library("ggthemes")
library("tidyverse")

# rootdir
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
results_dir <- file.path(root_dir,
                         "analyses",
                         "tp53_nf1_score",
                         "results")

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}


```

### Input files

`02-add-ploidy-consensus.Rmd` in focal-cn-file-preparation adds ploidy information to the consensus SEG file and adds a status column that defines gain and loss broadly.

```
Rscript -e "rmarkdown::render('analyses/focal-cn-file-preparation/02-add-ploidy-consensus.Rmd', clean = TRUE)"
```

consensus_seg_with_status.tsv was copied from scratch to input folder here so we can access the file in this notebook 

```{r}
# consensus seg for file for the location of CNVs
consensus_seg <- read_tsv(file.path(root_dir, 
                                    "analyses",
                                    "tp53_nf1_score",
                                    "input",
                                    "consensus_seg_with_status.tsv"))

# Gene location and domain overlap file 
bioMartDataPfamTp53 <- 
  readRDS(system.file("extdata", "pfamDataBioMart.RDS", package = "annoFuse")) %>%
  dplyr::filter(hgnc_symbol=="TP53")

# histology file
if ( params$base_run ==0 ){
  clinical<-read.delim(file.path(data_dir,"pbta-histologies.tsv"), stringsAsFactors = FALSE)
} else{
  clinical<-read.delim(file.path(data_dir,"pbta-histologies-base.tsv"), stringsAsFactors = FALSE)  
}

histologies_df <- clinical %>%
  dplyr::select("Kids_First_Biospecimen_ID",
                "sample_id")

# Classifier score 
# 1) Read in scores from tp53-nf1-classifier for stranded

score_stranded_df <- read_tsv(file.path(results_dir, "pbta-gene-expression-rsem-fpkm-collapsed.stranded_classifier_scores.tsv")) %>%
  dplyr::select(
    "tp53_score",
    "sample_id"
  ) %>% 
  dplyr::rename("Kids_First_Biospecimen_ID"="sample_id") %>%
  left_join(histologies_df,by=c("Kids_First_Biospecimen_ID")) %>%
  as.data.frame()

# 2) Read in scores from tp53-nf1-classifier for polya
score_polya_df <- read_tsv(file.path(results_dir, "pbta-gene-expression-rsem-fpkm-collapsed.polya_classifier_scores.tsv")) %>%
  dplyr::select(
    "tp53_score",
    "sample_id"
  ) %>%
  dplyr::rename("Kids_First_Biospecimen_ID"="sample_id") %>%
  left_join(histologies_df,by=c("Kids_First_Biospecimen_ID")) %>%
  as.data.frame()

# merge classifier scores
tp53_classifier_score <- rbind(score_stranded_df,score_polya_df)

```

Generating genomics ranges for cnv and domain dataframes

```{r}
# Genomic range for cnv seg file 
cnv_gr <- consensus_seg %>%
  dplyr::rename(chr = chrom, start = loc.start, end = loc.end,
                copy_number = copy.num) %>%
  dplyr::select(-num.mark, -seg.mean) %>%
  GenomicRanges::makeGRangesFromDataFrame(keep.extra.columns = TRUE,
                                          starts.in.df.are.0based = FALSE)

# Genomic range for gene location and domain overlap file 
domain_gr <- bioMartDataPfamTp53 %>%
  dplyr::filter(!is.na(domain_start),!is.na(domain_end)) %>%
  # formatting strand information
  dplyr::mutate(strand = if_else(strand=="-1","-","+"),
                chromosome_name = paste0("chr",chromosome_name)) %>%
  dplyr::rename(chr = chromosome_name, start = domain_start, end = domain_end) %>%
  GenomicRanges::makeGRangesFromDataFrame(keep.extra.columns = TRUE,
                                          starts.in.df.are.0based = FALSE)

```



### Overlap CNV and TP53 Domain ranges 

```{r}
# overlap cnv and domain
overlaps <- IRanges::mergeByOverlaps(cnv_gr, domain_gr)

# get CNV and domain overlap per BS id
annotated_cn <- data.frame(
  biospecimen_id = overlaps$Kids_First_Biospecimen_ID,
  status = overlaps$status,
  copy_number = overlaps$copy_number,
  ploidy = overlaps$tumor_ploidy,
  hgnc_symbol = overlaps$hgnc_symbol,
  pfam_id = overlaps$pfam_id,
  NAME = overlaps$NAME,
  stringsAsFactors = FALSE
) %>%
  dplyr::distinct() %>%
  # select loss that overlaps the TP53 core domains 
  dplyr::filter(status=="loss") %>%
  dplyr::group_by(biospecimen_id,copy_number,ploidy) %>%
  # summarise domain NAME per biospecimen_id TP53 loss
  dplyr::summarise(
    domain=toString(NAME)) %>%
  left_join(histologies_df,by=c("biospecimen_id"="Kids_First_Biospecimen_ID")) %>%
  left_join(tp53_classifier_score,by=c("sample_id"))
```

### Copy number overlapping TP53 domain

We want to check if classifier scores support CNV calls and if support from classifier (higher inactivation score) can be used to filter TP53 loss calls 

```{r}

ggplot(annotated_cn, aes(x = factor(copy_number), y = tp53_score)) +
  geom_violin()+
  geom_jitter(alpha = 0.5, width = 0.2) +
  stat_compare_means() +
  theme_bw() +
  ggtitle("Distribution of scores across aberrations") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  xlab("Copy number")

```


TP53 with copy_number <=1 show higher tp53 inactivation classifier scores, we will save copy_number <=1 calls for TP53 as high confidence losses.

### Save high confidence TP53 loss in file

```{r}
annotated_cn %>%
  # high confidence CNV losses
  # with high TP53 inactivation score
  dplyr::filter(copy_number<=1) %>%
  # remove columns 
  dplyr::select(-tp53_score,-Kids_First_Biospecimen_ID) %>%
  unique() %>%
  write_tsv( file.path(results_dir,"loss_overlap_domains_tp53.tsv"))

```


