---
title: "Find CNV losses that overlap with TP53 domains"
author: "K S Gaonkar (D3B)"
output: html_notebook
params:
  base_run:
    label: "1/0 to run with base histology"
    value: 0
    input: integer
---
    
In this script we will find CNV losses that overlap with TP53 domains:
 
 - TAD = trans-activating domain (essential for function)
 - DBD = DNA-binding domain (residues 102–292)
 - TD = tetramerization domain (residues 326–356)

We want to subset CNV calls where the domain are lost which will possibly	lead to loss of function to use for evaluation of TP53 inactivation score at a later step.

### Setup
```{r}
library("ggpubr")
library("ggthemes")
library("tidyverse")

# rootdir
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
results_dir <- file.path(root_dir,
                         "analyses",
                         "tp53_nf1_score",
                         "results")

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}


```

### Input files

`02-add-ploidy-consensus.Rmd` in focal-cn-file-preparation adds ploidy information to the consensus SEG file and adds a status column that defines gain and loss broadly.

```
Rscript -e "rmarkdown::render('analyses/focal-cn-file-preparation/02-add-ploidy-consensus.Rmd', clean = TRUE)"
```

`consensus_seg_with_status.tsv` was copied from scratch to input folder here so we can access the file in this notebook. Additionally, per our current workflow, we only run consensus CNV calling for WGS samples. And for WXS samples, we use CNVKIT results straight. Therefore, `cnvkit_with_status.tsv` is also moved and that is particularly for WXS samples. The following code needs to be run to get the cnvkit annotated file:

```
Rscript -e "rmarkdown::render('analyses/focal-cn-file-preparation/01-add-ploidy-cnvkit.Rmd', clean = TRUE)"
```

Read in CNV files with status and location-domain overlap file
```{r}
# consensus seg for file for the location of CNVs (WGS samples)
consensus_seg <- read_tsv(file.path(root_dir, 
                                    "analyses",
                                    "tp53_nf1_score",
                                    "input",
                                    "consensus_seg_with_status.tsv"))

# cnvkit seg for file for the location of CNVs (WXS samples)
cnvkit_seg<- read_tsv(file.path(root_dir, 
                                "analyses",
                                "tp53_nf1_score",
                                "input",
                                "cnvkit_with_status.tsv"))

# Gene location and domain overlap file 
bioMartDataPfamTp53 <- 
  readRDS(system.file("extdata", "pfamDataBioMart.RDS", package = "annoFuse")) %>%
  dplyr::filter(hgnc_symbol=="TP53")

```

Read in histology file. Subset based on experimental strategy so that all samples have the correct CNV status file to use.
```{r}
# histology file
if ( params$base_run ==0 ){
  clinical<-read.delim(file.path(data_dir,"histologies.tsv"), stringsAsFactors = FALSE)
} else{
  clinical<-read.delim(file.path(data_dir,"histologies-base.tsv"), stringsAsFactors = FALSE)  
}

histologies_df <- clinical %>% 
  dplyr::select(Kids_First_Biospecimen_ID, sample_id)

# find the WGS samples + WXS samples and subset "_with_status.tsv" file with that
wgs_samples <- histologies_wgs_df <- clinical %>% dplyr::filter(sample_type == "Tumor") %>%
  dplyr::filter(experimental_strategy == "WGS") %>% 
  dplyr::filter(cohort =="PBTA") %>% 
  dplyr::pull("Kids_First_Biospecimen_ID")

wxs_samples <- clinical %>% dplyr::filter(sample_type == "Tumor") %>%
  dplyr::filter(experimental_strategy == "WXS") %>% 
  dplyr::filter(cohort =="PBTA") %>% 
  dplyr::pull("Kids_First_Biospecimen_ID")
  
# Filter the consensus and the cnvkit seg files to contain only either WGS or WXS samples 
consensus_seg <- consensus_seg %>% dplyr::filter(Kids_First_Biospecimen_ID %in% wgs_samples)
cnvkit_seg <- cnvkit_seg %>% dplyr::filter(Kids_First_Biospecimen_ID %in% wxs_samples)

# merge the seg files together
combined_seg <- rbind(consensus_seg, cnvkit_seg)
  
# Classifier score 
# Read in scores from tp53-nf1-classifier 

tp53_classifier_score <- readr::read_tsv(file.path(results_dir, "gene-expression-rsem-tpm-collapsed_classifier_scores.tsv")) %>%
  dplyr::select(
    "tp53_score",
    "sample_id"
  ) %>% 
  dplyr::rename("Kids_First_Biospecimen_ID"="sample_id") %>%
  left_join(histologies_df,by=c("Kids_First_Biospecimen_ID")) %>%
  as.data.frame()

```

Generating genomics ranges for cnv and domain dataframes

```{r}
# Genomic range for cnv seg file 
cnv_gr <- combined_seg %>%
  dplyr::rename(chr = chrom, start = loc.start, end = loc.end,
                copy_number = copy.num) %>%
  dplyr::select(-num.mark, -seg.mean) %>%
  GenomicRanges::makeGRangesFromDataFrame(keep.extra.columns = TRUE,
                                          starts.in.df.are.0based = FALSE)

# Genomic range for gene location and domain overlap file 
domain_gr <- bioMartDataPfamTp53 %>%
  dplyr::filter(!is.na(domain_start),!is.na(domain_end)) %>%
  # formatting strand information
  dplyr::mutate(strand = if_else(strand=="-1","-","+"),
                chromosome_name = paste0("chr",chromosome_name)) %>%
  dplyr::rename(chr = chromosome_name, start = domain_start, end = domain_end) %>%
  GenomicRanges::makeGRangesFromDataFrame(keep.extra.columns = TRUE,
                                          starts.in.df.are.0based = FALSE)

```


### Overlap CNV and TP53 Domain ranges 

```{r}
# overlap cnv and domain
overlaps <- IRanges::mergeByOverlaps(cnv_gr, domain_gr)

# get CNV and domain overlap per BS id
annotated_cn <- data.frame(
  biospecimen_id = overlaps$Kids_First_Biospecimen_ID,
  status = overlaps$status,
  copy_number = overlaps$copy_number,
  ploidy = overlaps$tumor_ploidy,
  hgnc_symbol = overlaps$hgnc_symbol,
  pfam_id = overlaps$pfam_id,
  NAME = overlaps$NAME,
  stringsAsFactors = FALSE
) %>%
  dplyr::distinct() %>%
  # select loss that overlaps the TP53 core domains 
  dplyr::filter(status=="loss") %>%
  dplyr::group_by(biospecimen_id,copy_number,ploidy) %>%
  # summarise domain NAME per biospecimen_id TP53 loss
  dplyr::summarise(
    domain=toString(NAME)) %>%
  left_join(histologies_df,by=c("biospecimen_id"="Kids_First_Biospecimen_ID")) %>%
  left_join(tp53_classifier_score,by=c("sample_id"))
```

### Copy number overlapping TP53 domain

We want to check if classifier scores support CNV calls and if support from classifier (higher inactivation score) can be used to filter TP53 loss calls 

```{r}

ggplot(annotated_cn, aes(x = factor(copy_number), y = tp53_score)) +
  geom_violin()+
  geom_jitter(alpha = 0.5, width = 0.2) +
  stat_compare_means() +
  theme_bw() +
  ggtitle("Distribution of scores across aberrations") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  xlab("Copy number")

```


TP53 with copy_number <=1 show higher tp53 inactivation classifier scores, we will save copy_number <=1 calls for TP53 as high confidence losses.

### Save high confidence TP53 loss in file

```{r}
annotated_cn %>%
  # high confidence CNV losses
  # with high TP53 inactivation score
  dplyr::filter(copy_number<=1) %>%
  # remove columns 
  dplyr::select(-tp53_score,-Kids_First_Biospecimen_ID) %>%
  unique() %>%
  write_tsv( file.path(results_dir,"loss_overlap_domains_tp53.tsv"))

```


