---
title: "Find CNV losses that overlap with TP53 domains"
author: "K S Gaonkar (D3B)"
output: html_notebook
params:
  base_run:
    label: "1/0 to run with base histology"
    value: 0
    input: integer
  tumor_purity_threshold:
    label: Indicates whether this notebook should consume/output tumor purity thresholded or full (non-thresholded) results for stranded data specifically. 1 indicates thresholded, and 0 indicates full results.
    value: 0
    input: integer
---

In this script we will find CNV losses that overlap with TP53 domains:

- TAD = trans-activating domain (essential for function)
- DBD = DNA-binding domain (residues 102–292)
- TD = tetramerization domain (residues 326–356)

We want to subset CNV calls where the domain are lost which will possibly	lead to loss of function to use for evaluation of TP53 inactivation score at a later step.

### Setup
```{r}
suppressPackageStartupMessages({
  library(ggpubr)
  library(ggthemes)
  library(tidyverse)
  library(annoFuse)
})


# rootdir
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
results_dir <- file.path(root_dir,
                         "analyses",
                         "tp53_nf1_score",
                         "results")

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

# Save also as `polya_results_dir` to always read in polyA results properly
polya_results_dir <- results_dir
```

### Input files

```{r}
# consensus seg for file for the location of CNVs
consensus_seg <- read_tsv(file.path(data_dir,
                                    "consensus_seg_with_status.tsv"))

# Gene location and domain overlap file 
bioMartDataPfamTp53 <- 
  readRDS(system.file("extdata", "pfamDataBioMart.RDS", package = "annoFuse")) %>%
  dplyr::filter(hgnc_symbol=="TP53")

# histology file
if ( params$base_run ==0 ){
  clinical<-read.delim(file.path(data_dir,"pbta-histologies.tsv"), stringsAsFactors = FALSE)
} else{
  clinical<-read.delim(file.path(data_dir,"pbta-histologies-base.tsv"), stringsAsFactors = FALSE)  
}

histologies_df <- clinical %>%
  dplyr::select("Kids_First_Biospecimen_ID",
                "sample_id")

# Classifier score 
# 1) Read in scores from tp53-nf1-classifier for stranded

# Define a filename "suffix" that indicates tumor purity thresholded vs. regular files
# specifically relevant for **stranded** data
# We'll also want to update the output directory name for results
if (params$tumor_purity_threshold == 1) {
  
  tumor_purity_suffix <- "_tumor-purity-threshold"
  
  # Update results_dir name
  results_dir <- file.path(results_dir, "tumor-purity-threshold")
  if (!dir.exists(results_dir)) {
    dir.create(results_dir)
  }
  
} else {
  tumor_purity_suffix <- ""
}

score_stranded_df <- read_tsv(file.path(results_dir, 
                                        glue::glue("pbta-gene-expression-rsem-fpkm-collapsed.stranded_classifier_scores{tumor_purity_suffix}.tsv"))) %>%
  dplyr::select(
    "tp53_score",
    "sample_id"
  ) %>% 
  dplyr::rename("Kids_First_Biospecimen_ID"="sample_id") %>%
  left_join(histologies_df,by=c("Kids_First_Biospecimen_ID")) %>%
  as.data.frame()

# 2) Read in scores from tp53-nf1-classifier for polya
# Note that this path must "results" even if the `results_dir` was updated, so we use `polya_results_dir`
score_polya_df <- read_tsv(file.path(polya_results_dir, "pbta-gene-expression-rsem-fpkm-collapsed.polya_classifier_scores.tsv")) %>%
  dplyr::select(
    "tp53_score",
    "sample_id"
  ) %>%
  dplyr::rename("Kids_First_Biospecimen_ID"="sample_id") %>%
  left_join(histologies_df,by=c("Kids_First_Biospecimen_ID")) %>%
  as.data.frame()

# merge classifier scores
tp53_classifier_score <- rbind(score_stranded_df,score_polya_df)
```

Generating genomics ranges for cnv and domain dataframes

```{r}
# Genomic range for cnv seg file 
cnv_gr <- consensus_seg %>%
  dplyr::rename(chr = chrom, start = loc.start, end = loc.end,
                copy_number = copy.num) %>%
  dplyr::select(-num.mark, -seg.mean) %>%
  GenomicRanges::makeGRangesFromDataFrame(keep.extra.columns = TRUE,
                                          starts.in.df.are.0based = FALSE)

# Genomic range for gene location and domain overlap file 
domain_gr <- bioMartDataPfamTp53 %>%
  dplyr::filter(!is.na(domain_start),!is.na(domain_end)) %>%
  # formatting strand information
  dplyr::mutate(strand = if_else(strand=="-1","-","+"),
                chromosome_name = paste0("chr",chromosome_name)) %>%
  dplyr::rename(chr = chromosome_name, start = domain_start, end = domain_end) %>%
  GenomicRanges::makeGRangesFromDataFrame(keep.extra.columns = TRUE,
                                          starts.in.df.are.0based = FALSE)
```



### Overlap CNV and TP53 Domain ranges 

```{r}
# overlap cnv and domain
overlaps <- IRanges::mergeByOverlaps(cnv_gr, domain_gr)

# get CNV and domain overlap per BS id
annotated_cn <- data.frame(
  biospecimen_id = overlaps$Kids_First_Biospecimen_ID,
  status = overlaps$status,
  copy_number = overlaps$copy_number,
  ploidy = overlaps$tumor_ploidy,
  hgnc_symbol = overlaps$hgnc_symbol,
  pfam_id = overlaps$pfam_id,
  NAME = overlaps$NAME,
  stringsAsFactors = FALSE
) %>%
  dplyr::distinct() %>%
  # select loss that overlaps the TP53 core domains 
  dplyr::filter(status=="loss") %>%
  dplyr::group_by(biospecimen_id,copy_number,ploidy) %>%
  # summarise domain NAME per biospecimen_id TP53 loss
  dplyr::summarise(
    domain=toString(NAME)) %>%
  left_join(histologies_df,by=c("biospecimen_id"="Kids_First_Biospecimen_ID")) %>%
  left_join(tp53_classifier_score,by=c("sample_id"))
```

### Copy number overlapping TP53 domain

We want to check if classifier scores support CNV calls and if support from classifier (higher inactivation score) can be used to filter TP53 loss calls 

```{r}
ggplot(annotated_cn, aes(x = factor(copy_number), y = tp53_score)) +
  geom_violin()+
  geom_jitter(alpha = 0.5, width = 0.2) +
  stat_compare_means() +
  theme_bw() +
  ggtitle("Distribution of scores across aberrations") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  xlab("Copy number")
```

Majority of TP53 losses have high tp53 inactivation classifier, let's save to file.

### Save high confidence TP53 loss in file

```{r}
annotated_cn %>%
  # remove columns not required in output  
  dplyr::select(-tp53_score,-Kids_First_Biospecimen_ID) %>%
  unique() %>%
  write_tsv( file.path(results_dir,
                       glue::glue("loss_overlap_domains_tp53{tumor_purity_suffix}.tsv")))
```


## Session Info 

```{r}
sessionInfo()
```
