---
title: "Generate Fusion Summary Files"
output: html_notebook
author: Daniel Miller (D3b), Jaclyn Taroni (CCDL), Jo Lynne Rokita (D3b)
date: January 2020, November 2020
---

Generate fusion files specifically for consumption by molecular subtyping analyses

## Set up

### Libraries and functions

```{r}
library(tidyverse)
```

```{r}
#' Generate filtered fusion frame
#' @param df Unfiltered fusion data frame
#' @param bioid List of biospecimen IDs
#' @param fuses List of explicit fusion names
#' @param genes List of gene names
#' @return the filtered fusion data frame
filterFusion <- function(df, bioid, fuses, genes) {
  if (!missing(bioid)) {
    df <- filter(df, Sample %in% bioid)
  }
  if (!missing(fuses) & !missing(genes)) {
    df <- filter(df, FusionName %in% fuses |
                   Gene1A %in% genes |
                   Gene2A %in% genes |
                   Gene1B %in% genes |
                   Gene2B %in% genes)
  } else if (!missing(fuses)) {
    df <- filter(df, FusionName %in% fuses)
  } else if (!missing(genes)) {
    df <- filter(df,
                 Gene1A %in% genes |
                   Gene2A %in% genes |
                   Gene1B %in% genes |
                   Gene2B %in% genes)
  }
  return(df %>% select(Sample, FusionName))
}

#create function specifically for lgat, since we want more information from the putative oncogenic file
filterLGATFusion <- function(df, bioid, fuses, genes) {
  if (!missing(bioid)) {
    df <- filter(df, Sample %in% bioid)
  }
  if (!missing(fuses) & !missing(genes)) {
    df <- filter(df, FusionName %in% fuses |
                   Gene1A %in% genes |
                   Gene2A %in% genes |
                   Gene1B %in% genes |
                   Gene2B %in% genes)
  } else if (!missing(fuses)) {
    df <- filter(df, FusionName %in% fuses)
  } else if (!missing(genes)) {
    df <- filter(df,
                 Gene1A %in% genes |
                   Gene2A %in% genes |
                   Gene1B %in% genes |
                   Gene2B %in% genes)
  }
  return(df %>% 
           select(Sample, FusionName, Fusion_Type, Gene1A, Gene1B, Gene2A,  Gene2B, Gene1A_anno, Gene1B_anno, LeftBreakpoint, RightBreakpoint,      reciprocal_exists, DomainRetainedGene1A, DomainRetainedGene1B) %>%
           distinct())
}

#' Generate matrix with fusion counts
#' @param fuseDF Filtered fusion data frame
#' @param bioid List of biospecimen IDs that should be included in final table

#' @return Data frame that contains fusion counts
prepareOutput <- function(fuseDF, bioid) {
  fuseDF %>% 
    # some fusions have in-frame and frameshift fusion calls for a sample
    # this will make unique fusionName and Sample dataset to get 1/0 values
    dplyr::select(Sample,FusionName) %>%
    unique() %>%
    reshape2::dcast(Sample ~ FusionName,fun.aggregate = length) %>%
    right_join(data.frame(Sample = bioid)) %>%
    replace(is.na(.), 0) %>%
    rename(Kids_First_Biospecimen_ID = Sample)
}
```

### Read in data

```{r}
dataDir <- file.path("..", "..", "data")
fusDir <- file.path("..", "..", "analyses", "fusion_filtering", "results")
annotDir <- file.path("..", "..", "analyses", "fusion_filtering", "references")
#' The putative oncogenic fusion file is what we'll use to check for the 
#' annotation file
annot <- read.delim2(file.path(annotDir, "genelistreference.txt"), sep = "\t", header = T)
#' presence or absence of the fusions.
putativeOncogenicDF <- 
  read_tsv(file.path(fusDir, "pbta-fusion-putative-oncogenic.tsv"))
#' However, some biospecimens are not represented in this filtered, prioritized
#' file but *are* present in the original files -- this will cause them to be
#' "missing" in the final files for consumption which could mislead analysts.
arribaDF <- read_tsv(file.path(dataDir, "pbta-fusion-arriba.tsv.gz"))
starfusionDF <- read_tsv(file.path(dataDir, "pbta-fusion-starfusion.tsv.gz"))
```

### Output

```{r}
resultsDir <- "results"
if (!dir.exists(resultsDir)) {
  dir.create(resultsDir)
}
ependFile <- file.path(resultsDir, "fusion_summary_ependymoma_foi.tsv")
embryFile <- file.path(resultsDir, "fusion_summary_embryonal_foi.tsv")
ewingsFile <- file.path(resultsDir, "fusion_summary_ewings_foi.tsv")
lgatFile <- file.path(resultsDir, "fusion_summary_lgat_foi.tsv")
```

## Fusions and genes of interest

Taken from [`AlexsLemonade/OpenPBTA-analysis#245`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/245), [`AlexsLemonade/OpenPBTA-analysis#251`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/251), [`AlexsLemonade/OpenPBTA-analysis#623`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/623) respectively, and [`AlexsLemonade/OpenPBTA-analysis#808`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/808) 

```{r}
#' **Filters**
#'
#' *Fusions Filters*
#' 1: Exact match a list of fusions common in Ependymoma tumors
ependFuses <- c(
  "C11orf95--MAML2",
  "C11orf95--RELA",
  "C11orf95--YAP1",
  "LTBP3--RELA",
  "PTEN--TAS2R1",
  "YAP1--FAM118B",
  "YAP1--MAMLD1",
  "YAP1--MAMLD2"
)
ependGenes <- c(
  "RELA"
)
#' 2: Exact match a list of fusions common in Embryonal tumors
#' as well as fusions containing a particular gene with any other gene
embryFuses <- c(
  "CIC--NUTM1",
  "MN1--BEND2",
  "MN1--CXXC5"
)
embryGenes <- c(
  "FOXR2",
  "MN1",
  "TTYH1"
)

#' 3: Exact match a list of fusions which are hallmark fusions in Ewings Sarcoma
ewingsFuses<- c(
  "EWSR1--FLI1", 
  "EWSR1--ERG",
  "FUS--ERG"
)

#' 4: Exact match a list of fusions common in low-grade astrocytic (LGAT) tumors
#' as well as fusions containing a particular gene with any other gene
lgatFuses <- c(
  "KIAA1549--BRAF",
  "FGFR1--TACC1",
  "MYB--QKI"
)
lgatGenes <- c(
  "BRAF",
  "ALK",
  "ROS1",
  "NTRK1",
  "NTRK2",
  "NTRK3",
  "PDGFRA",
  "FGFR2",
  "FGFR1",
  "MYB",
  "MYBL1",
  "RAF1"
)

```

### Filter putative oncogenic fusions list

```{r}
allFuseEpend <- filterFusion(df = putativeOncogenicDF,
                             fuses = ependFuses,
                             genes = ependGenes)
allFuseEmbry <- filterFusion(df = putativeOncogenicDF,
                             fuses = embryFuses,
                             genes = embryGenes)
allFuseEwing <- filterFusion(df = putativeOncogenicDF,
                             fuses = ewingsFuses)
allFuseLGAT <- filterLGATFusion(df = putativeOncogenicDF,
                             fuses = lgatFuses,
                             genes = lgatGenes)

```

Get the biospecimen IDs that are present in *either* caller file (Arriba, STARFusion).
The fusions in the putative oncogenic fusion file can be retained even if they are not in both callers: https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/8fba1753608d8ac0aa3d5d7d63c480b8f00ff0e9/analyses/fusion_filtering/04-project-specific-filtering.Rmd#L242
We use the putative oncogenic file here, therefore any sample that is in either file but does not have a fusion that is relevant to the subtyping tickets is not _missing_ but instead has no evidence of the relevant fusions.

```{r}
specimensUnion<- union(arribaDF$tumor_id, starfusionDF$tumor_id)
```

#### Write non-MB, non-ATRT embryonal fusions to file

```{r}
allFuseEmbry <- allFuseEmbry %>%
  prepareOutput(specimensUnion)

# Which fusions of interest are not present in any samples?
missingEmbryFusion <- setdiff(embryFuses, colnames(allFuseEmbry)[-1])
# For the fusions that are not present, fill those columns with 0
allFuseEmbry[, missingEmbryFusion] <- 0

# Write to file
allFuseEmbry %>%
  write_tsv(embryFile)
```

#### Write ependymoma fusions to file

```{r}
allFuseEpend <- allFuseEpend %>%
  prepareOutput(specimensUnion) 

# Which fusions of interest are not present in any samples?
missingEpendFusion <- setdiff(ependFuses, colnames(allFuseEpend)[-1])
# For the fusions that are not present, fill those columns with 0
allFuseEpend[, missingEpendFusion] <- 0

allFuseEpend %>%
  write_tsv(ependFile)
```


#### Write ewings sarcoma fusions to file

```{r}
allFuseEwing <- allFuseEwing %>%
  prepareOutput(specimensUnion)

# Which fusions of interest are not present in any samples?
missingEwingFusion <- setdiff(ewingsFuses, colnames(allFuseEwing)[-1])
# For the fusions that are not present, fill those columns with 0
allFuseEwing[, missingEwingFusion] <- 0

allFuseEwing %>%
  write_tsv(ewingsFile)
```

#### Perform selection for LGAT fusions

```{r}
# Which genes/fusions are not kinases, but in the list?
# Separate LGAT fusions into genes, combine with gene list, check for not kinase
lgatFuses_df <- as.data.frame(lgatFuses) %>%
  separate(lgatFuses, into = c("Gene1A", "Gene1B"), remove = F)
kinases <- annot %>%
  filter(type == "Kinase")
lgatFuses_df$Gene1A_anno <- ifelse(lgatFuses_df$Gene1A %in% kinases$Gene_Symbol, "kinase", "")
lgatFuses_df$Gene1B_anno <- ifelse(lgatFuses_df$Gene1B %in% kinases$Gene_Symbol, "kinase", "")
# Only pull fusions that do not contain kinase genes, as ones with kinases will be dealt with sepaately later
lgatFuses_df <- lgatFuses_df %>%
  filter(lgatFuses_df$Gene1A_anno != "kinase" & lgatFuses_df$Gene1B_anno != "kinase") %>%
  select(lgatFuses)
nonkinase_lgatFuses <- lgatFuses_df$lgatFuses

# Identify non-kinase genes in LGAT goi list
nonkinase_lgatGenes <- setdiff(lgatGenes, kinases$Gene_Symbol)

# Pull LGAT non-kinase fusions
nonkinaseLGAT <- filterFusion(df = putativeOncogenicDF,
                             fuses = nonkinase_lgatFuses,
                             genes = nonkinase_lgatGenes) %>%
  distinct()

# First, filter all fusion dataframe for 3' kinases which are in-frame and retain the kinase domain - keep these
three_prime_kinase_inframe <- allFuseLGAT %>%
  filter(grepl("Kinase", Gene1B_anno) & Fusion_Type == "in-frame" & DomainRetainedGene1B == "Yes") %>%
  select(Sample, FusionName, Gene1A, Gene1B) %>%
  distinct()

# Which fusions are not in-frame?
three_prime_kinase_outframe <- allFuseLGAT %>%
  filter(grepl("Kinase", Gene1B_anno) & Fusion_Type != "in-frame") %>%
  distinct()
# Check they don't retain the kinase domain. They do not, so leave out.
table(three_prime_kinase_outframe$DomainRetainedGene1B)

# Filter all fusions for 5' kinase fusions which have reciprocal fusions and have lost the kinase domain. 
five_prime_domain_lost <- allFuseLGAT %>%
  filter(grepl("Kinase", Gene1A_anno) & reciprocal_exists == "TRUE") %>%
  select(Sample, FusionName, Gene1A, Gene1B, Fusion_Type, DomainRetainedGene1A) %>%
  filter(DomainRetainedGene1A == "No") %>%
  distinct()

# Keep those with kinase domain retained and fusion in-frame - keep this list
five_prime_domain_intact <- allFuseLGAT %>%
  filter(grepl("Kinase", Gene1A_anno) & reciprocal_exists == "TRUE") %>%
  select(Sample, FusionName, Gene1A, Gene1B, Fusion_Type, DomainRetainedGene1A) %>%
  filter(DomainRetainedGene1A == "Yes") %>%
  distinct()

# Identify reciprocal matched fusions which are in-frame
five_prime_domain_lost$reciprocal <- paste(five_prime_domain_lost$Gene1B, five_prime_domain_lost$Gene1A, sep ="--")
five_prime_reciprocals <- five_prime_domain_lost[,c("Sample", "reciprocal")] %>%
  rename(FusionName = reciprocal)
five_prime_reciprocals_full <- five_prime_reciprocals %>%
  left_join(allFuseLGAT, by = c("Sample", "FusionName")) %>%
  filter(Fusion_Type == "in-frame")
# Keep the list of the original 5' kinase fusions with matched reciprocals
five_prime_kinase_keep <- five_prime_reciprocals_full[,c("Sample", "Gene1A", "Gene1B")] %>%
  distinct()
  five_prime_kinase_keep$FusionName <- paste(five_prime_kinase_keep$Gene1B, five_prime_kinase_keep$Gene1A, sep ="--")

# Which ones are not in-frame? These are mostly genes fused to self
out_frame_reciprocals <- setdiff(five_prime_reciprocals[,c("Sample", "FusionName")], five_prime_reciprocals_full[,c("Sample", "FusionName")])
out_frame_reciprocals

#Check these do not retain the kinase domain. They do not, so will leave out
out_frame_reciprocals_full <- out_frame_reciprocals %>%
  left_join(allFuseLGAT, by = c("Sample", "FusionName")) %>%
  select(Sample, FusionName, Gene1A, Gene1B, Fusion_Type, DomainRetainedGene1A, DomainRetainedGene1B)

# Rbind lists for final table of LGAT fusions of interest
subsetFuseLGAT <- bind_rows(nonkinaseLGAT, three_prime_kinase_inframe[,c("Sample", "FusionName")],
                         five_prime_domain_intact[,c("Sample", "FusionName")],  
                         five_prime_kinase_keep[,c("Sample", "FusionName")]) %>%
  distinct()
```

#### Write LGAT fusions to file

```{r}
subFuseLGAT <- subFuseLGAT %>%
  prepareOutput(specimensUnion)

# Which fusions of interest are not present in any samples?
missingLgatFusion <- setdiff(lgatFuses, colnames(subFuseLGAT)[-1])
# For the fusions that are not present, fill those columns with 0
subFuseLGAT[, missingLgatFusion] <- 0

subFuseLGAT %>%
  write_tsv(lgatFile)
```

## Session Info

```{r}
sessionInfo()
```
