---
title: "Generate Fusion Summary Files"
output: html_notebook
author: Daniel Miller (D3b) and Jaclyn Taroni (CCDL)
date: January 2020
params:
  is_ci: 0
---

Generate fusion files specifically for consumption by molecular subtyping analyses

## Set up

```{r}
# if running in CI, we need to skip the EPN steps
if (params$is_ci == 0) running_in_ci <- FALSE
if (params$is_ci == 1) running_in_ci <- TRUE
```

### Libraries and functions

```{r}
library(tidyverse)
```

```{r}
#' Generate filtered fusion frame
#' @param df Unfiltered fusion data frame
#' @param bioid List of biospecimen IDs
#' @param fuses List of explicit fusion names
#' @param genes List of gene names
#' @return the filtered fusion data frame
filterFusion <- function(df, bioid, fuses, genes) {
  if (!missing(bioid)) {
    df <- filter(df, Sample %in% bioid)
  }
  if (!missing(fuses) & !missing(genes)) {
    df <- filter(df, FusionName %in% fuses |
                   Gene1A %in% genes |
                   Gene2A %in% genes |
                   Gene1B %in% genes |
                   Gene2B %in% genes)
  } else if (!missing(fuses)) {
    df <- filter(df, FusionName %in% fuses)
  } else if (!missing(genes)) {
    df <- filter(df,
                 Gene1A %in% genes |
                   Gene2A %in% genes |
                   Gene1B %in% genes |
                   Gene2B %in% genes)
  }
  return(df %>% select(Sample, FusionName))
}


#' Generate matrix with fusion counts
#' @param fuseDF Filtered fusion data frame
#' @param bioid List of biospecimen IDs that should be included in final table

#' @return Data frame that contains fusion counts
prepareOutput <- function(fuseDF, bioid) {
  fuseDF %>% 
    reshape2::dcast(Sample ~ FusionName) %>%
    right_join(data.frame(Sample = bioid)) %>%
    replace(is.na(.), 0) %>%
    rename(Kids_First_Biospecimen_ID = Sample)
}
```

### Read in data

```{r}
dataDir <- file.path("..", "..", "data")
#' The putative oncogenic fusion file is what we'll use to check for the 
#' presence or absence of the fusions.
putativeOncogenicDF <- 
  read_tsv(file.path(dataDir, "pbta-fusion-putative-oncogenic.tsv"))
#' However, some biospecimens are not represented in this filtered, prioritized
#' file but *are* present in the original files -- this will cause them to be
#' "missing" in the final files for consumption which could mislead analysts.
arribaDF <- read_tsv(file.path(dataDir, "pbta-fusion-arriba.tsv.gz"))
starfusionDF <- read_tsv(file.path(dataDir, "pbta-fusion-starfusion.tsv.gz"))
```

### Output

```{r}
resultsDir <- "results"
if (!dir.exists(resultsDir)) {
  dir.create(resultsDir)
}
ependFile <- file.path(resultsDir, "fusion_summary_ependymoma_foi.tsv")
embryFile <- file.path(resultsDir, "fusion_summary_embryonal_foi.tsv")
```

## Fusions and genes of interest

Taken from [`AlexsLemonade/OpenPBTA-analysis#245`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/245) and [`AlexsLemonade/OpenPBTA-analysis#251`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/251), respectively.

```{r}
#' **Filters**
#'
#' *Fusions Filters*
#' 1: Exact match a list of fusions common in Ependymoma tumors
ependFuses <- c(
  "C11orf95--MAML2",
  "C11orf95--RELA",
  "C11orf95--YAP1",
  "LTBP3--RELA",
  "PTEN--TAS2R1",
  "YAP1--FAM118B",
  "YAP1--MAMLD1",
  "YAP1--MAMLD2"
)
ependGenes <- c(
  "RELA"
)
#' 2: Exact match a list of fusions common in Embryonal tumors
#' as well as fusions containing a particular gene with any other gene
embryFuses <- c(
  "CIC--NUTM1",
  "MN1--BEND2",
  "MN1--CXXC5"
)
embryGenes <- c(
  "FOXR2",
  "MN1",
  "TTYH1"
)
```

### Filter putative oncogenic fusions list

```{r}
allFuseEpend <- filterFusion(df = putativeOncogenicDF,
                             fuses = ependFuses,
                             genes = ependGenes)
allFuseEmbry <- filterFusion(df = putativeOncogenicDF,
                             fuses = embryFuses,
                             genes = embryGenes)
```

Get the biospecimen IDs that are present in *either* caller file (Arriba, STARFusion).
The fusions in the putative oncogenic fusion file can be retained even if they are not in both callers: https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/8fba1753608d8ac0aa3d5d7d63c480b8f00ff0e9/analyses/fusion_filtering/04-project-specific-filtering.Rmd#L242
We use the putative oncogenic file here, therefore any sample that is in either file but does not have a fusion that is relevant to the subtyping tickets is not _missing_ but instead has no evidence of the relevant fusions.

```{r}
specimensUnion<- union(arribaDF$tumor_id, starfusionDF$tumor_id)
```

#### Write non-MB, non-ATRT embryonal fusions to file

```{r}
allFuseEmbry <- allFuseEmbry %>%
  prepareOutput(specimensUnion)
```

```{r}
# Are there any missing fusions?
setdiff(embryFuses, colnames(allFuseEmbry))
```

```{r}
allFuseEmbry %>%
  mutate(
    `CIC--NUTM1` = 0,
    `MN1--BEND2` = 0
  ) %>%
  write_tsv(embryFile)
```

#### Write ependymoma fusions to file

```{r}
if (!running_in_ci) {
  allFuseEpend %>%
    prepareOutput(specimensUnion) %>%
    mutate(
      `C11orf95--YAP1` = 0,
      `LTBP3--RELA` = 0,
      `PTEN--TAS2R1` = 0,
      `YAP1--MAMLD2` = 0
    ) %>%
    write_tsv(ependFile)
}
```
