---
title: "Project specific filtering"
output: html_notebook
params:
  histology:
    label: "Clinical file" 
    value: data/pbta-histologies.tsv
    input: file
  group:
    label: "Grouping variable"
    value: broad_histology
    input: string
  dataStranded:
    label: "Input filtered fusion dataframe"
    value: scratch/standardFusionStrandedExp_QC_expression_GTExComparison_annotated.RDS
    input: file
  dataPolya:
    label: "Input filtered fusion dataframe"
    value: scratch/standardFusionPolyaExp_QC_expression_GTExComparison_annotated.RDS
    input: file    
  numCaller:
    label: "Least Number of callers to have called fusion"
    value: 2
    input: integer
  numSample:
    label: "Least Number of samples to have fusion per group"
    value: 2
    input: integer
  numGroup:
    label: "Max number of groups found in"
    value: 1
    input: integer
  limitMultiFused:
    label: "Max number of times gene can be fused per sample"
    value: 5
    input: integer     

    
---

K S Gaonkar

Filtered Fusions:
1. In-frame fusions is called in atleast 2 samples per histology OR
2. In-frame fusions is called in atleast 2 callers 
AND
Filtered-fusions found in more than 1 histology OR
Filtered-fusion doesn't have multi-fused gene (more than 5 times in sample)

Putative Driver:
Filtering for general cancer specific genes
Fusions with genes in either onco

This notebook assumes you are in OpenPBTA-analysis project folder structure.


```{r}

#load required packages
suppressPackageStartupMessages(library("readr"))
suppressPackageStartupMessages(library("tidyverse"))
suppressPackageStartupMessages(library("reshape2"))
suppressPackageStartupMessages(library("rprojroot"))

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
setwd(root_dir)

#read filtFusion files
strandedQCGeneFiltered_filtFusion<-readRDS(params$dataStranded)
polyaQCGeneFiltered_filtFusion<-readRDS(params$dataPolya)

QCGeneFiltered_filtFusion<-rbind(strandedQCGeneFiltered_filtFusion,polyaQCGeneFiltered_filtFusion)

# subset for recurrent fusion detection and multifused genes QC
fusion_calls<-unique(QCGeneFiltered_filtFusion)
# remove distance from intergenic fusions
fusion_calls$FusionName<-unlist(lapply(fusion_calls$FusionName,function(x) rm_between(x, "(", ")", extract = F)))


# get grouping column id
group<-params$group

# get histology file
clinical<-read.delim(params$histology,stringsAsFactors = FALSE)

# Least number of callers
numCaller<-params$numCaller

# Least number of samples per group
numSample<-params$numSample

# Max number of groups 
numGroup<-params$numGroup

# Max number of times gene can be fused per sample
limitMultiFused<-params$limitMultiFused

print("Raw calls from STARfusion and Arriba for PBTA")
table(fusion_calls$Caller)
```


```{r}

# aggregate caller 
fusion_caller.summary <- fusion_calls %>%
  dplyr::filter(Fusion_Type != "other") %>%
  dplyr::select(Sample,FusionName,Caller) %>%
  group_by(FusionName, Sample ) %>%
  unique() %>%
  dplyr::mutate(CalledBy = toString(Caller), caller.count = n())
  
# aggregate by read count
 fusion_read.summary <- fusion_calls %>%
   dplyr::filter(Fusion_Type != "other") %>%
   dplyr::select(Sample,FusionName,JunctionReadCount,SpanningFragCount) %>%
   group_by(FusionName, Sample ) %>%
   unique() %>%
   dplyr::mutate(JunctionReadCountSum=sum(JunctionReadCount),SpanningFragCountSum=sum(SpanningFragCount))

fusion_summary<-left_join(fusion_caller.summary,fusion_read.summary,by=c("Sample","FusionName"))
 
fusion_calls<-fusion_calls %>% 
  dplyr::filter(Fusion_Type != "other") %>%
  full_join(fusion_summary,by=(c("Sample","FusionName","Caller","JunctionReadCount","SpanningFragCount")))

#merge with histology file
fusion_calls<-merge(fusion_calls,clinical,by.x="Sample",by.y="Kids_First_Biospecimen_ID")

# remove fusion within 50000 distance local rearrangement
fusion_calls <- fusion_calls %>%
  # remove local rearrangement/adjacent genes
  dplyr::filter(!grepl("LOCAL_REARRANGEMENT|LOCAL_INVERSION",annots))

```


```{r}
# Gene fusion should be in-frame
# AND
#
# 1. Called by at least n callers
fusion_calls.summary <- fusion_calls %>%
  dplyr::filter(caller.count >= numCaller) %>%
  unique() %>%
  mutate(note=paste0("Called by",numCaller, "callers")) %>%
  as.data.frame()

# OR
# 2. Found in at least n samples in each group
sample.count <- fusion_calls %>%
  dplyr::filter(Fusion_Type != "other") %>%
  dplyr::select(FusionName, Sample, !!as.name(group),-Fusion_Type) %>%
  unique() %>%
  group_by(FusionName, !!as.name(group)) %>%
  dplyr::mutate(sample.count = n(),Sample = toString(Sample)) %>%
  dplyr::filter(sample.count > numSample) %>%
  unique() %>%
  mutate(note=paste0("Found in at least ",numSample, " samples in a group")) %>%
  as.data.frame()


```



```{r}
#filter QCGeneFiltered_filtFusion to keep recurrent fusions from above sample.count and fusion_calls.summary

QCGeneFiltered_recFusion<-fusion_calls %>%
  dplyr::filter(FusionName %in% unique(c(sample.count$FusionName,fusion_calls.summary$FusionName)))

```





```{r}
# remove fusions that are in > numGroup
group.count <- fusion_calls %>%
  dplyr::select(FusionName, !!as.name(group)) %>%
  unique() %>%
  group_by(FusionName) %>%
  dplyr::mutate(group.ct = n(),Sample = toString(!!(as.name(group)))) %>%
  dplyr::filter(group.ct >numGroup)


# remove multi-fused genes
fusion_recurrent5_per_sample <- fusion_calls  %>%
  # We want to keep track of the gene symbols for each sample-fusion pair
  dplyr::select(Sample, FusionName, Gene1A, Gene1B, Gene2A, Gene2B) %>%
  # We want a single column that contains the gene symbols
  tidyr::gather(Gene1A, Gene1B, Gene2A, Gene2B,
                key = gene_position, value = GeneSymbol) %>%
  # Remove columns without gene symbols
  dplyr::filter(GeneSymbol != "") %>%
  dplyr::arrange(Sample, FusionName) %>%
  # Retain only distinct rows
  dplyr::distinct() %>%
  group_by(Sample,GeneSymbol) %>% 
  dplyr::summarise(Gene.ct = n()) %>% 
  dplyr::filter(Gene.ct>limitMultiFused) %>%
  mutate(note=paste0("multfused " ,limitMultiFused, " times per sample"))



```


```{r}


outputfolder<-"results"
setwd(outputfolder)
# filter QCGeneFiltered_recFusion to remove fusions found in more than 1 group 

QCGeneFiltered_recFusionUniq<-QCGeneFiltered_recFusion %>%
  dplyr::filter(!FusionName %in% group.count$FusionName) 


write.table(QCGeneFiltered_recFusionUniq,"FilteredFusionv2.tsv",sep="\t",quote=FALSE,row.names = FALSE)

```





```{r}
setwd(outputfolder)

# filter for putative driver genes and mutifused genes per sample
putative_fusions <- fusion_calls %>%
  dplyr::filter(!Gene1A %in% fusion_recurrent5_per_sample$GeneSymbol |
                  !Gene2A %in% fusion_recurrent5_per_sample$GeneSymbol |
                  !Gene1B %in% fusion_recurrent5_per_sample$GeneSymbol | 
                  !Gene2B %in% fusion_recurrent5_per_sample$GeneSymbol) %>%
  dplyr::filter(!is.na(Gene1A_anno) | !is.na(Gene1B_anno) | !is.na(Gene2A_anno) | !is.na(Gene2B_anno))
  
write.table(putative_fusions,"PutativeDriverFusionv2.tsv",sep="\t",quote=FALSE,row.names = FALSE)

# TO-DO
# maybe a more stringent inframe filter for putative drivers

```








