---
title: "Project specific filtering"
output: html_notebook
params:
  histology:
    label: "Clinical file" 
    value: data/pbta-histologies.tsv
    input: file
  group:
    label: "Grouping variable"
    value: broad_histology
    input: string
  dataStranded:
    label: "Input filtered fusion dataframe"
    value: scratch/standardFusionStrandedExp_QC_expression_GTExComparison_annotated.RDS
    input: file
  dataPolya:
    label: "Input filtered fusion dataframe"
    value: scratch/standardFusionPolyaExp_QC_expression_GTExComparison_annotated.RDS
    input: file    
  numCaller:
    label: "Least Number of callers to have called fusion"
    value: 2
    input: integer
  numSample:
    label: "Least Number of samples to have fusion per group"
    value: 2
    input: integer
  numGroup:
    label: "Max number of groups found in"
    value: 1
    input: integer
  limitMultiFused:
    label: "Max number of times gene can be fused per sample"
    value: 5
    input: integer  
  outputfolder:
    label: "results folder for *tsv files"
    value: results
    input: string     

    
---

K S Gaonkar

Filtered Fusions:
1. In-frame fusions is called in atleast 2 samples per histology OR
2. In-frame fusions is called in atleast 2 callers 
AND
Filtered-fusions found in more than 1 histology OR
Filtered-fusion doesn't have multi-fused gene (more than 5 times in sample)

Putative Driver:
Filtering for general cancer specific genes
Fusions with genes in either onco

This notebook assumes you are in OpenPBTA-analysis project folder structure.


```{r}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
#load required packages
suppressPackageStartupMessages(library("readr"))
suppressPackageStartupMessages(library("tidyverse"))
suppressPackageStartupMessages(library("reshape2"))
suppressPackageStartupMessages(library("qdapRegex"))

#read filtFusion files
strandedQCGeneFiltered_filtFusion<-readRDS(file.path(root_dir, params$dataStranded))
polyaQCGeneFiltered_filtFusion<-readRDS(file.path(root_dir, params$dataPolya))

# results folder
outputfolder<-params$outputfolder
QCGeneFiltered_filtFusion<-rbind(strandedQCGeneFiltered_filtFusion,polyaQCGeneFiltered_filtFusion)

write.table(QCGeneFiltered_filtFusion, file.path(outputfolder, "FilteredFusion.tsv"),sep="\t",quote=FALSE,row.names = FALSE)

# subset for recurrent fusion detection and multifused genes QC
fusion_calls<-unique(QCGeneFiltered_filtFusion)
# remove distance from intergenic fusions
fusion_calls$FusionName<-unlist(lapply(fusion_calls$FusionName,function(x) rm_between(x, "(", ")", extract = F)))


# get grouping column id
group<-params$group

# get histology file
clinical<-read.delim(file.path(root_dir, params$histology), stringsAsFactors = FALSE)
clinical<-clinical[,c("Kids_First_Biospecimen_ID","Kids_First_Participant_ID","broad_histology")]

# Least number of callers
numCaller<-params$numCaller

# Least number of samples per group
numSample<-params$numSample

# Max number of groups 
numGroup<-params$numGroup

# Max number of times gene can be fused per sample
limitMultiFused<-params$limitMultiFused


print("Raw calls from STARfusion and Arriba for PBTA")
table(fusion_calls$Caller)
```


```{r}

# aggregate caller 
fusion_caller.summary <- fusion_calls %>%
  dplyr::filter(Fusion_Type != "other") %>%
  dplyr::select(Sample,FusionName,Caller,Fusion_Type) %>%
  group_by(FusionName, Sample ,Fusion_Type) %>%
  unique() %>%
  dplyr::mutate(CalledBy = toString(Caller), caller.count = n()) %>%
  dplyr::select(-Caller)

# remove fusion within local rearrangement
fusion_calls <- fusion_calls %>%
  # remove local rearrangement/adjacent genes
  dplyr::filter(!grepl("LOCAL_REARRANGEMENT|LOCAL_INVERSION",annots))

#to add aggregated caller from fusion_caller.summary
fusion_calls<-fusion_calls %>% 
  dplyr::filter(Fusion_Type != "other") %>% dplyr::select(-Caller,-annots) %>%
  left_join(fusion_caller.summary,by=(c("Sample","FusionName","Fusion_Type"))) %>%
  dplyr::select(-JunctionReadCount,-SpanningFragCount,-Confidence,-LeftBreakpoint,-RightBreakpoint) %>% unique()

#merge with histology file
fusion_calls<-merge(fusion_calls,clinical,by.x="Sample",by.y="Kids_First_Biospecimen_ID")






```


```{r}
# Gene fusion should be in-frame
# AND
#
# 1. Called by at least n callers
fusion_calls.summary <- fusion_calls %>%
  dplyr::filter(caller.count >= numCaller) %>%
  unique() %>%
  mutate(note=paste0("Called by",numCaller, "callers")) %>%
  as.data.frame()

# OR
# 2. Found in at least n samples in each group
sample.count <- fusion_calls %>%
  dplyr::filter(Fusion_Type != "other") %>%
  dplyr::select(FusionName, Sample, !!as.name(group),-Fusion_Type) %>%
  unique() %>%
  group_by(FusionName, !!as.name(group)) %>%
  dplyr::mutate(sample.count = n(),Sample = toString(Sample)) %>%
  dplyr::filter(sample.count > numSample) %>%
  unique() %>%
  mutate(note=paste0("Found in atleast ",numSample, " samples in a group")) %>%
  as.data.frame()


```



```{r}
#filter QCGeneFiltered_filtFusion to keep recurrent fusions from above sample.count and fusion_calls.summary

QCGeneFiltered_recFusion<-fusion_calls %>%
  dplyr::filter(FusionName %in% unique(c(sample.count$FusionName,fusion_calls.summary$FusionName)))

```





```{r}
# remove fusions that are in > numGroup
group.count <- fusion_calls %>%
  dplyr::select(FusionName, !!as.name(group)) %>%
  unique() %>%
  group_by(FusionName) %>%
  dplyr::mutate(group.ct = n(),Sample = toString(!!(as.name(group)))) %>%
  dplyr::filter(group.ct >numGroup)


# remove multi-fused genes
fusion_recurrent5_per_sample <- fusion_calls  %>%
  # We want to keep track of the gene symbols for each sample-fusion pair
  dplyr::select(Sample, FusionName, Gene1A, Gene1B, Gene2A, Gene2B) %>%
  # We want a single column that contains the gene symbols
  tidyr::gather(Gene1A, Gene1B, Gene2A, Gene2B,
                key = gene_position, value = GeneSymbol) %>%
  # Remove columns without gene symbols
  dplyr::filter(GeneSymbol != "") %>%
  dplyr::arrange(Sample, FusionName) %>%
  # Retain only distinct rows
  dplyr::distinct() %>%
  group_by(Sample,GeneSymbol) %>% 
  dplyr::summarise(Gene.ct = n()) %>% 
  dplyr::filter(Gene.ct>limitMultiFused) %>%
  mutate(note=paste0("multfused " ,limitMultiFused, " times per sample"))



```


```{r}

# filter QCGeneFiltered_recFusion to remove fusions found in more than 1 group 

recurrent_symbols <- fusion_recurrent5_per_sample$GeneSymbol

QCGeneFiltered_recFusionUniq<-QCGeneFiltered_recFusion %>%
  dplyr::filter(!FusionName %in% group.count$FusionName) %>%
  dplyr::filter(!Gene1A %in% recurrent_symbols |
                  !Gene2A %in% recurrent_symbols |
                  !Gene1B %in% recurrent_symbols | 
                  !Gene2B %in% recurrent_symbols) %>%
  unique()




```





```{r}

# filter for putative driver genes and mutifused genes per sample
putative_driver_annotated_fusions <- fusion_calls %>%
  dplyr::filter(!is.na(Gene1A_anno) | !is.na(Gene1B_anno) | !is.na(Gene2A_anno) | !is.na(Gene2B_anno)) %>%
  unique()
  
# merge putative annotated oncogenic and scavenged back non-oncogenic annotated, recurrent fusions
putative_driver_fusions<-rbind(QCGeneFiltered_recFusionUniq,putative_driver_annotated_fusions) %>%
  unique() %>% select (-broad_histology) %>%
  as.data.frame()

write.table(putative_driver_fusions,file.path(outputfolder,"pbta-fusion-putative-oncogenic.tsv"),sep="\t",quote=FALSE,row.names = FALSE)

```



