---
title: "Project specific filtering"
output: html_notebook
params:
  histology:
    label: "Clinical file" 
    value: data/pbta-histologies.tsv
    input: file
  group:
    label: "Grouping variable"
    value: broad_histology
    input: string
  dataStrandedSTARFUSION:
    label: "Input filtered fusion dataframe"
    value: scratch/starfusionStranded_QC_expression_filtered_GTExComparison_annotated.RDS
    input: file
  dataStrandedARRIBA:
    label: "Input filtered fusion dataframe"
    value: scratch/arribaStranded_QC_expression_filtered_GTExComparison_annotated.RDS
    input: file
  dataPolyaSTARFUSION:
    label: "Input filtered fusion dataframe"
    value: scratch/starfusionPolya_QC_expression_filtered_GTExComparison_annotated.RDS
    input: file
  dataPolyaARRIBA:
    label: "Input filtered fusion dataframe"
    value: scratch/arribaPolya_QC_expression_filtered_GTExComparison_annotated.RDS
    input: file    
  numCaller:
    label: "Least Number of callers to have called fusion"
    value: 2
    input: integer
  numSample:
    label: "Least Number of samples to have fusion per group"
    value: 2
    input: integer
  numGroup:
    label: "Max number of groups found in"
    value: 5
    input: integer   
  expressionMatrix:
    label: "Expression matrix for the sample cohort"
    value: data/pbta-gene-expression-rsem-fpkm.stranded.rds
    input: file
    
---

K S Gaonkar

Filtered Fusions:
1. In-frame fusions is called in atleast 2 samples per histology OR
2. In-frame fusions is called in atleast 2 callers 
AND
Filtered-fusions found in more than 5 histology OR
Filtered-fusion doesn't have multi-fused gene (more than 5 times in sample)

Putative Driver:
Filtering for general cancer specific genes
Fusions with genes in either onco

This notebook assumes you are in OpenPBTA-analysis project folder structure.


```{r}

inputfolder<-file.path("..", "..")
setwd(inputfolder)
#load required packages
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("tidyr"))
suppressPackageStartupMessages(library("purrr"))
suppressPackageStartupMessages(library("reshape2"))

#read filtFusion files
arribaStrandedQCGeneFiltered_filtFusion<-readRDS(params$dataStrandedARRIBA)
starfusionStrandedQCGeneFiltered_filtFusion<-readRDS(params$dataStrandedSTARFUSION)
arribaPolyaQCGeneFiltered_filtFusion<-readRDS(params$dataPolyaARRIBA)
starfusionPolyaQCGeneFiltered_filtFusion<-readRDS(params$dataPolyaSTARFUSION)

QCGeneFiltered_filtFusion<-rbind(starfusionStrandedQCGeneFiltered_filtFusion,arribaStrandedQCGeneFiltered_filtFusion,starfusionPolyaQCGeneFiltered_filtFusion,arribaPolyaQCGeneFiltered_filtFusion)

# subset for recurrent fusion detection and multifused genes QC
fusion_calls<-unique(QCGeneFiltered_filtFusion[,c("FusionName","Sample","Caller","Fusion_Type","Gene1A","Gene2A","Gene1B","Gene2B")])

# get grouping column id
group<-params$group

# merge with histology file
clinical<-read.delim(params$histology,stringsAsFactors = FALSE)
fusion_calls<-merge(fusion_calls,clinical,by.x="Sample",by.y="Kids_First_Biospecimen_ID")

# Least number of callers
numCaller<-params$numCaller

# Least number of samples per group
numSample<-params$numSample

# Max number of groups 
numGroup<-params$numGroup

# Expression matrix
expressionMatrix<-params$expressionMatrix

# External matrix
extExpressionMatrix<-params$extExpressionMatrix

print("Raw calls from STARfusion and Arriba for PBTA")
table(fusion_calls$Caller)



```


```{r}
# Gene fusion should be in-frame
# AND
#
# 1. Called by at least n callers
fusion_calls.summary <- fusion_calls %>%
  dplyr::filter(Fusion_Type != "other") %>%
  dplyr::select(FusionName, Sample, !!as.name(group),Caller,-Fusion_Type) %>%
  group_by(FusionName, Sample, !!as.name(group) ) %>%
  unique() %>%
  dplyr::mutate(Caller = toString(Caller), caller.count = n()) %>%
  dplyr::filter(caller.count >= numCaller) %>%
  unique() %>%
  mutate(note=paste0("Called by",numCaller, "callers")) %>%
  as.data.frame()

# OR
# 2. Found in at least n samples in each group
sample.count <- fusion_calls %>%
  dplyr::filter(Fusion_Type != "other") %>%
  dplyr::select(FusionName, Sample, !!as.name(group),-Fusion_Type) %>%
  unique() %>%
  group_by(FusionName, !!as.name(group)) %>%
  dplyr::mutate(sample.count = n(),Sample = toString(Sample)) %>%
  dplyr::filter(sample.count > numSample) %>%
  unique() %>%
  mutate(note=paste0("Found in atleast ",numSample, " samples in a group")) %>%
  as.data.frame()


```



```{r}
#filter QCGeneFiltered_filtFusion to keep recurrent fusions from above sample.count and fusion_calls.summary

QCGeneFiltered_recFusion<-QCGeneFiltered_filtFusion %>%
  filter(FusionName %in% unique(sample.count$FusionName,fusion_calls.summary$FusionName))

```





```{r}
# remove fusions that are in > 5 group
group.count <- fusion_calls %>%
  dplyr::select(FusionName, !!as.name(group)) %>%
  unique() %>%
  group_by(FusionName) %>%
  dplyr::mutate(group.ct = n(),Sample = toString(!!(as.name(group)))) %>%
  dplyr::filter(group.ct >5)


# remove multi-fused genes
fusion_recurrent5_per_sample <- fusion_calls  %>%
  # We want to keep track of the gene symbols for each sample-fusion pair
  dplyr::select(Sample, FusionName, Gene1A, Gene1B, Gene2A, Gene2B) %>%
  # We want a single column that contains the gene symbols
  tidyr::gather(Gene1A, Gene1B, Gene2A, Gene2B,
                key = gene_position, value = GeneSymbol) %>%
  # Remove columns without gene symbols
  dplyr::filter(GeneSymbol != "") %>%
  dplyr::arrange(Sample, FusionName) %>%
  # Retain only distinct rows
  dplyr::distinct() %>%
  group_by(Sample,GeneSymbol) %>% 
  dplyr::summarise(Gene.ct = n()) %>% 
  dplyr::filter(Gene.ct>5) %>%
  mutate(note="multfused 5 times per sample")

```


```{r}

outputfolder<-"results"
setwd(outputfolder)
# filter QCGeneFiltered_recFusion to remove fusions found in more than 1 group and mutifused genes per sample

QCGeneFiltered_recFusionUniq<-QCGeneFiltered_recFusion %>%
  filter(!FusionName %in% group.count$FusionName) %>%
  filter(!Gene1A %in% fusion_recurrent5_per_sample$GeneSymbol,
         !Gene2A %in% fusion_recurrent5_per_sample$GeneSymbol,
         !Gene1B %in% fusion_recurrent5_per_sample$GeneSymbol,
         !Gene2B %in% fusion_recurrent5_per_sample$GeneSymbol)


write.table(QCGeneFiltered_recFusionUniq,"FilteredFusion.tsv",sep="\t",quote=FALSE,row.names = FALSE)

```


```{r}

setwd(outputfolder)
# putative driver genes
putative_driver_fusions <- QCGeneFiltered_filtFusion %>% 
  dplyr::filter(!is.na(Gene1A_anno) | !is.na(Gene1B_anno) | !is.na(Gene2A_anno)| !is.na(Gene2B_anno))

head(putative_driver_fusions)

write.table(putative_driver_fusions,"PutativeDriverFusion.tsv",sep="\t",quote=FALSE,row.names = FALSE)


```








