---
title: "Identify samples with _TP53_ and _NF1_ mutations in the stranded RNA-seq dataset"
output: html_notebook
author: Jaclyn Taroni for ALSF CCDL
date: 2020
---

To run and fully test the `tp53_nf1_score` module in continuous integration, we must ensure that there are positive examples of _TP53_ and _NF1_ mutations.
Furthermore, we need to include the same samples in the RNA-seq data.
Because we skip running the `tp53_nf1_score` classifier and evaluation steps on the poly-A dataset in CI, we can focus on the stranded data.

```{r}
library(tidyverse)
```

```{r}
set.seed(2020)
```

The table prepped in `tp53_nf1_score` contains all of the information we need regarding positive examples.
The histologies file contains the information we need for identifying samples that are in the stranded RNA-seq dataset.

```{r}
mutations_table <- read_tsv(file.path("..", "tp53_nf1_score", "results", 
                                      "TP53_NF1_snv_alteration.tsv"))
histologies_df <- read_tsv(file.path("..", "..", "data", 
                                     "pbta-histologies.tsv"),
                           col_types = cols(
                             molecular_subtype = col_character()
                           ))
```

Get biospecimen identifiers for samples with mutations of interest.

```{r}
tp53_examples <- mutations_table %>%
  filter(Hugo_Symbol == "TP53") %>%
  pull(Tumor_Sample_Barcode)
nf1_examples <- mutations_table %>%
  filter(Hugo_Symbol == "NF1") %>%
  pull(Tumor_Sample_Barcode)
```

### _TP53_ positive examples

```{r}
# randomly select samples that also have stranded RNA-seq data
tp53_samples <- histologies_df %>%
  filter(Kids_First_Biospecimen_ID %in% tp53_examples) %>%
  inner_join(histologies_df %>%
               filter(RNA_library == "stranded",
                      sample_type == "Tumor") %>%
               select(sample_id)) %>%
  distinct()

tp53_samples <- sample(tp53_samples$sample_id, size = 3)
```

```{r}
histologies_df %>%
  filter(sample_id %in% tp53_samples,
         sample_type == "Tumor") %>%
  arrange(sample_id)
```

### _NF1_ positive examples

```{r}
# randomly select samples that also have stranded RNA-seq data
nf1_samples <- histologies_df %>%
  filter(Kids_First_Biospecimen_ID %in% nf1_examples) %>%
  inner_join(histologies_df %>%
               filter(RNA_library == "stranded",
                      sample_type == "Tumor") %>%
               select(sample_id)) %>%
  distinct()

nf1_samples <- sample(nf1_samples$sample_id, size = 3)
```

```{r}
histologies_df %>%
  filter(sample_id %in% nf1_samples,
         sample_type == "Tumor") %>%
  arrange(sample_id)
```

### Session Info

```{r}
sessionInfo()
```

