---
title: "Subset MAF files: proof of concept"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: J. Taroni for ALSF CCDL
date: 2019
---

Does the MAF subsetting strategy in `01-create_subset_files` work?

## Setup

#### Packages and functions

```{r}
if (!("maftools" %in% installed.packages())) {
  BiocManager::install("maftools", update = FALSE)
}
```

`magrittr` pipe.

```{r}
`%>%` <- dplyr::`%>%`
```

#### Directories and files 

```{r}
data_dir <- file.path("..", "..", "data")
testing_data_dir <- file.path(data_dir, "testing")
mutect2_file <- "pbta-snv-mutect2.vep.maf.gz"
```

## Read in and compare

This is the file that was subset and written to file with a custom function,
`write_maf_file`.

```{r}
mutect2_subset <- maftools::read.maf(file.path(testing_data_dir, mutect2_file))
```

This is the original MuTect2 file.

```{r}
mutect2_full <- maftools::read.maf(file.path(data_dir, mutect2_file))
```

Select the tumor sample barcodes that are in the subset file.

```{r}

subset_tsb <- unique(mutect2_subset@data$Tumor_Sample_Barcode)
```

Use the tumor sample barcodes from the subset file to subset the full object.

```{r}
mutect2_post_subset <- maftools::subsetMaf(mutect2_full,
                                           tsb = subset_tsb)
```

Check if the object read in from the subset file and the MAF object that was
subset with `maftools::subsetMaf` are the same.

```{r}
all.equal(mutect2_subset, mutect2_post_subset)
```
