---
title: "Circos Plots Examples for Visualizing SV and CNV data"
output:   
  html_notebook: 
    toc: true
    toc_float: true
author: Candace Savonen for ALSF - CCDL
date: 2020
---

This notebook shows examples of how to use the circos_map_plot function for 
mapping data that corresponds to chromosomal coordinates.

### Usage

This notebook can be run via the command line from the top directory of the 
repository as follows:

```
Rscript -e "rmarkdown::render('analyses/chromosomal-instability/01b-visulization-cnv-sv.Rmd', 
                              clean = TRUE)"
```

### Set Up

```{r}
# Set seed so heatmaps turn out the same
set.seed(2020)

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

Read in the custom functions needed for this analysis. 

```{r}
source(file.path("util", "circos-plots.R"))
```

### Directories and Files

```{r}
# Path to data directory
data_dir <- file.path("..", "..", "data")
scratch_dir <- file.path("..", "..", "scratch")
sv_vcf_dir <- file.path(scratch_dir, "sv-vcf")

# Path to output directory
plots_dir <- "plots"

# Path to tumor type plots output directory
hist_plots_dir <- file.path(plots_dir, "tumor-type")

# Create the hist_plots_dir  if it does not exist
if (!dir.exists(hist_plots_dir)) {
  dir.create(hist_plots_dir, recursive = TRUE)
}
```

Run the SV process script if its output folder doesn't exist. 

```{r}
if (!dir.exists(sv_vcf_dir)) {
  source(file.path("..", "sv-analysis", "01-process-sv-file.R"))
}
```

### Read in data 

Set up metadata

```{r}
# Read in the metadata
metadata <- readr::read_tsv(file.path(data_dir, "pbta-histologies.tsv"))
```

Load in the previously formatted breakpoint data. 

```{r}
breaks_list <- readr::read_rds(file.path("breakpoint-data", "breaks_lists.RDS"))
```

Read in the CNV data. 

```{r}
cnv_df <- data.table::fread(
  file.path(data_dir, "pbta-cnv-cnvkit.seg.gz"), data.table = FALSE) 
```

Read in the formatted SV data. 

```{r}
# Get the file names for the 
sv_files <- list.files(file.path(scratch_dir, "sv-vcf"), 
                       pattern = "_withoutYandM.tsv", 
                       full.names = TRUE)

# Read in each of the sample's files
sv_data <- lapply(sv_files, function(sv_file){
                  suppressWarnings(readr::read_tsv(sv_file, 
                                  # Set default as character
                                  col_types = readr::cols(.default = "c")))
})

# Name the dataset by their biospecimen ID
names(sv_data) <- gsub("_withoutYandM.tsv", "", stringr::word(sv_files, sep = "/", 5))

# Collapse into one data.frame
sv_df <- dplyr::bind_rows(sv_data, .id = "biospecimen_id")
```

Make a translocation data.frame. 

```{r}
transloc_df <- sv_df %>% 
  dplyr::filter(SVtype == "TRA") %>% 
  dplyr::mutate(start1 = as.numeric(pos1), 
                start2 = as.numeric(pos2), 
                end2 = gsub("\\[|\\]|G|C|T|A", "", alt1), 
                end1 = gsub("\\[|\\]|G|C|T|A", "", alt2), 
                end2 = as.numeric(stringr::word(end2, sep = ":", 2)), 
                end1 = as.numeric(stringr::word(end1, sep = ":", 2)))
```

## Example uses of `circos_map_plot`

**Example 1** making a Circos plot that is color-coded circos by a factor.

```{r}
# Let's make a variable that specifies whether the data is a deletion or amplication. 
cnv_df <- cnv_df %>% 
  dplyr::mutate(direction = dplyr::case_when(seg.mean < 0 ~ "del", 
                                             seg.mean > 0 ~ "amp", 
                                             seg.mean == 0 ~"neutral"), 
                direction = as.factor(direction))
```

Create a Circos plot for a single sample.
For that same sample, let's add the translocations. 

```{r}
circos_map_plot(df = cnv_df, 
                add_track = FALSE,
                samples_col = "ID",
                sample_names = "BS_007JTNB8",
                chr_col = "chrom",
                start_col = "loc.start",
                end_col =  "loc.end",
                y_val = "copy.num",
                track_height = .15,
                type = "rect",
                colour_key = "direction",
                palette = "YlGnBu") 

# Now add the translocation bit. 
circos_map_transloc(transloc_df, 
                    add_track = TRUE, # We change this to true to add on to our already existing plot
                    sample_names = "BS_007JTNB8",
                    samples_col = "biospecimen_id",
                    chr_col_1 = "chrom1",
                    chr_col_2 = "chrom2",
                    start_col_1 = "start1",
                    start_col_2 = "start2",
                    end_col_1 =  "end1",
                    end_col_2 =  "end2")
```

### Session Info

```{r}
sessionInfo()
```

