---
title: "Chromosomal Instability: Breakpoints"
output:   
  html_notebook: 
    toc: true
    toc_float: true
author: Candace Savonen for ALSF - CCDL
date: 2020
---

This analysis evaluates chromosomal instability by using breakpoint SV and CNV 
data that was formatted co-localized by individual samples in 
`01-localization-of-breakpoints.Rmd`. 
This notebook returns chromosomal break heatmaps in the `plots` directory.

### Usage

This notebook can be run via the command line from the top directory of the 
repository as follows:

```
Rscript -e "rmarkdown::render('analyses/chromosomal-instability/02a-plot-chr-instability-by-sample.Rmd', 
                              clean = TRUE)"
```

### Set Up

```{r}
# Set seed so heatmaps turn out the same
set.seed(2020)

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

### Directories and Files

```{r}
# Path to data directory
data_dir <- file.path("..", "..", "data")
scratch_dir <- file.path("..", "..", "scratch")

# Path to output directory
plots_dir <- "plots"

# Create the hist_plots_dir  if it does not exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}
```

### Read in data 

Set up metadata

```{r}
# Read in the metadata
metadata <- readr::read_tsv(file.path(data_dir, "pbta-histologies.tsv"))
```

Load in the previously localized breakpoint data. 

```{r}
binned_counts_files <- list.files("breakpoint-data", 
                            pattern = "_binned_counts.tsv", 
                            full.names = TRUE)

# Read in each dataset
binned_counts <- lapply(binned_counts_files, readr::read_tsv)

# Name them
names(binned_counts) <- gsub("breakpoint-data/|_binned_counts.tsv", 
                             "", 
                             binned_counts_files)
```

Extract chromosome labels and make an alternating color key for them. 

```{r}
# Extract chromosome labels
chr_labels <- paste0("chr", binned_counts[[1]]$chr)

# Make chromosome labeling `HeatmapAnnotation` object.
chr_labels <- as.factor(chr_labels)

# Make a key for assigning alternating colors to the chromosomes
chr_colors <- rep(c("grey", "lightblue"), length.out = length(unique(chr_labels)))
names(chr_colors) <- unique(chr_labels)
```

### Set up for making heatmaps of the breakpoints

Given the `GenomicRanges` objects for each sample, create a combined plot for 
each. 

Make histology labeling `HeatmapAnnotation` object.

```{r}
# Get the histologies for the samples in this set and order them by histology
histologies <-
  data.frame(Kids_First_Biospecimen_ID = common_samples) %>%
  dplyr::inner_join(dplyr::select(metadata, Kids_First_Biospecimen_ID, short_histology)) %>%
  dplyr::mutate(short_histology = as.factor(short_histology)) %>%
  dplyr::filter(!is.na(short_histology)) %>% 
  tibble::column_to_rownames("Kids_First_Biospecimen_ID") 

# Get values that can be used to make colors equi distant hues away for the 
# number of histology groups we have
col_val <- seq(from = 0, to = 1, 
               length.out = length(unique(histologies$short_histology)))

# Translate into colors
col_key <- hsv(h = col_val, s = col_val, v = 1)

# Make this named based on histology
names(col_key) <- unique(histologies$short_histology)

# Make the same order as the data.frame
col_key <-  as.character(dplyr::recode(histologies$short_histology, !!!col_key))

# Make the names 
names(col_key) <- as.character(rownames(histologies))
```

Make a color function. 

```{r}
col_fun <- circlize::colorRamp2(
  c(0, .25, .5, 1, 3),
  c("#edf8fb", "#b2e2e2", "#66c2a4", "#2ca25f", "#006d2c")
)
```

Make a function for making the heatmaps. 

```{r}
breaks_heatmap <- function(data_name, chr_labels) {
  # A wrapper function for making a heatmap from the samples GenomicRanges list.
  #
  # Args:
  # data_name: a character string that matches a name in the list.

  # Returns:
  # A heatmap of the chromosomal breaks

  # Create the Heatmap annotation object
  chr_annot <- ComplexHeatmap::HeatmapAnnotation(
    df = data.frame(chr_labels), 
    col = list(chr_labels = c(chr_colors)),
    name = "",
    show_legend = FALSE,
    show_annotation_name = FALSE
  )
  # Create the Heatmap annotation object
  hist_annot <- ComplexHeatmap::HeatmapAnnotation(
    df = data.frame(histologies),
    col = list(histologies = col_key), 
    which = "row",
    show_annotation_name = FALSE
  )
  # Plot on a heatmap
  heatmap <- ComplexHeatmap::Heatmap(total_counts_df,
    col = col_fun,
    heatmap_legend_param = list(title = "Count of chr breaks"),
    cluster_columns = FALSE,
    cluster_rows = FALSE,
    show_column_names = FALSE,
    show_row_names = FALSE,
    bottom_annotation = chr_annot,
    left_annotation = hist_annot,
    na_col = "#f1f1f1"
  )
  # Return plot
  return(heatmap)
}
```

## Intersection of breaks heatmap

```{r}
intersection_of_heatmap <- breaks_heatmap(data_name = "intersection_of_breaks")

# Save plot as PNG
png(file.path(plots_dir, paste0("intersection_of_breaks_heatmap.png")),
  width = 1200, height = 900, units = "px"
)
intersection_of_heatmap
dev.off()

# Print out here
intersection_of_heatmap
```

## CNV breaks heatmap

```{r}
cnv_heatmap <- breaks_heatmap(data_name = "cnv_breaks")

# Save plot as PNG
png(file.path(plots_dir, paste0("cnv_breaks_heatmap.png")),
  width = 1200, height = 900, units = "px"
)
cnv_heatmap
dev.off()

# Print out here
cnv_heatmap
```

## SV breaks heatmap

```{r}
sv_heatmap <- breaks_heatmap(data_name = "sv_breaks")

# Save plot as PNG
png(file.path(plots_dir, paste0("sv_breaks_heatmap.png")),
  width = 1200, height = 900, units = "px"
)
sv_heatmap
dev.off()

# Print out here
sv_heatmap
```

### Session Info

```{r}
sessionInfo()
```

