---
title: "Chromosomal Instability: Heatmaps"
output:   
  html_notebook: 
    toc: true
    toc_float: true
author: Candace Savonen for ALSF - CCDL
date: 2020
---

This analysis evaluates chromosomal instability by using binned breakpoint counts 
for SV and CNVdata that was formatted co-localized by individual samples in 
`01-localization-of-breakpoints.Rmd`. 
This notebook returns chromosomal break heatmaps in the `plots` directory.

### Usage

This notebook can be run via the command line from the top directory of the 
repository as follows:

```
Rscript -e "rmarkdown::render('analyses/chromosomal-instability/02a-plot-chr-instability-heatmap.Rmd', 
                              clean = TRUE)"
```

### Set Up

```{r}
# Set seed so heatmaps turn out the same
set.seed(2020)

# Magrittr pipe
`%>%` <- dplyr::`%>%`

# Import standard color palette for project
source(file.path("..", "..", "scripts", "color_palettes.R"))
```

### Directories and Files

```{r}
# Path to data directory
data_dir <- file.path("..", "..", "data")
scratch_dir <- file.path("..", "..", "scratch")

# Path to output directory
plots_dir <- "plots"

# Create the hist_plots_dir  if it does not exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}
```

Here's all the input files we will need: 

```{r}
metadata_file <- file.path(data_dir, "pbta-histologies.tsv")
binned_counts_files <- list.files("breakpoint-data",
  pattern = "_binned_counts.tsv",
  full.names = TRUE
)
```

Output files:

```{r}
intersection_heatmap_file <- file.path(plots_dir, "intersection_breaks_heatmap.pdf")
cnv_heatmap_file <- file.path(plots_dir, "cnv_breaks_heatmap.pdf")
sv_heatmap_file <- file.path(plots_dir, "sv_breaks_heatmap.pdf")
```

Make a special function for making the heatmaps. 

```{r}
breaks_heatmap <- function(binned_counts_df,
                           chrs,
                           histologies,
                           chr_colors,
                           histologies_colors,
                           col_fun) {
  # A wrapper function for making a heatmap from the samples GenomicRanges list.
  #
  # Args:
  # binned_counts_df: a data.frame with the binned counts for each sample must 
  #                   have sample IDs in the row_ids and that the name of the 
  #                   histology column must match the histologies_colors vector
  #                   name
  # chrs: The chromosomes for each bin key.
  # histologies: The histologies to biospecimen's key.
  # chr_colors: A named vector to be used for coloring chromosomes.
  # histologies_colors: A named vector to be used for coloring histologies.
  # col_fun: a color key for the heatmap itself. Provided as a function.
  #
  # Returns:
  # A heatmap of the chromosomal breaks
  
  # Drop chr bin names
  binned_counts_mat <- binned_counts_df %>%
    dplyr::select(rownames(histologies)) %>%
    t()

  # Create the Heatmap annotation object
  chr_annot <- ComplexHeatmap::HeatmapAnnotation(
    df = data.frame(chrs),
    col = list(chrs = chr_colors),
    name = "",
    show_legend = FALSE,
    show_annotation_name = FALSE
  )
  # Create the Heatmap annotation object
  hist_annot <- ComplexHeatmap::HeatmapAnnotation(
    df = data.frame(histologies),
    col = list(short_histology = histologies_colors),
    which = "row",
    show_annotation_name = FALSE
  )
  # Plot on a heatmap
  heatmap <- ComplexHeatmap::Heatmap(binned_counts_mat,
    col = col_fun,
    heatmap_legend_param = list(title = "Count of chr breaks"),
    cluster_columns = FALSE,
    cluster_rows = FALSE,
    show_column_names = FALSE,
    show_row_names = FALSE,
    bottom_annotation = chr_annot,
    left_annotation = hist_annot,
    na_col = na_color
  )
  # Return plot
  return(heatmap)
}
```

### Read in data 

Set up metadata

```{r}
# Read in the metadata
metadata <- readr::read_tsv(metadata_file)
```

Load in the previously localized breakpoint data. 

```{r}
# Read in each dataset
binned_counts <- lapply(binned_counts_files, readr::read_tsv)

# Name them
names(binned_counts) <- gsub(
  "breakpoint-data/|_binned_counts.tsv",
  "",
  binned_counts_files
)
```

Extract chromosome labels and make an alternating color key for them. 

```{r}
# Extract chromosome labels
chrs <- paste0("chr", binned_counts[[1]]$chr_bin_names)

# Make chromosome labeling `HeatmapAnnotation` object.
chrs <- as.factor(chrs)

# Make a key for assigning alternating colors to the chromosomes
chr_colors <- rep(binary_col_palette, length.out = length(unique(chrs)))
names(chr_colors) <- unique(chrs)
```

Get sample names. 

```{r}
common_samples <- grep("chr_bin_names",
  colnames(binned_counts[[1]]),
  invert = TRUE,
  value = TRUE
)
```

### Set up for making heatmaps of the breakpoints

Given the `GenomicRanges` objects for each sample, create a combined plot for 
each. 

Make histology labeling `HeatmapAnnotation` object.

```{r}
# Get the histologies for the samples in this set and order them by histology
histologies <-
  data.frame(Kids_First_Biospecimen_ID = common_samples) %>%
  dplyr::inner_join(histology_col_by_sample) %>%
  dplyr::arrange(short_histology) %>% 
  tibble::column_to_rownames("Kids_First_Biospecimen_ID")

# Make color key specific to these samples
histologies_color_key_filtered <- histologies$sample_color
names(histologies_color_key_filtered) <- histologies$short_histology

# Drop this column so ComplexHeatmap isn't tempted to plot it
histologies <- dplyr::select(histologies, -sample_color)
```

Make a color function. 

```{r}
col_fun <- circlize::colorRamp2(
  c(0, .125, .25, .3, .5, 1, 1.5, 2, 3),
  gradient_col_palette
)
```

## Intersection of breaks heatmap

```{r}
intersection_of_heatmap <- breaks_heatmap(
  binned_counts_df = binned_counts$intersection_of_breaks,
  chrs,
  histologies,
  chr_colors,
  histologies_color_key_filtered,
  col_fun
)
# Print out here
intersection_of_heatmap

# Save plot as PNG
pdf(intersection_heatmap_file)
intersection_of_heatmap
dev.off()
```

## CNV breaks heatmap

```{r}
cnv_heatmap <- breaks_heatmap(
  binned_counts_df = binned_counts$cnv_breaks,
  chrs,
  histologies,
  chr_colors,
  histologies_color_key_filtered,
  col_fun
)
# Print out here
cnv_heatmap

# Save plot as PDF
pdf(cnv_heatmap_file)
cnv_heatmap
dev.off()
```

## SV breaks heatmap

```{r}
sv_heatmap <- breaks_heatmap(
  binned_counts_df = binned_counts$sv_breaks,
  chrs,
  histologies,
  chr_colors,
  histologies_color_key_filtered,
  col_fun
)
# Print out here
sv_heatmap

# Save plot as PDF
pdf(sv_heatmap_file)
sv_heatmap
dev.off()
```

### Session Info

```{r}
sessionInfo()
```

