---
title: "Repeated sample analysis"
output: 
  html_notebook:
    toc: true
    toc_float: true
---

## Purpose

There are many specimens in the full dataset that are repeated samples from the same participants.
This workbook is a quick summary of those samples, so that we can better select a unique set of specimens for downstream analysis. 
This is particularly important for analyses such as mutation co-occurence, where repeated sampling will inflate apparent co-ocurrence measures.


```{r setup}
# load libraries
library(dplyr)

# source sample selection funciton
source("independent-samples.R")

set.seed(201910)
```


```{r load files}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
sample_df <- readr::read_tsv(file.path(root_dir, "data", "pbta-histologies.tsv"))
```

## WGS and WXS samples
Sort to just sequencing samples of tumors
```{r}
seq_samples <- sample_df %>%
  filter(experimental_strategy %in% c("WGS", "WXS"), 
         sample_type == "Tumor")
```

How many WGS, how many WXS, summarize #s of samples and participants by 
```{r}
seq_samples %>%
  group_by(experimental_strategy, composition, tumor_descriptor) %>%
  summarise(samples = length(unique(Kids_First_Biospecimen_ID)),
            participants = length(unique(Kids_First_Participant_ID)), 
  ) 
```

What if we reduce to the earliest sample for each participant?

```{r}
early_samples <- seq_samples %>%
  group_by(Kids_First_Participant_ID) %>%
  summarize(age_at_diagnosis_days = min(age_at_diagnosis_days)) %>%
  left_join(seq_samples) %>%
  pull(Kids_First_Biospecimen_ID) %>%
  unique()

seq_samples %>% 
  filter(Kids_First_Biospecimen_ID %in% early_samples) %>%
  group_by(experimental_strategy, composition, tumor_descriptor) %>%
  summarise(samples = length(unique(Kids_First_Biospecimen_ID)),
            participants = length(unique(Kids_First_Participant_ID)), 
  )
```

There are still many samples for which there are more samples than participants, indicating duplicates, even among the earliest samples.


## Derived cell lines

Are there any samples for which we only have derived cell lines?

```{r}
derived_participants <- seq_samples %>%
  filter(composition == "Derived Cell Line") %>%
  pull(Kids_First_Participant_ID) %>%
  unique() 

non_derived_participants <- seq_samples %>% 
  filter(composition != "Derived Cell Line") %>%
  pull(Kids_First_Participant_ID) %>%
  unique() 

notissue_participants  <- derived_participants[!(derived_participants %in% non_derived_participants)]
length(notissue_participants)

```

Looks like there is a cell line sample with no corresponding tumor.

```{r}
seq_samples %>%
  filter(Kids_First_Participant_ID %in% notissue_participants)
```

Since this is only one sample it seems like the best move will be to remove all derived cell lines.

## Unlabeled composition samples

9 samples are listed as "Not Reported" for `composition`. 
What are these, and do we have correponding samples for which composition is known?

```{r}
notreported_participants <- seq_samples %>%
  filter(composition == "Not Reported") %>%
  pull(Kids_First_Participant_ID) %>%
  unique() 

seq_samples %>% 
  filter(Kids_First_Participant_ID %in% notreported_participants)
```

In one case (`PT_N8W26H19`), we do have Solid Tissue samples, but not for the others.
Should these samples be included in further analysis?

## WGS vs WXS
```{r}
WXS_participants <- seq_samples %>%
  filter(experimental_strategy == "WXS") %>%
  pull(Kids_First_Participant_ID) %>%
  unique() 

WGS_participants <- seq_samples %>%
  filter(experimental_strategy == "WGS") %>%
  pull(Kids_First_Participant_ID) %>%
  unique() 

No_WGS <- WXS_participants[!(WXS_participants %in% WGS_participants)]

length(No_WGS)
```

So there are currently 16 samples with missing WGS data, but that should change in a future release.

Just to check, is this still true when we restrict to early samples?
```{r}
WXS_participants <- seq_samples %>%
  filter(experimental_strategy == "WXS", 
         Kids_First_Biospecimen_ID %in% early_samples) %>%
  pull(Kids_First_Participant_ID) %>%
  unique() 

WGS_participants <- seq_samples %>%
  filter(experimental_strategy == "WGS", 
         Kids_First_Biospecimen_ID %in% early_samples) %>%
  pull(Kids_First_Participant_ID) %>%
  unique() 

No_WGS <- WXS_participants[!(WXS_participants %in% WGS_participants)]

length(No_WGS)
```

No change. For now, we will look only at WGS samples, but this could be adjusted in future analysis.



## Summarize repeated samples

```{r}
repeat_samples <- seq_samples %>% 
  dplyr::filter(Kids_First_Biospecimen_ID %in% early_samples, 
                experimental_strategy == "WGS") %>%
  dplyr::filter(composition != "Derived Cell Line") %>%
  dplyr::filter(Kids_First_Participant_ID %in% 
                  Kids_First_Participant_ID[duplicated(Kids_First_Participant_ID)])

repeat_samples %>% 
  group_by(Kids_First_Participant_ID) %>%
  summarise(samples = paste(Kids_First_Biospecimen_ID, collapse = ", "), 
         descriptors = paste(sort(unique(tumor_descriptor)),  collapse = ", "),
         age_diagnosis = paste(sort(unique(age_at_diagnosis_days)), collapse = ", "),
         experimental_stragegy = paste(sort(unique(experimental_strategy)), collapse = ", ")
  )
```


## Select independent samples

Using the information described above, I created a function `independent_samples()`
to select the earliest biospecimen from each participant, or in the case of more 
than one with the same date, randomly select among the available samples.

An example of this function is below, applied to the WGS Solid Tissue tumor samples.
Note that in this case, only tumors designated as primary will be included, 
though this can be adjusted with `tumor_types = "prefer_primary"` or 
`tumor_types = "any"` if we want to be less presecriptive. As of v5, primary tumors
are defined as those designated "Initial CNS Tumor" or "Diagnosis" in the
`tumor_descriptor` field.

In the accompanying script, I will generate both `primary_only` and `primary_plus` 
sets using the `"primary"` and `"prefer_primary"` resepectively.



```{r}
independents <- sample_df %>%
  filter(experimental_strategy == "WGS",
         sample_type == "Tumor", 
         composition == "Solid Tissue"
         ) %>%
  independent_samples(tumor_types = "primary")

nrow(independents)
```

Just for reference, lets see what tumor types we end up with in this set:

```{r}
sample_df %>% 
  filter(Kids_First_Biospecimen_ID %in% independents$Kids_First_Biospecimen_ID) %>%
  group_by(broad_histology) %>%
  summarize(n = n())
```



### Session info

```{r}
sessionInfo()
```
