---
title: "GISTIC entire PBTA cohort versus individual histology comparison of results"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2020
---

This notebook will look at the files in the `run-gistic/results` directory and compare those for the entire cohort to the three largest histology groups, namely, LGAT, HGAT and medulloblastoma.

The purpose is to identify, if any, disagreement between GISTIC results for the entire cohort versus the three individual histologies.

## Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

```
Rscript -e "rmarkdown::render('analyses/compare-gistic/01-GISTIC-cohort-vs-histology-comparison.Rmd', clean = TRUE)"
```

## Set up

```{r}
library(tidyverse)
```

### Files and Directories

```{r}
# Path to input directory
data_dir <- file.path("..", "run-gistic", "results")

# Path to output directory
plots_dir <- "plots"

# Create the plots_dir if it does not exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}
```

### Functions

```{r}
# Source `generate-multipanel-plot-functions.R` script
source(file.path("..",
                 "..",
                 "analyses",
                 "transcriptomic-dimension-reduction",
                 "util",
                 "generate-multipanel-plot-functions.R"))

# Source this module's custom functions R script
source(file.path("util", "GISTIC-comparison-functions.R"))
```

## Compare "scores.gistic" files

### Unzip GISTIC results folders

```{r}
# Unzip and set up GISTIC results folders and files.
cohort_gistic_dir <- file.path(data_dir, "pbta-cnv-consensus-gistic")
cohort_gistic_zip <- file.path(data_dir, "pbta-cnv-consensus-gistic.zip")

if (!dir.exists(cohort_gistic_dir)) {
  unzip(cohort_gistic_zip, 
        exdir = data_dir)
}

lgat_gistic_dir <- file.path(data_dir, "pbta-cnv-consensus-lgat-gistic")
lgat_gistic_zip <- file.path(data_dir, "pbta-cnv-consensus-lgat-gistic.zip")

if (!dir.exists(lgat_gistic_dir)) {
  unzip(lgat_gistic_zip, 
        exdir = data_dir)
}

hgat_gistic_dir <- file.path(data_dir, "pbta-cnv-consensus-hgat-gistic")
hgat_gistic_zip <- file.path(data_dir, "pbta-cnv-consensus-hgat-gistic.zip")

if (!dir.exists(hgat_gistic_dir)) {
  unzip(hgat_gistic_zip, 
        exdir = data_dir)
}

medulloblastoma_gistic_dir <- file.path(data_dir, "pbta-cnv-consensus-medulloblastoma-gistic")
medulloblastoma_gistic_zip <- file.path(data_dir, "pbta-cnv-consensus-medulloblastoma-gistic.zip")

if (!dir.exists(medulloblastoma_gistic_dir)) {
  unzip(medulloblastoma_gistic_zip, 
        exdir = data_dir)
}

```

### Run `plot_gistic_scores` for each file

```{r}
# Define file paths to `scores.gistic` files and run `plot_gistic_scores`
# function for each file
cohort_scores <-
  file.path(cohort_gistic_dir, "scores.gistic")
cohort_scores_plot <- plot_gistic_scores(cohort_scores)

lgat_scores <-
  file.path(lgat_gistic_dir, "scores.gistic")
lgat_scores_plot <- plot_gistic_scores(lgat_scores)

hgat_scores <-
  file.path(hgat_gistic_dir, "scores.gistic")
hgat_scores_plot <- plot_gistic_scores(hgat_scores)

medulloblastoma_scores <-
  file.path(medulloblastoma_gistic_dir, "scores.gistic")
medulloblastoma_scores_plot <-
  plot_gistic_scores(medulloblastoma_scores)
```

### Generate multipanel plots 

```{r fig.height = 10, fig.width = 24}
lgat_plot_list <- list(lgat_scores_plot, cohort_scores_plot)
generate_multipanel_plot(plot_list = lgat_plot_list,
                         plot_title = "LGAT scores versus entire cohort",
                         output_directory = plots_dir,
                         output_filename = "lgat_gistic_scores_multipanel_plot.png")

hgat_plot_list <- list(hgat_scores_plot, cohort_scores_plot)
generate_multipanel_plot(plot_list = hgat_plot_list,
                         plot_title = "HGAT scores versus entire cohort",
                         output_directory = plots_dir,
                         output_filename = "hgat_gistic_scores_multipanel_plot.png")

medulloblastoma_plot_list <- list(medulloblastoma_scores_plot, cohort_scores_plot)
generate_multipanel_plot(plot_list = medulloblastoma_plot_list,
                         plot_title = "Medulloblastoma scores versus entire cohort",
                         output_directory = plots_dir,
                         output_filename = "medulloblastoma_gistic_scores_multipanel_plot.png")
```

## Compare `amp_genes.conf_90.txt` and `del_genes.conf_90.txt` files

### Run `plot_genes_venn_diagram_wrapper` on `del_genes.conf_90.txt` files

```{r warning = FALSE, message = FALSE}
# Define GISTIC del genes file paths
cohort_del_genes <- file.path(cohort_gistic_dir, "del_genes.conf_90.txt")
lgat_del_genes <- file.path(lgat_gistic_dir, "del_genes.conf_90.txt")
hgat_del_genes <- file.path(hgat_gistic_dir, "del_genes.conf_90.txt")
medulloblastoma_del_genes <- file.path(medulloblastoma_gistic_dir, "del_genes.conf_90.txt")

# Run `plot_genes_venn_diagram_wrapper` function on GISTIC del genes files
del_plot_list <- plot_genes_venn_diagram_wrapper(cohort_del_genes,
                                                 lgat_del_genes,
                                                 hgat_del_genes,
                                                 medulloblastoma_del_genes)
```

## Session Info

```{r}
sessionInfo()
```
