---
title: "GISTIC tidy data preparation"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2020
---

This notebook will tidy the GISTIC data files required to compare the GISTIC's gene level calls with our gene level calls in `analyses/focal-cn-file-prepartion/results`. The required GISTIC files include `all_lesions.conf_90.txt`, `amp_genes.conf_90.txt`, and `del_genes.conf_90.txt` (these files are stored in `analyses/run-gistic/results`). 

The purpose of this notebook is to prepare GISTIC data in a format that can be compared to our focal CN files (both on a gene level and cytoband level). 

_Note_: This notebook will focus on preparing the GISTIC result files for the LGAT histology (the largest histology group in the cohort) and for the entire cohort. This decision was made for the purpose of developing this analysis in both a time efficient and informative manner.

## Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

```
Rscript -e "rmarkdown::render('analyses/compare-gistic/02-GISTIC-tidy-data-prep.Rmd', clean = TRUE)"
```

## Set up

```{r}
# Load in the tidyverse library of packages and functions
library(tidyverse)

# Source this module's custom functions R script
source(file.path("util", "GISTIC-comparison-functions.R"))
```

### Files and Directories

```{r}
# Path to input directory
data_dir <- file.path("..", "run-gistic", "results")

# Path to output directory
results_dir <- "results"

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

### Unzip GISTIC results

```{r}
# Unzip and set up GISTIC results folders and files.
cohort_gistic_dir <-
  file.path(data_dir, "pbta-cnv-consensus-gistic")
cohort_gistic_zip <-
  file.path(data_dir, "pbta-cnv-consensus-gistic.zip")

if (!dir.exists(cohort_gistic_dir)) {
  unzip(cohort_gistic_zip,
    exdir = data_dir
  )
}

lgat_gistic_dir <-
  file.path(data_dir, "pbta-cnv-consensus-lgat-gistic")
lgat_gistic_zip <-
  file.path(data_dir, "pbta-cnv-consensus-lgat-gistic.zip")

if (!dir.exists(lgat_gistic_dir)) {
  unzip(lgat_gistic_zip,
    exdir = data_dir
  )
}
```

### Define file paths for GISTIC `all_lesions.conf_90.txt` files

```{r}
cohort_all_lesions_file <- file.path(cohort_gistic_dir, "all_lesions.conf_90.txt")

lgat_all_lesions_file <- file.path(lgat_gistic_dir, "all_lesions.conf_90.txt")
```

### Define file path for GISTIC `amp_genes.conf_90.txt` and `del_genes.conf_90.txt` result files

```{r}
cohort_amp_genes_file <- file.path(cohort_gistic_dir, "amp_genes.conf_90.txt")

lgat_amp_genes_file <- file.path(lgat_gistic_dir, "amp_genes.conf_90.txt")

cohort_del_genes_file <- file.path(cohort_gistic_dir, "del_genes.conf_90.txt")

lgat_del_genes_file <- file.path(lgat_gistic_dir, "del_genes.conf_90.txt")
```

## Generate gene-level tidy tables

### Explore amp and del genes data 
This section is testing the logic that was updated in the `format_gistic_genes` function of the `GISTIC-comparison-functions.R` script.

```{r}
#### Amplification genes file -------------------------------------------------
 amp_genes_ragged_list <- data.table::fread(lgat_amp_genes_file,
    data.table = FALSE
  )

# Transpose data
amp_genes_transposed <- amp_genes_ragged_list %>%
  t()

# Make the header information in the first row the column names
colnames(amp_genes_transposed) <- amp_genes_transposed[1, ]

amp_genes_df <- amp_genes_transposed %>%
  # Make into a data.frame object
  as.data.frame(stringsAsFactors = FALSE) %>%
  # Remove the row with header information
  dplyr::filter(cytoband != "cytoband")

# Gather the gene data
amp_genes_gathered <- amp_genes_df %>%
  tidyr::gather("wide_peak",
                "gene",
                -cytoband,
                -`q value`,
                -`residual q value`,
                -`wide peak boundaries`) %>%
  dplyr::select(-wide_peak) %>%
  # Remove the blanks in the `gene` column that result from the gathering step
  dplyr::filter(gene != "")

# Display the amp genes gathered data.frame
amp_genes_gathered %>%
  arrange(`wide peak boundaries`)

#### Deletion genes file ------------------------------------------------------
del_genes_ragged_list <- data.table::fread(lgat_del_genes_file,
    data.table = FALSE
  )

# Transpose data
del_genes_transposed <- del_genes_ragged_list %>%
  t()

colnames(del_genes_transposed) <- del_genes_transposed[1, ]
del_genes_df <- del_genes_transposed %>%
  # Make into a data.frame object
  as.data.frame(stringsAsFactors = FALSE) %>%
  # Remove the row with header information
  dplyr::filter(cytoband != "cytoband")

del_genes_gathered <- del_genes_df %>%
  
  tidyr::gather("wide_peak",
                "gene",
                -cytoband,
                -`q value`,
                -`residual q value`,
                -`wide peak boundaries`) %>%
  dplyr::select(-wide_peak) %>%
  dplyr::distinct() %>%
  dplyr::filter(gene != "")

# Display the del genes data.frame
del_genes_gathered

# Combine amp and del genes data.frames
all_genes_gathered <- rbind(amp_genes_gathered, del_genes_gathered) %>%
  dplyr::select(-c("q value", "residual q value"))

# Display the data.frame combining the data from both the amp and del genes
# files
all_genes_gathered %>% 
  arrange(`wide peak boundaries`)
```

### Use GISTIC results for LGAT histology

```{r warning = FALSE, message = FALSE}
lgat_merged_calls_table <- prepare_gene_level_gistic(
  lgat_all_lesions_file,
  lgat_amp_genes_file,
  lgat_del_genes_file,
  output_filename = "lgat_gistic_gene_cn_status_table.tsv.gz"
)

lgat_merged_calls_table %>%
  dplyr::filter(detection_peak == "Deletion Peak 1")
```

### Use GISTIC results for entire cohort

```{r warning = FALSE, message = FALSE}
cohort_merged_calls_table <- prepare_gene_level_gistic(
  cohort_all_lesions_file,
  cohort_amp_genes_file,
  cohort_del_genes_file,
  output_filename = "cohort_gistic_gene_cn_status_table.tsv.gz"
)

cohort_merged_calls_table %>%
  dplyr::filter(detection_peak == "Deletion Peak  1")
```

## Generate cytoband-level tidy tables

### Use GISTIC results for LGAT histology

```{r warning = FALSE, message = FALSE}
lgat_cytoband_status_table <- prepare_cytoband_level_gistic(
  lgat_all_lesions_file,
  output_filename = "lgat_gistic_cytoband_cn_status_table.tsv.gz"
)

lgat_cytoband_status_table %>%
  dplyr::filter(cytoband == "3q29")
```

### Use GISTIC results for entire cohort

```{r warning = FALSE, message = FALSE}
cohort_cytoband_status_table <-
  prepare_cytoband_level_gistic(cohort_all_lesions_file,
                                output_filename = "cohort_gistic_cytoband_cn_status_table.tsv.gz")

cohort_cytoband_status_table %>%
  dplyr::filter(cytoband == "3q29")
```

## Session Info

```{r}
sessionInfo()
```

