---
title: "GISTIC copy number calls comparison at the gene level"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2020
---

This notebook will use the files prepared in `02-GISTIC-tidy-data-prep.Rmd` and stored in this module's `results` directory to compare the copy number calls made by GISTIC to the copy number calls we have prepared in the `analyses/focal-cn-file-preparation` module.

The purpose of this notebook is to compare GISTIC's copy number calls to our copy number calls at a gene level. 

## Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

```
Rscript -e "rmarkdown::render('analyses/compare-gistic/03-GISTIC-gene-level-cn-status-comparison.Rmd', clean = TRUE)"
```

## Set up

```{r}
library(tidyverse)
```

### Files and Directories

```{r}
# Path to results directory -- this is the input and output directory for this
# notebook
results_dir <- "results"

# Path to our focal CN files directory
focal_dir <- file.path("..", "focal-cn-file-preparation", "results")
```

#### Read in files

```{r}
# Read in the gene level files prepared in `02-GISTIC-tidy-data-prep.Rmd`
lgat_gene_level_calls <-
  readr::read_tsv(file.path(results_dir, "lgat_gistic_gene_cn_status_table.tsv.gz"))

cohort_gene_level_calls <-
  readr::read_tsv(file.path(results_dir, "cohort_gistic_gene_cn_status_table.tsv.gz"))

# Read in the consensus focal CN calls file
consensus_focal_cn_calls <-
  readr::read_tsv(file.path(focal_dir, "consensus_seg_annotated_cn_autosomes.tsv.gz"))
```

## Wrangle Data

### Function to wrangle calls data and tally the number of samples per gene for each call

```{r}
tally_calls <- function(gistic_calls_df,
                        focal_calls_df,
                        focal_or_gistic,
                        output_filename) {
  # Given a data.frame with calls data, tally the number of samples per gene
  # that fall under each call category and return the table of counts.
  #
  # Args:
  #   gistic_calls_df: the data.frame containing prepared GISTIC calls
  #   focal_calls_df: the data.frame containing prepared focal calls
  #   focal_or_gistic: the string indicating whether or not the function should
  #                    tally the GISTIC calls or the focal calls
  #   output_filename: the string to save the output table as
  #
  # Return:
  #   final_df: the data.frame containing the final counts of samples per gene
  #             for each call category

  # Merge the GISTIC and focal CN calls
  final_df <- gistic_calls_df %>%
    dplyr::rename(gistic_status = status) %>%
    left_join(focal_calls_df,
              by = c("Kids_First_Biospecimen_ID" = "biospecimen_id", "gene_symbol"))

  # Add loss, gain, amplification, and neutral columns to final table --
  # these columns will have the values "gistic" if the call came from gistic
  # and "focal" if the call came from our `focal-cn-file-preparation` module
  final_df <- final_df %>%
    mutate(
      loss = case_when(
        gistic_status == "loss" ~ "gistic",
        focal_status == "loss" ~ "focal",
        TRUE ~ "NA"
      ),
      gain = case_when(
        gistic_status == "gain" ~ "gistic",
        focal_status == "gain" ~ "focal",
        TRUE ~ "NA"
      ),
      amplification = case_when(
        gistic_status == "amplification" ~ "gistic",
        focal_status == "amplification" ~ "focal",
        TRUE ~ "NA"
      ),
      neutral = case_when(
        gistic_status == "neutral" ~ "gistic",
        focal_status == "neutral" ~ "focal",
        TRUE ~ "NA"
      )
    ) %>%
    select(-c("gistic_status", "focal_status"))
  
  final_loss_df <- final_df %>%
    group_by(gene_symbol, detection_peak, loss) %>%
    tally()
  
  final_gain_df <- final_df %>%
    group_by(gene_symbol, detection_peak, gain) %>%
    tally()
  
  final_amplification_df <- final_df %>%
    group_by(gene_symbol, detection_peak, amplification) %>%
    tally()
  
  final_neutral_df <- final_df %>%
    group_by(gene_symbol, detection_peak, neutral) %>%
    tally()
  
  if (focal_or_gistic == "focal") {
    # We do not join the neutral data here because we filter neutral calls out
    # of the focal data in the `focal-cn-file-preparation` module
    final_df <- final_loss_df %>%
      inner_join(final_gain_df, by = c("gene_symbol", "detection_peak")) %>%
      inner_join(final_amplification_df,
                 by = c("gene_symbol", "detection_peak")) %>%
      # Filter out non-focal calls
      filter(loss == "focal",
             gain == "focal",
             amplification == "focal") %>%
      mutate(loss = n.x,
             gain = n.y,
             amplification = n) %>%
      select(-c("n.x", "n.y", "n")) %>%
      distinct()
  } else {
    # We do not join the amplification data here because there are no
    # amplification calls in the GISTIC data
    final_df <- final_loss_df %>%
      inner_join(final_gain_df, by = c("gene_symbol", "detection_peak")) %>%
      inner_join(final_neutral_df,
                 by = c("gene_symbol", "detection_peak")) %>%
      # Filter out non-gistic calls
      filter(loss == "gistic",
             gain == "gistic",
             neutral == "gistic") %>%
      mutate(loss = n.x,
             gain = n.y,
             neutral = n) %>%
      select(-c("n.x", "n.y", "n"))
  }

  # Save final table
  write_tsv(final_df, file.path(results_dir, output_filename))

  # Display final table -- arrange by loss count
  final_df %>%
    arrange(detection_peak)
}
```

### Select relevant fields from consensus focal data

```{r}
# Keep only the relevant variables from the consensus focal CN file
consensus_focal_cn_calls <- consensus_focal_cn_calls %>%
  select(focal_status = status, gene_symbol, biospecimen_id) %>%
  distinct()
```

### Run `tally_calls` function on LGAT GISTIC calls

```{r}
# Run `tally_calls` function on LGAT GISTIC calls and focal consensus calls
# data -- specify that we want to tally the number of focal calls per gene and
# peak
lgat_focal <- tally_calls(
  gistic_calls_df = lgat_gene_level_calls,
  focal_calls_df = consensus_focal_cn_calls,
  focal_or_gistic = "focal",
  output_filename = "lgat_gene_level_focal_cn_counts.tsv"
)

# Run `tally_calls` function on LGAT GISTIC calls and focal consensus calls
# data -- specify that we want to tally the number of gistic calls per gene and
# peak
tally_calls(
  gistic_calls_df = lgat_gene_level_calls,
  focal_calls_df = consensus_focal_cn_calls,
  focal_or_gistic = "gistic",
  output_filename = "lgat_gene_level_gistic_cn_counts.tsv"
)
```

### Run `tally_calls` function on entire cohort GISTIC calls

```{r}
# Run `tally_calls` function on entire cohort GISTIC calls and focal consensus calls
# data -- specify that we want to tally the number of focal calls per gene and
# peak
tally_calls(
  gistic_calls_df = cohort_gene_level_calls,
  focal_calls_df = consensus_focal_cn_calls,
  focal_or_gistic = "focal",
  output_filename = "cohort_gene_level_focal_cn_counts.tsv"
)

# Run `tally_calls` function on entire cohort GISTIC calls and focal consensus calls
# data -- specify that we want to tally the number of gistic calls per gene and
# peak
tally_calls(
  gistic_calls_df = cohort_gene_level_calls,
  focal_calls_df = consensus_focal_cn_calls,
  focal_or_gistic = "gistic",
  output_filename = "cohort_gene_level_gistic_cn_counts.tsv"
)
```

## Session Info

```{r}
sessionInfo()
```


