---
title: "Tally copy number calls at the gene-level"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2020
params:
  is_ci: 0
---

This notebook will tally the number of samples with a `loss`, `gain`, or `neutral` status for an individual gene in two sets of files:

* Files prepared in `02-GISTIC-tidy-data-prep.Rmd` and stored in this module's `results` directory (referred to as "gistic" throughout)
* Files from `analyses/focal-cn-file-preparation/results` (referred to as "exon" throughout)

Note that we will prepare only files derived from the consensus SEG file.

## Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

```
Rscript -e "rmarkdown::render('analyses/compare-gistic/03-GISTIC-gene-level-cn-status-comparison.Rmd', clean = TRUE)"
```

## Set up

```{r}
library(tidyverse)
```

If this is running in CI, skip the LGAT steps because there will not be enough samples.

```{r}
if (params$is_ci == 0) {
  run_lgat <- TRUE
} else {
  run_lgat <- FALSE
}
```

### Functions

For tallying up copy number status

```{r}
get_gene_tally <- function(gene_status_df,
                           gene_mapping_df,
                           method,
                           biospecimens_included = NULL,
                           genes_included = NULL) {
  # This function takes two data.frames: 1) either an annotated seg file 
  # (method = "exon") from `focal-cn-file-preparation` or the gene status 
  # data.frame from prepare_gene_level_gistic (method = "gistic") and 2) the 
  # peak assignment output of prepare_gene_level_gistic and returns a tally of 
  # copy number status for individual genes
  # 
  # Args:
  #   gene_status_df: either an annotated seg file (set method = "exon") from 
  #                   `focal-cn-file-preparation` or the gene status data.frame 
  #                   from prepare_gene_level_gistic (set method = "gistic")
  #   gene_mapping_df: the peak assignment output of prepare_gene_level_gistic 
  #   method: "gistic" or "exon", see above
  #   biospecimens_included: a vector of biospecimen identifiers to be used
  #                          to filter gene_status_df; only used when
  #                          method = "exon"
  #   genes_included: a vector of gene symbols to be used to filter 
  #                   gene_status_df; only used when method = "exon"
  # 
  # Returns:
  #   gene_tally_df: a data.frame with the following columns
  #                  1. gene: gene symbol
  #                  2. gain: number of gains tallied
  #                  3. loss: number of losses tallied
  #                  4. neutral: number of copy neutral events tallied
  #                  5. detection_peak: all peaks a gene symbol mapped to 
  #                                     via GISTIC, comma-separated
  
  # Handles the case where someone writes GISTIC
  method <- tolower(method)
  # Handles unsupported methods
  if (!(method %in% c("gistic", "exon"))) {
    stop("method must be 'gistic' or 'exon'")
  }
  
  # Check if filtering arguments are used or not
  null_filtering_args <- c(is.null(biospecimens_included),
                           is.null(genes_included))
  
  # If we're using the exon method, we need the biospecimens_included and
  # genes_included arguments
  if ((method == "exon") & any(null_filtering_args)) {
    stop("The 'exon' method is intended to be used with the biospecimens_included and genes_included arguments")
  }
  
  # If the method is gistic and one of biospecimens_included or genes_included
  # was used, warn the user
  if ((method == "gistic") & !all(null_filtering_args)) {
    warning("biospecimen_included and genes_included arguments are only used when method = 'exon', they will be ignored")
  }
  
  # Filtering by biospecimen and gene are only necessary when method = "exon"
  if (method == "exon") {
     gene_status_df <- gene_status_df %>%
        filter(Kids_First_Biospecimen_ID %in% biospecimens_included,
                gene %in% genes_included)
  }
  
  gene_tally_df <- gene_status_df %>%
    # Remove any genes that are "nearest genes." We don't expect them to be *in*
    # the peak and therefore do not expect them to be comparable to the other
    # method that we used.
    filter(str_detect(gene, "\\[", negate = TRUE)) %>%
    # Count the number of times a copy number alteration occurs in an individual
    # gene
    group_by(gene, status) %>%
    tally() %>%
    spread(status, n) %>%
    # Any instances where a copy number alteration is NA are equivalent to a 
    # count of zero
    replace_na(list(loss = 0, 
                    gain = 0))
  
  # In the GISTIC case, missing values in the neutral column are the same as
  # zero counts
  if (method == "gistic") {
    gene_tally_df <- gene_tally_df %>%
      replace_na(list(neutral = 0))
  }
  
  # Because neutral calls are not included in the annotated consensus file,
  # we can assume the number of neutral calls is the number of samples included 
  # minus any alterations (gain + loss)
  if (method == "exon") {
    gene_tally_df <- gene_tally_df %>%
       mutate(neutral = length(biospecimens_included) - (gain + loss))
  }
  
  gene_tally_df %>%
    # Add detection peak information to this data.frame. Peaks can be 
    # overlapping in GISTIC and contain the same gene symbols, so we need to 
    # summarize the detection_peak column
    inner_join(gene_mapping_df %>%
                 select(gene,
                        detection_peak),
               by = "gene") %>%
    group_by(gene, gain, loss, neutral) %>%
    summarize(detection_peak = paste(detection_peak, collapse = ", ")) %>%
    ungroup()
  
}
```

<!-- TODO: one way to compare the different outcomes is to use a visualization
that demonstrates what proportion of samples have a particular status for an 
individual gene (a heatmap) and because of the large number of genes we can
make panels on a per detection peak basis.

For creating a heatmap that shows the proportion of samples with a copy number status for an individual gene

```{r eval=FALSE}
create_heatmap <- function(gene_tally_df,
                           detection_peak_filter,
                           plot_title = "EXON OVERLAPS",
                           high_color = "red") {
  # For a given detection peak in the output data.frame of get_gene_tally,
  # filter to only those genes and create a heatmap with cells that are filled
  # by the proportion of samples that have the corresponding status value.
  # 
  # Args:
  #   gene_tally_df: output of get_gene_tally
  #   detection_peak_filter: character value supplied to dplyr::filter to filter
  #                          to rows that are in a given detection peak
  #   high_color: the high value supplied to scale_fill_gradient; low = "white"
  #   
  # Returns:
  #   A heatmap (geom_tile) where the rows are genes and the columns are status
  
  gene_tally_df %>%
    filter(detection_peak == detection_peak_filter) %>%
    group_by(gene) %>%
    summarize(gain = gain / sum(gain, loss, neutral),
           loss = loss / sum(gain, loss, neutral),
           neutral = neutral / sum(gain, loss, neutral)) %>%
    reshape2::melt(variable.name = "status",
                   value.name = "proportion_of_samples") %>%
    ggplot(aes(status, gene)) +
    geom_tile(aes(fill = proportion_of_samples)) +  
    scale_fill_gradient(low = "white", high = high_color) +
    labs(title = plot_title)
}

plot_detection_peak <- function(exon_tally_df,
                                gistic_tally_df,
                                peak_filter) {
  # Make a panel (row) of heatmaps for a given detection peak
  # 
  # Arguments:
  #   exon_tally_df: output of get_gene_tally using `focal-cn-file-preparation`
  #                  results
  #   gistic_tally_df: output of get_gene_tally using prepare_gene_level_gistic
  #   peak_filter: will be supplied as detection_peak_filter to create_heatmap
  # 
  # Returns:
  #   Two heatmaps titled with peak_filter
  
  exon_heatmap <- create_heatmap(gene_tally_df = exon_tally_df,
                                 detection_peak_filter = peak_filter,
                                 plot_title = "EXON OVERLAPS",
                                 high_color = "red")
  gistic_heatmap <- create_heatmap(gene_tally_df = gistic_tally_df,
                                   detection_peak_filter = peak_filter,
                                   plot_title = "GISTIC",
                                   high_color = "navy")
  
  plot_row <- cowplot::plot_grid(exon_heatmap + 
                                   theme(legend.position = "none"),
                                 gistic_heatmap + 
                                   theme(legend.position = "none"))
  title_pane <- cowplot::ggdraw() + 
    cowplot::draw_label(
      peak_filter,
      fontface = 'bold',
      x = 0,
      hjust = 0
    ) 
  
  cowplot::plot_grid(title_pane, 
                     plot_row,
                     ncol = 1,
                     rel_heights = c(0.1, 1))
}
```
-->

### Files and Directories

```{r}
# Path to results directory -- this is the input and output directory for this
# notebook
results_dir <- "results"

# Comment out because heatmap functionality is commented out
# Plots
# plots_dir <- "plots"
# if (!dir.exists(plots_dir)) {
#  dir.create(plots_dir)
# }

# Path to our focal CN files directory
focal_dir <- file.path("..", "focal-cn-file-preparation", "results")
# Path to GISTIC results
gistic_dir <- file.path("..", "run-gistic", "results")
```

#### Input

```{r}
# Files related to the entire cohort
cohort_mapping_file <- file.path(results_dir, 
                                 "cohort_gistic_peak_assignment.tsv.gz")
cohort_gene_status <- file.path(results_dir, 
                                "cohort_gistic_gene_cn_status_table.tsv.gz")
cohort_seg_counts <- file.path(gistic_dir, 
                               "pbta-cnv-consensus-gistic",
                               "sample_seg_counts.txt")

# Files related to LGAT samples only
lgat_mapping_file <- file.path(results_dir, 
                               "lgat_gistic_peak_assignment.tsv.gz")
lgat_gene_status <- file.path(results_dir, 
                              "lgat_gistic_gene_cn_status_table.tsv.gz")
lgat_seg_counts <- file.path(gistic_dir, 
                             "pbta-cnv-consensus-lgat-gistic",
                             "sample_seg_counts.txt")

# File generated via the overlapping exons method 
annotated_cn_file <- file.path(focal_dir, 
                               "consensus_seg_annotated_cn_autosomes.tsv.gz")
```

#### Output

```{r}
cohort_exon_file <- file.path(results_dir,
                              "cohort_gene_level_focal_cn_counts.tsv")
cohort_gistic_file <- file.path(results_dir,
                                "cohort_gene_level_gistic_cn_counts.tsv")
lgat_exon_file <- file.path(results_dir,
                            "lgat_gene_level_focal_cn_counts.tsv")
lgat_gistic_file <- file.path(results_dir,
                              "lgat_gene_level_gistic_cn_counts.tsv")
```

## Read in files

```{r}
# Read in the gene level files prepared in `02-GISTIC-tidy-data-prep.Rmd`
cohort_gene_mapping <- read_tsv(cohort_mapping_file)
cohort_gene_level_calls <- read_tsv(cohort_gene_status)
lgat_gene_mapping <- read_tsv(lgat_mapping_file)
lgat_gene_level_calls <- read_tsv(lgat_gene_status)

# Get the biospecimen identifiers for samples that were included in the GISTIC
# run
all_ids_in_consensus <- read_tsv(cohort_seg_counts) %>%
  filter(included == "yes") %>%
  pull(`sample`)

lgat_ids_in_consensus <- read_tsv(lgat_seg_counts) %>%
  filter(included == "yes") %>%
  pull(`sample`)

# Read in the consensus focal CN calls file and rename gene and biospecimen
# columns
consensus_focal_cn_calls <- read_tsv(annotated_cn_file) %>%
  rename(gene = gene_symbol,
         Kids_First_Biospecimen_ID = biospecimen_id) %>%
  mutate(status = dplyr::recode(status, amplification = "gain"))
```

## Count copy number alterations

### Entire cohort

```{r}
cohort_gistic_tally <- get_gene_tally(gene_status_df = cohort_gene_level_calls,
                                      gene_mapping_df = cohort_gene_mapping,
                                      method = "gistic") %>%
  write_tsv(cohort_gistic_file)
cohort_exon_tally <- get_gene_tally(gene_status_df = consensus_focal_cn_calls,
                                    gene_mapping_df = cohort_gene_mapping,
                                    method = "exon",
                                    biospecimens_included = all_ids_in_consensus,
                                    genes_included = cohort_gistic_tally$gene) %>%
  write_tsv(cohort_exon_file)
```

### LGAT only

```{r}
# only run when params$is_ci == 0, the default
if (run_lgat) {
  lgat_gistic_tally <- get_gene_tally(gene_status_df = lgat_gene_level_calls,
                                      gene_mapping_df = lgat_gene_mapping,
                                      method = "gistic") %>%
    write_tsv(lgat_gistic_file)
  lgat_exon_tally <- get_gene_tally(gene_status_df = consensus_focal_cn_calls,
                                    gene_mapping_df = lgat_gene_mapping,
                                    method = "exon",
                                    biospecimens_included = lgat_ids_in_consensus,
                                    genes_included = lgat_gistic_tally$gene) %>%
    write_tsv(lgat_exon_file)
}
```

<!--This is using the heatmap code that is commented out above.

## Heatmap

First, only retain genes that are in the exon files.

```{r eval=FALSE}
cohort_gistic_tally <- cohort_gistic_tally %>%
  filter(gene %in% cohort_exon_tally$gene)
lgat_gistic_tally <- lgat_gistic_tally %>%
  filter(gene %in% lgat_exon_tally$gene)
```

### LGAT only

```{r eval=FALSE}
lgat_heatmaps <- 
  purrr::map(unique(lgat_exon_tally$detection_peak),
             ~ plot_detection_peak(exon_tally_df = lgat_exon_tally,
                                   gistic_tally_df = lgat_gistic_tally,
                                   peak_filter = .x))

lgat_hm_panels <- cowplot::plot_grid(plotlist = lgat_heatmaps,
                                     nrow = length(lgat_heatmaps))

ggsave(lgat_hm_file, plot = lgat_hm_panels, width = 8.5, height = 14)
```

-->

## Session Info

```{r}
sessionInfo()
```

