---
title: "ATRT Molecular Subtyping - Data Prep"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2019
---

This notebook addresses the issue of molecular subtyping ATRT samples. 

# Usage

This script is intended to be run via the command line from the top directory
of the repository as follows:

`Rscript 'analyses/molecular-subtyping-ATRT/ATRT-molecular-subtyping.R'`

# Set Up

```{r}
# Get `magrittr` pipe
`%>%` <- dplyr::`%>%`

# Install `flextable` package for displaying tables
if (!("flextable" %in% installed.packages())) {
  install.packages("flextable")
}
library(flextable)
```

## Directories and Files

```{r}
# Detect the ".git" folder -- this will in the project root directory.
# Use this as the root directory to ensure proper sourcing of functions no
# matter where this is called from
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

# File path to plots directory
plots_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-ATRT", "plots")

if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}

# File path to results directory
results_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-ATRT", "results")

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

# Read in metadata
metadata <-
  readr::read_tsv(file.path(root_dir, "data", "pbta-histologies.tsv"))

# Select wanted columns in metadata for merging and assign to a new object
select_metadata <- metadata %>%
  dplyr::select(sample_id, Kids_First_Participant_ID, Kids_First_Biospecimen_ID)

# Read in ssGSEA pathway information
ssGSEA <-
  as.data.frame(readr::read_rds(
    file.path(
      root_dir,
      "analyses",
      "ssgsea-hallmark",
      "results",
      "GeneSetExpressionMatrix.RDS"
    )
  ))

# Read in RNA expression data
stranded_expression <-
  readr::read_rds(file.path(
    root_dir,
    "data",
    "pbta-gene-expression-rsem-fpkm.stranded.rds"
  ))

# Read in focal CN data
## TODO: This section will be updated to read in focal CN data derived from 
##       copy number consensus calls.
cn_df <- readr::read_tsv(
  file.path(
    root_dir,
    "analyses",
    "focal-cn-file-preparation",
    "results",
    "controlfreec_annotated_cn_autosomes.tsv.gz"
  )
)

# Read in consensus mutation data
tmb_df <-
  data.table::fread(
    file.path(
      root_dir,
      "data",
      "snv-consensus_11122019",
      "consensus_mutation_tmb.tsv"
    )
  )
```

# Filter Data

## Filter metadata

```{r}
# Define regions of the brain (using Anatomy of the Brain figure found at
# https://www.ncbi.nlm.nih.gov/books/NBK65903/figure/CDR0000574573__205/)
supratentorial <-
  c(
    "Skull",
    "Thalamus",
    "Temporal Lobe",
    "Frontal Lobe",
    "Parietal Lobe",
    "Cerebrum",
    "Basal Ganglia",
    "Cranial Nerves NOS",
    "Basal Ganglia;Temporal Lobe",
    "Frontal Lobe;Parietal Lobe;Temporal Lobe",
    "Parietal Lobe;Temporal Lobe",
    "Frontal Lobe;Parietal Lobe",
    "Frontal Lobe;Meninges/Dura;Parietal Lobe;Spinal Cord- Cervical;Spinal Cord- Lumbar/Thecal Sac;Spinal Cord- Thoracic;Temporal Lobe;Ventricles"
    # Not sure how the line above should be broken up as `Frontal Lobe` is
    # supratentorial, while `Spinal Cord` is not
  )

infratentorial <-
  c(
    "Cerebellum/Posterior Fossa",
    "Brain Stem- Pons;Cerebellum/Posterior Fossa",
    "Cerebellum/Posterior Fossa;Other locations NOS",
    "Brain Stem",
    "Brain Stem- Midbrain/Tectum;Ventricles",
    "Cerebellum/Posterior Fossa;Ventricles",
    "Cerebellum/Posterior Fossa;Spinal Cord- Cervical;Spinal Cord- Lumbar/Thecal Sac;Spinal Cord- Thoracic",
    "Other locations NOS;Spinal Cord- Lumbar/Thecal Sac;Spinal Cord- Thoracic;Ventricles",
    "Other locations NOS;Spinal Cord- Lumbar/Thecal Sac;Spinal Cord- Thoracic;Temporal Lobe"
    # Not sure how the line above should be broken up as `Temporal Lobe` is
    # supratentorial, while the `Spinal Cord` is not
  )

# Filter metadata for `ATRT` and define `location_summary` based on values in
# `primary_site`
atrt_df <- metadata %>%
  dplyr::filter(short_histology == "ATRT",
                experimental_strategy == "RNA-Seq") %>%
  dplyr::mutate(
    location_summary = dplyr::case_when(
      primary_site %in% infratentorial ~ "infratentorial",
      primary_site %in% supratentorial ~ "supratentorial",
      TRUE ~ as.character(primary_site)
    )
  ) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::summarize(Kids_First_Biospecimen_ID = paste(sort(unique(
    Kids_First_Biospecimen_ID
  )),
  collapse = ", "),
  Kids_First_Participant_ID,
  location_summary,
  age_at_diagnosis_days,
  germline_sex_estimate, 
  primary_site) 

# Display `atrt_df`
atrt_df %>%
  head(n = 15) %>%
  regulartable() %>%
  fontsize(size = 12, part = "all")
```

## Filter RNA expression, CN, TMB, and ssGSEA data by merging with `sample_id`

### RNA expression data 

```{r}
# Define target overexpressed gene vectors
tyr_genes <- c("TYR", "MITF", "DCT", "VEGFA", "DNAH11", "SPEF1")
shh_genes <- c("MYCN", "GLI2", "CDK6", "ASCL1", "HES5/6", "DLL1/3")
myc_genes <- c("MYC", "HOTAIR", "HOX")

# Filter expression data for target overexpressed genes
stranded_expression_wide_df <- stranded_expression %>%
  as.data.frame() %>%
  dplyr::mutate(gene_id = gsub(".*\\_", "", gene_id)) %>% # Trim the gene ids to include only the gene symbol
  dplyr::filter(gene_id %in% tyr_genes | gene_id %in% shh_genes | gene_id %in% myc_genes) %>%
  tibble::column_to_rownames("gene_id")

# Log2 transformation
norm_expression <- log2(stranded_expression_wide_df + 1)

# normData mean and sd
norm_expression_means <- rowMeans(norm_expression, na.rm = TRUE)
norm_expression_sd <- apply(norm_expression, 1, sd, na.rm = TRUE)

# Subtract mean
expression_zscored <- sweep(stranded_expression_wide_df, 1, norm_expression_means, FUN = "-")

# Divide by SD remove NAs and Inf values from zscore for genes with 0 in normData
expression_zscored <- sweep(expression_zscored, 1, norm_expression_sd, FUN = "/") %>% 
  dplyr::na_if(Inf) %>% 
  na.omit()

# Transpose
long_stranded_expression <- t(expression_zscored)

# Merge metadata with expression data 
expression_metadata <- long_stranded_expression %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Kids_First_Biospecimen_ID") %>%
  dplyr::left_join(select_metadata, by = "Kids_First_Biospecimen_ID")

# Display `expression_metadata`
expression_metadata %>%
  head(n = 15) %>%
  regulartable() %>%
  fontsize(size = 12, part = "all")

# Join expression data with metadata filtered for `ATRT`
atrt_expression_df <- atrt_df %>%
  dplyr::left_join(expression_metadata,
                    by = "sample_id")
```

### CN data

```{r}
# Collapse target overexpressed gene vectors
tyr_genes <- paste(tyr_genes, collapse = "|")
shh_genes <- paste(shh_genes, collapse = "|")
myc_genes <- paste(myc_genes, collapse = "|")

# Join focal CN data with metadata
cn_metadata <- cn_df %>%
  dplyr::left_join(metadata,
                   by = c("biospecimen_id" = "Kids_First_Biospecimen_ID")) %>%
  dplyr::select(gene_symbol,
                sample_id,
                Kids_First_Participant_ID,
                biospecimen_id,
                status) %>%
  dplyr::filter(sample_id %in% atrt_expression_df$sample_id) %>%
  dplyr::mutate(
    SMARCB1_focal_status = dplyr::case_when(gene_symbol == "SMARCB1" ~ status,
                                            TRUE ~ "neutral"),
    SMARCA4_focal_status = dplyr::case_when(gene_symbol == "SMARCA4" ~ status,
                                            TRUE ~ "neutral"),
    Overexpressed_gene_sets = dplyr::case_when(
      stringr::str_detect(gene_symbol, tyr_genes) &
        status != "loss" ~ "TYR_genes",
      stringr::str_detect(gene_symbol, shh_genes) &
        status != "loss" ~ "SHH_genes",
      stringr::str_detect(gene_symbol, myc_genes) &
        status != "loss" ~ "MYC_genes",
      TRUE ~ "NA"
    )
  ) %>%
  dplyr::select(-gene_symbol) %>%
  dplyr::distinct()

#Display `cn_metadata`
cn_metadata %>%
  head(n = 15) %>%
  regulartable() %>%
  fontsize(size = 12, part = "all")

# Join ATRT expression data with focal CN data
atrt_expression_cn_df <- atrt_expression_df %>%
  dplyr::left_join(cn_metadata, by = "sample_id") %>%
  dplyr::select(-status)
```

### ssGSEA data

```{r}
# Calculate ssGSEA mean and sd
ssGSEA_means <- rowMeans(ssGSEA, na.rm = TRUE)
ssGSEA_sd <- apply(ssGSEA, 1, sd, na.rm = TRUE)

# Subtract mean
ssGSEA_zscored <- sweep(ssGSEA, 1, ssGSEA_means, FUN = "-")

# Divide by SD remove NAs and Inf values from zscore for genes with 0 in normData
ssGSEA_zscored <- sweep(ssGSEA_zscored, 1, ssGSEA_sd, FUN = "/") %>% 
  dplyr::na_if(Inf) %>% 
  na.omit()

# Transpose
transposed_ssGSEA <- t(ssGSEA_zscored)

# Select wanted pathways and merge metadata 
transposed_ssGSEA <- transposed_ssGSEA %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Kids_First_Biospecimen_ID") %>%
  dplyr::select(
    Kids_First_Biospecimen_ID,
    HALLMARK_MYC_TARGETS_V1,
    HALLMARK_MYC_TARGETS_V2,
    HALLMARK_NOTCH_SIGNALING
  ) %>%
  dplyr::left_join(select_metadata, by = "Kids_First_Biospecimen_ID")

#Display `transposed_ssGSEA`
transposed_ssGSEA %>%
  head(n = 15) %>%
  regulartable() %>%
  fontsize(size = 12, part = "all")

# Join ATRT expression and focal CN data with transposed ssGSEA data
atrt_expression_cn_df <- atrt_expression_cn_df %>%
  dplyr::left_join(transposed_ssGSEA,
                    by = "sample_id")
```

### Tumor mutation burden data

```{r}
# Join tumor mutuation data with metadata
tmb_df <- tmb_df %>%
  dplyr::select(Tumor_Sample_Barcode, tmb) %>%
  dplyr::inner_join(select_metadata,
                   by = c("Tumor_Sample_Barcode" = "Kids_First_Biospecimen_ID"))

#Display `tmb_df`
tmb_df %>%
  head(n = 15) %>%
  regulartable() %>%
  fontsize(size = 12, part = "all")

# Join ATRT expression, focal CN data and transposed ssGSEA data with tumor 
# mutation burden data
atrt_expression_cn_tmb_df <- atrt_expression_cn_df %>%
  dplyr::left_join(tmb_df, by = "sample_id")

## TODO: Add a column to this data.frame denoting `chr22q` loss using the SV
# data.
```

# Save results

```{r}
# Save final data.frame
final_df <- atrt_expression_cn_tmb_df %>%
  dplyr::group_by(sample_id) %>%
  dplyr::summarise(Kids_First_Biospecimen_ID = paste(sort(unique(
    Kids_First_Biospecimen_ID.x
  )), collapse = ", "),
  Kids_First_Participant_ID = paste(sort(unique(Kids_First_Participant_ID.x)), collapse = ", "),
  location_summary = paste(sort(unique(location_summary)), collapse = ", "),
  age_at_diagnosis_days = paste(sort(unique(age_at_diagnosis_days)), collapse = ", "),
  germline_sex_estimate = paste(sort(unique(germline_sex_estimate)), collapse = ", "),
  SMARCB1_focal_status = paste(sort(unique(SMARCB1_focal_status)), collapse = ", "),
  SMARCA4_focal_status = paste(sort(unique(SMARCA4_focal_status)), collapse = ", "),
  Overexpressed_gene_sets = paste(sort(unique(Overexpressed_gene_sets)), collapse = ", "),
  HALLMARK_MYC_TARGETS_V1 = mean(HALLMARK_MYC_TARGETS_V1),
  HALLMARK_MYC_TARGETS_V2 = mean(HALLMARK_MYC_TARGETS_V2),
  HALLMARK_NOTCH_SIGNALLING = mean(HALLMARK_NOTCH_SIGNALING),
  tmb = mean(tmb),
  GLI2 = mean(GLI2),
  TYR = mean(TYR),
  DCT = mean(DCT),
  SPEF1 = mean(SPEF1),
  CDK6 = mean(CDK6),
  DNAH11 = mean(DNAH11),
  VEGFA = mean(VEGFA),
  MYCN = mean(MYCN),
  MYC = mean(MYC),
  ASCL1 = mean(ASCL1),
  MITF = mean(MITF),
  HOTAIR = mean(HOTAIR)
  ) %>%
  dplyr::distinct()

readr::write_tsv(final_df,
                 file.path(results_dir, "ATRT_molecular_subtypes.tsv.gz"))

# Display `final_df`
final_df %>%
  regulartable() %>%
  fontsize(size = 12, part = "all")
```

