---
title: "ATRT Molecular Subtyping - Data Prep"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2019
---

This notebook addresses the issue of molecular subtyping ATRT samples. 

# Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

`Rscript -e "rmarkdown::render('analyses/molecular-subtyping-ATRT/01-ATRT-molecular-subtyping-data-prep.Rmd', clean = TRUE)"`

# Set Up

```{r}
# Get `magrittr` pipe
`%>%` <- dplyr::`%>%`

# Install `DT` and `htmlwidgets` packages for displaying tables
if (!("DT" %in% installed.packages())) {
  install.packages("DT")
}
if (!("htmlwidgets" %in% installed.packages())) {
  install.packages("htmlwidgets")
}
```

## Directories and Files

```{r}
# Detect the ".git" folder -- this will in the project root directory.
# Use this as the root directory to ensure proper sourcing of functions no
# matter where this is called from
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

# File path to results directory
input_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-ATRT", "atrt-subset")

# File path to results directory
results_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-ATRT", "results")

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

# Read in non-subsetted metadata
metadata <- readr::read_tsv(file.path(root_dir, "data", "pbta-histologies.tsv"))

# Read in ATRT subset metadata
subset_metadata <- metadata %>%
  dplyr::filter(short_histology == "ATRT",
                sample_type == "Tumor",
                composition == "Solid Tissue")

# Select wanted columns in metadata for merging and assign to a new object
select_metadata <- metadata %>%
  dplyr::select(sample_id,
                Kids_First_Biospecimen_ID)

# Read in ATRT subset GSVA pathway scores
gsva_subset <-
  as.data.frame(readr::read_tsv(file.path(input_dir, "atrt_gsva.tsv")))

# Read in ATRT subset z-scored RNA expression data
stranded_expression <-
  readr::read_rds(
    file.path(
      input_dir,
      "atrt_log_expression.RDS"
    )
  )

# Read in ATRT subset focal CN data
cn_df <- readr::read_tsv(
  file.path(
    input_dir,
    "atrt_focal_cn.tsv.gz"
  )
)

# Read in ATRT subset consensus mutation data
tmb_df <-
  data.table::fread(file.path(input_dir,
                              "atrt_tmb.tsv"))

# Read in ATRT subset GISTIC data
gistic_df <-
  readr::read_tsv(file.path(input_dir, "atrt_gistic_broad_values.tsv"))

```

## Custom Function

```{r}
# Custom datatable function
# Function code adapted from: https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/49acc98f5ffd86853fc70f220623311e13e3ca9f/analyses/collapse-rnaseq/02-analyze-drops.Rmd#L23
viewDataTable <- function(data) {
  DT::datatable(
    data,
    rownames = FALSE,
    filter = "bottom",
    class = 'cell-border stripe',
    options = list(
      pageLength = 5,
      searchHighlight = TRUE,
      scrollX = TRUE,
      dom = 'tpi',
      initComplete = htmlwidgets::JS(
        "function(settings, json) {",
        "$(this.api().table().header()).css({'background-color':
                                            '#004467', 'color': '#fff'});",
        "}"
      )
    )
  )
}
```

# Prepare Data

## Metadata

```{r}
# Define regions of the brain (using Anatomy of the Brain figure found at
# https://www.ncbi.nlm.nih.gov/books/NBK65903/figure/CDR0000574573__205/)
supratentorial <-
  c(
    "Skull",
    "Thalamus",
    "Temporal Lobe",
    "Frontal Lobe",
    "Parietal Lobe",
    "Cerebrum",
    "Basal Ganglia",
    "Cranial Nerves NOS",
    "Basal Ganglia;Temporal Lobe",
    "Frontal Lobe;Parietal Lobe;Temporal Lobe",
    "Parietal Lobe;Temporal Lobe",
    "Frontal Lobe;Parietal Lobe"
  )

infratentorial <-
  c(
    "Cerebellum/Posterior Fossa",
    "Brain Stem- Pons;Cerebellum/Posterior Fossa",
    "Cerebellum/Posterior Fossa;Other locations NOS",
    "Brain Stem",
    "Brain Stem- Midbrain/Tectum;Ventricles",
    "Cerebellum/Posterior Fossa;Ventricles",
    "Cerebellum/Posterior Fossa;Spinal Cord- Cervical;Spinal Cord- Lumbar/Thecal Sac;Spinal Cord- Thoracic",
    "Other locations NOS;Spinal Cord- Lumbar/Thecal Sac;Spinal Cord- Thoracic;Ventricles"
  )

collapsed_metadata <- subset_metadata %>%
  dplyr::mutate(
    location_summary = dplyr::case_when(
      primary_site %in% infratentorial ~ "infratentorial",
      primary_site %in% supratentorial ~ "supratentorial",
      TRUE ~ "NA"
    ),
    # Convert `OS_days` values to months and assign new values to `OS_months`
    # column
    OS_months = OS_days / (365 / 12)
  ) %>%
  dplyr::select(
    sample_id,
    Kids_First_Biospecimen_ID,
    Kids_First_Participant_ID,
    location_summary,
    age_at_diagnosis_days,
    OS_months,
    germline_sex_estimate,
    primary_site
  ) %>%
  dplyr::group_by(
    sample_id,
    Kids_First_Participant_ID
  ) %>%
  dplyr::summarize_all(function(x) paste(sort(unique(x)), collapse = ", "))

# Display metadata subsetted for ATRT samples
collapsed_metadata %>%
  head(n = 15)
```

## Filter and join RNA expression, CN, TMB, ssGSVA and GISTIC data

### RNA expression data 

```{r}
# Define target overexpressed gene vectors
tyr_genes <-
  c("TYR",
    "MITF",
    "DCT",
    "VEGFA",
    "DNAH11",
    "SPEF1",
    "POU3F4",
    "POU3F2",
    "PBX1")
shh_genes <-
  c(
    "MYCN",
    "GLI2",
    "CDK6",
    "ASCL1",
    "HES5/6",
    "DLL1/3",
    "ZBTB7A",
    "RXF3",
    "RXF2",
    "MYBL2",
    "MXI1",
    "MEIS3",
    "MEIS2",
    "MAX",
    "INSM1",
    "FOXK1"
  )
myc_genes <-
  c(
    "MYC",
    "HOTAIR",
    "HOX",
    "TCF7L2",
    "STAT1",
    "REST",
    "RARG",
    "RAD21",
    "NR4A2",
    "IRF9",
    "IRF8",
    "FOXC1",
    "CEBPB",
    "ATF4"
  )

# Filter to only the genes of interest
filtered_expression <- stranded_expression[which(
   rownames(stranded_expression) %in% c(tyr_genes, shh_genes, myc_genes)
), ]

# scale does column centering, so we transpose first
long_stranded_expression <- scale(t(filtered_expression), 
                                  center = TRUE,
                                  scale = TRUE)

# Merge metadata with expression data
expression_metadata <- long_stranded_expression %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Kids_First_Biospecimen_ID") %>% 
  dplyr::left_join(select_metadata, by = "Kids_First_Biospecimen_ID") %>%
  dplyr::select(-Kids_First_Biospecimen_ID)

# Display `expression_metadata`
expression_metadata %>%
  head(n = 15)

# Join expression data with metadata filtered for `ATRT`
atrt_expression_df <- collapsed_metadata %>%
  dplyr::left_join(expression_metadata,
                   by = "sample_id")

# Remove data we no longer need
rm(filtered_expression, long_stranded_expression, expression_metadata)
```

### CN data

```{r}
# Filter focal CN data for SMARCB1 and SMARCA4 status
genes_cn_df <- cn_df %>%
  dplyr::filter(gene_symbol %in% c("SMARCB1", "SMARCA4")) %>%
  tidyr::spread(gene_symbol, status, fill = "neutral") %>%
  dplyr::rename(SMARCB1_focal_status = SMARCB1,
                SMARCA4_focal_status = SMARCA4)

# add in the samples that did not have any copy number changes for either
# of these genes
missing_cn_samples <- setdiff(unique(cn_df$sample_id), genes_cn_df$sample_id)
genes_cn_df <- subset_metadata %>%
  dplyr::filter(sample_id %in% missing_cn_samples) %>%
  dplyr::select(sample_id,
                Kids_First_Participant_ID,
                Kids_First_Biospecimen_ID) %>%
  dplyr::rename(biospecimen_id = Kids_First_Biospecimen_ID) %>%
  # bind rows fills with NAs -- if something is NA, there is no evidence for
  # a copy number alteration in that gene and we will call it neutral
  dplyr::bind_rows(genes_cn_df) %>%
  dplyr::mutate_all(function(x) tidyr::replace_na(x, "neutral"))

# Display `genes_cn_df`
genes_cn_df %>%
  head(n = 15)

# Join ATRT expression data with focal CN data
atrt_expression_cn_df <- atrt_expression_df %>%
  dplyr::left_join(dplyr::select(genes_cn_df,
                                 -biospecimen_id,
                                 -Kids_First_Participant_ID), 
                   by = "sample_id")

# Remove data we don't need 
rm(missing_cn_samples, cn_df, atrt_expression_df, genes_cn_df)
```

### GSVA data

```{r}
# Filter to only the hallmark gene sets that were named in the issue
relevant_genesets <- c("HALLMARK_MYC_TARGETS_V1",
                       "HALLMARK_MYC_TARGETS_V2",
                       "HALLMARK_NOTCH_SIGNALING")
gsva_filtered <- gsva_subset %>%
  dplyr::filter(hallmark_name %in% relevant_genesets)

# Matrix with gene set columns that can be scaled
gsva_mat <- gsva_filtered %>% 
  dplyr::select(-Kids_First_Biospecimen_ID) %>% 
  tidyr::spread(hallmark_name, gsea_score) %>%
  tibble::column_to_rownames("sample_id") %>%
  as.matrix()

# z-score scores (here they are columns)
scaled_gsva <- scale(gsva_mat, center = TRUE, scale = TRUE) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("sample_id")

# Display scaled_gsva
scaled_gsva %>%
  head(n = 15)

# Join ATRT expression and focal CN data with GSVA data
atrt_expression_cn_df <- atrt_expression_cn_df %>%
  dplyr::left_join(scaled_gsva,
                   by = "sample_id")

# Remove data we no longer need
rm(gsva_filtered, gsva_mat, scaled_gsva)
```

### Tumor mutation burden data

```{r}
# Display `tmb_df`
tmb_df %>%
  head(n = 15)

# Join ATRT expression, focal CN data and transposed ssGSEA data with tumor
# mutation burden data
atrt_expression_cn_tmb_df <- atrt_expression_cn_df %>%
  dplyr::left_join(dplyr::select(tmb_df,
                                 -Tumor_Sample_Barcode,
                                 -Kids_First_Participant_ID), 
                   by = "sample_id")

# Remove data we no longer need
rm(tmb_df, atrt_expression_cn_df)
```

### GISTIC data

```{r}
gistic_df <- gistic_df %>%
  dplyr::select(-Kids_First_Biospecimen_ID) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(chr_22q_loss = dplyr::case_when(`22q` == -1 ~ "Yes",
                                                TRUE ~ "No")) %>%
  dplyr::select(-`22q`)

# Display `gistic_df`
gistic_df %>%
  head(n = 15)

# Join GISTIC data with the running final data.frame
final_df <- atrt_expression_cn_tmb_df %>%
  dplyr::left_join(gistic_df, by = "sample_id") %>%
  dplyr::distinct()

# Remove data we no longer need
rm(gistic_df, atrt_expression_cn_tmb_df)
```

# Save final table of results

```{r}
# Save final data.frame
final_df <- final_df %>%
  dplyr::select(
    sample_id,
    Kids_First_Biospecimen_ID,
    Kids_First_Participant_ID,
    age_at_diagnosis_days,
    OS_months,
    germline_sex_estimate,
    primary_site,
    location_summary,
    chr_22q_loss,
    SMARCB1_focal_status,
    TYR,
    MITF,
    DCT,
    VEGFA,
    DNAH11,
    SPEF1,
    POU3F4,
    POU3F2,
    PBX1,
    SMARCA4_focal_status,
    HALLMARK_NOTCH_SIGNALING,
    MYCN,
    GLI2,
    CDK6,
    ASCL1,
    ZBTB7A,
    MYBL2,
    MXI1,
    MEIS3,
    MEIS2,
    MAX,
    INSM1,
    FOXK1,
    HALLMARK_MYC_TARGETS_V1,
    HALLMARK_MYC_TARGETS_V2,
    dplyr::everything()
  )

# Write final table to file
readr::write_tsv(final_df,
                 file.path(results_dir, "ATRT_molecular_subtypes.tsv"))

# Display `final_df`
viewDataTable(final_df) 
```

# Session Info

```{r}
sessionInfo()
```

