---
title: "SHH TP53 Molecular Subtyping"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Candace Savonen for ALSF CCDL
date: 2019
---

This notebook [molecularly subtypes Medulloblastoma samples into their TP53 status](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/247). 

## Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

`Rscript -e "rmarkdown::render('analyses/molecular-subtyping-SHH-tp53/SHH-tp53-molecular-subtyping-data-prep.Rmd', clean = TRUE)"`

## Set Up

```{r}
# Get `magrittr` pipe
`%>%` <- dplyr::`%>%`
```

### Directories and Files

```{r}
# File path to data directory
data_dir <- file.path("..", "..", "data")

# File path to results directory
results_dir <- "results"

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

## Read in the data

Read in metadata and isolate to medulloblastoma samples.

```{r}
metadata <-
  readr::read_tsv(file.path(data_dir, "pbta-histologies.tsv"),
                  col_types = readr::cols(
                    molecular_subtype = readr::col_character() 
                  )) %>%
  dplyr::filter(disease_type_new == "Medulloblastoma")

# Check summary of subtypes here
summary(as.factor(metadata$molecular_subtype))
```

Propagate the `molecular_subtype` labels to biospecimens of the same `sample_id`.
First make a key of molecular subtype to `sample_id`.

```{r}
key_df <- metadata %>%
  dplyr::filter(!is.na(molecular_subtype)) %>%
  dplyr::select(molecular_subtype, sample_id)

# Make into list form so it can be used in dplyr::recode
key <- key_df$molecular_subtype
names(key) <- key_df$sample_id
```

Recode and filter to all `sample_id`s that are `SHH` and have `WGS` or `WXS` data.

```{r}
shh_metadata <- metadata %>%
  dplyr::mutate(molecular_subtype = dplyr::recode(sample_id, 
                                                  !!!key, 
                                                  .default = "NA")) %>%
  dplyr::filter(molecular_subtype == "SHH")

# Check summary after recoding
summary(as.factor(shh_metadata$molecular_subtype))
```

Read in the SNV consensus calls. 

```{r}
snv_df <-
  # Read in the consensus file
  data.table::fread(file.path(data_dir, "pbta-snv-consensus-mutation.maf.tsv.gz"),
    data.table = FALSE
  ) %>%
  # Filter to only Medulloblastoma samples
  dplyr::filter(Tumor_Sample_Barcode %in% shh_metadata$Kids_First_Biospecimen_ID)
```

## Identify TP53 SNVs.

Get a list of samples that are positive for TP53 mutations in the SNV consensus data.

```{r}
tp53_mut_biospecimens <- snv_df %>%
  # Filter to only SNV's in TP53 gene
  dplyr::filter(Hugo_Symbol == "TP53") %>%
  dplyr::pull(Tumor_Sample_Barcode)
```

Now recode their TP53 status to be associated with their `sample_id`.

```{r}
# Make into list form so it can be used in dplyr::recode
key <- shh_metadata$sample_id
names(key) <- shh_metadata$Kids_First_Biospecimen_ID

# List of tp53 postive samples
tp53_mut_samples <- dplyr::recode(tp53_mut_biospecimens, !!!key)
```

Summarize TP53 status to a table with other sample identifiers.

```{r}
tp53_status_table <- shh_metadata %>%
  # Narrow down to sample identifier columns
  dplyr::select("Kids_First_Participant_ID", 
                "sample_id", 
                "Kids_First_Biospecimen_ID", 
                "experimental_strategy") %>%
  # Collapse Biospecimen ids for each sample
  dplyr::group_by(Kids_First_Participant_ID, sample_id) %>%
  dplyr::summarize(Kids_First_Biospecimen_IDs = paste(Kids_First_Biospecimen_ID, 
                                                      collapse = ", "), 
                   experimental_strategies = paste(sort(experimental_strategy), 
                                                   collapse = ", ")) %>%
  # Make a variable that says whether or not there is a TP53 SNV mutation
  # Add an exception where NA is placed if there are no DNA-seq samples
  dplyr::mutate(TP53_mutation =  sample_id %in% tp53_mut_samples,
                TP53_mutation = dplyr::case_when(
                  !grepl("WXS|WGS", experimental_strategies) ~ NA,
                  TRUE ~ TP53_mutation
                ))

# Print out this table
tp53_status_table
```

Save table to a TSV file. 

```{r}
readr::write_tsv(tp53_status_table, file.path(results_dir, 
                                              "tp53-shh-samples-status.tsv"))
```

## Session Info

```{r}
sessionInfo()
```
