---
title: "Format analyses directories contributions"
output: html_notebook
author: JN Taroni for ALSF CCDL
date: 2022
---

## Set up

```{r}
library(tidyverse)
```

In the first script, we created TXT files that contain the authors of contributions to each module in `analyses/`.
Those files live in a directory called `scratch/count-contributions/` (relative to the root of the repo).
We're going to save the wrangled results to a `results/` directory

```{r}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git")) 
components_dir <- file.path(root_dir, "scratch", "count-contributions")
results_dir <- "results"
dir.create(results_dir, showWarnings = FALSE)
```

We're going save the file that maps contributors to modules _and_ a file that counts the number of modules someone has contributed to.

```{r}
module_contributions_file <- file.path(results_dir, 
                                       "module_contributors.tsv")
count_modules_file <- file.path(results_dir,
                                "module_contribution_counts.tsv")
total_countributions_file <- file.path(results_dir,
                                       "total_contributions_clean.tsv")
```

## Mapping different IDs to the same contributor

Some folks contributed under two "names" (usually their GH handle and their name), or just their first name, or just their GH handle.
So let's clean this up in the interest of readability in the final data frames.

```{r}
name_recoding <- c(
  jashapiro = "Joshua A. Shapiro",
  `Joshua Shapiro` = "Joshua A. Shapiro",
  Stephanie = "Stephanie J. Spielman",
  Bailey = "Bailey K. Farrow",
  runjin326 = "Run Jin",
  Mateusz = "Mateusz P. Koptyra",
  tkoganti = "Tejaswi Koganti",
  NNoureen = "Nighat Noureen",
  `Jo Lynne` = "Jo Lynne Rokita",
  jharenza = "Jo Lynne Rokita",
  `e-t-k` = "Ellen Kephardt",
  yangyangclover = "Yang Yang",
  PichaiRaman = "Pichai Raman",
  hbeale = "Holly Beale",
  `Jaclyn Taroni` = "Jaclyn N. Taroni",
  `Krutika Gaonkar` = "Krutika S. Gaonkar",
  `Candace Savonen` = "Candace L. Savonen",
  `Chante Bethell` = "Chante J. Bethell",
  `Komal Rathi` = "Komal S. Rathi",
  `Laura Egolf` = "Laura E. Egolf",
  `Dan Miller` = "Daniel P. Miller",
  `Casey Greene` = "Casey S. Greene",
  `Teja Koganti` = "Tejaswi Koganti"
)
```

## Removing bot contributors

We have a bot `alexslemonade-docs-bot` that may make commits, including via GHA (GitHub Actions) to run this very module and produce the final author contribution lists.
We should not include the bot itself as a contributor.

```{r}
remove_names <- c("alexslemonade-docs-bot")
```

## Module contributions

There are some modules that did not generate results that are included in the manuscript.
For our purposes, we want to skip those lists of contributors.
The not included list is shorter than the included list, so that's why we'll take this approach!

```{r}
# We're not counting these because they didn't make it to the manuscript
not_included_analyses <- c(
  "cnv-comparison",
  "comparative-RNASeq-analysis",
  "compare-gistic",
  "count-contributions",
  "create-subset-files",
  "molecular-subtyping-ATRT",
  "molecular-subtyping-SHH-tp53",
  "mutect2-vs-strelka2",
  "rna-seq-composition",
  "selection-strategy-comparison",
  "sex-prediction-from-RNASeq",
  "ssgsea-hallmark",
  "tcga-capture-kit-investigation",
  "sample-distribution-analysis"
)
```

All of the _relevant files_ are those TXT files in `components/` that do not match one of the not included modules listed above.
We're also going to skip `components/total_contributions.txt` for now. 

```{r}
# All of the files that were included in components directory
contribution_files <- list.files(components_dir)

# We'll count the files that come from modules that are included in the 
# manuscript by removing the modules that *are not* included
count_files <- contribution_files[str_detect(contribution_files, 
                                             pattern = paste(
                                               not_included_analyses, 
                                               collapse = "|"),
                                             negate = TRUE)]

# We're also not going to look at total contributions here
count_files <- setdiff(count_files, "total_contributions.tsv")

count_files
```

Now read in the relevant contribution files to a data frame that we can use to count.

```{r}
count_contributions_df <- suppressMessages(  # This is chatty!
  # For all the files we'll consider
  file.path(components_dir, count_files) %>%
    # Read them in, but skip the first (blank) line and set the (only) column name
    # This step produces a list
    map(read_tsv, skip = 1, col_names = "author") %>%
    # Use the file names themselves to set the names of the list
    # This will be the name of the module
    set_names(str_remove(count_files, "_contributors.txt")) %>% 
    # Create a data frame where we use the module names to keep track of where
    # the data frame of contributors came from
    bind_rows(.id = "module") %>% 
    distinct()
)

colnames(count_contributions_df) <- c("module", "author")
```

Recode names/handles to be consistent and more readable.

```{r}
# Recode GitHub names to be more readable and de-duplicated in some cases
count_contributions_df <- count_contributions_df %>%
  mutate(author = recode_factor(author, !!!name_recoding)) %>%
# Add Yang Yang to the list for contributions to the `chromothripsis` module. This was renamed from
# `sv-analysis` since those contributions, so those were missed. [Ticket](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/1599) and [rename PR](https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/1069)
  add_row(module = "chromothripsis", author = "Yang Yang") %>%
  # If people had two names/IDs, this will de-duplicate now that things have 
  # been recoded
  distinct() %>%
  arrange(module)

# Write this less summarized data to file
write_tsv(count_contributions_df, module_contributions_file)
```

Now we can count the number of modules each person contributed to.

```{r}
# Count the number of modules each person contributed to, which we can simply
# do with count
counts_df <- count_contributions_df %>%
  count(author, name = "num_modules") %>%
  arrange(desc(num_modules))

# Write the summarized modules counted
write_tsv(counts_df, count_modules_file)
```

## Total contributions

We'll clean up the total contributions generated with `git shortlog` by summarizing commits by a single person and generally making things more readable.

```{r}
total_contributions_df <- read_tsv(file.path(components_dir,
                                             "total_contributions.tsv"),
                                   col_names = c("num_commits", "author")) 

# Update with new column names
total_contributions_df <- total_contributions_df %>%
  # Reorder
  select(author, num_commits) %>%
  # Remove authors which we should not keep, namely bots %>%
  filter(!(author %in% remove_names)) %>%
  # Recode names/IDs to be consistent
  mutate(author = recode_factor(author, !!!name_recoding)) %>%
  # For folks that had multiple names included, summarize the total commits
  group_by(author) %>%
  summarize(num_commits = sum(num_commits)) %>%
  # Restore descending order for number of commits
  arrange(desc(num_commits))
```

Write to file.

```{r}
write_tsv(total_contributions_df, path = total_countributions_file)
```


Finally, let's print session info:

```{r}
sessionInfo()
```

