---
title: "Creating a TSV of author information for AlexsLemonade/OpenPBTA-manuscript"
output: 
  html_notebook:
    toc: true
    toc_float: true
author: Stephanie Spielman for ALSF CCDL
date: 2022
---

```{r setup}
library(magrittr) # load for piping
```


This notebook parses the `metadata.yaml` file to obtain a TSV of author information relevant for manuscript submission, including:

- Name
- Affiliation(s)
- ORCID
- Email 


## Read in metadata 

The previously-run notebook `03-set-authorship-order.Rmd` will have created an updated metadata YAML file in `results/metadata.yaml`, so we can directly read it in and grab the author list.


```{r}
manuscript_metadata <- yaml::read_yaml(
  file.path("results", "metadata.yaml")
)
author_list <- manuscript_metadata$author
```

## Parse the metadata

First, we'll define a function that is useful for parsing fields out of the metadata. 
This function allows for the fact that some keys may be missing for some authors (e.g. `ORCID` or `email`), and the function will fill these in as `NA`s.

```{r}
# Function to parse out a given field
# This function allows us to extract fields
#  that might missing, which get assigned NA
extract_yaml_value <- function(x, name) {
  
  # If the name is not present, return NA
  if (!(name %in% names(x))) {
    return_list <- list(NA)
    names(return_list) <- name
    return(return_list)
  } else {
    # Process affiliations specifically since there can be >1
    #  and at least 1 has superscripts that should be removed
    if (name == "affiliations") {
      list("affiliations" =
             paste(x[["affiliations"]], collapse = "; ") %>%
             # Remove any superscripts
             stringr::str_replace_all(., "<sup>.+<\\/sup>", "")
      )
    } else {
      # Extract the value directly
      return(magrittr::extract(x, name))
    }
  }
  
}
```


Now, we can use this function to parse the metadata into a data frame.


```{r}

# Extract info
author_info <- dplyr::bind_cols(
 purrr::map_dfr(author_list, extract_yaml_value, "name"),
 purrr::map_dfr(author_list, extract_yaml_value, "affiliations"),
 purrr::map_dfr(author_list, extract_yaml_value, "orcid"),
 purrr::map_dfr(author_list, extract_yaml_value, "email")
)


# Make names titlecase, except ORCID which should be all uppercase
names(author_info) <- stringr::str_to_title(names(author_info))
author_info <- dplyr::rename(author_info, ORCID = Orcid)
```

### Save

And we're ready to save!

```{r}
readr::write_tsv(
  author_info,
  "author_information.tsv"
)
```

## Session Info

```{r}
sessionInfo()
```

