---
title: "Compile molecular subtyping results"
output: 
  html_notebook:
    toc: true
    toc_float: true
author: Jaclyn Taroni for CCDL, Jo Lynne Rokita for D3b
date: 2020
params:
   is_ci: FALSE
---

The purpose of this notebook is to aggregate molecular subtyping results from the following mature analysis modules:

* [`molecular-subtyping-EWS`](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/molecular-subtyping-EWS)
* [`molecular-subtyping-HGG`](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/molecular-subtyping-HGG)
* [`molecular-subtyping-LGAT`](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/molecular-subtyping-LGAT)
* [`molecular-subtyping-embryonal`](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/molecular-subtyping-embryonal)
* [`molecular-subtyping-CRANIO`](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/molecular-subtyping-CRANIO)
* [`molecular-subtyping-EPN`](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/molecular-subtyping-EPN\)
* [`molecular-subtyping-MB`](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/molecular-subtyping-MB)
* [`molecular-subtyping-neurocytoma`](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/molecular-subtyping-neurocytoma)
* [`molecular-subtyping-chordoma`](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/molecular-subtyping-chordoma)


## Set up

```{r}
# This logic is the same as analyses/gene-set-enrichment-analysis/02-model-gsea.Rmd
# Assigning params$is_ci to running_in_ci avoids a locked binding error
running_in_ci <- params$is_ci

# Are we testing? In case of a non 0/1 number, we recast as logical, and then 
# ensure logical.
if (running_in_ci %in% c(0,1)) running_in_ci <- as.logical(running_in_ci)
if (!(is.logical(running_in_ci)))
{
  stop("\n\nERROR: The parameter `is_ci` should be FALSE/TRUE (or 0/1).")
}
```

### Libraries and functions

```{r}
library(tidyverse)
```

```{r}
bind_exp_strategies <- function(results_df) {
  # Many of the molecular subtyping tables contain a 
  # Kids_First_Biospecimen_ID_RNA and Kids_First_Biospecimen_ID_DNA column,
  # this function separates those out and binds the rows to mirror how 
  # pbta-histologies.tsv is organized
  #
  # Args:
  #   results_df: a molecular subtyping table with Kids_First_Biospecimen_ID_RNA
  #               and Kids_First_Biospecimen_ID_DNA columns
  # Returns:
  #   A data.frame where each biospecimen is a row
  rna_df <- results_df %>%
    dplyr::select(-Kids_First_Biospecimen_ID_DNA) %>%
    dplyr::filter(!is.na(Kids_First_Biospecimen_ID_RNA)) %>%
    dplyr::rename(Kids_First_Biospecimen_ID = Kids_First_Biospecimen_ID_RNA)
  
  dna_df <- results_df %>%
    dplyr::select(-Kids_First_Biospecimen_ID_RNA) %>%
    dplyr::filter(!is.na(Kids_First_Biospecimen_ID_DNA)) %>%
    dplyr::rename(Kids_First_Biospecimen_ID = Kids_First_Biospecimen_ID_DNA)
  
  # sometimes there will be multiple RNA-seq assays per single DNA-seq assay
  # or vice versa, which is why this distinct is necessary
  return(bind_rows(rna_df, dna_df) %>% distinct())
}
```

### Directories and files

```{r}
data_dir <- file.path("..", "..", "data")
analyses_dir <- ".."

# directories for upstream subtyping modules
cranio_dir <- file.path(analyses_dir, "molecular-subtyping-CRANIO")
ews_dir <- file.path(analyses_dir, "molecular-subtyping-EWS")
epn_dir <- file.path(analyses_dir, "molecular-subtyping-EPN")
hgg_dir <- file.path(analyses_dir, "molecular-subtyping-HGG")
lgat_dir <- file.path(analyses_dir, "molecular-subtyping-LGAT")
mb_dir <- file.path(analyses_dir, "molecular-subtyping-MB")
embryonal_dir <- file.path(analyses_dir, "molecular-subtyping-embryonal")
neurocytoma_dir <- file.path(analyses_dir, "molecular-subtyping-neurocytoma")
chordoma_dir <- file.path(analyses_dir, "molecular-subtyping-chordoma")

# the folder that contains the tabular results is standardized across modules
results_dir <- "results"
# create the results directory for this module if it does not yet exist
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

#### Input files

When we run this locally, we want to tie it to a specific version of the histologies file _prior_ to feedback we received from pathology that is recorded on [#609](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/609).

```{r}
if (running_in_ci) {
  histologies_file <- file.path(data_dir, "pbta-histologies-base.tsv")
} else {
  histologies_file <- file.path(data_dir, "release-v18-20201123",
                                "pbta-histologies-base.tsv")
}

join_cols <- c("Kids_First_Biospecimen_ID",
               "short_histology",
               "broad_histology",
               "Notes")

```

Results files from individual modules.

```{r}
cranio_results_file <- file.path(cranio_dir, results_dir, "CRANIO_molecular_subtype.tsv")
ews_results_file <- file.path(ews_dir, results_dir, "EWS_results.tsv")
epn_results_file <- file.path(epn_dir, results_dir, "EPN_all_data_withsubgroup.tsv")
hgg_results_file <- file.path(hgg_dir, results_dir, "HGG_molecular_subtype.tsv")
lgat_results_file <- file.path(lgat_dir, results_dir, "lgat_subtyping.tsv")
mb_results_file <- file.path(mb_dir, results_dir, "MB_molecular_subtype.tsv")
embryonal_results_file <- file.path(embryonal_dir, results_dir,
                                    "embryonal_tumor_molecular_subtypes.tsv")
neurocytoma_results_file <- file.path(neurocytoma_dir, results_dir, "neurocytoma_subtyping.tsv")
# reading in smarcb1 status only, subtyping will come through pathology review in step 3
chordoma_results_file <- file.path(chordoma_dir, results_dir, "chordoma_smarcb1_status.tsv")

```

#### Output file

```{r}
output_file <- file.path(results_dir, "compiled_molecular_subtypes.tsv")
```

## Read in data

```{r message=FALSE}
# split 
histologies_df <- read_tsv(histologies_file, guess_max = 10000,col_types = cols(.default = "c"))
cranio_results_df <- read_tsv(cranio_results_file)
ews_results_df <- read_tsv(ews_results_file)
epn_results_df <- read_tsv(epn_results_file)
hgg_results_df <- read_tsv(hgg_results_file)
lgat_results_df <- read_tsv(lgat_results_file)
mb_results_df <- read_tsv(mb_results_file)
neurocytoma_results_df <- read_tsv(neurocytoma_results_file)
embryonal_results_df <- read_tsv(embryonal_results_file)
chordoma_results_df <- read_tsv(chordoma_results_file)
```

## Compile the subtyping resutls

### Handling non-ATRT/non-MB embryonal tumors

The molecular subtyping information from these tumors will go into the v18 release, and we will update `integrated_diagnosis`, `short_histology`, `broad_histology`, and `Notes` columns now, until SQL rules PR [#748](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/748) goes in later.

```{r}
embryonal_results_df <- bind_exp_strategies(embryonal_results_df) %>%
  inner_join(select(histologies_df,join_cols),
             by = "Kids_First_Biospecimen_ID") %>%
  mutate(integrated_diagnosis = case_when(molecular_subtype == "CNS Embryonal, NOS"~ "CNS Embryonal tumor, NOS",
                                          molecular_subtype == "CNS HGNET-MN1"~ "CNS Embryonal tumor, HGNET-MN1",
                                          molecular_subtype == "CNS NB-FOXR2" ~ "CNS neuroblastoma",
                                          molecular_subtype == "ETMR, C19MC-altered"~ "Embryonal tumor with multilayer rosettes, C19MC-altered",
                                          molecular_subtype == "ETMR, NOS"~"Embryonal tumor with multilayer rosettes, NOS",
                                          TRUE ~ NA_character_), 
        short_histology = 
          if_else(molecular_subtype %in% c("ETMR, C19MC-altered", "ETMR, NOS"),
                "ETMR", "Embryonal tumor"),
        broad_histology = "Embryonal tumor",
        Notes = if_else(!is.na(integrated_diagnosis), "Updated via OpenPBTA subtyping", Notes))

```

### Handling EWS

The EWS results were updated in V18. Adding integrated dx, broad hist, short hist here for now.

```{r}
# Add EWS integrated diagnosis, broad histology, short histology
ews_results_df <- bind_exp_strategies(ews_results_df) %>%
  inner_join(select(histologies_df,join_cols),
             by = "Kids_First_Biospecimen_ID") %>%
  mutate(integrated_diagnosis = "Ewing sarcoma",
         broad_histology = "Mesenchymal non-meningothelial tumor",
         short_histology = "EWS",
         Notes = if_else(!is.na(integrated_diagnosis), "Updated via OpenPBTA subtyping", Notes))
```

### Handling HGG

HGG subtyping was updated with V18.

```{r}
hgg_results_df <- bind_exp_strategies(hgg_results_df) %>%
  inner_join(select(histologies_df,join_cols),
             by = "Kids_First_Biospecimen_ID") %>%
  mutate(integrated_diagnosis = case_when(molecular_subtype == "DMG, H3 K28"~ "Diffuse midline glioma, H3 K28-mutant",
                                          molecular_subtype == "DMG, H3 K28, TP53 activated"~ "Diffuse midline glioma, H3 K28-mutant",
                                          molecular_subtype == "DMG, H3 K28, TP53 loss"~ "Diffuse midline glioma, H3 K28-mutant",
                                          molecular_subtype == "HGG, H3 G35" ~ "High-grade glioma/astrocytoma, H3 G35-mutant",
                                          molecular_subtype == "HGG, H3 G35, TP53 loss" ~ "High-grade glioma/astrocytoma, H3 G35-mutant",
                                          molecular_subtype == "HGG, H3 wildtype" ~ "High-grade glioma/astrocytoma, H3 wildtype",
                                          molecular_subtype == "HGG, H3 wildtype, TP53 activated" ~ "High-grade glioma/astrocytoma, H3 wildtype",
                                          molecular_subtype == "HGG, H3 wildtype, TP53 loss" ~ "High-grade glioma/astrocytoma, H3 wildtype",
                                          molecular_subtype == "HGG, IDH"~ "High-grade glioma/astrocytoma, IDH-mutant",
                                          molecular_subtype == "HGG, IDH, TP53 loss"~ "High-grade glioma/astrocytoma, IDH-mutant",
                                          molecular_subtype == "HGG, IDH, TP53 activated"~ "High-grade glioma/astrocytoma, IDH-mutant",
                                          TRUE~ NA_character_),
                  broad_histology = "Diffuse astrocytic and oligodendroglial tumor",
                  short_histology = "HGAT",
         Notes = if_else(!is.na(integrated_diagnosis), "Updated via OpenPBTA subtyping", Notes))
```

### Handling LGAT


```{r}
lgat_results_df <- bind_exp_strategies(lgat_results_df) %>%
  inner_join(select(histologies_df,c(join_cols,
                    pathology_diagnosis)),
             by = "Kids_First_Biospecimen_ID") %>%
mutate(integrated_diagnosis = case_when(
            str_detect(molecular_subtype, "To be classified") ~ NA_character_,
            pathology_diagnosis == "Ganglioglioma" ~ str_replace(molecular_subtype, "LGG, ", "Ganglioglioma, "),
            TRUE ~ str_replace(molecular_subtype, "LGG, ", "Low-grade glioma/astrocytoma, ")
          ),
         broad_histology = "Low-grade astrocytic tumor",
         short_histology = if_else(pathology_diagnosis == "Ganglioglioma", "Ganglioglioma", "LGAT"),
         Notes = if_else(!is.na(integrated_diagnosis), "Updated via OpenPBTA subtyping", Notes)
  )

```

### Handling EPN 

```{r}
epn_results_df <- bind_exp_strategies(epn_results_df) %>%
  inner_join(select(histologies_df,join_cols),
           by = "Kids_First_Biospecimen_ID") %>%
  mutate(molecular_subtype = subgroup) %>%
  mutate(molecular_subtype = if_else(is.na(molecular_subtype), "EPN, To be classified", molecular_subtype), 
         integrated_diagnosis = case_when(molecular_subtype == "EPN, PF A" ~ "Ependymoma, Posterior Fossa, Type A",
                                          molecular_subtype == "EPN, ST RELA" ~ "Ependymoma, Supratentorial, RELA fusion positive",
                                          molecular_subtype == "EPN, ST YAP1" ~ "Ependymoma, Supratentorial, YAP1 fusion positive", 
                                          TRUE~ NA_character_),
         broad_histology = "Ependymal tumor",
         short_histology = "Ependymoma",
         Notes = if_else(!is.na(integrated_diagnosis), "Updated via OpenPBTA subtyping", Notes))
```

### Handling MB

```{r}
mb_results_df <- bind_exp_strategies(mb_results_df) %>%
  inner_join(select(histologies_df,join_cols),
            by = "Kids_First_Biospecimen_ID") %>%
  mutate(integrated_diagnosis = case_when(molecular_subtype == "MB, SHH"~"Medulloblastoma, SHH-activated",
                                          molecular_subtype == "MB, WNT"~"Medulloblastoma, WNT-activated",
                                          molecular_subtype == "MB, Group3"~"Medulloblastoma, group 3",
                                          molecular_subtype == "MB, Group4"~ "Medulloblastoma, group 4", 
                                          TRUE ~NA_character_),
         broad_histology = "Embryonal tumor",
         short_histology = "Medulloblastoma",
         Notes = if_else(!is.na(integrated_diagnosis), "Subtype based on prediction;Updated via OpenPBTA subtyping", Notes))
```


### Handling CRANIO

```{r}
cranio_results_df <- bind_exp_strategies(cranio_results_df) %>%
  inner_join(select(histologies_df,join_cols),
             by = "Kids_First_Biospecimen_ID") %>%
  mutate(integrated_diagnosis = case_when(molecular_subtype == "CRANIO, ADAM" ~"Adamantinomatous craniopharyngioma",
                                          molecular_subtype == "CRANIO, PAP" ~"Papillary craniopharyngioma", 
                                          TRUE ~ NA_character_),
         broad_histology = "Tumors of sellar region",
         short_histology = "Craniopharyngioma",
         Notes = if_else(!is.na(integrated_diagnosis), "Updated via OpenPBTA subtyping", Notes))
```

### Handling Neurocytoma
```{r}
neurocytoma_results_df <- neurocytoma_results_df %>%
  inner_join(select(histologies_df,join_cols),
            by = "Kids_First_Biospecimen_ID") %>%
  mutate(harmonized_diagnosis = case_when(molecular_subtype == "CNC"~ "Central Neurocytoma",
                                          molecular_subtype == "EVN" ~"Extraventricular Neurocytoma", 
                                        TRUE ~ NA_character_),
         broad_histology = "Neuronal and mixed neuronal-glial tumor",
         short_histology = "Neurocytoma",
         Notes = if_else(!is.na(harmonized_diagnosis), "Updated via OpenPBTA subtyping", Notes))
```

### Handling Chordoma
```{r}

chordoma_results_df <- bind_exp_strategies(chordoma_results_df) %>%
  inner_join(select(histologies_df,join_cols),
             by = "Kids_First_Biospecimen_ID") %>%
  # will update subtype and integrated diagnosis in step 3
  mutate(molecular_subtype = NA_character_,
         integrated_diagnosis = NA_character_,
         broad_histology = "Chordoma",
         short_histology = "Chordoma",
         Notes = if_else(!is.na(integrated_diagnosis), "Updated via OpenPBTA subtyping", Notes))

```




### All results

Compile results, sort, and write to file

```{r}
all_results_df <- bind_rows(embryonal_results_df,
                            ews_results_df,
                            hgg_results_df,
                            lgat_results_df,
                            epn_results_df,
                            cranio_results_df,
                            mb_results_df,
                            neurocytoma_results_df,
                            chordoma_results_df) %>%
  select(Kids_First_Participant_ID, sample_id, Kids_First_Biospecimen_ID, molecular_subtype,
         integrated_diagnosis, short_histology, broad_histology, Notes) %>%
  # Cleanup a few Notes which have changed since last time
  # Remove this because those were taken out of EWS
  mutate(Notes = case_when(Notes == "Reclassified due to presence of hallmark EWS fusions"~ NA_character_,
                          Notes == "Subtype based on prediction"~ NA_character_,
                          TRUE ~ Notes)) %>%
  arrange(Kids_First_Participant_ID, sample_id) %>%
  write_tsv(output_file)
```

## Session Info

```{r}
sessionInfo()
```
