---
title: "Compile molecular subtyping results"
output: 
  html_notebook:
    toc: true
    toc_float: true
author: Jaclyn Taroni for CCDL
date: 2020
params:
   is_ci: FALSE
---

The purpose of this notebook is to aggregate molecular subtyping results from the following mature analysis modules:

* [`molecular-subtyping-EWS`](https://github.com/jaclyn-taroni/OpenPBTA-analysis/tree/645-pathology-feedback/analyses/molecular-subtyping-EWS)
* [`molecular-subtyping-HGG`](https://github.com/jaclyn-taroni/OpenPBTA-analysis/tree/645-pathology-feedback/analyses/molecular-subtyping-HGG)
* [`molecular-subtyping-LGAT`](https://github.com/jaclyn-taroni/OpenPBTA-analysis/tree/645-pathology-feedback/analyses/molecular-subtyping-LGAT)
* [`molecular-subtyping-embryonal`](https://github.com/jaclyn-taroni/OpenPBTA-analysis/tree/645-pathology-feedback/analyses/molecular-subtyping-embryonal)

## Set up

```{r}
# This logic is the same as analyses/gene-set-enrichment-analysis/02-model-gsea.Rmd
# Assigning params$is_ci to running_in_ci avoids a locked binding error
running_in_ci <- params$is_ci

# Are we testing? In case of a non 0/1 number, we recast as logical, and then 
# ensure logical.
if (running_in_ci %in% c(0,1)) running_in_ci <- as.logical(running_in_ci)
if (!(is.logical(running_in_ci)))
{
  stop("\n\nERROR: The parameter `is_ci` should be FALSE/TRUE (or 0/1).")
}
```

### Libraries and functions

```{r}
library(tidyverse)
```

```{r}
bind_exp_strategies <- function(results_df) {
  # Many of the molecular subtyping tables contain a 
  # Kids_First_Biospecimen_ID_RNA and Kids_First_Biospecimen_ID_DNA column,
  # this function separates those out and binds the rows to mirror how 
  # pbta-histologies.tsv is organized
  #
  # Args:
  #   results_df: a molecular subtyping table with Kids_First_Biospecimen_ID_RNA
  #               and Kids_First_Biospecimen_ID_DNA columns
  # Returns:
  #   A data.frame where each biospecimen is a row
  rna_df <- results_df %>%
    select(-Kids_First_Biospecimen_ID_DNA) %>%
    filter(!is.na(Kids_First_Biospecimen_ID_RNA)) %>%
    rename(Kids_First_Biospecimen_ID = Kids_First_Biospecimen_ID_RNA)
  
  dna_df <- results_df %>%
    select(-Kids_First_Biospecimen_ID_RNA) %>%
    filter(!is.na(Kids_First_Biospecimen_ID_DNA)) %>%
    rename(Kids_First_Biospecimen_ID = Kids_First_Biospecimen_ID_DNA)
  
  # sometimes there will be multiple RNA-seq assays per single DNA-seq assay
  # or vice versa, which is why this distinct is necessary
  return(bind_rows(rna_df, dna_df) %>% distinct())
}
```

### Directories and files

```{r}
data_dir <- file.path("..", "..", "data")
analyses_dir <- ".."

# directories for upstream subtyping modules
ews_dir <- file.path(analyses_dir, "molecular-subtyping-EWS")
hgg_dir <- file.path(analyses_dir, "molecular-subtyping-HGG")
lgat_dir <- file.path(analyses_dir, "molecular-subtyping-LGAT")
embryonal_dir <- file.path(analyses_dir, "molecular-subtyping-embryonal")

# the folder that contains the tabular results is standardized across modules
results_dir <- "results"
# create the results directory for this module if it does not yet exist
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

#### Input files

When we run this locally, we want to tie it to a specific version of the histologies file _prior_ to feedback we received from pathology that is recorded on [#609](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/609).

```{r}
if (running_in_ci) {
  histologies_file <- file.path(data_dir, "pbta-histologies.tsv")
} else {
  histologies_file <- file.path(data_dir, "release-v15-20200228",
                                "pbta-histologies.tsv")
}
```

Results files from individual modules.

```{r}
ews_results_file <- file.path(ews_dir, results_dir, "EWS_results.tsv")
hgg_results_file <- file.path(hgg_dir, results_dir, "HGG_molecular_subtype.tsv")
lgat_results_file <- file.path(lgat_dir, results_dir, "lgat_subtyping.tsv")
embryonal_results_file <- file.path(embryonal_dir, results_dir,
                                    "embryonal_tumor_molecular_subtypes.tsv")
```

#### Output file

```{r}
output_file <- file.path(results_dir, "compiled_molecular_subtypes.tsv")
```

## Read in data

```{r message=FALSE}
histologies_df <- read_tsv(histologies_file, guess_max = 10000)
ews_results_df <- read_tsv(ews_results_file)
hgg_results_df <- read_tsv(hgg_results_file)
lgat_results_df <- read_tsv(lgat_results_file)
embryonal_results_df <- read_tsv(embryonal_results_file)
```

## Compile the subtyping resutls

### Handling non-ATRT/non-MB embryonal tumors

The molecular subtyping information from these tumors went into the v15 release, so we can use the `integrated_diagnosis`, `short_histology`, `broad_histology`, and `Notes` columns from the histologies file from that release.

```{r}
embryonal_results_df <- bind_exp_strategies(embryonal_results_df) %>%
  inner_join(select(histologies_df,
                    Kids_First_Biospecimen_ID,
                    integrated_diagnosis,
                    short_histology,
                    broad_histology,
                    Notes),
             by = "Kids_First_Biospecimen_ID")
```

### Handling EWS

The EWS results were post-v15 and come with their own `Notes` column.

```{r}
ews_results_df <- bind_exp_strategies(ews_results_df) %>%
  rename(integrated_diagnosis = integrated_diagnosis_reclassified,
         short_histology = short_histology_reclassified,
         broad_histology = broad_histology_reclassified) %>%
  select(-sample_type)
```

### Handling HGG

Like the non-ATRT/non-MB embryonal tumors, HGG subtyping was performed prior to v15.

```{r}
hgg_results_df <- bind_exp_strategies(hgg_results_df) %>%
  inner_join(select(histologies_df,
                    Kids_First_Biospecimen_ID,
                    integrated_diagnosis,
                    short_histology,
                    broad_histology,
                    Notes),
             by = "Kids_First_Biospecimen_ID")
```

### Handling LGAT

No columns that are disease labels have been changed yet.

```{r}
lgat_results_df <- bind_exp_strategies(lgat_results_df) %>%
  inner_join(select(histologies_df,
                    Kids_First_Biospecimen_ID,
                    integrated_diagnosis,
                    short_histology,
                    broad_histology,
                    Notes),
             by = "Kids_First_Biospecimen_ID")
```

### Handling EPN 

```
# TODO
```

### All results

Compile results and write to file

```{r}
all_results_df <- bind_rows(embryonal_results_df,
                            ews_results_df,
                            hgg_results_df,
                            lgat_results_df) %>%
  
  write_tsv(output_file)
```

## Session Info

```{r}
sessionInfo()
```

