---
title: "Update subtypes using pathology-free-text-diagnosis"
output: html_notebook
---

The samples in the files below have molecular-subtyping results which are already part of the compile file `analyses/molecular-subtyping-pathology/results/compiled_molecular_subtypes.ts` so we will be updating the values for these samples using outputs from the pathology-free-text-diagnosis terms.

File | Logic to include as results | Previous subtype | Description ticket 
-- | -- | -- | -- 
analyses/molecular-subtyping-pathology/results/lgat-pathology-free-text-subtypes.tsv #1060  | Update existing values | LGAT | #1000 
analyses/molecular-subtyping-pathology/results/cranio_adam_subtypes.tsv #823  | Update existing values | CRANIO | #994
analyses/molecular-subtyping-pathology/results/glialneuronal_tumor_subtypes.tsv  | Update existing values | LGAT | #1017 

The following subtypes can be directly added to `analyses/molecular-subtyping-pathology/results/compiled_molecular_subtypes.tsv` as well since these subtypes don't have molecular-subtyping results
 
File | Logic to include as results |  Description ticket 
 -- | -- | -- 
analyses/molecular-subtyping-pathology/results/cns-lymphoma-subtypes.tsv | Append | #1057  
analyses/molecular-subtyping-pathology/results/juvenile-xanthogranuloma-subtypes.tsv | Append | #1056 
analyses/molecular-subtyping-pathology/results/choroid_plexus_papilloma_subtypes.tsv  | Append | #1065

## Set Up

### Libraries

```{r}
# Pipes
library(magrittr)
library(tidyverse)
```

### Input

```{r}
data_dir <- file.path("..", "..", "data")
results_dir <- file.path("results")
histologies_file <- file.path(data_dir, "pbta-histologies-base.tsv")
compiled_subtypes_file <- file.path(results_dir, "compiled_molecular_subtypes.tsv")
meningioma_subtypes_file <- file.path(results_dir, "meningioma_subtypes.tsv")
lgat_recode_file <- file.path(results_dir, "lgat-pathology-free-text-subtypes.tsv")
glialnueronal_recode_file <- file.path(results_dir, "glialneuronal_tumor_subtypes.tsv")
cranio_adam_recode_file <- file.path(results_dir, "cranio_adam_subtypes.tsv")
cns_lymphoma_subtypes_file <- file.path(results_dir, "cns-lymphoma-subtypes.tsv")
jxg_subtypes_file <- file.path(results_dir, "juvenile-xanthogranuloma-subtypes.tsv")
choroid_plexus_papilloma_subtypes_file <- file.path(results_dir, "choroid_plexus_papilloma_subtypes.tsv")
```

### Output

```{r}
results_dir <- "results"
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
output_file <- file.path(results_dir, "compiled_molecular_subtypes_and_report_info.tsv")

```

## Read in data

```{r}
# clinical pbta histology
clinical <- readr::read_tsv(file.path(data_dir,
                                "pbta-histologies-base.tsv")) 
compiled_subtypes_df <- readr::read_tsv(compiled_subtypes_file) %>%  
  # ids to match
  left_join(clinical[,c("Kids_First_Participant_ID",
                        "Kids_First_Biospecimen_ID",
                        # need tumor_descriptor to match to clinical review
                        "tumor_descriptor")],by=c("Kids_First_Biospecimen_ID",
                                       "Kids_First_Participant_ID")) 

lgat_recode_df <- readr::read_tsv(lgat_recode_file)
meningioma_subtypes_df <- readr::read_tsv(meningioma_subtypes_file)
glialnueronal_recode_df <- readr::read_tsv(glialnueronal_recode_file)
cranio_adam_recode_df <- readr::read_tsv(cranio_adam_recode_file)
cns_lymphoma_subtypes_df <- readr::read_tsv(cns_lymphoma_subtypes_file) 
jxg_subtypes_df <- readr::read_tsv(jxg_subtypes_file)
choroid_plexus_papilloma_subtypes_df <- readr::read_tsv(choroid_plexus_papilloma_subtypes_file)

```

## Update subtypes

### Add meningioma subtypes
```{r}
compiled_subtypes_df <- compiled_subtypes_df %>%
  # add
  bind_rows(meningioma_subtypes_df)


```


### Recoded LGAT integrated_diagnosis 
```{r}
compiled_subtypes_df <- compiled_subtypes_df %>% 
  # remove old data
  dplyr::filter(!Kids_First_Biospecimen_ID %in% lgat_recode_df$Kids_First_Biospecimen_ID) %>%
  # replace 
  bind_rows(lgat_recode_df) 


```

### Neuronal and mixed neuronal-glial tumors updates in broad_histology, short_histology and harmonized_diagnosis
```{r}

compiled_subtypes_df <- compiled_subtypes_df %>% 
  # remove old data
  dplyr::filter(!Kids_First_Biospecimen_ID %in% glialnueronal_recode_df$Kids_First_Biospecimen_ID) %>%
  # replace 
  bind_rows(glialnueronal_recode_df)  %>%
  select(-pathology_diagnosis,-pathology_free_text_diagnosis)

```

### Recoding adamantinomatous craniopharyngiomas
```{r}

compiled_subtypes_df <- compiled_subtypes_df %>% 
  # remove old data
  dplyr::filter(!Kids_First_Biospecimen_ID %in% cranio_adam_recode_df$Kids_First_Biospecimen_ID) %>%
  # replace 
  bind_rows(cranio_adam_recode_df) 

```

## Append new subtypes from pathology-free-text-diagnosis

### CNS lymphoma harmonized_diagnosis update
```{r}

compiled_subtypes_df <- compiled_subtypes_df %>% 
  # append
  bind_rows(cns_lymphoma_subtypes_df) 

```

### Juvenile xanthogranuloma harmonized_diagnosis update
```{r}

compiled_subtypes_df <- compiled_subtypes_df %>% 
  # append
  bind_rows(jxg_subtypes_df) 

```


### Atypical choroid plexus papilloma harmonized_diagnosis update
```{r}

compiled_subtypes_df <- compiled_subtypes_df %>% 
  # append
  bind_rows(choroid_plexus_papilloma_subtypes_df) %>%
  select(-pathology_diagnosis,-pathology_free_text_diagnosis)

```


## Save to file
```{r}
compiled_subtypes_df %>% 
  arrange(Kids_First_Biospecimen_ID)%>%
  write_tsv(output_file)

```

