---
title: "Incorporating CHOP pathology input to molecular subtyping calls"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Jaclyn Taroni for ALSF CCDL
date: 2020
params:
  is_ci: FALSE
---

_Authorship above refers to this notebook **only**._

## Background

As part of this project, we have undertaken several analyses that use molecular data to subtype or reclassify biospecimens. 
In some cases, our analyses have resulted in an update of the `integrated_diagnosis` field included in the clinical file (`pbta-histologies.tsv` [[doc](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/master/doc/data-formats.md#data-caveats)]).
However, not all logic included in the code of the repository was ultimately in agreement with feedback from pathologists at CHOP.
In this notebook, we will make changes and document the cases where the logic in the modules of the repository deviate from the final labels.

## Pathology Feedback

Here, we're copying the table from [#606 (comment)](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/609#issuecomment-602821376) here:

| Kids_First_Biospecimen_ID | glioma_brain_region | sample_id | aliquot_id | Kids_First_Participant_ID | experimental_strategy | pathology_diagnosis                              | integrated_diagnosis                     | Notes from Path inspection                                                                  | Action                                                                                 | Given to Rita 3/13/20                                                                                      | short_histology | broad_histology                                                          | broad_composition | Notes                                                                 | cancer_predispositions | molecular_subtype   |
|---------------------------|---------------------|-----------|------------|---------------------------|-----------------------|--------------------------------------------------|------------------------------------------|---------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------|------------------------------------------------------------------------------------------------------------|-----------------|--------------------------------------------------------------------------|-------------------|-----------------------------------------------------------------------|------------------------|---------------------|
| BS_C6ZZ3FY5               | hemispheric         | 7316-1105 | 470017     | PT_X71D7G5S               | RNA-Seq               | Supratentorial or Spinal Cord PNET               | High-grade glioma                        | No path report available                                                                    | reviewed - OK with new dx                                                              | 1 slide, pathology report                                                                                  | HGAT            | Diffuse astrocytic and oligodendroglial tumor                            | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | HGG, H3 G35         |
| BS_P8WN8XEQ               | hemispheric         | 7316-1105 | 549712     | PT_X71D7G5S               | WGS                   | Supratentorial or Spinal Cord PNET               | High-grade glioma                        | No path report available                                                                    | reviewed - OK with new dx                                                              | 1 slide, pathology report                                                                                  | HGAT            | Diffuse astrocytic and oligodendroglial tumor                            | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | HGG, H3 G35         |
| BS_23M72ABG               | NA                  | 7316-523  | 588343     | PT_AQWDQW27               | RNA-Seq               | Meningioma                                       | </s>Ependymoma</s> Meningioma                   | No path report available                                                                    | reviewed - keep as meningioma                                                          | 2 slides, pathology report; email request for radiology                                                    | tumor           | <s>Ependymal tumor</s>                                                          | tumor             |                                                                       | None documented        | <s>ST-EPN-YAP</s>          |
| BS_5JM573JC               | hemispheric         | 7316-2255 | 717166     | PT_6Q0NPVP3               | RNA-Seq               | Ganglioglioma                                    | Diffuse midline glioma                   | Path report ganglioglioma with extensive chondroid metaplasia                               | need to review radiology report; if not diffuse, ganglioglioma, H3 mut - cannot review | pathology report (no slides - from Pitt); email request for radiology - will not receive in the short term | HGAT            | Diffuse astrocytic and oligodendroglial tumor                            | tumor             | Updated via OpenPBTA subtyping                         | None documented        | DMG, H3 K28         |
| BS_E5H6CFYT               | hemispheric         | 7316-2255 | 711134     | PT_6Q0NPVP3               | WGS                   | Ganglioglioma                                    | Diffuse midline glioma                   | Path report ganglioglioma with extensive chondroid metaplasia                               | need to review radiology report; if not diffuse, ganglioglioma, H3 mut - cannot review | pathology report (no slides - from Pitt); email request for radiology - will not receive in the short term | HGAT            | Diffuse astrocytic and oligodendroglial tumor                            | tumor             | Updated via OpenPBTA subtyping                         | None documented        | DMG, H3 K28         |
| BS_BFDEZK1C               | other               | 7316-1102 | 549714     | PT_5BWZA0NT               | WGS                   | Supratentorial or Spinal Cord PNET               | Diffuse midline glioma                   | No path report available                                                                    | reviewed - OK with new dx                                                              | 3 slides, pathology report                                                                                 | HGAT            | Diffuse astrocytic and oligodendroglial tumor                            | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | DMG, H3 K28         |
| BS_MB7WN0ZB               | other               | 7316-1102 | 470031     | PT_5BWZA0NT               | RNA-Seq               | Supratentorial or Spinal Cord PNET               | Diffuse midline glioma                   | No path report available                                                                    | reviewed - OK with new dx                                                              | 3 slides, pathology report                                                                                 | HGAT            | Diffuse astrocytic and oligodendroglial tumor                            | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | DMG, H3 K28         |
| BS_H1XPVS9A               | hemispheric         | 7316-1106 | 549575     | PT_T8V9ES93               | RNA-Seq               | High-grade glioma/astrocytoma (WHO grade III/IV) | <s>High-grade glioma</s> Low-grade glioma       | No path report available                                                                    | reviewed - OK with new dx (PXA/LGG)                                                    | 2 slides, pathology report                                                                                 | HGAT            | <s>Diffuse astrocytic and oligodendroglial tumor</s> Low-grade astrocytic tumor | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | LGG, BRAF V600E          |
| BS_J4E9SW51               | hemispheric         | 7316-1106 | 549708     | PT_T8V9ES93               | WGS                   | High-grade glioma/astrocytoma (WHO grade III/IV) | <s>High-grade glioma</s> Low-grade glioma       | No path report available                                                                    | reviewed - OK with new dx (PXA/LGG)                                                    | 2 slides, pathology report                                                                                 | HGAT            | <s>Diffuse astrocytic and oligodendroglial tumor</s> Low-grade astrocytic tumor | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | LGG, BRAF V600E          |
| BS_J8VX4D17               | mixed               | 7316-506  | 588299     | PT_7E3V3JFX               | RNA-Seq               | Ependymoma                                       | <s>Diffuse midline glioma</s> Ependymoma        | Path report posterior fossa ependymoma                                                      | reviewed - keep EPN with H3 mut- rare                                                  |                                                                                                            | HGAT            | <s>Diffuse astrocytic and oligodendroglial tumor</s> Ependymal tumor            | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | <s>DMG, H3 K28</s>         |
| BS_G16NM5WJ               | mixed               | 7316-506  | 588011     | PT_7E3V3JFX               | WGS                   | Ependymoma                                       | <s>Diffuse midline glioma</s> Ependymoma        | Path report posterior fossa ependymoma                                                      | reviewed - keep EPN with H3 mut- rare                                                  |                                                                                                            | HGAT            | <s>Diffuse astrocytic and oligodendroglial tumor</s> Ependymal tumor            | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | <s>DMG, H3 K28</s>         |
| BS_KSKZ9J7J               | NA                  | 7316-765  | 470731     | PT_80NVYCBS               | RNA-Seq               | Ependymoma                                       | CNS Embryonal Tumor                      | Path report anaplastic ependymoma                                                           | reviewed - OK with new dx                                                              |                                                                                                            | Embryonal Tumor | Embryonal tumor                                                          | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | CNS HGNET-MN1       |
| BS_K07KNTFY               | NA                  | 7316-272  | 588001     | PT_00G007DM               | WGS                   | Medulloblastoma                                  | Embryonal tumor with multilayer rosettes | Path report had "medulloblastoma" handwritten, but final dx ETMR with multilayer rosettes   | reviewed - OK with new dx                                                              |                                                                                                            | ETMR            | Embryonal tumor                                                          | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | ETMR, C19MC-altered |
| BS_QWNBZ9RJ               | NA                  | 7316-272  | 588287     | PT_00G007DM               | RNA-Seq               | Medulloblastoma                                  | Embryonal tumor with multilayer rosettes | Path report had "medulloblastoma" handwritten, but final dx ETMR with multilayer rosettes   | reviewed - OK with new dx                                                              |                                                                                                            | ETMR            | Embryonal tumor                                                          | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | ETMR, C19MC-altered |
| BS_P39SQPTS               | NA                  | 7316-447  | 588300     | PT_01MZ62KG               | RNA-Seq               | Pineoblastoma                                    | Embryonal tumor with multilayer rosettes | Path report PNET/pineoblastoma; "there are scattered true rosettes (Flexner-Wintersteiner)" | reviewed - OK with new dx                                                              |                                                                                                            | ETMR            | Embryonal tumor                                                          | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | ETMR, C19MC-altered |
| BS_TE8QFF7T               | NA                  | 7316-447  | 588012     | PT_01MZ62KG               | WGS                   | Pineoblastoma                                    | Embryonal tumor with multilayer rosettes | Path report PNET/pineoblastoma; "there are scattered true rosettes (Flexner-Wintersteiner)" | reviewed - OK with new dx                                                              |                                                                                                            | ETMR            | Embryonal tumor                                                          | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | ETMR, C19MC-altered |
| BS_N6N147BY               | NA                  | 7316-238  | 232096     | PT_X648RVMK               | RNA-Seq               | High-grade glioma/astrocytoma (WHO grade III/IV) | Embryonal tumor with multilayer rosettes | Path report handwritten "high grade glioma"; final dx infiltrating glioneuronal neoplasm    | reviewed - OK with new dx (ETMR original tumor 7316-2975 and this one consistent)      |                                                                                                            | ETMR            | Embryonal tumor                                                          | tumor             | Updated via OpenPBTA subtyping                                        | None documented        | ETMR, C19MC-altered |
| BS_5P1TN10Z               | hemispheric         | 7316-158  | 746061     | PT_9BZETM0M               | RNA-Seq               | Supratentorial or Spinal Cord PNET               | High-grade glioma                        | Path report PNET                                                                            | reviewed - OK with new dx grade III-IV                                                 |                                                                                                            | HGAT            | Diffuse astrocytic and oligodendroglial tumor                            | tumor             | from PNET to HGG due to H3F3A mutation;Updated via OpenPBTA subtyping | None documented        | HGG, H3 G35         |
| BS_STNH7YSX               | hemispheric         | 7316-158  | 746056     | PT_9BZETM0M               | WGS                   | Supratentorial or Spinal Cord PNET               | High-grade glioma                        | Path report PNET                                                                            | reviewed - OK with new dx grade III-IV                                                 |                                                                                                            | HGAT            | Diffuse astrocytic and oligodendroglial tumor                            | tumor             | from PNET to HGG due to H3F3A mutation;Updated via OpenPBTA subtyping | None documented        | HGG, H3 G35         |
| BS_TV5B86ZD               | hemispheric         | 7316-158  | 150593     | PT_9BZETM0M               | RNA-Seq               | Supratentorial or Spinal Cord PNET               | High-grade glioma                        | Path report PNET                                                                            | reviewed - OK with new dx grade III-IV                                                 |                                                                                                            | HGAT            | Diffuse astrocytic and oligodendroglial tumor                            | tumor             | from PNET to HGG due to H3F3A mutation;Updated via OpenPBTA subtyping | None documented        | HGG, H3 G35         |

And the notes:

> #### Few notes:
> 1. `PT_7E3V3JFX` specimens were consistent with the original EPN dx, so pathology would call this a rare EPN, H3 K28 mutated tumor, rather than DMG.
> 2. `PT_AQWDQW27` specimen was consistent with meningioma, even though it has a hallmark EPN fusion, so pathology would also call this a rare meningioma with a _YAP1_ fusion.
> 3. Because 1 is a rare tumor (maybe first seen), the logic of searching for all H3 K28 mutations in [HGG subtyping](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/249) would convert this sample by default - how to handle this? 
>  4. Pathology confirmed this HGG BRAF V600E mutant tumor, [`BS_H1XPVS9A`](https://cbethell.github.io/open-pbta-output/09-HGG-with-braf-clustering.nb.html#identify_sample_that_clusters_with_lgat), to be a LGAT (PXA). I updated `molecular_subtype` here based on what it would look like, but this should come through via the LGAT [subtyping](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/631) ticket. How should we add this info?

## Set up 

```{r}
# This logic is the same as analyses/gene-set-enrichment-analysis/02-model-gsea.Rmd
# Assigning params$is_ci to running_in_ci avoids a locked binding error
running_in_ci <- params$is_ci

# Are we testing? In case of a non 0/1 number, we recast as logical, and then 
# ensure logical.
if (running_in_ci %in% c(0,1)) running_in_ci <- as.logical(running_in_ci)
if (!(is.logical(running_in_ci)))
{
  stop("\n\nERROR: The parameter `is_ci` should be FALSE/TRUE (or 0/1).")
}
```

```{r warning=FALSE}
library(tidyverse)
```

### Directory and files

#### Directories

```{r}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
results_dir <- "results"
```

#### Input

When we run this locally, we want to tie it to a specific version of the histologies file _prior_ to feedback we received from pathology that is recorded on [#609](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/609).

```{r}
if (running_in_ci) {
  histologies_file <- file.path(data_dir, "pbta-histologies.tsv")
} else {
  histologies_file <- file.path(data_dir, "release-v15-20200228",
                                "pbta-histologies.tsv")
}
```

```{r}
compiled_results_file <- file.path(results_dir, 
                                   "compiled_molecular_subtypes.tsv")
fusions_file <- file.path(data_dir,
                          "pbta-fusion-putative-oncogenic.tsv")
```

#### Output

```{r}
output_file <- file.path(results_dir,
                         "compiled_molecular_subtypes_with_pathology_feedback.tsv")
```

## Read in data

```{r message=FALSE}
histologies_df <- read_tsv(histologies_file, guess_max = 10000)
compiled_df <- read_tsv(compiled_results_file)
fusions_df <- read_tsv(fusions_file)
```

## Revise subtyping calls based on pathology feedback

### `PT_7E3V3JFX`

`PT_7E3V3JFX` was classified in `molecular-subtyping-HGG` as `DMG, H3 K28` because the first step in that module was to identify any samples in the cohort with "defining lesions:" H3 K28 or H3 G35 (original subtyping issue: [#249](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/249)).

```{r}
compiled_df %>%
  filter(Kids_First_Participant_ID == "PT_7E3V3JFX")
```

The relevant notes above are

> 1. `PT_7E3V3JFX` specimens were consistent with the original EPN dx, so pathology would call this a rare EPN, H3 K28 mutated tumor, rather than DMG.
> 3. Because 1 is a rare tumor (maybe first seen), the logic of searching for all H3 K28 mutations in [HGG subtyping](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/249) would convert this sample by default - how to handle this? 

So we will revise the `molecular_subtype`, `integrated_diagnosis`, `short_histology`, and `broad_histology` accordingly.

```{r}
compiled_df <- compiled_df %>%
  mutate(
    integrated_diagnosis = case_when(
      Kids_First_Participant_ID == "PT_7E3V3JFX" ~ "Ependymoma",
      TRUE ~ integrated_diagnosis 
    ),
    short_histology = case_when(
      Kids_First_Participant_ID == "PT_7E3V3JFX" ~ "Ependymoma",
      TRUE ~ short_histology
    ),
    broad_histology = case_when(
      Kids_First_Participant_ID == "PT_7E3V3JFX" ~ "Ependymal tumor",
      TRUE ~ broad_histology
    ),
    molecular_subtype = case_when(
      Kids_First_Participant_ID == "PT_7E3V3JFX" ~ "EPN, H3 K28",
      TRUE ~ molecular_subtype
    )
  )
```

### `BS_H1XPVS9A`

`BS_H1XPVS9A` is from a sample with a BRAF V600E mutation that is H3 wildtype.
This sample was originally labeled as a HGG.
In low-dimensional transcriptomic space (e.g., UMAP, t-SNE), `BS_H1XPVS9A` clusters with LGG samples.
(See `molecular-subtyping-HGG`.)

The relevant note from above:

>  4. Pathology confirmed this HGG BRAF V600E mutant tumor, `BS_H1XPVS9A`
to be a LGAT (PXA). I updated `molecular_subtype` here based on what it would look like, but this should come through via the LGAT [subtyping](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/631) ticket. How should we add this info?

The steps for LGAT subtyping are to classify LGG samples on the basis of the presence/absence of _BRAF_ fusions and BRAF V600E mutations.

```{r}
compiled_df %>%
  filter(Kids_First_Biospecimen_ID == "BS_H1XPVS9A") %>%
  select(Kids_First_Participant_ID, sample_id) %>%
  left_join(compiled_df)
```

We know from `molecular-subtyping-HGG` that a BRAF V600E mutation is present.
Now we have to check for the presence of _BRAF_ fusions.

```{r}
fusions_df %>%
  filter(str_detect(FusionName, "BRAF"),
         Sample == "BS_H1XPVS9A")
```

There are no _BRAF_ fusions in `BS_H1XPVS9A`, so we are able to classify this sample (`7316-1106`) as `LGG, BRAF V600E`.

```{r}
compiled_df <- compiled_df %>%
  mutate(
    integrated_diagnosis = case_when(
      sample_id == "7316-1106" ~ "Low-grade astrocytic tumor",
      TRUE ~ integrated_diagnosis
    ),
    short_histology = case_when(
      sample_id == "7316-1106" ~ "LGAT",
      TRUE ~ short_histology
    ),
    broad_histology = case_when(
      sample_id == "7316-1106" ~ "Low-grade astrocytic tumor",
      TRUE ~ broad_histology
    ),
    molecular_subtype = case_when(
      sample_id == "7316-1106" ~ "LGG, BRAF V600E",
      TRUE ~ molecular_subtype
    ),
    Notes = case_when(
      sample_id == "7316-1106" ~ "Updated via OpenPBTA subtyping",
      TRUE ~ Notes
    )
  )
```

### HGG BRAF V600E

The follow point comes from another issue [#627 (comment)](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/627#issuecomment-598789232):

> We have a standalone BRAF V600E subtype, but this is HGG only, and should be labeled as such (will create an update HGG analysis ticket).

```{r}
compiled_df %>%
  filter(molecular_subtype == "BRAF V600E")
```

Now that we have addressed the one sample that should have been reclassified as LGG, we are able to update `molecular_subtype` to `HGG, BRAF V600E` for the remaining samples.

```{r}
compiled_df <- compiled_df %>%
  mutate(molecular_subtype = case_when(
    molecular_subtype == "BRAF V600E" ~ "HGG, BRAF V600E",
    TRUE ~ molecular_subtype
  ))
```

### `PT_AQWDQW27`

> 2. `PT_AQWDQW27` specimen was consistent with meningioma, even though it has a hallmark EPN fusion, so pathology would also call this a rare meningioma with a _YAP1_ fusion.

The `molecular-subtyping-EPN` module has not been completed yet, but the logic that is in that module may mean that we need to include revising the labels of `PT_AQWDQW27`.

```
# TODO: do we need to update PT_AQWDQW27 once molecular-subtyping-EPN is 
# complete?
```

### Write revised table to file

```{r}
write_tsv(compiled_df, output_file)
```

## Session Info

```{r}
sessionInfo()
```

