---
title: "Using WHO 2016 CNS subtypes to improve CNS lymphoma harmonized diagnosis"
output: 
  html_notebook:
    toc: true
    toc_float: true
author: JN Taroni for ALSF CCDL (code) ; K Gaonkar updated for CNS lymphoma
date: 2021
---

CNS lymphoma have subtypes per the [WHO 2016 CNS subtypes](https://link.springer.com/content/pdf/10.1007/s00401-016-1545-1.pdf). 
However, these are not captured in our molecular data.
Instead, we can use the pathology free text information in the histologies file to further classify the CNS lymphoma harmonized diagnosis information.
We will use this notebook to do so; see [#999](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/999) for more information.

## Set Up

### Libraries

```{r}
# Pipes
library(magrittr)
library(tidyverse)
```

### Input

```{r}
data_dir <- file.path("..", "..", "data")
results_dir <- file.path("results")
histologies_file <- file.path(data_dir, "pbta-histologies-base.tsv")
```

### Output

```{r}
results_dir <- "results"
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
output_file <- file.path(results_dir, "cns-lymphoma-subtypes.tsv")
```

## Read in data

```{r}
histologies_df <- readr::read_tsv(histologies_file, guess_max = 10000)
```

### Display `pathology_free_text_diagnosis` values

```{r}
histologies_df %>%
  dplyr::filter(pathology_diagnosis == "Primary CNS lymphoma") %>%
  dplyr::select(pathology_free_text_diagnosis) %>%
  dplyr::distinct()
```

## Recoding `harmonized_diagnosis` 

We use the following table from [#999](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/999) to recode the `harmonized_diagnosis` values for samples with a `pathology_diagnosis` of CNS lymphoma:

pathology_diagnosis | subtyping module | pathology_free_text_diagnosis | broad_histology | short_histology | harmonized_diagnosis
-- | -- | -- | -- | -- | --
Primary CNS lymphoma | NA | burkitt's lymphoma | Lymphoma | CNS lymphoma | CNS Burkitt's lymphoma




```{r}
lymphoma_df <- histologies_df %>%
  # We are only concerned with samples where the pathology_free_text_diagnosis
  # contains Primary CNS lymphoma
  dplyr::filter(pathology_diagnosis == "Primary CNS lymphoma") %>%
  # Retain only relevant identifier and disease label columns
  dplyr::select(Kids_First_Biospecimen_ID,
                Kids_First_Participant_ID,
                sample_id,
                tumor_descriptor,
                pathology_free_text_diagnosis,
                broad_histology,
                short_histology) %>%
  # To smooth the way for string detection for the pathology free text, we
  # add a column where all of the text is converted to lowercase
  dplyr::mutate(pathology_free_text_dx_lower = 
                  stringr::str_to_lower(pathology_free_text_diagnosis)) %>%
  # String detection in pathology free text per the table
  dplyr::mutate(
    harmonized_diagnosis = dplyr::case_when(
      stringr::str_detect(pathology_free_text_dx_lower,
                          "burkitt's lymphoma") ~ "CNS Burkitt's lymphoma"
    ),
    broad_histology = dplyr::case_when(
      stringr::str_detect(pathology_free_text_dx_lower,
                          "burkitt's lymphoma") ~ "Lymphoma",
      TRUE ~ broad_histology
    ),
    short_histology = dplyr::case_when(
      stringr::str_detect(pathology_free_text_dx_lower,
                          "burkitt's lymphomag") ~ "CNS lymphoma",
      TRUE ~ short_histology
    ),
    Notes = "Updated via OpenPBTA subtyping from pathology_free_text_diagnosis"
  ) %>%
  # Drop the column we added for convenience of string detection
  # and to format to match compiled_mol_subtypes_pathology_clinical_df
  dplyr::select(-pathology_free_text_dx_lower,
                -pathology_free_text_diagnosis)
```

Write to file!

```{r}
lymphoma_df %>%
  arrange(Kids_First_Biospecimen_ID) %>%
  readr::write_tsv(output_file)
```

## Session Info

```{r}
sessionInfo()
```