---
title: "Recoding adamantinomatous craniopharyngiomas"
output: 
  html_notebook:
    toc: true
    toc_float: true
author: JN Taroni for ALSF CCDL (code)
date: 2021
---

_Background adapted from [#994](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/994)_

There are Craniopharyngioma samples which may have been annotated as "To be classified" in `molecular-subtyping-CRANIO` because they lack canonical mutations.
However, for adamantinomatous craniopharyngioma, the b-catenin SNV is not present in all samples ([ref](https://doi.org/10.1093/jnen/nlw116)):

> In our cohort of [adamantinomatous craniopharyngiomas] specimens from 117 patients we found _CTNNB1_ mutations in 89 cases (76.1%).

There are samples described as `Adamantinomatous` in pathology reports, so we can update the `harmonized_diagnosis` and `molecular_subtype` information accordingly.

## Set up

### Libraries

```{r}
library(tidyverse)
```

### Input

```{r}
data_dir <- file.path("..", "..", "data")
results_dir <- "results"
histologies_file <- file.path(data_dir, "pbta-histologies-base.tsv")
compiled_subtypes_file <- file.path(results_dir, "compiled_molecular_subtypes.tsv")
```

### Output

```{r}
output_file <- file.path(results_dir, "cranio_adam_subtypes.tsv")
```

## Read in data

```{r}
histologies_df <- read_tsv(histologies_file)
```

```{r}
subtypes_df <- read_tsv(compiled_subtypes_file,
                        guess_max = 10000)
```

### Samples to be reclassified

The instructions on [#994](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/994) are to use specific sample identifiers to do the reclassification because it is on the basis of pathology report review.

```{r}
joined_df <- histologies_df %>%
  select(Kids_First_Biospecimen_ID,
         Kids_First_Participant_ID,
         sample_id,
         sample_type,  # required to filter out normal samples
         pathology_diagnosis,
         pathology_free_text_diagnosis,
         tumor_descriptor) %>%
  inner_join(subtypes_df,
             by = c("Kids_First_Biospecimen_ID", 
                    "Kids_First_Participant_ID",
                    "sample_id"))
```

Here's the same filtering steps that are performed on the issue itself that we'll save in a new data frame.

```{r}
acp_df <- joined_df %>%
  # Same logic as on the issue!
  filter(pathology_diagnosis == "Craniopharyngioma",
         molecular_subtype == "CRANIO, To be classified",
         str_detect(str_to_lower(pathology_free_text_diagnosis), 
                    "adamantinomatous")) %>%
  select(sample_id,
         pathology_free_text_diagnosis,
         molecular_subtype) %>%
  distinct()

acp_df
```

We can pull the `sample_id` values out and use that in our next steps.

```{r}
sample_ids_reclassification <- acp_df %>%
  pull(sample_id)
```

## Recode adamantinomatous craniopharyngiomas

We can use the following table from [#994](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/994) to guide how we recode the labels for these samples.

| broad_histology         | short_histology   | harmonized_diagnosis               | molecular_subtype |
|-------------------------|-------------------|------------------------------------|-------------------|
| Tumors of sellar region | Craniopharyngioma | Adamantinomatous craniopharyngioma | CRANIO, ADAM      |

```{r}
cranio_adam_df <- joined_df %>% 
  filter(sample_id %in% sample_ids_reclassification,
         # Exclude normal samples
         sample_type != "Normal") %>%
  # Filter to relevant ID and disease type label columns
  select(Kids_First_Biospecimen_ID,
         Kids_First_Participant_ID,
         sample_id,
         broad_histology,
         short_histology,
         molecular_subtype,
         tumor_descriptor) %>%
  # Code the values that are in the table above
  mutate(
    broad_histology = "Tumors of sellar region",
    short_histology = "Craniopharyngioma",
    harmonized_diagnosis = "Adamantinomatous craniopharyngioma",
    molecular_subtype = "CRANIO, ADAM",
    Notes = "Updated via OpenPBTA subtyping from pathology_free_text_diagnosis"
  )
```

Write to file!

```{r}
write_tsv(cranio_adam_df, output_file)
```

## Session Info

```{r}
sessionInfo()
```

