---
title: "Using WHO 2016 CNS subtypes to improve LGAT harmonized diagnosis"
output: 
  html_notebook:
    toc: true
    toc_float: true
author: JN Taroni for ALSF CCDL (code) ; K Gaonkar updated for LGAT
date: 2021
---

CNS lgat have subtypes per the [WHO 2016 CNS subtypes](https://link.springer.com/content/pdf/10.1007/s00401-016-1545-1.pdf). 
However, these are not captured in our molecular data.
Instead, we can use the pathology free text information in the histologies file to further classify the CNS lgat harmonized diagnosis information.
We will use this notebook to do so; see [#1000](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/1000) for more information.

## Set Up

### Libraries

```{r}
# Pipes
library(magrittr)
library(tidyverse)
```

### Input

```{r}
data_dir <- file.path("..", "..", "data")
results_dir <- file.path("results")
histologies_file <- file.path(data_dir, "pbta-histologies-base.tsv")
compiled_subtypes_file <- file.path(results_dir, "compiled_molecular_subtypes.tsv")
```

### Output

```{r}
results_dir <- "results"
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
output_file <- file.path(results_dir, "lgat-pathology-free-text-subtypes.tsv")
```

## Read in data

```{r}
histologies_df <- readr::read_tsv(histologies_file, guess_max = 10000,
                                  col_types = readr::cols(
                                    molecular_subtype = readr::col_character(),
                                    integrated_diagnosis = readr::col_character(),
                                    Notes = readr::col_character())) %>%
  filter(sample_type != "Normal") %>%
    select(Kids_First_Biospecimen_ID,
         Kids_First_Participant_ID,
         sample_id, 
         pathology_diagnosis,
         pathology_free_text_diagnosis,
         tumor_descriptor,
         Notes) 
  
compiled_subtypes_df <- readr::read_tsv(compiled_subtypes_file)
```

### Display `pathology_free_text_diagnosis` values

```{r}
compiled_subtypes_df <- compiled_subtypes_df %>%
  left_join(histologies_df) %>%
  dplyr::filter(pathology_diagnosis == "Low-grade glioma/astrocytoma (WHO grade I/II)") 

compiled_subtypes_df %>%
  dplyr::select(pathology_free_text_diagnosis,integrated_diagnosis) %>%
  dplyr::distinct()
```

## Recoding `integrated_diagnosis` 

We use the following table from [#1000](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/1000) to recode the `harmonized_diagnosis` values for samples with a `pathology_diagnosis` of LGAT:

pathology_diagnosis | subtyping module | pathology_free_text_diagnosis | broad_histology | short_histology | integrated_diagnosis
-- | -- | -- | -- | -- | --
Low-grade glioma/astrocytoma (WHO grade I/II) | LGAT | contains "sega or subependymal giant cell astrocytoma" | Low grade astrocytic tumor | LGAT | Subependymal Giant Cell Astrocytoma, `molecular_subtype`
Low-grade glioma/astrocytoma (WHO grade I/II) | LGAT | contains "fibrillary" | Low grade astrocytic tumor | LGAT | Diffuse fibrillary astrocytoma, `molecular_subtype`
Low-grade glioma/astrocytoma (WHO grade I/II) | LGAT | gliomatosis cerebri, type 1, ia | Low grade astrocytic tumor | LGAT | Gliomatosis cerebri, `molecular_subtype`
Low-grade glioma/astrocytoma (WHO grade I/II) | LGAT | if contains "jpa" or "juvenile astrocytoma" or "pilocytic" or "pilocystic" (typo) or "pilomyxoid" but does not contain "fibrillary" | Low grade astrocytic tumor | LGAT | Pilocytic astrocytoma, `molecular_subtype`
Low-grade glioma/astrocytoma (WHO grade I/II) | LGAT | oligodendroglioma who ii | Diffuse astrocytic and oligodendroglial tumor | Oligodendroglioma | Oligodendroglioma, `molecular_subtype`
Low-grade glioma/astrocytoma (WHO grade I/II) | LGAT | if contains "pxa" or "pleomorphic xanthoastrocytoma" | Low grade astrocytic tumor | LGAT | Pleomorphic xanthoastrocytoma, `molecular_subtype`

```{r}

#' @description CNS tumors have subtypes as per the [WHO 2016 CNS subtypes](https://link.springer.com/content/pdf/10.1007/s00401-016-1545-1.pdf). 
#' However, these are not captured in our molecular data so would need to be updated by
#' searching for terms in the reported pathology_free_text_diagnosis column in OpenPBTA histology file.
#' recode_integrated_diagnosis() can be used to recode a integreated_diagnosis based on the
#' pathology_free_text_diagnosis in addition to molecular subtypes.
#' @param histologies_df : Dataframe with Kids_First_Biospecimen_ID,Kids_First_Participant_ID,
#' sample_id,tumor_descriptor,pathology_free_text_diagnosis,broad_histology,short_histology,
#' molecular_subtype,integrated_diagnosis
#' @param pathology_diagnois : A term to filter histologies_df to samples that will be recoded
#' @param include_path_free_text_dx_terms : Regex term to match with values 
#' in pathology_free_text_diagnosis in histologies_df to be included in the recoding analysis
#' @param exclude_path_free_text_dx_terms : Default = NULL; Regex term to match with values in 
#' pathology_free_text_diagnosis in histologies_df to be excluded from the recoding analysis
#' @param old_integrated_diagnosis_term : Old term in integrated_diagnosis that should be replaced 
#' by replace_integrated_diagnosis_term.
#' @param replace_integrated_diagnosis_term : New term for integrated_diagnosis that will replace 
#' an old integrated_term, if value in integrated_diagnosis in input histologies_df is NA then 
#' replace_integrated_diagnosis_term will replace NA.


source(file.path("utils","recode_integrated_diagnosis.R"))

```

### Recoding sample with "sega|subependymal giant cell astrocytoma" in pathology_free_text_diagnosis

```{r}

lgat_sega_df <- recode_integrated_diagnosis(histologies_df = compiled_subtypes_df,
                                   pathology_diagnosis = "Low-grade glioma/astrocytoma (WHO grade I/II)",
                                   include_path_free_text_dx_terms = "sega|subependymal giant cell astrocytoma",
                                   old_integrated_diagnosis_term = "Low-grade glioma/astrocytoma",
                                   replace_integrated_diagnosis_term = "Subependymal Giant Cell Astrocytoma")
lgat_sega_df
```

### Recoding samples with "fibrillary" in pathology_free_text_diagnosis

```{r}
lgat_fibrillary_df <- recode_integrated_diagnosis(histologies_df = compiled_subtypes_df,
                                   pathology_diagnosis = "Low-grade glioma/astrocytoma (WHO grade I/II)",
                                   include_path_free_text_dx_terms = "fibrillary",
                                   old_integrated_diagnosis_term = "Low-grade glioma/astrocytoma",
                                   replace_integrated_diagnosis_term = "Diffuse fibrillary astrocytoma")
lgat_fibrillary_df
```


### Recoding samples with "gliomatosis cerebri, type 1, ia" in pathology_free_text_diagnosis
```{r}
lgat_gliomatosis_cerebri_df <- recode_integrated_diagnosis(histologies_df = compiled_subtypes_df,
                                   pathology_diagnosis = "Low-grade glioma/astrocytoma (WHO grade I/II)",
                                   include_path_free_text_dx_terms = "gliomatosis cerebri, type 1, ia",
                                   old_integrated_diagnosis_term = "Low-grade glioma/astrocytoma",
                                   replace_integrated_diagnosis_term = "Gliomatosis cerebri")

lgat_gliomatosis_cerebri_df
```

### Recoding samples with ""jpa" or "juvenile astrocytoma" or "pilocytic" or "pilocystic" (typo) or "pilomyxoid" but does not contain "fibrillary"" in pathology_free_text_diagnosis
```{r}
lgat_pilocytic_df <- recode_integrated_diagnosis(histologies_df = compiled_subtypes_df,
                                   pathology_diagnosis = "Low-grade glioma/astrocytoma (WHO grade I/II)",
                                   include_path_free_text_dx_terms = "jpa|juvenile astrocytoma|juvenile astocytoma|pilocytic|pilocystic|pilocytioc|pilomyxoid|piloxymoid",
                                   exclude_path_free_text_dx_term = "fibrillary",
                                   old_integrated_diagnosis_term = "Low-grade glioma/astrocytoma",
                                   replace_integrated_diagnosis_term = "Pilocytic astrocytoma")

lgat_pilocytic_df
```


### Recoding samples with "oligodendroglioma who ii" in pathology_free_text_diagnosis
```{r}
lgat_oligodendroglioma_df <- recode_integrated_diagnosis(histologies_df = compiled_subtypes_df,
                                   pathology_diagnosis = "Low-grade glioma/astrocytoma (WHO grade I/II)",
                                   include_path_free_text_dx_terms = "oligodendroglioma who ii",
                                   old_integrated_diagnosis_term = "Low-grade glioma/astrocytoma",
                                   replace_integrated_diagnosis_term = "Oligodendroglioma")

lgat_oligodendroglioma_df
```

### Recoding samples with "pxa" or "pleomorphic xanthoastrocytoma" in pathology_free_text_diagnosis
```{r}
lgat_pleomorphic_xanthoastrocytoma_df <- recode_integrated_diagnosis(histologies_df = compiled_subtypes_df,
                                   pathology_diagnosis = "Low-grade glioma/astrocytoma (WHO grade I/II)",
                                   include_path_free_text_dx_terms = "pleomorphic xanthroastrocytoma|pleomorphic xantoastrocytoma|pleomorphic xanthoastrocytoma|pxa",
                                   old_integrated_diagnosis_term = "Low-grade glioma/astrocytoma",
                                   replace_integrated_diagnosis_term = "Pleomorphic xanthoastrocytoma")

lgat_pleomorphic_xanthoastrocytoma_df
```


Add recoded LGAT to a new file

```{r}
bind_rows(lgat_sega_df,
          lgat_fibrillary_df,
          lgat_gliomatosis_cerebri_df,
          lgat_pilocytic_df,
          lgat_oligodendroglioma_df,
          lgat_pleomorphic_xanthoastrocytoma_df) %>%
  arrange(Kids_First_Biospecimen_ID) %>%
  readr::write_tsv(output_file)
```

## Session Info

```{r}
sessionInfo()
```