---
title: "Using WHO 2016 CNS subtypes to improve meningioma harmonized diagnosis"
output: 
  html_notebook:
    toc: true
    toc_float: true
author: JN Taroni for ALSF CCDL (code)
date: 2021
---

Meningiomas have subtypes per the [WHO 2016 CNS subtypes](https://link.springer.com/content/pdf/10.1007/s00401-016-1545-1.pdf). 
However, these are not captured in our molecular data.
Instead, we can use the pathology free text information in the histologies file to further classify the meningioma harmonized diagnosis information.
We will use this notebook to do so; see [#993](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/993) for more information.

## Set Up

### Libraries

```{r}
# Pipes
library(magrittr)
library(tidyverse)
```

### Input

```{r}
data_dir <- file.path("..", "..", "data")
results_dir <- file.path("results")
histologies_file <- file.path(data_dir, "pbta-histologies-base.tsv")
compiled_mol_subtypes_pathology_clinical_file <- file.path(results_dir,
                         "compiled_molecular_subtypes_with_clinical_pathology_feedback.tsv")
```

### Output

```{r}
results_dir <- "results"
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

output_file <- file.path(results_dir, "compiled_molecular_subtypes_with_clinical_pathology_feedback_and_report_info.tsv")
```

## Read in data

```{r}
histologies_df <- readr::read_tsv(histologies_file, guess_max = 10000)
compiled_mol_subtypes_pathology_clinical_df <- read_tsv(compiled_mol_subtypes_pathology_clinical_file, guess = 10000)
```

### Display `pathology_free_text_diagnosis` values

```{r}
histologies_df %>%
  dplyr::filter(pathology_diagnosis == "Meningioma") %>%
  dplyr::select(pathology_free_text_diagnosis) %>%
  dplyr::distinct()
```

## Recoding `harmonized_diagnosis` 

We use the following table from [#993](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/993) to recode the `harmonized_diagnosis` values for samples with a `pathology_diagnosis` of Meningioma:

| pathology_diagnosis | subtyping module | pathology_free_text_diagnosis                                                | broad_histology | short_histology | harmonized_diagnosis              |
|---------------------|------------------|------------------------------------------------------------------------------|-----------------|-----------------|-----------------------------------|
| Meningioma          | NA               | contains "atypical"                                                          | Meningioma      | Meningioma      | Atypical meningioma               |
| Meningioma          | NA               | contains "anaplastic"                                                        | Meningioma      | Meningioma      | Anaplastic (malignant) meningioma |
| Meningioma          | NA               | contains "clear cell meningioma"                                             | Meningioma      | Meningioma      | Clear cell meningioma             |
| Meningioma          | NA               | contains "meningothelial"                                                    | Meningioma      | Meningioma      | Meningothelial meningioma         |
| Meningioma          | NA               | does not contain "atypical", "anaplastic", "clear cell", or "meningothelial" | Meningioma      | Meningioma      | Meningioma                        |


```{r}
meningioma_df <- histologies_df %>%
  # We are only concerned with samples where the pathology diagnosis is
  # meningioma
  dplyr::filter(pathology_diagnosis == "Meningioma") %>%
  # Retain only relevant identifier and disease label columns
  dplyr::select(Kids_First_Biospecimen_ID,
                Kids_First_Participant_ID,
                sample_id,
                tumor_descriptor,
                pathology_free_text_diagnosis,
                broad_histology,
                short_histology) %>%
  # To smooth the way for string detection for the pathology free text, we
  # add a column where all of the text is converted to lowercase
  dplyr::mutate(pathology_free_text_dx_lower = 
                  stringr::str_to_lower(pathology_free_text_diagnosis)) %>%
  # String detection in pathology free text per the table
  dplyr::mutate(
    harmonized_diagnosis = dplyr::case_when(
      stringr::str_detect(pathology_free_text_dx_lower,
                          "atypical") ~ "Atypical meningioma",
      stringr::str_detect(pathology_free_text_dx_lower,
                          "anaplastic") ~ "Anaplastic (malignant) meningioma",
      stringr::str_detect(pathology_free_text_dx_lower,
                          "clear cell meningioma") ~ "Clear cell meningioma",
      stringr::str_detect(pathology_free_text_dx_lower,
                          "meningothelial") ~ "Meningothelial meningioma",
      # This will be true when none of the conditions (e.g., strings are 
      # detected) above
      TRUE ~ "Meningioma"
    ),
    Notes = "Updated via OpenPBTA subtyping from pathology_free_text_diagnosis"
  ) %>%
  # Drop the column we added for convenience of string detection
  # and to format to match compiled_mol_subtypes_pathology_clinical_df
  dplyr::select(-pathology_free_text_dx_lower,
                -pathology_free_text_diagnosis)
```

Write to file!
Since Meningioma was not subtyped as part of OpenPBTA molecular subtying we can append the meningioma results to `compiled_mol_subtypes_pathology_clinical_df`.

```{r}

meningioma_df %>%
  bind_rows(compiled_mol_subtypes_pathology_clinical_df) %>%
  arrange(Kids_First_Biospecimen_ID) %>%
  readr::write_tsv(output_file)

```

## Session Info

```{r}
sessionInfo()
```
