---
title: "Atypical choroid plexus papilloma"
output: 
  html_notebook:
    toc: true
    toc_float: true
author: JN Taroni for ALSF CCDL (code)
date: 2021
---

_Background adapted from [#997](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/997)_

Atypical choroid plexus papilloma is in the WHO 2016 CNS subtypes, but is not yet reflected in the `harmonized_diagnosis` in this project. 
We will use the `pathology_free_text_diagnosis` field to identify atypical choroid plexus papillomas.

## Set up

### Libraries

```{r}
library(tidyverse)
```

### Input

```{r}
data_dir <- file.path("..", "..", "data")
results_dir <- "results"
histologies_file <- file.path(data_dir, "pbta-histologies-base.tsv")
```

### Output

```{r}
output_file <- file.path(results_dir, "choroid_plexus_papilloma_subtypes.tsv")
```

## Read in data

```{r}
histologies_df <- read_tsv(histologies_file)
```

## Subtype choroid plexus papillomas

From [#997](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/997):

> If `pathology_diagnosis == Choroid plexus papilloma` and `pathology_free_text_diagnosis` contains "atypical", then annotate as below:

pathology_diagnosis | subtyping module | pathology_free_text_diagnosis | broad_histology | short_histology | harmonized_diagnosis
-- | -- | -- | -- | -- | --
Choroid plexus papilloma | NA | if contains "atypical" | Choroid plexus tumor | Choroid plexus tumor | Atypical choroid plexus papilloma

```{r}
cpp_df <- histologies_df %>%
  filter(pathology_diagnosis == "Choroid plexus papilloma") %>%
  # Subset to relevant ID and disease labels column
  select(Kids_First_Biospecimen_ID,
         Kids_First_Participant_ID,
         sample_id,
         pathology_diagnosis,
         pathology_free_text_diagnosis,
         broad_histology,
         short_histology,
         tumor_descriptor) %>%
  # For convenience, let's add a column where pathology_free_text_diagnosis
  # is all lower case out of an abundance of caution
  mutate(pathology_free_text_dx_lower = str_to_lower(pathology_free_text_diagnosis)) %>%
  # When we detect "atypical" in the free text, set harmonized diagnosis to 
  # Atypical choroid plexus papilloma
  mutate(harmonized_diagnosis = case_when(
    str_detect(pathology_free_text_dx_lower, "atypical") ~ "Atypical choroid plexus papilloma",
    TRUE ~ "Choroid plexus papilloma"
  )) %>%
  select(-pathology_free_text_dx_lower)
```

Write to file.

```{r}
write_tsv(cpp_df, output_file)
```

## Session Info

```{r}
sessionInfo()
```

