---
title: "Subtype Neurocytoma tumors as central or extra-ventricular"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Krutika Gaonkar for D3b
date: 2020
---

For, "Neurocytoma" samples detected in pathology_diagnosis, subtyping conditions as per issue [#805](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/805) are as follows :

 - If primary_site == "Ventricles", then molecular_subtype == CNC (central neurocytoma) 
 - If primary_site != "Ventricles", then molecular_subtype == EVN (extraventricular neurocytoma)

## Set up 

```{r}
library(tidyverse)
```

## Input directories and files

```{r}
data_dir <- file.path("..", "..", "data")
histologies_df <- read_tsv(file.path(data_dir, "pbta-histologies.tsv"))

```

## Output directory and file

```{r}
results_dir <- "results"
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
outputfile <- file.path(results_dir,"neurocytoma_subtyping.tsv")
```

Are there any Neurocytoma samples in `pathology_free_text_diagnosis` that are not listed in pathology_diagnosis?

```{r}

neurocytoma_from_pathology_free_text_diagnosis <- histologies_df %>%
  filter(str_detect(pathology_free_text_diagnosis, "neurocytoma")
                    ) 

neurocytoma_from_pathology_diagnosis <- histologies_df %>%
  filter(str_detect(pathology_diagnosis, "Neurocytoma")
                    ) 

neurocytoma_from_pathology_free_text_diagnosis %>%
  dplyr::filter(!Kids_First_Biospecimen_ID %in% neurocytoma_from_pathology_diagnosis$Kids_First_Biospecimen_ID )

```
There are no Kids_First_Biospecimen_ID in `pathology_free_text_diagnosis` that are not listed as "Neurocytoma" in  `pathology_diagnosis`. So will only use  `pathology_diagnosis` to look up "Neurocytoma" samples for subtyping


Checking if `pathology_diagnosis` has "Neurocytoma" what are all primary_site associated with it
```{r}
neurocytoma_from_pathology_diagnosis$primary_site %>% unique()
```

The values are only "Ventricles" and "Parietal Lobe" so we can directly check if primary_site== "Ventricles" ~ "CNC" and if primary_site != "Ventricles" ~ "EVN"

```{r}

neurocytoma_subtype <- neurocytoma_from_pathology_diagnosis %>% 
  dplyr::mutate(molecular_subtype= 
                  # primary_site == "Ventricles", then molecular_subtype == CNC (central neurocytoma) 
                  case_when(primary_site == "Ventricles" ~ "CNC",
                  # primary_site != "Ventricles", then molecular_subtype == EVN (extraventricular neurocytoma)       
                            primary_site != "Ventricles" ~ "EVN"
                            )
  ) %>%
  dplyr::select(Kids_First_Participant_ID,
                sample_id,
                Kids_First_Biospecimen_ID,
                molecular_subtype) %>%
  write_tsv(outputfile)

```
