---
title: 'Chromothripsis Plots by Histology'
author: "Laura Egolf"
date: "2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This notebook plots the proportion of tumors with chromothripsis across histology groups, with a breakdown by confidence level (high or low confidence chromothripsis) and by the number of chromothripsis regions observed. 

### Define directory paths and load packages

```{r}
# Set directory paths
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
analysis_dir <- file.path(root_dir, "analyses", "chromothripsis")
plots_dir <- file.path(analysis_dir, "plots", "03-histology")

# Create plots_dir if it doesn't exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}

# Load packages
library(ggplot2)
library(ggpubr)
library(RColorBrewer)
library(readr)
library(dplyr)
```


### Load chromothripsis data and histology display groups

```{r}
# Read in summary of chromothripsis regions per sample (number of regions for each confidence level)
chromoth_per_sample <- read_tsv(file.path(analysis_dir, "results", "chromothripsis_summary_per_sample.txt"))

# Import standard color palettes & histology display groups for project
histology_label_mapping <- read_tsv(
  file.path(root_dir, "figures", "palettes", "histology_label_color_table.tsv")) %>% 
  # Select just the columns we will need for plotting
  select(Kids_First_Biospecimen_ID, display_group, hex_codes, cancer_group, cancer_group_hex_codes)

# Merge chromothripsis data and histology display groups
chromoth_histology <- inner_join(chromoth_per_sample, histology_label_mapping, by="Kids_First_Biospecimen_ID")
```


### Summarize chromothripsis data by histology (display_group and cancer_group)
```{r}
## For each histology (display_group), calculate proportion of tumors with at least 
## one chromothripsis region (`prop`)
## This will be used in first plot, and to reorder display_group in subsequent plots
chromoth_histology_summary <- chromoth_histology %>%
  count(any_regions_logical, display_group, hex_codes) %>%
  tidyr::spread(key = any_regions_logical, value = n, fill=0) %>%
  group_by(display_group) %>%
  mutate(display_group_size = sum(`TRUE`, `FALSE`)) %>%
  mutate(prop = `TRUE` / display_group_size) %>%
  mutate(labels = paste0(`TRUE`, " / ", display_group_size)) %>%
  ungroup(display_group, hex_codes)

## Reorder display_group by proportion of tumors with chromothripsis:
# Reorder factor levels of `display_group` by `prop` in chromoth_histology_summary
chromoth_histology_summary <- chromoth_histology_summary %>%
  mutate(display_group = forcats::fct_reorder(display_group, prop, .desc = TRUE))
# Assign the same ordering of factor levels to chromoth_histology 
chromoth_histology <- chromoth_histology %>%
  mutate(display_group = factor(display_group, levels=levels(chromoth_histology_summary$display_group)))

### Repeat all of these steps for cancer_group instead of display_group
chromoth_histology_summary_cg <- chromoth_histology %>%
  count(any_regions_logical, cancer_group, cancer_group_hex_codes) %>%
  tidyr::spread(key = any_regions_logical, value = n, fill=0) %>%
  group_by(cancer_group) %>%
  mutate(cancer_group_size = sum(`TRUE`, `FALSE`)) %>%
  mutate(prop = `TRUE` / cancer_group_size) %>%
  mutate(labels = paste0(`TRUE`, " / ", cancer_group_size)) %>%
  ungroup(cancer_group, cancer_group_hex_codes) %>%
  filter(cancer_group_size >= 3 & !is.na(cancer_group)) # Only keep groups with >=3 tumors
# Reorder cancer_group based on proportion
chromoth_histology_summary_cg <- chromoth_histology_summary_cg %>%
  mutate(cancer_group = forcats::fct_reorder(cancer_group, prop))

```

### Define shared plot parameters
```{r}
# Define ylim for display_group plots based on the max value of `prop`
prop_max <- max(chromoth_histology_summary$prop)
prop_ylim <- ceiling(prop_max * 10) / 10  # Round up to nearest 0.1

# Set ggplot2 theme
theme_set(theme_pubr())

# Save ggplot2 options
plot_options <- list(
  ylim(c(0, prop_ylim)),
  xlab(NULL),
  ylab("Proportion of Tumors"),
  theme(axis.text.x = element_text(angle = 90, hjust=0.95))
)
```

### Plot proportion of tumors that have at least one chromothripsis region of any confidence level, broken down by display_group

Annotate with sample sizes for each group (number of tumors with chromothripsis / total number of tumors in that display_group)

```{r}
p <- ggplot(chromoth_histology_summary, aes(x = display_group, y = prop, fill = hex_codes)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label=labels), vjust=-0.2, size=2.5) + 
  scale_fill_identity() +
  plot_options
p
ggsave(file.path(plots_dir, "chromothripsis_proportion_by_displayGroup.png"), p, width=7, height=7)
```

### Plot proportion of tumors by cancer_group

Annotate with sample sizes for each group (number of tumors with chromothripsis / total number of tumors in that cancer_group). Only groups with >=3 total tumors are shown.

```{r}
p <- ggplot(chromoth_histology_summary_cg, aes(x = cancer_group, y = prop, fill = cancer_group_hex_codes)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label=labels), hjust=-0.2, size=2) + 
  scale_fill_identity() +
  coord_flip() + 
  xlab(NULL) + 
  ylab("Proportion of Tumors")
p
ggsave(file.path(plots_dir, "chromothripsis_proportion_by_cancerGroup.png"), p, width=7, height=7)
```

### Plot proportion of tumors by display_group, using fill to show the cancer_groups

This plot may be a little confusing since the proportions are calculated based on the total for each display_group - so the total proportion for the display_group is accurate, but the proportions for the cancer_groups are relative.

```{r}
# Calculate proportion of tumors with chromothripsis for each cancer_group and display_group, 
# out of the display_group total
chromoth_histology_summary_cg_by_dg <- chromoth_histology %>%
  count(any_regions_logical, display_group, cancer_group, cancer_group_hex_codes) %>%
  tidyr::spread(key = any_regions_logical, value = n, fill=0) %>%
  group_by(display_group) %>%
  mutate(display_group_size = sum(`TRUE`, `FALSE`)) %>%
  mutate(display_group_true_total = sum(`TRUE`)) %>%
  mutate(prop_cancer_group = `TRUE` / display_group_size) %>%
  mutate(prop_display_group = display_group_true_total / display_group_size) %>%
  mutate(labels = paste0(display_group_true_total, " / ", display_group_size)) %>%
  ungroup(display_group) %>%
  # Reorder factor levels based on the original proportion for display_group (calculated above)
  mutate(display_group = factor(display_group, levels=levels(chromoth_histology_summary$display_group)))

p <- ggplot(chromoth_histology_summary_cg_by_dg, aes(x = display_group, y = prop_cancer_group, 
                                                     fill = cancer_group_hex_codes)) + 
  geom_bar(position = "stack", stat="identity") +
  geom_text(aes(x=display_group, y=prop_display_group, label=labels), vjust=-0.2, size=2.5) + 
  scale_fill_identity() +
  plot_options
p
ggsave(file.path(plots_dir, "chromothripsis_proportion_by_displayGroup_fillByCancerGroup.png"), p, width=7, height=7)
```


### Plot proportion of tumors by display_group, using fill to show low- or high-confidence chromothripsis 

Proportion of tumors with at least one low- or high-confidence chromothripsis region (if a tumor has both low- and high-confidence regions, it will be grouped with high-confidence)

Note ggplot will output warning for 17 rows missing - these are the bars for "No Calls" which don't display anyway, so it doesn't matter

```{r}
# Scale alpha with confidence level - set "No Calls" as alpha=0 so the bar doesn't show
p <- ggplot(chromoth_histology, aes(x = display_group, 
                                   fill = hex_codes, 
                                   alpha = factor(any_regions, 
                                                  levels=c("No Calls", "Low Confidence", "High Confidence")))) +
  geom_bar(position = "fill") +
  scale_fill_identity() +
  scale_alpha_manual(values=c(0, 0.5, 1), name="Chromothripsis\nConfidence Level") +
  plot_options
p
ggsave(file.path(plots_dir, "chromothripsis_proportion_by_displayGroup_fillByConfidence.png"), p, width=7, height=7)
```

### Plot proportion of tumors by display_group, using fill to show number of chromothripsis regions per sample

Keep in mind that ShatterSeek only reports one chromothripsis region per chromosome, which is why the number of events per sample may seem low.

Note ggplot will output warning for 17 rows missing - these are the bars for "0" which don't display anyway, so it doesn't matter

```{r}
# Define color scale for # chromothripsis regions, but set "0" as transparent so the bar doesn't show
# Pick number of colors based on max number of chromothripsis regions observed
max_regions <- max(chromoth_per_sample$count_regions_any_conf)
colors <- colorRampPalette(brewer.pal(9,"YlOrRd"))(max_regions+1)
colors[1] <- "#1C00ff00"

p <- ggplot(chromoth_histology, aes(x = display_group, fill = as.factor(count_regions_any_conf))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values=colors, name="# Chromothripsis\nRegions") +
  plot_options
p
ggsave(file.path(plots_dir, "chromothripsis_proportion_by_displayGroup_fillByCount.png"), p, width=7, height=7)
```

