---
title: "Add ploidy column, status to CNVkit output"
output: html_notebook
author: J. Taroni for ALSF CCDL
date: 2019
---

The `pbta-histologies.tsv` file contains a `tumor_ploidy` column, which is tumor ploidy as inferred by ControlFreeC.
The copy number information should be interpreted in the light of this information (see: [current version of Data Formats section of README](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/390f1e08e481da5ec0b2c62d886d5fd298bbf017#data-formats)).

This notebook adds ploidy information to the CNVkit results and adds a status column that defines gain and loss broadly.

```{r}
library(dplyr)
```

### Read in data

```{r}
cnvkit_file <- file.path("..", "..", "data", "pbta-cnv-cnvkit.seg.gz")
cnvkit_df <- readr::read_tsv(cnvkit_file)
```

```{r}
histologies_file <- file.path("..", "..", "data", "pbta-histologies.tsv")
histologies_df <- readr::read_tsv(histologies_file)
```

### Add inferred ploidy information to CNVkit results

```{r}
add_ploidy_df <- histologies_df %>%
  select(Kids_First_Biospecimen_ID, tumor_ploidy) %>%
  inner_join(cnvkit_df, by = c("Kids_First_Biospecimen_ID" = "ID")) %>%
  select(-tumor_ploidy, everything())
```

### Add status column

This is intended to mirror the information contained in the ControlFreeC output.

```{r}
add_ploidy_df <- add_ploidy_df %>%
  mutate(status = case_when(
    # when the copy number is less than inferred ploidy, mark this as a loss
    copy.num < tumor_ploidy ~ "loss",
    # if copy number is higher than ploidy, mark as a gain
    copy.num > tumor_ploidy ~ "gain",
    copy.num == tumor_ploidy ~ "neutral"
  ))

head(add_ploidy_df, 10)
```

### Write to `scratch`

```{r}
output_file <- file.path("..", "..", "scratch", "cnvkit_with_status.tsv")
readr::write_tsv(add_ploidy_df, output_file)
```
