---
title: "Add ploidy column, status to CNVkit output"
output: html_notebook
author: J. Taroni for ALSF CCDL
date: 2019
---

The `histologies.tsv` file contains a `tumor_ploidy` column, which is tumor ploidy as inferred by ControlFreeC.
The copy number information should be interpreted in the light of this information (see: [current version of Data Formats section of README](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/390f1e08e481da5ec0b2c62d886d5fd298bbf017#data-formats)).

This notebook adds ploidy information to the CNVkit results and adds a status column that defines gain and loss broadly.

```{r}
library(dplyr)
```

### Read in data

```{r}
cnvkit_file <- file.path("..", "..", "data", "v8", "cnv-cnvkit.seg.gz")
cnvkit_df <- readr::read_tsv(cnvkit_file)
```

```{r}
histologies_file <- file.path("..", "..", "data", "v8", "histologies.tsv")
histologies_df <- readr::read_tsv(histologies_file, guess_max = 100000)
```


# Edits for https://github.com/PediatricOpenTargets/ticket-tracker/issues/177
# For PBTA, histologies file has germline_sex_estimate already assigned to WXS patients which have had WGS performed
# Updating here germline_sex_estimate for cohort=="TARGET" where is.na(germline_sex_estimate)
#Isolate IDs where germline_sex_estimate is NA and cohort == "TARGET"
germline_sex_estimate_na <- histologies_df %>% filter(cohort == "TARGET") %>% filter(is.na(germline_sex_estimate))
na_ids <- germline_sex_estimate_na %>% pull(Kids_First_Biospecimen_ID)

#Split the dataset into two groups, fix data for select IDs
histology_subset1 <- histologies_df %>% filter(!Kids_First_Biospecimen_ID %in% na_ids)

histology_subset2 <- germline_sex_estimate_na %>% mutate(germline_sex_estimate = reported_gender)

#Merge data after fixing germline_estimate with new value

histologies_df_final <- rbind(histology_subset1,histology_subset2)

#Verify counts match
#length(histologies_df$Kids_First_Biospecimen_ID)
#length(histology_subset1$Kids_First_Biospecimen_ID)
#length(histology_subset2$Kids_First_Biospecimen_ID)
#length(histologies_df_final$Kids_First_Biospecimen_ID)

### Add inferred ploidy information to CNVkit results

```{r}
add_ploidy_df <- histologies_df_final  %>%
  select(Kids_First_Biospecimen_ID, tumor_ploidy, germline_sex_estimate) %>%
  inner_join(cnvkit_df, by = c("Kids_First_Biospecimen_ID" = "ID")) %>%
  select(-tumor_ploidy, -germline_sex_estimate, everything())
```

### Add status column

This is intended to mirror the information contained in the ControlFreeC output.

```{r}
add_autosomes_df <- add_ploidy_df %>%
  # x and y chromosomes must be handled differently
  filter(!(chrom %in% c("chrX", "chrY"))) %>%
  mutate(status = case_when(
    # when the copy number is less than inferred ploidy, mark this as a loss
    copy.num < tumor_ploidy ~ "loss",
    # if copy number is higher than ploidy, mark as a gain
    copy.num > tumor_ploidy ~ "gain",
    copy.num == tumor_ploidy ~ "neutral"
  ))


# this logic is consistent with what is observed in the controlfreec file
# specifically, in samples where germline sex estimate = Female, X chromosomes
# appear to be treated the same way as autosomes

add_xy_df <- add_ploidy_df %>%
  filter(chrom %in% c("chrX", "chrY")) %>%  
  mutate(status = case_when(
    germline_sex_estimate == "Male" & copy.num < (0.5 * tumor_ploidy) ~ "loss",
    germline_sex_estimate == "Male" & copy.num > (0.5 * tumor_ploidy) ~ "gain",
    germline_sex_estimate == "Male" & copy.num == (0.5 * tumor_ploidy) ~ "neutral",
    # there are some instances where there are chromosome Y segments are in
    # samples where the germline_sex_estimate is Female
    chrom == "chrY" & germline_sex_estimate == "Female" & copy.num > 0 ~ "unknown",
    chrom == "chrY" & germline_sex_estimate == "Female" & copy.num == 0 ~ "neutral",
    # mirroring controlfreec, X treated same as autosomes
    chrom == "chrX" & germline_sex_estimate == "Female" & copy.num < tumor_ploidy ~ "loss",
    chrom == "chrX" & germline_sex_estimate == "Female" & copy.num > tumor_ploidy ~ "gain",
    chrom == "chrX" & germline_sex_estimate == "Female" & copy.num == tumor_ploidy ~ "neutral"
  ))



add_status_df <- dplyr::bind_rows(add_autosomes_df, add_xy_df) %>%
  dplyr::select(-germline_sex_estimate)
```

### Write to `scratch`

```{r}
output_file <- file.path("..", "..", "scratch", "cnvkit_with_status.tsv")
readr::write_tsv(add_status_df, output_file)
```
