---
title: "Add ploidy column, status to CNVkit output"
output: html_notebook
author: J. Taroni for ALSF CCDL
date: 2019
---

The `histologies.tsv` file contains a `tumor_ploidy` column, which is tumor ploidy as inferred by ControlFreeC.
The copy number information should be interpreted in the light of this information (see: [current version of Data Formats section of README](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/390f1e08e481da5ec0b2c62d886d5fd298bbf017#data-formats)).

This notebook adds ploidy information to the CNVkit results and adds a status column that defines gain and loss broadly.

```{r}
library(dplyr)
```

### Read in data

```{r}
cnvkit_file <- file.path("..", "..", "data","v8", "cnv-cnvkit.seg.gz")
cnvkit_df <- readr::read_tsv(cnvkit_file)
```

```{r}
histologies_file <- file.path("..", "..", "data","v8", "histologies.tsv")
histologies_df <- readr::read_tsv(histologies_file, guess_max = 100000)
```

### Add inferred ploidy information to CNVkit results

```{r}
add_ploidy_df <- histologies_df %>%
  select(Kids_First_Biospecimen_ID, tumor_ploidy, germline_sex_estimate) %>%
  inner_join(cnvkit_df, by = c("Kids_First_Biospecimen_ID" = "ID")) %>%
  select(-tumor_ploidy, -germline_sex_estimate, everything())
```

### Add status column

This is intended to mirror the information contained in the ControlFreeC output.

```{r}
add_autosomes_df <- add_ploidy_df %>%
  # x and y chromosomes must be handled differently
  filter(!(chrom %in% c("chrX", "chrY"))) %>%
  mutate(status = case_when(
    # when the copy number is less than inferred ploidy, mark this as a loss
    copy.num < tumor_ploidy ~ "loss",
    # if copy number is higher than ploidy, mark as a gain
    copy.num > tumor_ploidy ~ "gain",
    copy.num == tumor_ploidy ~ "neutral"
  ))

# this logic is consistent with what is observed in the controlfreec file
# specifically, in samples where germline sex estimate = Female, X chromosomes
# appear to be treated the same way as autosomes

samples_WGS <- histologies_df %>% filter(experimental_strategy == "WGS")
samples_WXS <- histologies_df %>% filter(experimental_strategy == "WXS")
samples_superset <- inner_join(samples_WGS, samples_WXS, by = "cohort_participant_id", copy = FALSE)


add_xy_df <- add_ploidy_df %>%
  filter(chrom %in% c("chrX", "chrY")) %>%  
  mutate(status = case_when(

  copy.num < (0.5 * tumor_ploidy) & case_when ( cohort == "PBTA" & germline_sex_estimate == NA ~ samples_superset$germline_sex_estimate.x , TRUE ~ germline_sex_estimate) == "Male" ~ "loss",

  copy.num > (0.5 * tumor_ploidy) & case_when ( cohort == "PBTA" & germline_sex_estimate == NA ~ samples_superset$germline_sex_estimate.x , TRUE ~ germline_sex_estimate) == "Male" ~ "gain",

  copy.num == (0.5 * tumor_ploidy) & case_when ( cohort == "PBTA" & germline_sex_estimate == NA ~ samples_superset$germline_sex_estimate.x , TRUE ~ germline_sex_estimate) == "Male"  ~ "neutral",


  copy.num < (0.5 * tumor_ploidy) & case_when (cohort == 'TARGET' & germline_sex_estimate == NA ~ case_when (samples_superset$germline_sex_estimate.x == NA ~ reported_gender, TRUE ~ samples_superset$germline_sex_estimate.x), TRUE ~ germline_sex_estimate) == "Male" ~ "loss",


  copy.num > (0.5 * tumor_ploidy) & case_when (cohort == 'TARGET' & germline_sex_estimate == NA ~ case_when (samples_superset$germline_sex_estimate.x == NA ~ reported_gender, TRUE ~ samples_superset$germline_sex_estimate.x), TRUE ~ germline_sex_estimate) == "Male" ~ "gain",

  copy.num == (0.5 * tumor_ploidy) & case_when (cohort == 'TARGET' & germline_sex_estimate == NA ~ case_when (samples_superset$germline_sex_estimate.x == NA ~ reported_gender, TRUE ~ samples_superset$germline_sex_estimate.x), TRUE ~ germline_sex_estimate) == "Male" ~ "neutral",


  

  # there are some instances where there are chromosome Y segments are in
  # samples where the germline_sex_estimate is Female

  chrom == "chrY" & copy.num > 0 & case_when ( cohort == "PBTA" & germline_sex_estimate == NA ~ samples_superset$germline_sex_estimate.x , TRUE ~ germline_sex_estimate) == "Female" ~ "unknown",

  chrom == "chrY" & copy.num > 0 & case_when (cohort == 'TARGET' & germline_sex_estimate == NA ~ case_when (samples_superset$germline_sex_estimate.x == NA ~ reported_gender, TRUE ~ samples_superset$germline_sex_estimate.x), TRUE ~ germline_sex_estimate) == "Female" ~ "unknown",

  chrom == "chrY" & copy.num == 0 & case_when ( cohort == "PBTA" & germline_sex_estimate == NA ~ samples_superset$germline_sex_estimate.x , TRUE ~ germline_sex_estimate) == "Female"  ~ "neutral",

  chrom == "chrY" & copy.num == 0 & case_when (cohort == 'TARGET' & germline_sex_estimate == NA ~ case_when (samples_superset$germline_sex_estimate.x == NA ~ reported_gender, TRUE ~ samples_superset$germline_sex_estimate.x), TRUE ~ germline_sex_estimate) == "Female" ~ "neutral",




  # mirroring controlfreec, X treated same as autosomes

  chrom == "chrX" & copy.num < tumor_ploidy & case_when ( cohort == "PBTA" & germline_sex_estimate == NA ~ samples_superset$germline_sex_estimate.x , TRUE ~ germline_sex_estimate) == "Female" ~ "loss",
  chrom == "chrX" & copy.num < tumor_ploidy & case_when (cohort == 'TARGET' & germline_sex_estimate == NA ~ case_when (samples_superset$germline_sex_estimate.x == NA ~ reported_gender, TRUE ~ samples_superset$germline_sex_estimate.x), TRUE ~ germline_sex_estimate) == "Female" ~ "loss",

  chrom == "chrX" & copy.num > tumor_ploidy & case_when ( cohort == "PBTA" & germline_sex_estimate == NA ~ samples_superset$germline_sex_estimate.x , TRUE ~ germline_sex_estimate) == "Female" ~ "gain",
  chrom == "chrX" & copy.num > tumor_ploidy & case_when (cohort == 'TARGET' & germline_sex_estimate == NA ~ case_when (samples_superset$germline_sex_estimate.x == NA ~ reported_gender, TRUE ~ samples_superset$germline_sex_estimate.x), TRUE ~ germline_sex_estimate) == "Female" ~ "gain",

  chrom == "chrX" & copy.num == tumor_ploidy & case_when ( cohort == "PBTA" & germline_sex_estimate == NA ~ samples_superset$germline_sex_estimate.x , TRUE ~ germline_sex_estimate) == "Female" ~ "neutral"
  chrom == "chrX" & copy.num == tumor_ploidy & case_when (cohort == 'TARGET' & germline_sex_estimate == NA ~ case_when (samples_superset$germline_sex_estimate.x == NA ~ reported_gender, TRUE ~ samples_superset$germline_sex_estimate.x), TRUE ~ germline_sex_estimate) == "Female" ~ "neutral"
  
))

add_status_df <- dplyr::bind_rows(add_autosomes_df, add_xy_df) %>%
  dplyr::select(-germline_sex_estimate)
```

### Write to `scratch`

```{r}
output_file <- file.path("..", "..", "scratch", "cnvkit_with_status.tsv")
readr::write_tsv(add_status_df, output_file)
```
