---
title: "Find most focal recurrent copy number units"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2020
---

This notebook defines the most focal recurrent copy number units by separating out entire chromsome arm losses and gains.

## Usage

This notebook is intended to be run from the command line with the following (assumes you are in the root directory of the repository):

```
Rscript -e "rmarkdown::render('analyses/focal-cn-file-preparation/05-define-most-focal-cn-units.Rmd', clean = TRUE)"
```

## Set up

### Libraries and functions

```{r}
library(tidyverse)
```

### Files and directories

```{r}
results_dir <- "results"
```

### Read in files

```{r}
# Read in the file with consensus CN status data and the UCSC cytoband data --
# generated in `03-add-cytoband-status-consensus.Rmd`
consensus_seg_cytoband_status_df <-
  read_tsv(file.path("results", "consensus_seg_with_ucsc_cytoband_status.tsv.gz"))

# Read in the annotated CN file (without the UCSC data)
consensus_seg_autosomes_df <-
  read_tsv(file.path(results_dir, "consensus_seg_annotated_cn_autosomes.tsv.gz"))
```

## Define most focal units

### First, we want find entire chromosome arm losses/gains.

```{r}
consensus_seg_cytoband_status_df <-
  consensus_seg_cytoband_status_df %>%
  select(chromosome_arm,
         dominant_status,
         cytoband,
         Kids_First_Biospecimen_ID)

status_count <- consensus_seg_cytoband_status_df %>%
  # Count the amount og times each arm has a specific `dominant_status` value
  count(chromosome_arm, dominant_status) %>%
  # Spread the data -- each row represents a unique chromosome arm
  spread(dominant_status, n) %>%
  select(-c("uncallable", "unstable")) %>%
  replace_na(list(
    gain = 0,
    loss = 0,
    neutral = 0
  )) %>%
  group_by(chromosome_arm) %>%
  # The logic here is to define the arm status based on the majority of calls
  # in the chromosome arm -- if the number of calls for a specific status is
  # responsible for more than 50% of the total calls in that arm, then it is
  # used to define the arm status
  mutate(
    dominant = case_when(
      gain > (0.5 * (gain + loss + neutral)) ~ "arm_gain",
      loss > (0.5 * (gain + loss + neutral)) ~ "arm_loss",
      neutral > (0.5 * (gain + loss + neutral)) ~ "arm_neutral",
      TRUE ~ "uncallable"
    )
  )

# Display table
status_count
```

### Next we want to use the annotated CN data to join the cytoband calls

We want to include cytoband calls for chromosome arms that have not been defined as a gain or loss.

## Session Info

```{r}
sessionInfo()
```

