---
title: "Find most focal recurrent copy number units"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2020
params:
  is_ci: FALSE
---

This notebook defines the most focal recurrent copy number units by removing focal changes that are within entire chromosome arm losses and gains.
_Most focal_ here meaning:

- If a chromosome arm is not clearly defined as a gain or loss (and is callable) we look to define the cytoband level status
- If a cytoband is not clearly defined as a gain or loss (and is callable) we then look to define the gene level status

## Usage

This notebook is intended to be run from the command line with the following (assumes you are in the root directory of the repository):

```
Rscript -e "rmarkdown::render('analyses/focal-cn-file-preparation/05-define-most-focal-cn-units.Rmd', clean = TRUE)"
```

### Cutoffs: 

```{r}
# The percentage of calls a particular status needs to be 
# above to be called the majority status -- the decision
# for a cutoff of 90% here was made to ensure that the status
# is not only the majority status but it is also significantly
# called more than the other status values in the region
percent_threshold <- 0.9

# The percentage threshold for determining if enough of a region
# (arm, cytoband, or gene) is callable to determine its status --
# the decision for a cutoff of 50% here was made as it seems reasonable
# to expect a region to be more than 50% callable for a dominant status
# call to be made
uncallable_threshold <- 0.5
```

## Set up

### Libraries and functions

```{r}
library(tidyverse)
```

### Custom Function

```{r}
status_majority_caller <- function(status_df,
                                   region_variable,
                                   status_column_name,
                                   percent_threshold_value, 
                                   uncallable_threshold_value = uncallable_threshold) {
  # Given a data.frame with cytoband/gene level copy number status data,
  # find the dominant status of the cytoband/gene region by calculating
  # the percentages of the region that each call represents.
  #
  # Args:
  #   status_df: data.frame with cytoband/gene level copy number status data
  #   region_variable: string of the column name/region to calculate copy number
  #                    status percentages for
  #   status_column_name: string of the column name that holds the relevant copy
  #                       number status data
  #   percent_threshold_value: What percent of calls a particular status needs to be 
  #                      above to be called the majority. 
  #   uncallable_threshold_value: a threshold for determining if enough of a region is 
  #                         callable to determine its status. 
  #   
  # Return:
  #   status_count: data.frame with percentage values for each unique status in
  #                 each unique region (arm/cytoband/gene) and the dominant
  #                 status for that region
  
  # Tidyeval for these columns 
  region_sym <- rlang::sym(region_variable)
  status_sym <- rlang::sym(status_column_name)
  
  # Format the data and group it
  status_count <- status_df %>%
    count(!!region_sym, !!status_sym) %>%
    # Spread the data -- each row represents a unique chromosome arm
    spread(!!status_sym, n) %>%
    # Turn NAs into 0s
    replace_na(list(
      gain = 0,
      loss = 0,
      neutral = 0,
      amplification = 0,
      uncallable = 0,
      unstable = 0
    )) %>%
    # Getting counts by region 
    group_by(!!region_sym) 
  
  # Let's store the regions separately so as to avoid weird coercions
  region_vector <- 
    status_count %>% 
    dplyr::pull(!!region_sym)
  
  # Calculate percent 
  status_count <- status_count %>%
    # Store region variable column out of the way as rownames
    tibble::column_to_rownames(region_variable) %>% 
    # Obtain a total counts variable column
    dplyr::mutate(total = apply(., 1, sum)) %>% 
    # Get the ratio of each status count to the total
    dplyr::mutate_at(dplyr::vars(-total), dplyr::funs(. / total)) %>% 
    # Bring back our region variable as its own column
    dplyr::mutate(!!region_variable := region_vector)

  # The logic here is to define the region status based on the majority of calls
  # in the region -- if the number of calls for a specific status is 
  # responsible for more than `percent_threshold` value of the total calls in that
  # region, then it is used to define the region's status (exception is for the 
  # uncallable status where we define the region as uncallable when more than the
  # `uncallable_threshold` of the calls in that region are uncallable)
  if ((region_variable == "chromosome_arm") | (region_variable == "gene_symbol")) {
    status_count <- status_count %>%
      mutate(
        dominant_status = case_when(
          gain > percent_threshold ~ "gain",
          loss > percent_threshold ~ "loss",
          amplification > percent_threshold ~ "amplification",
          TRUE ~ "neutral"
        )
      )
  } else if (region_variable == "cytoband") {
    status_count <- status_count %>%
      mutate(
        dominant_status = case_when(
          uncallable > uncallable_threshold ~ "uncallable",
          gain > percent_threshold~ "gain",
          loss > percent_threshold ~ "loss",
          neutral > percent_threshold ~ "neutral",
          TRUE ~ "unstable"
        )
      )
  }

  return(status_count)
}
```

```{r}
plot_dominant_status_calls <- function(status_count_df,
                                       region_variable) {
  # Given a data.frame with the percentage values for each region and the
  # dominant status for that region, plot the dominant status call on the
  # x-axis with the percentage values on the y-axis.
  #
  # Args:
  #   status_count_df: data.frame with percentage values for each unique status
  #                    in each unique region (arm/cytoband/gene) and the
  #                    dominant status for that region
  #   region_variable: string of the region (which will also be a column name)
  #                    that the data.frame holds percentage values for
  # Return:
  #   status_plot: plot representing the dominant status call on the x-axis and
  #                the percentage values on the y-axis
  
  status_count_df %>%
    select(-total) %>%
    dplyr::ungroup() %>%
    tibble::column_to_rownames(region_variable) %>%
    tidyr::gather(status, percent,-dominant_status) %>%
    ggplot2::ggplot(ggplot2::aes(x = status,
                                 y = percent)) +
    ggplot2::geom_jitter() +
    ggplot2::facet_wrap( ~ dominant_status)
}
```


### Files and directories

```{r}
results_dir <- "results"

# Define a logical object for running in CI
running_in_ci <- params$is_ci
```

### Read in files

Read in cytoband status file and format it for what we will need in this notebook. 

```{r}
# Read in the file with consensus CN status data and the UCSC cytoband data --
# generated in `03-add-cytoband-status-consensus.Rmd`
consensus_seg_cytoband_status_df <-
  read_tsv(file.path("results", "consensus_seg_with_ucsc_cytoband_status.tsv.gz")) %>%
  # Need this to not have `chr`
  mutate(chr = gsub("chr", "", chr),
         cytoband = paste0(chr, cytoband)) %>%
  select(
    chromosome_arm,
    # Distinguish this dominant status that is based on cytobands, from the status 
    dominant_cytoband_status = dominant_status,
    cytoband,
    Kids_First_Biospecimen_ID
  )
```

Read in the chromosome-level and gene-level data. 

```{r}
# Read in the annotated CN file (without the UCSC data)
consensus_seg_autosomes_df <-
  read_tsv(file.path(results_dir, "consensus_seg_annotated_cn_autosomes.tsv.gz"))
```

Joining the gene-level, cytoband-level, and arm-level data into one data frame.

```{r}
combined_status_df <- consensus_seg_autosomes_df %>%
  inner_join(
    consensus_seg_cytoband_status_df,
    by = c("biospecimen_id" = "Kids_First_Biospecimen_ID", "cytoband")
  )
```

## Define most focal units

### Determine chromosome arm status

```{r}
# Use our custom function to make the status calls
arm_status_count <- status_majority_caller(
  combined_status_df,
  "chromosome_arm",
  "status", 
  percent_threshold = percent_threshold
)

# Display table
arm_status_count %>%
  group_by(dominant_status)
```

These are the chromosome arms that have not been defined as gain or loss -- we want to define their cytoband/gene level status

```{r}
# Let's get a vector of the neutral arms
neutral_arms <- arm_status_count %>%
  filter(dominant_status == "neutral") %>% 
  dplyr::pull("chromosome_arm")
```

### Determine cytoband status

We want to include cytoband and gene level calls for chromosome arms that have not been defined as a gain or loss.

```{r}
# Filter the annotated CN data to include only neutral chromosome arms
neutral_status_arm_df <- combined_status_df %>%
  filter(chromosome_arm %in% neutral_arms)
```

Making cytoband-level majority calls. 

```{r}
if (!(running_in_ci)) {
# Now count the cytoband level calls (for each status call) and define
# the cytoband as that status if more than 50% of the total counts are
# for that particular status
cytoband_status_count <- status_majority_caller(
  neutral_status_arm_df,
  "cytoband",
  "dominant_cytoband_status",
  percent_threshold = percent_threshold
)

# Display table
cytoband_status_count
}
```

### Determine gene-level status

```{r}
if (!(running_in_ci)) {
  # These are the cytobands that have not been defined as gain or loss --
  # we want to define their gene level status
  neutral_cytobands <- cytoband_status_count %>%
    filter(dominant_status %in% c("unstable", "neutral")) %>%
    dplyr::pull("cytoband")

  # Filter the annotated CN data to include only these cytobands
  neutral_status_cytoband_df <- combined_status_df %>%
    filter(cytoband %in% neutral_cytobands)
  
  # Now count the gene level calls (for each status call) and define
  # the gene as that status if more than 50% of the total counts are
  # for that particular status
  gene_status_count <- status_majority_caller(neutral_status_cytoband_df,
                                              "gene_symbol",
                                              "status", 
                                              percent_threshold = percent_threshold)
  
  # Display table
  gene_status_count
}
```

## Plot calls

Plot the final dominant status call on the x-axis and the percent of each status on the y-axis.

### Plot chromosome arm status calls

```{r}
# Run `plot_dominant_status_calls` function for the chromosome arm calls
plot_dominant_status_calls(
  arm_status_count,
  "chromosome_arm"
)
```

### Plot cytoband status calls

```{r fig.width = 10}
# Run `plot_dominant_status_calls` function for the cytoband calls if not
# running in CI
if (!(running_in_ci)) {
  plot_dominant_status_calls(cytoband_status_count,
                             "cytoband")
}
```

### Plot gene status calls

```{r}
# Plot if not running in circleCI
if (!(running_in_ci)) {
  plot_dominant_status_calls(gene_status_count,
                             "gene_symbol")
}
```

## Combine arm, cytoband, and gene level status data

```{r}
# The logic variable `running_in_ci` is needed here because the CI testing
# files do not contain any of the genes in the `gene_status_count` data.frame
# generated above (when `running_in_ci` == FALSE)
if (!(running_in_ci)) {
  final_df <- consensus_seg_autosomes_df %>%
    mutate(chromosome_arm = gsub("(p|q).*", "\\1", cytoband)) %>%
    inner_join(arm_status_count,
               by = "chromosome_arm") %>%
    left_join(cytoband_status_count,
              by = "cytoband",
              suffix = c(".arm", ".cytoband")) %>%
    left_join(gene_status_count,
              by = "gene_symbol",
              suffix = c(".arm", ".gene")) %>%
    mutate(
      focal_call = paste0(
        dominant_status.arm,
        ", ",
        dominant_status.cytoband,
        ", ",
        dominant_status
      ),
      # Here we want to define the most focal call based on the arm, cytoband,
      # and gene status information -- if a loss/gain is defined at the arm
      # level then the focal call will be "arm_loss" or "arm_gain" respectively,
      # and so on.
      focal_call = case_when(focal_call == "loss, NA, NA" ~ "arm_loss",
                             focal_call == "neutral, uncallable, NA" ~ "uncallable",
                             focal_call == "neutral, loss, NA" ~ "cytoband_loss",
                             focal_call == "neutral, NA, NA" ~ "arm_neutral",
                             focal_call == "neutral, unstable, neutral" ~ "gene_neutral",
                             focal_call == "neutral, unstable, loss" ~ "gene_loss",
                             TRUE ~ "Other")
    )
} else {
  final_df <- consensus_seg_autosomes_df %>%
    mutate(chromosome_arm = gsub("(p|q).*", "\\1", cytoband)) %>%
    inner_join(arm_status_count, by = "chromosome_arm") %>%
    mutate(focal_call = dominant_status)
}

# Write final table to file
write_tsv(final_df, file.path(results_dir, "consensus_seg_most_focal_cn_status.tsv.gz"))

# Display final table
final_df %>%
  arrange(focal_call) %>%
  group_by(biospecimen_id, focal_call)
```

### Find recurrent calls

```{r}
# Tally recurrent calls
final_df %>%
  group_by(chromosome_arm, focal_call) %>%
  tally() %>%
  arrange(chromosome_arm)
```

## Session Info

```{r}
sessionInfo()
```

