---
title: "Find most focal recurrent copy number units"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2020
params:
  is_ci: FALSE
---

This notebook defines the most focal recurrent copy number units by removing focal changes that are within entire chromosome arm losses and gains.
_Most focal_ here meaning:

- If a chromosome arm is not clearly defined as a gain or loss (and is callable) we look to define the cytoband level status
- If a cytoband is not clearly defined as a gain or loss (and is callable) we then look to define the gene level status

## Usage

This notebook is intended to be run from the command line with the following (assumes you are in the root directory of the repository):

```
Rscript -e "rmarkdown::render('analyses/focal-cn-file-preparation/05-define-most-focal-cn-units.Rmd', clean = TRUE)"
```

## Set up

### Libraries and functions

```{r}
library(tidyverse)
```

### Custom Function

```{r}
status_majority_caller <- function(status_df,
                                   region_variable,
                                   status_column_name,
                                   percent_threshold, 
                                   uncallable_threshold = 0.5) {
  # Given a data.frame with cytoband/gene level copy number status data,
  # find the dominant status of the cytoband/gene region by calculating
  # the percentages of the region that each call represents.
  #
  # Args:
  #   status_df: data.frame with cytoband/gene level copy number status data
  #   region_variable: string of the column name/region to calculate copy number
  #                    status percentages for
  #   status_column_name: string of the column name that holds the relevant copy
  #                       number status data
  #   percent_threshold: What percent of calls a particular status needs to be 
  #                      above to be called the majority. 
  #   uncallable_threshold: a threshold for determining if enough of a region is 
  #                         callable to determine its status. 
  #   
  # Return:
  #   status_count: data.frame with percentage values for each unique status in
  #                 each unique region (arm/cytoband/gene) and the dominant
  #                 status for that region
  
  # Tidyeval for these columns 
  region_sym <- rlang::sym(region_variable)
  status_sym <- rlang::sym(status_column_name)
  
  # Format the data and group it
  status_count <- status_df %>%
    count(!!region_sym, !!status_sym) %>%
    # Spread the data -- each row represents a unique chromosome arm
    spread(!!status_sym, n) %>%
    # Turn NAs into 0s
    replace_na(list(
      gain = 0,
      loss = 0,
      neutral = 0,
      amplification = 0,
      uncallable = 0,
      unstable = 0
    )) %>%
    # Getting counts by region 
    group_by(!!region_sym) 
  
  # Let's store the regions separately so as to avoid weird coercions
  region_vector <- 
    status_count %>% 
    dplyr::pull(!!region_sym)
  
  # Calculate percent 
  status_count <- status_count %>%
    # Store region variable column out of the way as rownames
    tibble::column_to_rownames(region_variable) %>% 
    # Obtain a total counts variable column
    dplyr::mutate(total = apply(., 1, sum)) %>% 
    # Get the ratio of each status count to the total
    dplyr::mutate_at(dplyr::vars(-total), dplyr::funs(. / total)) %>% 
    # Bring back our region variable as its own column
    dplyr::mutate(!!region_variable := region_vector)

      # The logic here is to define the region status based on the majority of calls
      # in the region -- if the number of calls for a specific status is 
      # responsible for more than 90% of the total calls in that region, then it is
      # used to define the region's status (exception is for the uncallable status
      # where we define the region as uncallable when more than 50% of the calls
      # in that region are uncallable)
  if ((region_variable == "chromosome_arm") | (region_variable == "gene_symbol")) {
    status_count <- status_count %>%
      mutate(
        dominant_status = case_when(
          gain > percent_threshold ~ "gain",
          loss > percent_threshold ~ "loss",
          amplification > percent_threshold ~ "amplification",
          TRUE ~ "neutral"
        )
      )
  } else if (region_variable == "cytoband") {
    status_count <- status_count %>%
      mutate(
        dominant_status = case_when(
          uncallable > uncallable_threshold ~ "uncallable",
          gain > percent_threshold~ "gain",
          loss > percent_threshold ~ "loss",
          neutral > percent_threshold ~ "neutral",
          TRUE ~ "unstable"
        )
      )
  }

  return(status_count)
}
```

```{r}
plot_dominant_status_calls <- function(status_count_df,
                                       status_variable) {
  # Given a data.frame with the percentage values for each unique status in
  # each unique region (arm/cytoband/gene) and the dominant status for that region,
  # plot the dominant status call on the x-axis with the percentage values for a
  # specified status variable on the y-axis.
  #
  # Args:
  #   status_count_df: data.frame with percentage values for each unique status
  #                    in each unique region (arm/cytoband/gene) and the
  #                    dominant status for that region
  #   status_variable: string of the column/status that contains the percentage
  #                    values that should be plotted on the y-axis
  # Return:
  #   status_plot: plot representing the dominant status call on the x-axis and
  #                the specified status coverage percentage values on the y-axis

  # Make variable a symbol for plotting
  status_sym <- rlang::sym(status_variable)

  status_plot <- ggplot2::ggplot(
    status_count_df,
    ggplot2::aes(
      x = dominant_status,
      y = !!status_sym
    )
  ) +
    ggplot2::geom_jitter() +
    ggplot2::ylab(status_variable)
}
```


### Files and directories

```{r}
results_dir <- "results"

# Define a logical object for running in CI
running_in_ci <- params$is_ci
```

### Read in files

Read in cytoband status file and format it for what we will need in this notebook. 

```{r}
# Read in the file with consensus CN status data and the UCSC cytoband data --
# generated in `03-add-cytoband-status-consensus.Rmd`
consensus_seg_cytoband_status_df <-
  read_tsv(file.path("results", "consensus_seg_with_ucsc_cytoband_status.tsv.gz")) %>%
  # Need this to not have `chr`
  mutate(chr = gsub("chr", "", chr),
         cytoband = paste0(chr, cytoband)) %>%
  select(
    chromosome_arm,
    # Distinguish this dominant status that is based on cytobands, from the status 
    dominant_cytoband_status = dominant_status,
    cytoband,
    Kids_First_Biospecimen_ID
  )
```

Read in the chromosome-level and gene-level data. 

```{r}
# Read in the annotated CN file (without the UCSC data)
consensus_seg_autosomes_df <-
  read_tsv(file.path(results_dir, "consensus_seg_annotated_cn_autosomes.tsv.gz"))
```

Joining the gene-level, cytoband-level, and arm-level data into one data frame.

```{r}
combined_status_df <- consensus_seg_autosomes_df %>%
  inner_join(
    consensus_seg_cytoband_status_df,
    by = c("biospecimen_id" = "Kids_First_Biospecimen_ID", "cytoband")
  )
```

## Define most focal units

### Determine chromosome arm status

```{r}
# Use our custom function to make the status calls
arm_status_count <- status_majority_caller(
  combined_status_df,
  "chromosome_arm",
  "status", 
  percent_threshold = .9
)

# Display table
arm_status_count %>%
  group_by(dominant_status)
```

These are the chromosome arms that have not been defined as gain or loss -- we want to define their cytoband/gene level status

```{r}
# Let's get a vector of the neutral arms
neutral_arms <- arm_status_count %>%
  filter(dominant_status == "neutral") %>% 
  dplyr::pull("chromosome_arm")
```

### Determine cytoband status

We want to include cytoband and gene level calls for chromosome arms that have not been defined as a gain or loss.

```{r}
# Filter the annotated CN data to include only neutral chromosome arms
neutral_status_arm_df <- combined_status_df %>%
  filter(chromosome_arm %in% neutral_arms)
```

Making cytoband-level majority calls. 

```{r}
if (!(running_in_ci)) {
# Now count the cytoband level calls (for each status call) and define
# the cytoband as that status if more than 50% of the total counts are
# for that particular status
cytoband_status_count <- status_majority_caller(
  combined_status_df_arm,
  "cytoband",
  "dominant_cytoband_status"
)

# Display table
cytoband_status_count
}
```

### Determine gene-level status

```{r}
if (!(running_in_ci)) {
  # These are the cytobands that have not been defined as gain or loss --
  # we want to define their gene level status
  neutral_cytobands <- cytoband_status_count %>%
    filter(dominant_status %in% c("unstable", "neutral")) %>%
    dplyr::pull("cytoband")


  # Filter the annotated CN data to include only these cytobands
  neutral_status_cytoband_df <- combined_status_df %>%
    filter(cytoband %in% neutral_cytobands)
  
  # Now count the gene level calls (for each status call) and define
  # the gene as that status if more than 50% of the total counts are
  # for that particular status
  gene_status_count <- status_majority_caller(combined_status_df_cytoband,
                                              "gene_symbol",
                                              "status", 
                                              percent_threshold = .9)
  
  # Display table
  gene_status_count
}
```

## Plot calls

Plot the final dominant status call on the x-axis and the percent of each status on the y-axis.

### Plot chromosome arm status calls

```{r}
#TODO: Functionalize this plotting thingy
arm_status_count %>% 
  dplyr::ungroup() %>% 
  tibble::column_to_rownames("chromosome_arm") %>%
  tidyr::gather(status, percent, -dominant_status) %>% 
    ggplot2::ggplot(ggplot2::aes(
      x = status,
      y = percent
    )
  ) +
    ggplot2::geom_jitter() +
  ggplot2::facet_wrap(~dominant_status)
```

```{r}
# Run `plot_dominant_status_calls` function for each specific status' set of
# percentage values
arm_gain_status_plot <- plot_dominant_status_calls(
  arm_status_count,
  "percent_gain"
)

arm_loss_status_plot <- plot_dominant_status_calls(
  arm_status_count,
  "percent_loss"
)

arm_amplification_status_plot <- plot_dominant_status_calls(
  arm_status_count,
  "percent_amplification"
)

# Display plots
arm_gain_status_plot + arm_loss_status_plot + arm_amplification_status_plot + plot_layout(ncol = 2, nrow = 2)
```

### Plot cytoband status calls

```{r}
# Run `plot_dominant_status_calls` function for each specific status' set of
# percentage values
cytoband_gain_status_plot <- plot_dominant_status_calls(
  cytoband_status_count,
  "percent_gain"
)

cytoband_loss_status_plot <- plot_dominant_status_calls(
  cytoband_status_count,
  "percent_loss"
)

cytoband_neutral_status_plot <- plot_dominant_status_calls(
  cytoband_status_count,
  "percent_neutral"
)

cytoband_uncallable_status_plot <- plot_dominant_status_calls(
  cytoband_status_count,
  "percent_uncallable"
)

cytoband_unstable_status_plot <- plot_dominant_status_calls(
  cytoband_status_count,
  "percent_unstable"
)

cytoband_gain_status_plot + cytoband_loss_status_plot + cytoband_neutral_status_plot +
  cytoband_uncallable_status_plot + cytoband_unstable_status_plot + plot_layout(ncol = 2, nrow = 3)
```

### Plot gene status calls

```{r}
# Plot if not running in circleCI
if (!(running_in_ci)) {
  gene_gain_status_plot <- plot_dominant_status_calls(gene_status_count,
                                                      "percent_gain")
  
  gene_loss_status_plot <- plot_dominant_status_calls(gene_status_count,
                                                      "percent_loss")
  
  gene_amplification_status_plot <- plot_dominant_status_calls(gene_status_count,
                                                               "percent_amplification")
  
  # Display plots using the `list` function
  gene_gain_status_plot + gene_loss_status_plot + gene_amplification_status_plot +
    plot_layout(ncol = 2, nrow = 2)
}
```

## Combine arm, cytoband, and gene level status data

```{r}
# The logic variable `running_in_ci` is needed here because the CI testing
# files do not contain any of the genes in the `gene_status_count` data.frame
# generated above (when `running_in_ci` == FALSE)
if (!(running_in_ci)) {
  final_df <- consensus_seg_autosomes_df %>%
    mutate(chromosome_arm = gsub("(p|q).*", "\\1", cytoband)) %>%
    inner_join(arm_status_count,
               by = "chromosome_arm") %>%
    left_join(cytoband_status_count,
              by = "cytoband",
              suffix = c(".arm", ".cytoband")) %>%
    left_join(gene_status_count,
              by = "gene_symbol",
              suffix = c(".arm", ".gene")) %>%
    mutate(
      focal_call = paste0(
        dominant_status.arm,
        ", ",
        dominant_status.cytoband,
        ", ",
        dominant_status
      ),
      # Here we want to define the most focal call based on the arm, cytoband,
      # and gene status information -- if a loss/gain is defined at the arm
      # level then the focal call will be "arm_loss" or "arm_gain" respectively,
      # and so on.
      focal_call = case_when(focal_call == "loss, NA, NA" ~ "arm_loss",
                             focal_call == "neutral, uncallable, NA" ~ "uncallable",
                             focal_call == "neutral, loss, NA" ~ "cytoband_loss",
                             focal_call == "neutral, NA, NA" ~ "arm_neutral",
                             focal_call == "neutral, unstable, neutral" ~ "gene_neutral",
                             focal_call == "neutral, unstable, loss" ~ "gene_loss",
                             TRUE ~ "Other")
    )
} else {
  final_df <- consensus_seg_autosomes_df %>%
    mutate(chromosome_arm = gsub("(p|q).*", "\\1", cytoband)) %>%
    inner_join(arm_status_count, by = "chromosome_arm") %>%
    left_join(cytoband_status_count,
              by = "cytoband",
              suffix = c(".arm", ".cytoband")) %>%
    mutate(focal_call = paste0(dominant_status.arm, ", ", dominant_status.cytoband),
           # Here we want to define the most focal call based on the arm, cytoband,
           # and gene status information -- if a loss/gain is defined at the arm
           # level then the focal call will be "arm_loss" or "arm_gain" respectively,
           # and so on.
           focal_call = case_when(focal_call == "loss, NA" ~ "arm_loss",
                             focal_call == "neutral, uncallable" ~ "uncallable",
                             focal_call == "neutral, loss" ~ "cytoband_loss",
                             focal_call == "neutral, NA" ~ "arm_neutral",
                             focal_call == "neutral, unstable" ~ "cytoband_unstable",
                             TRUE ~ "Other"))
}

# Write final table to file
write_tsv(final_df, file.path(results_dir, "consensus_seg_most_focal_cn_status.tsv.gz"))

# Display final table
final_df %>%
  arrange(focal_call)
```

## Session Info

```{r}
sessionInfo()
```

