---
title: "Determine the recurrent focal CN dominant status calls"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2020
---

This notebook determines the recurrent focal copy number dominant status calls by region using the output of `05-define-most-focal-cn-units.Rmd`.

Recurrence here has been arbitrarily defined based on the plotting of the distribution of status calls and a [similar decision](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/66bb67a7bf29aad4510a0913a2dbc88da0013be8/analyses/fusion_filtering/06-recurrent-fusions-per-histology.R#L152) made in `analyses/fusion_filtering/06-recurrent-fusions-per-histology.R` to make the cutoff for recurrence to be greater than a count of 3 samples that have the same CN status call in the same region.

## Usage

This notebook is intended to be run from the command line with the following (assumes you are in the root directory of the repository):

```
Rscript -e "rmarkdown::render('analyses/focal-cn-file-preparation/06-find-recurrent-calls.Rmd', clean = TRUE)"
```

## Set up

### Cutoffs

```{r}
# Here, we arbitrarily define recurrence to be greater than 3 based on 
# the plot above and on a similar decision made in
# `analyses/fusion_filtering/06-recurrent-fusions-per-histology.R`
# https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/66bb67a7bf29aad4510a0913a2dbc88da0013be8/analyses/fusion_filtering/06-recurrent-fusions-per-histology.R#L152
recurrence_cutoff <- 3

# Define the cutoff for the number of genes to keep after counting and
# arranging by copy number status recurrence in all regions -- these genes
# will be considered the genes with the most recurrent CNVs
top_genes_cutoff <- 20
```


### Libraries and functions

```{r}
library(tidyverse)
```

### Files and directories

```{r}
results_dir <- "results"

plots_dir <- "plots"
```

### Read in files

Read in focal CN dominant status calls prepared in `05-define-most-focal-cn-units.Rmd`.

```{r}
focal_cn_df <-
  read_tsv(file.path(results_dir, "consensus_seg_most_focal_cn_status.tsv.gz")) %>%
  # Filter out the `uncallable` statuses
  filter(status != "uncallable")

metadata <- read_tsv(file.path("..", "..", "data", "histologies.tsv"), guess_max = 100000)
```

## Find recurrent calls

### Tally the sample counts of each unique status call by region.

```{r}
# Tally the number of samples that have the same status call by region
final_df_tally <- focal_cn_df %>%
  group_by(status, region, region_type) %>%
  tally(name = "count") %>%
  # Add column named `short_histology` for joining later
  mutate(short_histology = "All")

# Display table with counts of each unique status call by region
final_df_tally %>%
  arrange(desc(count))
```

### Tally the sample counts of each unique status call by region and short histology.

```{r}
# Tally the number of samples that have the same status call by region
final_df_tally_histology <- focal_cn_df %>%
  # Join the metadata by biospecimen IDs to get short histology data
  inner_join(select(metadata,
                    c(
                      "Kids_First_Biospecimen_ID", "short_histology"
                    )),
             by = "Kids_First_Biospecimen_ID") %>%
  group_by(status, region, region_type, short_histology) %>%
  tally(name = "count")

# Display table with counts of each unique status call by region and short
# histology
final_df_tally_histology %>%
  arrange(desc(count))
```

### Plot the distribution of counts

```{r}
# Plot the distribution of counts for all histologies
ggplot(final_df_tally, aes(x = count)) +
  geom_histogram(bins = 100) +
  geom_vline(
    aes(xintercept = recurrence_cutoff),
    linetype = "longdash",
    color = "green",
    size = .4
  ) +
  xlab("number of samples with a copy number alteration") +
  ggtitle("Distribution of copy number status counts by region across all histologies") +
  geom_text(
    aes(x = recurrence_cutoff, label = "recurrence cutoff", y = 5000),
    colour = "darkgreen",
    angle = 90,
    vjust = 1.2,
    size = 4
  )

# Save plot
ggsave(file.path(plots_dir, "all_histology_cn_counts.png"), plot = last_plot())
```

```{r}
# Plot the distribution of counts for histologies separated
ggplot(final_df_tally_histology, aes(x = count)) +
  geom_histogram(bins = 100) +
  geom_vline(
    aes(xintercept = recurrence_cutoff),
    linetype = "longdash",
    color = "green",
    size = .4
  ) +
  xlab("number of samples with a copy number alteration") +
  ggtitle("Distribution of copy number status counts by region and histology") +
  geom_text(
    aes(x = recurrence_cutoff, label = "recurrence cutoff", y = 5000),
    colour = "darkgreen",
    angle = 90,
    vjust = 1.2,
    size = 4
  )

# Save plot
ggsave(file.path(plots_dir, "separate_histology_cn_counts.png"), plot = last_plot())
```

The `recurrence_cutoff` defined as 3 appears to be reasonable as seen in the distribution of counts plot above.

### Isolate recurrent calls

#### All histologies represented

```{r}
# Create a data frame of only observations that have counts greater than the `recurrence_cutoff`
recurrent_df <- final_df_tally %>%
  filter(count > recurrence_cutoff)

# Display table
recurrent_df
```

#### Separated by histology

```{r}
# Create a data frame of only observations that have counts greater than the `recurrence_cutoff`
recurrent_df_histology <- final_df_tally_histology %>%
  filter(count > recurrence_cutoff)

# Display table
recurrent_df_histology
```

#### Combine all histology recurrent data with the data separated by histology 

```{r}
final_df_recurrent <- recurrent_df %>%
  bind_rows(recurrent_df_histology) %>%
  arrange(region) %>%
  select(region, status, short_histology, count, region_type)

# Display table
final_df_recurrent
```

### Save list of genes that have `recurrent` calls

```{r}
# Save list of genes that have `recurrent` status calls for use in downstream
# analyses
top_recurrent_genes <- final_df_recurrent %>%
  # Ungroup data from previous step
  ungroup() %>%
  # Arrange by descending count
  arrange(desc(count)) %>%
  # Filter to only gene regions
  filter(region_type == "gene") %>%
  # Select the top genes by count
  top_n(top_genes_cutoff, count) %>%
  # Select only the gene names
  select(gene = region)

# Save list of recurrent genes to file
write_tsv(
  top_recurrent_genes,
  file.path(results_dir, "consensus_seg_focal_cn_recurrent_genes.tsv")
)
```


### Write final data frame to file

```{r}
# Write the recurrent focal CN calls by region and histology data frame to
# file
write_tsv(final_df_recurrent, file.path(results_dir, "consensus_seg_recurrent_focal_cn_units.tsv"))
```

## Session Info

```{r}
sessionInfo()
```
