---
title: "Determine the recurrent focal CN dominant status calls"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2020
---

This notebook determines the recurrent focal copy number dominant status calls by region using the output of `05-define-most-focal-cn-units.Rmd`.

Recurrence here has been arbitrarily defined based on the plotting of the distribution of status calls and a [similar decision](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/66bb67a7bf29aad4510a0913a2dbc88da0013be8/analyses/fusion_filtering/06-recurrent-fusions-per-histology.R#L152) made in `analyses/fusion_filtering/06-recurrent-fusions-per-histology.R` to make the cutoff for recurrence to be greater than a count of 3 samples that have the same CN status call in the same region.

## Usage

This notebook is intended to be run from the command line with the following (assumes you are in the root directory of the repository):

```
Rscript -e "rmarkdown::render('analyses/focal-cn-file-preparation/06-find-recurrent-calls.Rmd', clean = TRUE)"
```

## Set up

### Cutoffs

```{r}
# Here, we arbitrarily define recurrence to be greater than 3 based on 
# the plot above and on a similar decision made in
# `analyses/fusion_filtering/06-recurrent-fusions-per-histology.R`
# https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/66bb67a7bf29aad4510a0913a2dbc88da0013be8/analyses/fusion_filtering/06-recurrent-fusions-per-histology.R#L152
recurrence_cutoff <- 3
```


### Libraries and functions

```{r}
library(tidyverse)
```

### Files and directories

```{r}
results_dir <- "results"
```

### Read in files

Read in focal CN dominant status calls prepared in `05-define-most-focal-cn-units.Rmd`.

```{r}
focal_cn_df <- read_tsv(file.path(results_dir, "consensus_seg_most_focal_cn_status.tsv.gz"))
```

## Find recurrent calls

### Tally the sample counts of each unique status call by region.

```{r}
# Tally the number of samples that have the same status call by region
final_df_tally <- focal_cn_df %>%
  group_by(status, region, region_type) %>%
  tally(name = "count")

# Display table with counts of each unique status call by region
final_df_tally %>%
  arrange(desc(count))
```

### Plot the distribution of counts

```{r}
# Plot the distribution of counts
ggplot(final_df_tally, aes(x = count)) +
  geom_histogram(bins = 100) +
  geom_vline(aes(xintercept = recurrence_cutoff), linetype = "longdash", color = "green", size = .4)
```

The `recurrence_cutoff` defined as 3 appears to be reasonable as seen in the distribution of counts plot above.

### Isolate recurrent calls

```{r}
# Create a data frame of only observations that have counts greater than the `recurrence_cutoff`
recurrent_df <- final_df_tally %>%
  filter(count > recurrence_cutoff)

# Display table
recurrent_df
```

### Filter focal CN data frame to include only recurrent observations

```{r}
# Filter the focal CN data frame to include only the recurrent observations
final_df_recurrent <- focal_cn_df %>%
  # Remove the `region_type` column before joining the `recurrent_df` object to
  # avoid a duplicate `region_type` column
  select(-region_type) %>%
  inner_join(recurrent_df, by = c("region", "status"))
```

### Write to file

```{r}
# Write the recurrent focal CN calls by region at a sample level data frame to
# file
write_tsv(final_df_recurrent, file.path(results_dir, "consensus_seg_recurrent_focal_cn_units.tsv"))
```

## Session Info

```{r}
sessionInfo()
```
