---
title: "GSVA Score Modeling"
author: "Stephanie J. Spielman for ALSF CCDL"
date: '2020'
output:
  html_document:
    df_print: paged
    toc: yes
  html_notebook:
    df_print: paged
    toc: yes
    toc_float: yes
params:
  plot_ci: yes
  is_ci: FALSE
---

### Purpose

The purpose of this analysis is to assess significant differences in GSVA scores for each hallmark pathways. Using ANOVA and subsequent Tukey tests, we ask:

+ For each pathway, are GSVA scores significantly different across `display_group`? If so, which histologies are significantly different?

+ For each pathway, are GSVA scores significantly different across `harmonized_diagnosis`? If so, which types are significantly different?

We perform this using both GSVA scores calculated from the stranded and polyA RNA-seq libraries. Code is also flexible enough to test a different variable besides `display_group`, etc.

### Usage

To run this from the command line, use:
```
Rscript -e "rmarkdown::render('analyses/gene-set-enrichment-analysis/02-model-gsea.Rmd', clean = TRUE)" 
```
_This assumes you are in the top directory of the repository._

### Setup

Load libraries and define certain constants:

```{r, lib-load, warning=FALSE, message=FALSE}
library(tidyverse)
library(broom)
`%>%` <- dplyr::`%>%`

# This script contains functions used to modeling GSVA scores
source(file.path("util", "hallmark_models.R"))

# Significance testing universal threshold
SIGNIFICANCE_THRESHOLD <- 0.01

# Assigning params$is_ci to running_in_ci avoids a locked binding error
running_in_ci <- params$is_ci

# Are we testing? In case of a non 0/1 number, we recast as logical, and then ensure logical.
if (running_in_ci %in% c(0,1)) running_in_ci <- as.logical(running_in_ci)
if (!(is.logical(running_in_ci)))
{
  stop("\n\nERROR: The parameter `is_ci` should be FALSE/TRUE (or 0/1).")
}

```


<br>
Next, define directories and load data files:
```{r, data-load, message=FALSE, warning=FALSE}
### Define directories
# Establish base dir
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir    <- file.path(root_dir, "data") 
results_dir <- "results"

######### Define input files
## Metadata file (histologies/clinical data)
metadata_file <- file.path(data_dir, "pbta-histologies.tsv")
histology_color_file <- file.path(root_dir, "figures", "palettes", "histology_label_color_table.tsv")

## GSEA scores
scores_stranded_file <- file.path(results_dir, "gsva_scores_stranded.tsv")
scores_polya_file    <- file.path(results_dir, "gsva_scores_polya.tsv")

######## Define output files
file_anova_stranded_display_group  <- file.path(results_dir, "gsva_anova_stranded_display_group.tsv")
file_anova_polya_display_group     <- file.path(results_dir, "gsva_anova_polya_display_group.tsv")
file_anova_stranded_harmonized_diagnosis <- file.path(results_dir, "gsva_anova_stranded_harmonized_diagnosis.tsv")
file_anova_polya_harmonized_diagnosis    <- file.path(results_dir, "gsva_anova_polya_harmonized_diagnosis.tsv")

file_tukey_stranded_display_group  <- file.path(results_dir, "gsva_tukey_stranded_display_group.tsv")
file_tukey_polya_display_group     <- file.path(results_dir, "gsva_tukey_polya_display_group.tsv")
file_tukey_stranded_harmonized_diagnosis <- file.path(results_dir, "gsva_tukey_stranded_harmonized_diagnosis.tsv")
file_tukey_polya_harmonized_diagnosis    <- file.path(results_dir, "gsva_tukey_polya_harmonized_diagnosis.tsv")


######## Load input files
metadata        <- readr::read_tsv(metadata_file, guess_max = 10000)
scores_stranded <- readr::read_tsv(scores_stranded_file)
scores_polya    <- readr::read_tsv(scores_polya_file)
```



### ANOVA and Tukey analysis of GSVA scores

Here we perform a series of ANOVAs, for polyA and stranded libraries separately, to determine whether mean GSVA scores for a given grouping are significantly different across hallmarks (pathways). The given groupings examined here are `display_group` and `harmonized_diagnosis`. 
In other words, we perform an ANOVA (and associated posthoc Tukey test) for each hallmark as, `hallmark ~ grouping`. Users can specify the grouping variable.

First, prepare the data for modeling:

```{r aov-prep}

### Merge histology metadata with each set of gsea scores
scores_stranded  <- scores_stranded %>% mutate(data_type = "stranded")
scores_polya     <- scores_polya %>% mutate(data_type = "polya")
all_scores      <- bind_rows(scores_stranded, scores_polya) %>%
                        mutate(data_type    = factor(data_type),
                        hallmark_name = factor(hallmark_name))

# Import standard groupings from the color palette
histology_label_mapping <- readr::read_tsv(histology_color_file) %>% 
  # Select just the columns we will need for plotting
  dplyr::select(Kids_First_Biospecimen_ID, display_group, display_order, hex_codes) %>% 
  # Reorder display_group based on display_order
  dplyr::mutate(display_group = forcats::fct_reorder(display_group, display_order))

metadata_with_gsva <- metadata %>%
  filter(experimental_strategy == "RNA-Seq") %>%
  inner_join(all_scores, by = "Kids_First_Biospecimen_ID") %>% 
  # Join on color palette groupings 
  inner_join(histology_label_mapping)
```


Now, model:

```{r, aov-perform}
display_group_stranded_model_results <- gsva_anova_tukey(metadata_with_gsva, display_group, "stranded", SIGNIFICANCE_THRESHOLD) 
harmonized_diagnosis_stranded_model_results <- gsva_anova_tukey(metadata_with_gsva, harmonized_diagnosis, "stranded", SIGNIFICANCE_THRESHOLD) 

head(display_group_stranded_model_results)
head(harmonized_diagnosis_stranded_model_results)

## Save stranded library results
write_tsv(display_group_stranded_model_results[["anova"]],  file_anova_stranded_display_group)
write_tsv(harmonized_diagnosis_stranded_model_results[["anova"]], file_anova_stranded_harmonized_diagnosis)
write_tsv(display_group_stranded_model_results[["tukey"]],  file_tukey_stranded_display_group)
write_tsv(harmonized_diagnosis_stranded_model_results[["tukey"]], file_tukey_stranded_harmonized_diagnosis)


### Don't run polyA samples in CI due to data limitations, won't be enough levels for ANOVA
if (!(running_in_ci)){
  display_group_polya_model_results <- gsva_anova_tukey(metadata_with_gsva, display_group, "polya", SIGNIFICANCE_THRESHOLD)
  harmonized_diagnosis_polya_model_results <- gsva_anova_tukey(metadata_with_gsva, harmonized_diagnosis, "polya", SIGNIFICANCE_THRESHOLD)
  head(display_group_polya_model_results)
  head(harmonized_diagnosis_polya_model_results)

  ### Save polya library results
  write_tsv(display_group_polya_model_results[["anova"]],     file_anova_polya_display_group)
  write_tsv(harmonized_diagnosis_polya_model_results[["anova"]],    file_anova_polya_harmonized_diagnosis)
  write_tsv(display_group_polya_model_results[["tukey"]],     file_tukey_polya_display_group)
  write_tsv(harmonized_diagnosis_polya_model_results[["tukey"]],    file_tukey_polya_harmonized_diagnosis)

}
```

**How many `display_group` have significant ANOVAs across hallmark pathways, for each library?**
```{r}
display_group_stranded_model_results[["anova"]] %>% count(significant_anova)
if (!(running_in_ci))
{
  display_group_polya_model_results[["anova"]] %>% count(significant_anova)
}
```

> All are significantly different for stranded, none for polya. Likely due to data/power limitations.

**How many `harmonized_diagnosis` have significant ANOVAs across hallmark pathways, for each library?**
```{r}
harmonized_diagnosis_stranded_model_results[["anova"]] %>% count(significant_anova)
if (!(running_in_ci))
{
  harmonized_diagnosis_polya_model_results[["anova"]] %>% count(significant_anova)
}
```

> Again, all are significantly different for stranded, none for polya. Likely due to data/power limitations.
