---
title: "GSVA Score Modeling"
author: "Stephanie J. Spielman for ALSF CCDL"
date: '2020'
output:
  html_document:
    df_print: paged
    toc: yes
  html_notebook:
    df_print: paged
    toc: yes
    toc_float: yes
params:
  plot_ci: yes
  is_ci: FALSE
---

### Purpose

The purpose of this analysis is to assess significant differences in GSVA scores for each hallmark pathways. Using ANOVA and subsequent Tukey tests, we ask:

+ For each pathway, are GSVA scores significantly different across `short_histology`? If so, which histologies are significantly different?

+ For each pathway, are GSVA scores significantly different across `disease_type_new`? If so, which types are significantly different?

We perform this using both GSVA scores calculated from the stranded and polyA RNA-seq libraries. Code is also flexible enough to test a different variable besides `short_histology`, etc.

### Usage

To run this from the command line, use:
```
Rscript -e "rmarkdown::render('analyses/gene-set-enrichment-analysis/02-model-gsea.Rmd', clean = TRUE)" 
```
_This assumes you are in the top directory of the repository._

### Setup

Load libraries and define certain constants:

```{r, lib-load, warning=FALSE, message=FALSE}
library(tidyverse)
library(broom)
`%>%` <- dplyr::`%>%`

# This script contains functions used to modeling GSVA scores
source(file.path("util", "hallmark_models.R"))

# Significance testing universal threshold
SIGNIFICANCE_THRESHOLD <- 0.01

# Assigning params$is_ci to running_in_ci avoids a locked binding error
running_in_ci <- params$is_ci

# Are we testing? In case of a non 0/1 number, we recast as logical, and then ensure logical.
if (running_in_ci %in% c(0,1)) running_in_ci <- as.logical(running_in_ci)
if (!(is.logical(running_in_ci)))
{
  stop("\n\nERROR: The parameter `is_ci` should be FALSE/TRUE (or 0/1).")
}

```


<br>
Next, define directories and load data files:
```{r, data-load, message=FALSE, warning=FALSE}
### Define directories
data_dir    <- file.path("..", "..", "data") 
results_dir <- "results"

######### Define input files
## Metadata file (histologies/clinical data)
metadata_file <- file.path(data_dir, "pbta-histologies.tsv")

## GSEA scores
scores_stranded_file <- file.path(results_dir, "gsva_scores_stranded.tsv")
scores_polya_file    <- file.path(results_dir, "gsva_scores_polya.tsv")

######## Define output files
file_anova_stranded_short_histology  <- file.path(results_dir, "gsva_anova_stranded_short_histology.tsv")
file_anova_polya_short_histology     <- file.path(results_dir, "gsva_anova_polya_short_histology.tsv")
file_anova_stranded_disease_type_new <- file.path(results_dir, "gsva_anova_stranded_disease_type_new.tsv")
file_anova_polya_disease_type_new    <- file.path(results_dir, "gsva_anova_polya_disease_type_new.tsv")

file_tukey_stranded_short_histology  <- file.path(results_dir, "gsva_tukey_stranded_short_histology.tsv")
file_tukey_polya_short_histology     <- file.path(results_dir, "gsva_tukey_polya_short_histology.tsv")
file_tukey_stranded_disease_type_new <- file.path(results_dir, "gsva_tukey_stranded_disease_type_new.tsv")
file_tukey_polya_disease_type_new    <- file.path(results_dir, "gsva_tukey_polya_disease_type_new.tsv")


######## Load input files
metadata        <- readr::read_tsv(metadata_file)
scores_stranded <- readr::read_tsv(scores_stranded_file)
scores_polya    <- readr::read_tsv(scores_polya_file)
```



### ANOVA and Tukey analysis of GSVA scores

Here we perform a series of ANOVAs, for polyA and stranded libraries separately, to determine whether mean GSVA scores for a given grouping are significantly different across hallmarks (pathways). The given groupings examined here are `short_histology` and `disease_type_new`. 
In other words, we perform an ANOVA (and associated posthoc Tukey test) for each hallmark as, `hallmark ~ grouping`. Users can specify the grouping variable.

First, prepare the data for modeling:

```{r aov-prep}

### Merge histology metadata with each set of gsea scores
scores_stranded  <- scores_stranded %>% mutate(data_type = "stranded")
scores_polya     <- scores_polya %>% mutate(data_type = "polya")
all_scores      <- bind_rows(scores_stranded, scores_polya) %>%
                        mutate(data_type    = factor(data_type),
                        hallmark_name = factor(hallmark_name))

metadata_with_gsva <- metadata %>%
                        filter(experimental_strategy == "RNA-Seq") %>%
                        inner_join(all_scores, by = "Kids_First_Biospecimen_ID" )
```


Now, model:

```{r, aov-perform}
short_histology_stranded_model_results <- gsva_anova_tukey(metadata_with_gsva, short_histology, "stranded", SIGNIFICANCE_THRESHOLD) 
disease_type_new_stranded_model_results <- gsva_anova_tukey(metadata_with_gsva, disease_type_new, "stranded", SIGNIFICANCE_THRESHOLD) 

head(short_histology_stranded_model_results)
head(disease_type_new_stranded_model_results)

## Save stranded library results
write_tsv(short_histology_stranded_model_results[["anova"]],  file_anova_stranded_short_histology)
write_tsv(disease_type_new_stranded_model_results[["anova"]], file_anova_stranded_disease_type_new)
write_tsv(short_histology_stranded_model_results[["tukey"]],  file_tukey_stranded_short_histology)
write_tsv(disease_type_new_stranded_model_results[["tukey"]], file_tukey_stranded_disease_type_new)


### Don't run polyA samples in CI due to data limitations, won't be enough levels for ANOVA
if (!(running_in_ci)){
  short_histology_polya_model_results <- gsva_anova_tukey(metadata_with_gsva, short_histology, "polya", SIGNIFICANCE_THRESHOLD)
  disease_type_new_polya_model_results <- gsva_anova_tukey(metadata_with_gsva, disease_type_new, "polya", SIGNIFICANCE_THRESHOLD)
  head(short_histology_polya_model_results)
  head(disease_type_new_polya_model_results)

  ### Save polya library results
  write_tsv(short_histology_polya_model_results[["anova"]],     file_anova_polya_short_histology)
  write_tsv(disease_type_new_polya_model_results[["anova"]],    file_anova_polya_disease_type_new)
  write_tsv(short_histology_polya_model_results[["tukey"]],     file_tukey_polya_short_histology)
  write_tsv(disease_type_new_polya_model_results[["tukey"]],    file_tukey_polya_disease_type_new)

}
```

**How many `short_histology` have significant ANOVAs across hallmark pathways, for each library?**
```{r}
short_histology_stranded_model_results[["anova"]] %>% count(significant_anova)
if (!(running_in_ci))
{
  short_histology_polya_model_results[["anova"]] %>% count(significant_anova)
}
```

> All are significantly different for stranded, none for polya. Likely due to data/power limitations.

**How many `disease_type_new` have significant ANOVAs across hallmark pathways, for each library?**
```{r}
disease_type_new_stranded_model_results[["anova"]] %>% count(significant_anova)
if (!(running_in_ci))
{
  disease_type_new_polya_model_results[["anova"]] %>% count(significant_anova)
}
```

> Again, all are significantly different for stranded, none for polya. Likely due to data/power limitations.
