---
title: "GSEA scores for samples passing threshold"
author: "SJ Spielman for CCDL"
date: "2023"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
params:
  release: release-v23-20230115
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_bw())

# anova/tukey test functions
source(file.path("util", "hallmark_models.R"))
```
This notebook was written as part of manuscript revisions.
Here, we explore how GSEA-related results might change if only tumors at a certain level of known tumor purity (`tumor_fraction` metadata variable) are considered.
Only stranded RNA data is considered here, as in the manuscript.
This notebook consumes result file `gsva_scores_stranded_thresholded.tsv` produced by this module's script `01-conduct-gsea-analysis` run with the `--apply_tumor_purity_threshold` flag, and compares back to results in `gsva_scores_stranded.tsv` where that flag was not used.

The associated issue for this exploration is here: https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/1626


## Functions

This section defines some helper functions used in _this_ notebook.


```{r}
# This function prepares a data frame of GSEA results for modeling
prepare_gsea_for_model <- function(gsea_df, metadata_df, palette_df) {
  gsea_df %>%
    # join with cancer_group_display information
    inner_join(
      select(
        metadata_df, 
        Kids_First_Biospecimen_ID, 
        cancer_group
      )
    ) %>%
    inner_join(
      select(
        palette_df, 
        cancer_group,
        cancer_group_display
      )
    ) %>%
    # add column for modeling function compatibility
    mutate(data_type = "stranded") %>%
    # remove plain cancer_group column
    select(-cancer_group) %>%
    # remove underscores from cancer group names to facilitate parsing
    mutate(cancer_group_display_underscore = stringr::str_replace(cancer_group_display, "-", ""))
}
```



```{r}
# This function assesses certain pathways are significant for a given set of results
# The function return a table with three sets of counts
### Total number of significantly upregulated comparisons
### Total number of significantly downregulated comparisons
### Total number of non-significant comparisons
# The total count in the result _IS_ the total number of comparisons
# Note that if there is a 4th count called "error in logic," something went wrong!
count_tukey_results <- function(df, term_vector, cancer_group) {
  df %>%
    filter(hallmark_name %in% term_vector, 
           (cancer_group1 == cancer_group | cancer_group2 == cancer_group)) %>%
    # Assign comparison as upregulated, downregulated, or not significant
    mutate(comparison = case_when(
      # significant upregulation where cancer_group is in group 1 variable
      cancer_group1 == cancer_group & significant_tukey_bonf & pathway_score_difference > 0 ~ "upregulated",
      # significant upregulation where cancer_group is in group 2 variable
      cancer_group2 == cancer_group & significant_tukey_bonf & pathway_score_difference < 0 ~ "upregulated",
      # significant downregulation where cancer_group is in group 1 variable
      cancer_group1 == cancer_group & significant_tukey_bonf & pathway_score_difference < 0 ~ "downregulated",
      # significant downregulation where cancer_group is in group 2 variable
      cancer_group2 == cancer_group & significant_tukey_bonf & pathway_score_difference > 0 ~ "downregulated",
      # Not significant
      !(significant_tukey_bonf) ~ "not significant", 
      # did anything go wrong?
      TRUE ~ "error in logic"
    )) %>%
    count(comparison)
}

# This function combines counted original and thresholded results into a single data frame
compare_original_threshold <- function(df_original, df_thresholded, term_vector, cancer_group) {
  original <- count_tukey_results(df_original, term_vector, cancer_group) %>% mutate(type = "original")
  count_tukey_results(df_thresholded, term_vector, cancer_group) %>% mutate(type = "thresholded") %>%
    bind_rows(original) %>%
    arrange(type)
}



# This function obtains results for a given SINGLE pathway comparison 
#  among a given set of cancer groups of interest
get_tukey_comparisons <- function(df, pathway_of_interest, groups_of_interest) {
  df %>%
    filter(hallmark_name == pathway_of_interest, 
           cancer_group1 %in% groups_of_interest,
           cancer_group2 %in% groups_of_interest)
}

# This function wraps the above function to return results for both original and thresholded
combine_tukey_comparisons <- function(df_original, df_thresholded, pathway_of_interest, groups_of_interest) {
  original <- get_tukey_comparisons(df_original, pathway_of_interest, groups_of_interest) %>% 
    mutate(type = "original")
  
  get_tukey_comparisons(df_thresholded, pathway_of_interest, groups_of_interest) %>% 
    mutate(type = "thresholded") %>%
    bind_rows(original) %>%
    # select only rows needed for interpretation
    # since only 1 pathway, no need to keep hallmark name
    select(type, cancer_group1, cancer_group2, pathway_score_difference, significant_tukey_bonf) %>%
    arrange(type)
}
```

## Set directories and files

```{r}
# Input
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data", params$release)
pal_dir <- file.path(root_dir, "figures", "palettes")
metadata_file <- file.path(data_dir, "pbta-histologies.tsv")
palette_file <- file.path(root_dir, 
                          "figures", 
                          "palettes", 
                          "broad_histology_cancer_group_palette.tsv")


gsea_results_original_file <- file.path(root_dir, 
                          "analyses",
                          "gene-set-enrichment-analysis",
                          "results",
                          "gsva_scores_stranded.tsv")

gsea_results_thresholded_file <- file.path(root_dir, 
                          "analyses",
                          "gene-set-enrichment-analysis",
                          "results",
                          "gsva_scores_stranded_thresholded.tsv")
```

Read in and prepare data:

```{r}
gsea_df_thresholded <- read_tsv(gsea_results_thresholded_file)
gsea_df_original <- read_tsv(gsea_results_original_file)
metadata_df <- read_tsv(metadata_file, guess_max = 10000)
palette_df <- read_tsv(palette_file)

gsea_df_thresholded <- prepare_gsea_for_model(gsea_df_thresholded, metadata_df, palette_df)
gsea_df_original <- prepare_gsea_for_model(gsea_df_original, metadata_df, palette_df)
```

## Re-analyze GSEA scores

We'll model across cancer groups for each GSEA version of results.


First, the original:
```{r}
# Perform anova and post-hoc tukey tests on original GSEA results
modeled_original <- gsva_anova_tukey(gsea_df_original, 
                                     cancer_group_display_underscore, 
                                     "stranded", 
                                     # p-value threshold used in manuscript
                                     0.05) 

# Obtain tukey results
tukey_original <- modeled_original$tukey %>%
  # separate the comparisons back out into two variables
  separate(comparison, into = c("cancer_group1", "cancer_group2"), sep = "-") 
```

Next, the thresholded:
```{r}
# Perform anova and post-hoc tukey tests on thresholded GSEA results
modeled_thresholded <- gsva_anova_tukey(gsea_df_thresholded, 
                                        cancer_group_display_underscore, 
                                        "stranded", 
                                        # p-value threshold used in manuscript
                                        0.05) 

# Obtain tukey results
tukey_thresholded <- modeled_thresholded$tukey %>%
  # separate the comparisons back out into two variables
  separate(comparison, into = c("cancer_group1", "cancer_group2"), sep = "-") 
```

  
## Assessing specific comparisons: Part 1

We'll now assess specific results from the Tukey tests. 
In this section, we'll check individual trends as stated here in the manuscript:


https://github.com/AlexsLemonade/OpenPBTA-manuscript/blob/b0fdf6f24fe9c021f22e8b7da348708f68877ee5/content/03.results.md?plain=1#L275-L276

> Indeed, we detected significant (Bonferroni-corrected p < 0.05) upregulation of _MYC_ and _E2F_ targets, as well as G2M (cell cycle phase following S phase) in MBs, ATRTs, and HGGs compared to several other cancer groups.
> In contrast, LGGs showed significant downregulation (Bonferroni-corrected p < 0.05) of these pathways.



### MYC target pathways

#### MB

> Originally, 20/40 pathways were upregulated. After thresholding, 17/38 are upregulated.
> This is broadly consistent.

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, 
                           c("HALLMARK_MYC_TARGETS_V1", "HALLMARK_MYC_TARGETS_V2"), "Medulloblastoma")
```


#### ATRT

> Originally, 24/40 pathways were upregulated. After thresholding, 13/38 are upregulated.
> Many comparisons are no longer significant.

```{r}
compare_original_threshold(tukey_original, tukey_thresholded,  c("HALLMARK_MYC_TARGETS_V1", "HALLMARK_MYC_TARGETS_V2"), "Atypical Teratoid Rhabdoid Tumor")
```

#### Other HGGs

> Originally, 12/40 pathways were upregulated. After thresholding, 10/38 are upregulated.
> This is broadly consistent.

```{r}
compare_original_threshold(tukey_original, tukey_thresholded,  c("HALLMARK_MYC_TARGETS_V1", "HALLMARK_MYC_TARGETS_V2"), "Other highgrade glioma")
```


#### DMGs

> Originally, 12/40 pathways were upregulated. After thresholding, 3/38 are upregulated.
> These results are substantially different.


```{r}
compare_original_threshold(tukey_original, tukey_thresholded, c("HALLMARK_MYC_TARGETS_V1", "HALLMARK_MYC_TARGETS_V2"), "Diffuse midline glioma")
```

#### Other LGGs


> Originally, 15/40 pathways were downregulated. After thresholding, 7/38 are downregulated.
> This is broadly consistent though with some loss of power.

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, c("HALLMARK_MYC_TARGETS_V1", "HALLMARK_MYC_TARGETS_V2"), "Other lowgrade glioma")
```


#### Pilocytic astrocytoma

> Originally, 18/40 pathways were downregulated. After thresholding, 11/38 are downregulated.
> This is broadly consistent though with some loss of power.

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, c("HALLMARK_MYC_TARGETS_V1", "HALLMARK_MYC_TARGETS_V2"), "Pilocytic astrocytoma")
```

#### Subependymal Giant Cell Astrocytoma

> Originally, 1/40 pathways was downregulated. After thresholding, none are downregulated
> This is broadly consistent.

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, c("HALLMARK_MYC_TARGETS_V1", "HALLMARK_MYC_TARGETS_V2"), "Subependymal Giant Cell Astrocytoma")
```



#### Pleomorphic xanthoastrocytoma

> Originally, 4/40 pathways were downregulated. After thresholding, none are downregulated.
> This is broadly consistent.

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, c("HALLMARK_MYC_TARGETS_V1", "HALLMARK_MYC_TARGETS_V2"), "Pleomorphic xanthoastrocytoma")
```

### E2F target pathways


#### MB

> Originally, 15/20 pathways were upregulated. After thresholding, 13/19 are upregulated.
> This is broadly consistent.

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_E2F_TARGETS", "Medulloblastoma")
```

#### ATRT

> Originally, 13/20 pathways were upregulated. After thresholding, 11/19 are upregulated.
> This is broadly consistent.

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_E2F_TARGETS", "Atypical Teratoid Rhabdoid Tumor")
```

#### Other HGGs


> Originally, 13/20 pathways were upregulated and 1/20 was downregulated. After thresholding, 6/19 are upregulated.
> These results have somewhat shifted

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_E2F_TARGETS", "Other highgrade glioma")
```


#### DMGs

> Originally, 13/20 pathways were upregulated. After thresholding, 6/19 are upregulated.
> These results are different.


```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_E2F_TARGETS", "Diffuse midline glioma")
```



#### Other LGGs


> Originally, 8/20 pathways were downregulated. After thresholding, 7/19 are downregulated.
> This is broadly consistent

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_E2F_TARGETS", "Other lowgrade glioma")
```


#### Pilocytic astrocytoma

> Originally, 8/20 pathways were downregulated. After thresholding, 9/19 are downregulated.
> This is broadly consistent

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_E2F_TARGETS", "Pilocytic astrocytoma")
```

#### Subependymal Giant Cell Astrocytoma

> Originally, 6/20 pathways were downregulated. After thresholding, 3/16 are downregulated.
> This is broadly consistent though with some loss of power

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_E2F_TARGETS", "Subependymal Giant Cell Astrocytoma")
```

#### Pleomorphic xanthoastrocytoma

> Originally, 6/20 pathways were downregulated. After thresholding, 3/19 are downregulated.
> This is broadly consistent.

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_E2F_TARGETS", "Pleomorphic xanthoastrocytoma")
```



### G2M checkpoint 

#### MB

> Originally, 15/20 pathways were upregulated. After thresholding, 13/19 are upregulated.
> This is broadly consistent.

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_G2M_CHECKPOINT", "Medulloblastoma")
```


#### ATRT

> Originally, 13/20 pathways were upregulated. After thresholding, 9/19 are upregulated.
> This is broadly consistent, although with some minor loss of power

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_G2M_CHECKPOINT", "Atypical Teratoid Rhabdoid Tumor")
```


#### Other HGGs

> Originally, 13/20 pathways were upregulated and 1/20 was downregulated. After thresholding, 6/19 are upregulated.
> These results have somewhat shifted


```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_G2M_CHECKPOINT", "Other highgrade glioma")
```

#### DMGs

> Originally, 13/20 pathways were upregulated. After thresholding, 5/19 are upregulated.
> These results are different.


```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_G2M_CHECKPOINT", "Diffuse midline glioma")
```



#### Other LGGs


> Originally, 9/20 pathways were downregulated. After thresholding, 7/19 are downregulated.
> This is broadly consistent

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_G2M_CHECKPOINT", "Other lowgrade glioma")
```


#### Pilocytic astrocytoma

> Originally, 10/20 pathways were downregulated. After thresholding, 9/19 are downregulated.
> This is broadly consistent

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_G2M_CHECKPOINT", "Pilocytic astrocytoma")
```

#### Subependymal Giant Cell Astrocytoma

> Originally, 6/20 pathways were downregulated. After thresholding, 1/19 are downregulated.
> This is broadly consistent though with some loss of power

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_G2M_CHECKPOINT", "Subependymal Giant Cell Astrocytoma")
```

#### Pleomorphic xanthoastrocytoma

> Originally, 7/20 pathways were downregulated. After thresholding, 3/19 are downregulated.
> This is broadly consistent.

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_G2M_CHECKPOINT", "Pleomorphic xanthoastrocytoma")
```


  
## Assessing specific comparisons: Part 2

In this section, we'll check individual trends as stated here in the manuscript:


https://github.com/AlexsLemonade/OpenPBTA-manuscript/blob/b0fdf6f24fe9c021f22e8b7da348708f68877ee5/content/03.results.md?plain=1#L277

> Schwannomas and neurofibromas, which have a documented inflammatory immune microenvironment of T and B lymphocytes as well as tumor-associated macrophages (TAMs), are driven by upregulation of cytokines such as IFN$\gamma$, IL-1, and IL-6, and TNF$\alpha$ [@doi:10.1093/noajnl/vdaa023].
> Indeed, we observed significant upregulation of these cytokines in GSVA hallmark pathways (Bonferroni-corrected p < 0.05) (**Figure {@fig:Fig5}B**)...


### IFN-gamma 


#### Schwannomas

> Originally, 13/20 pathways were upregulated. After thresholding, 9/19 are upregulated.
> Results are broadly consistent

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_INTERFERON_GAMMA_RESPONSE", "Schwannoma")
```

#### Neurofibromas

> Originally, 9/20 pathways were upregulated. After thresholding, 6/19 are upregulated.
> Results are broadly consistent

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_INTERFERON_GAMMA_RESPONSE", "Neurofibroma Plexiform")
```

### IL-6

#### Schwannomas


> Originally, 10/20 pathways were upregulated. After thresholding, 7/19 are upregulated.
> Results are broadly consistent

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_IL6_JAK_STAT3_SIGNALING", "Schwannoma")
```

#### Neurofibromas

> Originally, 8/20 pathways were upregulated. After thresholding, 7/19 are upregulated.
> Results are broadly consistent

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_IL6_JAK_STAT3_SIGNALING", "Neurofibroma Plexiform")
```


### TNF-alpha
 
#### Schwannomas

> Originally, 9/20 pathways were upregulated. After thresholding, 4/19 are upregulated.
> Results are broadly consistent but there is a bigger loss of power

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_TNFA_SIGNALING_VIA_NFKB", "Schwannoma")
```

#### Neurofibromas

> Originally, 9/20 pathways were upregulated. After thresholding, 4/19 are upregulated.
> Results are broadly consistent but there is a bigger loss of power

```{r}
compare_original_threshold(tukey_original, tukey_thresholded, "HALLMARK_TNFA_SIGNALING_VIA_NFKB", "Neurofibroma Plexiform")
```

## Assessing specific comparisons: Part 3

In this section, we'll check individual trends as stated here in the manuscript:

https://github.com/AlexsLemonade/OpenPBTA-manuscript/blob/b0fdf6f24fe9c021f22e8b7da348708f68877ee5/content/03.results.md?plain=1#L279

> We also observed significant upregulation of pro-inflammatory cytokines IFN$\alpha$ and IFN$\gamma$ in LGGs and craniopharyngiomas compared to medulloblastoma and ependymoma tumors (Bonferroni-corrected p < 0.05), both of which showed significant down-regulation of these cytokines (**Figure {@fig:Fig5}B**).

### IFN-alpha

#### LGG compared to MB

> Both original and thresholded results show that LGGs are significantly upregulated compared to MB.
> Results are consistent.

```{r}
# vector of all the LGG cancer groups
lgg_groups <- c("Other lowgrade glioma", "Pilocytic astrocytoma", "Pleomorphic xanthoastrocytoma", "Subependymal Giant Cell Astrocytoma")


combine_tukey_comparisons(tukey_original, tukey_thresholded, 
                          "HALLMARK_INTERFERON_ALPHA_RESPONSE", 
                          # Compare across the following groups:
                          c(lgg_groups, "Medulloblastoma")) %>%
  # filter to only rows where MB is a cancer_group (rather than an LGG compared to an LGG)
  filter("Medulloblastoma" == cancer_group1 | "Medulloblastoma" == cancer_group2)
                          
```


#### LGG compared to EPN

> Original results show that "Pilocytic astrocytoma" and "Subependymal Giant Cell Astrocytoma" are upregulated compared to EPN.
> Thresholded results show that "Pilocytic astrocytoma" and "Other low-grade glioma" are upregulated compared to EPN.
> Results are not identical but there is conceptual overlap.

```{r}
combine_tukey_comparisons(tukey_original, tukey_thresholded, 
                          "HALLMARK_INTERFERON_ALPHA_RESPONSE", 
                          # Compare across the following groups:
                          c(lgg_groups, "Ependymoma")) %>%
  # filter to only rows where EPN is a cancer_group (rather than an LGG compared to an LGG)
  filter("Ependymoma" == cancer_group1 | "Ependymoma" == cancer_group2)
                          
```



#### Craniopharyngioma compared to MB

> Both original and thresholded results show that cranio is upregulated compared to MB.
> Results are consistent.

```{r}
combine_tukey_comparisons(tukey_original, tukey_thresholded, 
                          "HALLMARK_INTERFERON_ALPHA_RESPONSE", 
                          # Compare across the following groups:
                          c("Medulloblastoma", "Craniopharyngioma")) 
                          
```



#### Craniopharyngioma compared to EPN

> Both original and thresholded results show that cranio is upregulated compared to EPN.
> Results are consistent.

```{r}
combine_tukey_comparisons(tukey_original, tukey_thresholded, 
                          "HALLMARK_INTERFERON_ALPHA_RESPONSE", 
                          # Compare across the following groups:
                          c("Ependymoma", "Craniopharyngioma")) 
                          
```





### IFN-gamma

#### LGG compared to MB

> Both original and thresholded results show that LGGs are significantly upregulated compared to MB.
> Results are consistent.

```{r}
combine_tukey_comparisons(tukey_original, tukey_thresholded, 
                          "HALLMARK_INTERFERON_GAMMA_RESPONSE", 
                          # Compare across the following groups:
                          c(lgg_groups, "Medulloblastoma")) %>%
  # filter to only rows where MB is a cancer_group (rather than an LGG compared to an LGG)
  filter("Medulloblastoma" == cancer_group1 | "Medulloblastoma" == cancer_group2)
                          
```


#### LGG compared to EPN

> Both original and thresholded results show that LGGs are significantly upregulated compared to EPN, _except_ for the thresholded "Pleomorphic xanthoastrocytoma" comparison which is not signficant.
> Results are consistent, with some loss of power.


```{r}
combine_tukey_comparisons(tukey_original, tukey_thresholded, 
                          "HALLMARK_INTERFERON_GAMMA_RESPONSE", 
                          # Compare across the following groups:
                          c(lgg_groups, "Ependymoma")) %>%
  # filter to only rows where EPN is a cancer_group (rather than an LGG compared to an LGG)
  filter("Ependymoma" == cancer_group1 | "Ependymoma" == cancer_group2)
                          
```



#### Craniopharyngioma compared to MB

> Both original and thresholded results show that cranio is upregulated compared to MB.
> Results are consistent.

```{r}
combine_tukey_comparisons(tukey_original, tukey_thresholded, 
                          "HALLMARK_INTERFERON_GAMMA_RESPONSE", 
                          # Compare across the following groups:
                          c("Medulloblastoma", "Craniopharyngioma")) 
                          
```



#### Craniopharyngioma compared to EPN

> Both original and thresholded results show that cranio is upregulated compared to EPN
> Results are consistent.

```{r}
combine_tukey_comparisons(tukey_original, tukey_thresholded, 
                          "HALLMARK_INTERFERON_GAMMA_RESPONSE", 
                          # Compare across the following groups:
                          c("Ependymoma", "Craniopharyngioma")) 
                          
```

## Conclusions


Previously-identified upregulated and downregulated pathways still generally follow the same trends for the relevant cancer groups, with some (not unexpected) power loss for some comparisons.
On the whole, results remain as we have reported them in the manuscript.

## Session info

```{r}
sessionInfo()
```
