---
title: "Add molecular subtype for samples that have hallmark Ewings Sarcoma fusions"
output: html_notebook
author: K S Gaonkar for D3b
date: January 2020
---

Identify sample IDs with hallmark _EWSR1_ fusions and subtype as `EWS` 

```{r}
library("tidyverse")
```

### Read in data

```{r}
# to get root directory
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

# data directory
dataDir <- file.path(root_dir, "data")

# histology file
clinical <- readr::read_tsv(file.path(dataDir, "pbta-histologies-base.tsv"),
                            guess_max = 10000) %>%
  select(Kids_First_Biospecimen_ID,
         Kids_First_Participant_ID, 
         experimental_strategy, 
         pathology_diagnosis,
         sample_id,
         sample_type)

# get ewings fusion calls from fusion-summary results file
ewings_foi_bsids <- readr::read_tsv(
  file.path(dataDir,
            "fusion_summary_ewings_foi.tsv")
  ) 

# get subset folder
results_dir <- "results"

# create if doesn't exist
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

### Get sample_id from RNA-Seq samples that have ewings sarcoma EWSR1 fusion

```{r}
# get a vector of biospecimen IDs for those biospecimens that should be subtyped
has_ews_fusion <- rowSums(ewings_foi_bsids[, -1]) > 0

# get rna sample_id which have hallmark fusion 
has_ews_fusion_sampleid <- ewings_foi_bsids %>%
  filter(has_ews_fusion) %>%
  left_join(clinical) %>%
  pull(sample_id)

```

### Get RNA-Seq where pathology_diagnosis is Ewings Sarcoma OR has Ewings Sarcoma _EWSR1_ fusion

```{r}

ews_rnaseq <- clinical %>%
  # pathology_diagnosis is Ewings Sarcoma
  filter(str_detect(pathology_diagnosis,"Ewings Sarcoma")|
           # OR sample_id associated with the RNA sample with hallmark fusion
         sample_id %in% has_ews_fusion_sampleid,
         experimental_strategy == "RNA-Seq") %>%
  select(Kids_First_Biospecimen_ID, 
         Kids_First_Participant_ID, 
         sample_id) 
```

### Get WGS samples where pathology_diagnosis is Ewings Sarcoma OR has sample_id associated with rna-seq sample with Ewings Sarcoma _EWSR1_ fusion

```{r}

ews_wgs <- clinical %>%
  # pathology_diagnosis is Ewings Sarcoma
  filter(str_detect(pathology_diagnosis,"Ewings Sarcoma")|
         # OR sample_id associated with the RNA sample with hallmark fusion
         sample_id %in% has_ews_fusion_sampleid,
         # filter to Tumor since we have Normal WGS as well
         sample_type == "Tumor",
         experimental_strategy == "WGS") %>%
  select(Kids_First_Biospecimen_ID, 
         Kids_First_Participant_ID, 
         sample_id) 

```


### Join DNA and RNA sample together to have molecular subtyping per "Kids_First_Participant_ID", and "sample_id"

```{r}

ews_subtype <- ews_rnaseq %>% 
  # merge DNA and RNA samples by sample_id
  full_join(ews_wgs,
             by = c("Kids_First_Participant_ID",
                    "sample_id"),
             suffix = c("_RNA", "_DNA")) %>%
  select(Kids_First_Participant_ID,
         Kids_First_Biospecimen_ID_RNA,
         Kids_First_Biospecimen_ID_DNA,
         sample_id,
         everything()) %>%
  dplyr::mutate(molecular_subtype="EWS")
```

### List which samples are molecular_subtype == "EWS" or "EWS, To be classified"

```{r}
ews_subtype 
```

#### Save to file

```{r}
ews_subtype %>%
  readr::write_tsv(file.path(results_dir, "EWS_results.tsv"))
```

