---
title: "Reclassify integrated_diagnosis, short_histology metadata column for samples that have hallmark Ewings Sarcoma fusions"
output: html_notebook
author: K S Gaonkar for D3b
date: January 2020
---

Identify `integrated_diagnosis`, `short_histology` from clinical file for sample IDs with hallmark _EWSR1_ fusions and reclassify as `Ewings Sarcoma` (integrated_diagnosis) and `EWS` (short_histology) and relabel `CNS EFT-CIC` tumors as `EWS`

```{r}
library("tidyverse")
```

### Read in data

```{r}
# to get root directory
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

# data directory
dataDir <- file.path(root_dir, "data")

# histology file
clinical <- readr::read_tsv(file.path(dataDir, "pbta-histologies.tsv"),
                            guess_max = 10000) %>%
  select(Kids_First_Biospecimen_ID,
         integrated_diagnosis,
         short_histology, 
         broad_histology, 
         Kids_First_Participant_ID, 
         experimental_strategy, 
         sample_id,
         sample_type)

# get ewings fusion calls from fusion-summary results file
ewings_foi_bsids <- readr::read_tsv(
  file.path(root_dir, 
            "analyses",
            "fusion-summary",
            "results",
            "fusion_summary_ewings_foi.tsv")
  ) 

# get subset folder
results_dir <- "results"

# create if doesn't exist
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

### Recode `CNS EFT-CIC` as `EWS`

According to [#624](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/624), tumors where `short_histology` is currently `CNS EFT-CIC` should be recoded to `EWS`.

```{r}
clinical <- clinical  %>%
  mutate(short_histology = case_when(
    short_histology == "CNS EFT-CIC" ~ "EWS",
    TRUE ~ short_histology
  ))
```


### Reclassify integrated diagnosis and short_histology of samples that have hallmark ewings sarcoma EWSR1 fusion

```{r}
# get a vector of biospecimen IDs for those biospecimens that should be reclassified
have_ews_fusion <- rowSums(ewings_foi_bsids[, -1]) > 0
ews_rnaseq_bsids <- ewings_foi_bsids %>%
  filter(have_ews_fusion) %>%
  pull(Kids_First_Biospecimen_ID)

# recode the different disease labels for RNA-seq samples only
# because we have fusion information for RNA-seq samples
clinical_rnaseq <- clinical %>%
  filter(experimental_strategy == "RNA-Seq") %>%
  select(-experimental_strategy) %>%
  # the `TRUE ~` steps leave the original disease label 
  mutate(
    integrated_diagnosis_reclassified = case_when(
      Kids_First_Biospecimen_ID %in% ews_rnaseq_bsids ~ "Ewings Sarcoma",
      TRUE ~ integrated_diagnosis
    ),
    short_histology_reclassified = case_when(
      Kids_First_Biospecimen_ID %in% ews_rnaseq_bsids ~ "EWS",
      TRUE ~ short_histology
    ),
    broad_histology_reclassified = case_when(
      Kids_First_Biospecimen_ID %in% ews_rnaseq_bsids ~ "Mesenchymal non-meningothelial tumor",
      TRUE ~ broad_histology
    ),
    Notes = if_else(
      Kids_First_Biospecimen_ID %in% ews_rnaseq_bsids,
      "Reclassified due to presence of hallmark EWS fusions",
      NA_character_
    )
  ) %>%
  select(-integrated_diagnosis, 
         -short_histology, 
         -broad_histology)

# we're going to join the WGS samples to RNA-seq
# so that the WGS biospecimens that correspond to RNA-seq specimens 
# are reclassified as well
clinical_wgs <- clinical %>%
  filter(experimental_strategy == "WGS" & sample_type == "Tumor") %>%
  select(Kids_First_Biospecimen_ID, 
         Kids_First_Participant_ID, 
         sample_id)

# Join together
clinical_reclassified <- clinical_rnaseq %>% 
  full_join(clinical_wgs,
             by = c("Kids_First_Participant_ID",
                    "sample_id"),
             suffix = c("_RNA", "_DNA")) %>%
  select(Kids_First_Participant_ID,
         Kids_First_Biospecimen_ID_RNA,
         Kids_First_Biospecimen_ID_DNA,
         sample_id,
         everything())
```


#### Save to file

```{r}
clinical_reclassified %>%
  filter(short_histology_reclassified == "EWS") %>%
  readr::write_tsv(file.path(results_dir, "EWS_results.tsv"))
```

### List which samples were reclassified as `Ewings Sarcoma`

```{r}
clinical_reclassified %>% 
  dplyr::filter(!is.na(Notes))
```
