---
title: "BS_F0GNWEJJ hypermutator genomic investigation"
author: "Jo Lynne Rokita (D3b)"
date: "2022"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
library(tidyverse)
```

## Investigation of BS_F0GNWEJJ for TP53 or MYCN genomic alterations for subtyping purposes

This Rmarkdown investigates specific genomic alterations in metastatic neuroblastoma (NBL) tumor specimen BS_F0GNWEJJ for Table 2. All other hypermutated tumors were HGG and have undergone molecular subtyping _and_ are _TP53_ altered. Here, we are investigating whether this NBL sample, which has not undergone subtyping, harbors any _TP53_ alteration or _MYCN_ amplification.

## Define directories and files
```{r define directories and input files}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
analysis_dir <- file.path(root_dir, "analyses")
data_dir <- file.path(root_dir, "data")

# Define input files
focal_cn_file <- file.path(data_dir, "consensus_seg_annotated_cn_autosomes.tsv.gz")
tp53_file <- file.path(analysis_dir, "tp53_nf1_score", "results", "tp53_altered_status.tsv")
```

## Read in files
```{r read in files}
tp53 <- read_tsv(tp53_file) 
focal_cn <- read_tsv(focal_cn_file) 
```

## Does this sample have a high TP53 score and/or TP53 copy loss
> No evidence for TP53 CNV, SV, Fusion, or hotspot SNV, but high TP53 score

```{r TP53 investigation}
id_of_interest <- "BS_F0GNWEJJ"

tp53 %>%
  filter(Kids_First_Biospecimen_ID_DNA %in% id_of_interest) %>%
  select(tp53_score, CNV_loss_evidence, SV_type, Fusion_evidence, hotspot, activating, tp53_altered)
```

## Does this sample have MYCN amplification?
> No evidence of MYCN amplification

```{r MYCN investigation}
focal_cn %>%
  filter(biospecimen_id %in% id_of_interest,
         gene_symbol == "MYCN")
```

## Session Info
```{r session info}
sessionInfo()
```