---
title: "Example queries for finding the consensus calls from the snv database"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

##Setup
```{r}
library(dplyr)
```

# Files and directories
```{r}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
db_file <- file.path(root_dir, "scratch", "snv_db.sqlite")
```

### Connect to the database

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), db_file)
DBI::dbListTables(con)
```

## Using dbplyr

### Set up tables
```{r}
meta <- tbl(con, "samples")
strelka <- tbl(con, "strelka")
lancet <- tbl(con, "lancet")
mutect <- tbl(con, "mutect")
vardict <- tbl(con, "vardict")
```

### Find intersection of all
Note that running the block below does no actual processing, it just sets up the query.
Processing will be only be done when we try to look at it.

```{r}
join_cols = c("Chromosome",
              "Start_Position",
              "Reference_Allele",
              "Allele",
              "Tumor_Sample_Barcode")

full_consensus <- strelka %>%
  inner_join(lancet,
             by = join_cols,
             suffix = c("s", "l")) %>%
  inner_join(mutect,
             by = join_cols,
             suffix = c("s", "m")) %>%
  inner_join(vardict, by = join_cols,
             suffix = c("s", "v")) %>%
  select(join_cols) 
```


Lets check that SQLite is doing what we might expect (using indexes)
```{r, results = 'hide'}
full_consensus %>% explain()
```
It is using the SNV indexes, so that is good. 


Print part of the query we just prepared. 
**Now** we are hitting the db and actually doing the processing.

```{r}
full_consensus %>% head(20)
```


### SQL equivalent

The full query is pretty long and ugly in SQL, but a two sample consensus is shown below.

```{r}
consensus_query <- "
  SELECT `S`.`Chromosome`,
    `S`.`Start_Position`,
    `S`.`Reference_Allele`,
    `S`.`Allele`,
    `S`.`Tumor_Sample_Barcode`
  FROM `strelka` as `S`
  INNER JOIN `lancet` AS `L`
  ON `S`.`Chromosome` = `L`.`Chromosome`
      AND `S`.`Start_Position` = `L`.`Start_Position`
      AND `S`.`Reference_Allele` = `L`.`Reference_Allele`
      AND `S`.`Allele` = `L`.`Allele`
      AND `S`.`Tumor_Sample_Barcode` = `L`.`Tumor_Sample_Barcode`
  LIMIT 20
  "

consensus_result <- DBI::dbGetQuery(con, consensus_query)
consensus_result
```

If you want to see the full query created by `dplyr`, you can use `dplyr::show_query()` to see the very long (but correct!) queries that it creates.

Note that if you want to update the database (i.e. to store the results of a query for posterity), 
the best way to do so is likely to write the SQL. 
That could be something like the following (not run):

```
CREATE TABLE `consensus` AS (
  SELECT `S`.`Chromosome`,
    `S`.`Start_Position`,
    `S`.`Reference_Allele`,
    `S`.`Allele`,
    `S`.`Tumor_Sample_Barcode`
  FROM `strelka` as `S`
  INNER JOIN `lancet` AS `L`
  ON `S`.`Chromosome` = `L`.`Chromosome`
      AND `S`.`Start_Position` = `L`.`Start_Position`
      AND `S`.`Reference_Allele` = `L`.`Reference_Allele`
      AND `S`.`Allele` = `L`.`Allele`
      AND `S`.`Tumor_Sample_Barcode` = `L`.`Tumor_Sample_Barcode`
)
```

To run this in R, you could use `DBI::dbExecute()`

Note that for the query above, it would probably be a good idea to add on an index equivalent to the ones present when the MAF tables were created:

```
CREATE INDEX `consensus_SNV` 
  ON consensus(
    `Chromosome`,
    `Start_Position`,
    `Reference_Allele`,
    `Allele`
  )
```