---
title: "Comparison of Expected and Observed MB Subtype Classification"
output:
  html_document:
    df_print: paged
params:
  expected_input:
    value: input/expected_class.rds
  observed_input:
    value: results/mb-classified.rds
---
  
```{r include = FALSE}
knitr::opts_chunk$set(comment = NA)
getOption(x = 'DT.warn.size', default = FALSE)
```

#### Code:

<details><summary>Load libraries</summary>
```{r load_packages, echo = TRUE}
# load libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(DT))
```
</details>

<details><summary>Directories</summary>
```{r}
# set data directory
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")

# set results directory
results_dir <- file.path(root_dir, "analyses", "molecular-subtyping-MB", "results")
```
</details>

<details><summary>Read input</summary>
```{r read_data, echo = TRUE}
# read input
clin <- read.delim(file.path(data_dir, "pbta-histologies.tsv"))
exp.class <- readRDS(params$expected_input)
obs.class <- readRDS(params$observed_input)

# format ambiguous expected classes 
exp.class <- exp.class %>%
  mutate(pathology_subtype = replace(pathology_subtype, 
                                     pathology_subtype == "Group 3 or 4", "Group3, Group4")) %>%
  mutate(pathology_subtype = gsub(" ", "", pathology_subtype))

# format clinical file
# RNA MB samples
clin.rna <- clin  %>%
  filter(short_histology == "Medulloblastoma",
         experimental_strategy == "RNA-Seq")  %>%
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = Kids_First_Biospecimen_ID) %>%
  dplyr::select(Kids_First_Participant_ID, sample_id, Kids_First_Biospecimen_ID_RNA)
# DNA MB samples
clin.wgs <- clin  %>%
  filter(short_histology == "Medulloblastoma",
         experimental_strategy == "WGS")  %>%
  dplyr::rename(Kids_First_Biospecimen_ID_DNA = Kids_First_Biospecimen_ID) %>%
  dplyr::select(Kids_First_Participant_ID, sample_id, Kids_First_Biospecimen_ID_DNA)
# combine both
clin.merged <- clin.wgs %>%
  right_join(clin.rna, by = c('sample_id','Kids_First_Participant_ID'))
```
</details>

<details><summary>Data Table</summary>
```{r data_table, echo = TRUE}
# custom datatable function
viewDataTable <- function(dat){
  DT::datatable(dat,
                rownames = FALSE,
                filter = "bottom",
                class = 'cell-border stripe',
                options = list(pageLength = 5,
                               searchHighlight = TRUE,
                               scrollX = TRUE,
                               dom = 'tpi',
                               initComplete = JS("function(settings, json) {",
                                                 "$(this.api().table().header()).css({'background-color':
                                            '#004467', 'color': '#fff'});","}"))
  )
}
```
</details>

<details><summary>Merge expected and observed class + compute accuracy </summary>
```{r compute_accuracy, echo = TRUE}
# function for merging expected and observed subtypes and returning accuracy
compute.accuracy <- function(expected.input, observed.input){
  # merge expected and observed subtypes
  merged.table <- expected.input %>%
    left_join(observed.input, by = c('Kids_First_Biospecimen_ID' = 'sample')) %>%
    mutate(match = str_detect(pathology_subtype, best.fit))
  
  # compute accuracy
  accuracy <- merged.table %>%
    filter(!is.na(pathology_subtype)) %>%
    group_by(match) %>%
    summarise(n = n()) %>%
    mutate(Accuracy = paste0(n/sum(n)*100, '%')) %>%
    filter(match) %>%
    .$Accuracy
  
  # return both as list
  output <- list(merged.table = merged.table, accuracy = accuracy)
  return(output)
}
```
</details>

<details><summary>Create consensus output + compute accuracy </summary>
```{r consensus_table, echo = TRUE}
# function for merging outputs from both classifiers
calc.consensus <- function(mm2s, medullo.classifier){
  # create df with common cols
  mm2s <- mm2s %>%
    dplyr::rename(MM2S_best_fit = best.fit) %>%
    dplyr::select(Kids_First_Biospecimen_ID, sample_id, 
           pathology_subtype, other_molecular_findings, MM2S_best_fit)
  medullo.classifier <- medullo.classifier %>%
    dplyr::rename(medulloPackage_best_fit = best.fit) %>%
    dplyr::select(Kids_First_Biospecimen_ID, sample_id, 
           pathology_subtype, other_molecular_findings, medulloPackage_best_fit)
  
  # join both
  consensus.mat <- mm2s %>%
    inner_join(medullo.classifier, 
               by = c('Kids_First_Biospecimen_ID', 'sample_id',
                      'pathology_subtype', 'other_molecular_findings'))

  # create consensus molecular subtype if both classifiers agree 
  # match consensus molecular subtype to the pathology report
  consensus.mat  <- consensus.mat %>%
    mutate(molecular_subtype = ifelse(MM2S_best_fit == medulloPackage_best_fit, MM2S_best_fit, NA)) %>%
    mutate(match = str_detect(pathology_subtype, molecular_subtype))
  
  # % accuracy
  consensus.accuracy <- consensus.mat %>%
    filter(!is.na(pathology_subtype)) %>%
    group_by(match) %>%
    summarise(n = n()) %>%
    mutate(Accuracy = paste0(n/sum(n)*100, '%')) %>%
    filter(match) %>%
    .$Accuracy
  
  # return both as list
  output <- list(consensus.mat = consensus.mat, consensus.accuracy = consensus.accuracy)
  return(output)
}
```
</details>

#### Details:

* % Accuracy is currently being calculated by matching observed and expected subtypes where expected subtype info is available. In case of ambiguous subtypes, we treat it as a match if the observed subtype matches with any one of the expected subtypes.

* Consensus tabs: Molecular subtype labels are assigned only when the two classifiers agree, leaving all other samples as unclassified. 

* Pathology report has subtype information on 32/122 (26.2%) samples. Following is the breakdown of pathology identified subtypes: 

```{r, echo=FALSE}
exp.count <- exp.class %>% 
  group_by(pathology_subtype) %>% 
  tally()
viewDataTable(exp.count)
```

#### Individual classifier output:

## {.tabset .tabset-fade}

### MM2S (batch-corrected)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- compute.accuracy(expected.input = exp.class, 
                        observed.input = obs.class$corrected_mat_MM2S)

# accuracy
mm2s.corrected.acc <- out$accuracy
print(paste0("Accuracy: ", mm2s.corrected.acc))

# output table
mm2s.corrected <- out$merged.table
viewDataTable(mm2s.corrected)
```

### MM2S (uncorrected)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- compute.accuracy(expected.input = exp.class, 
                        observed.input = obs.class$uncorrected_mat_MM2S)

# accuracy
mm2s.uncorrected.acc <- out$accuracy
print(paste0("Accuracy: ", mm2s.corrected.acc))

# output table
mm2s.uncorrected <- out$merged.table
viewDataTable(mm2s.uncorrected)
```

### medulloPackage (batch-corrected)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- compute.accuracy(expected.input = exp.class, 
                        observed.input = obs.class$`corrected_mat_medullo-classifier`)

# accuracy
medullo.classifier.corrected.acc <- out$accuracy
print(paste0("Accuracy: ", medullo.classifier.corrected.acc))

# output table
medullo.classifier.corrected <- out$merged.table
viewDataTable(medullo.classifier.corrected)
```

### medulloPackage (uncorrected)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- compute.accuracy(expected.input = exp.class, 
                        observed.input = obs.class$`uncorrected_mat_medullo-classifier`)

# accuracy
medullo.classifier.uncorrected.acc <- out$accuracy
print(paste0("Accuracy: ", medullo.classifier.uncorrected.acc))

# output table
medullo.classifier.uncorrected <- out$merged.table
viewDataTable(medullo.classifier.uncorrected)
```

#### Consensus output:

## {.tabset .tabset-fade}

### Consensus (batch-corrected)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- calc.consensus(mm2s = mm2s.corrected, 
                      medullo.classifier = medullo.classifier.corrected)

# accuracy
consensus.corrected.acc <- out$consensus.accuracy
print(paste0("Accuracy: ", consensus.corrected.acc))

# output table
consensus.corrected <- out$consensus.mat
viewDataTable(consensus.corrected)

# merge with clinical ids and save output
consensus.corrected.out <- clin.merged %>%
  inner_join(consensus.corrected %>% 
               dplyr::select(sample_id, molecular_subtype), by = 'sample_id')
write.table(consensus.corrected.out, 
            file = file.path(results_dir, 'MB_batchcorrected_molecular_subtype.tsv'), 
            row.names = F, quote = F, sep = "\t")
```

### Consensus (uncorrected)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- calc.consensus(mm2s = mm2s.uncorrected, 
                      medullo.classifier = medullo.classifier.uncorrected)

# accuracy
consensus.uncorrected.acc <- out$consensus.accuracy
print(paste0("Accuracy: ", consensus.uncorrected.acc))

# output table
consensus.uncorrected <- out$consensus.mat
viewDataTable(consensus.uncorrected)

# merge with clinical ids and save output
consensus.uncorrected.out <- clin.merged %>%
  inner_join(consensus.uncorrected %>% 
               dplyr::select(sample_id, molecular_subtype), by = 'sample_id')
write.table(consensus.uncorrected.out, 
            file = file.path(results_dir, 'MB_molecular_subtype.tsv'), 
            row.names = F, quote = F, sep = "\t")
```
