---
title: "Comparison of Expected and Observed MB Subtype Classification"
output:
  html_document:
    df_print: paged
params:
  expected_input:
    value: input/openPBTA-mb-pathology-subtypes.rds
  observed_input:
    value: results/mb-classified.rds
---
  
```{r include = FALSE}
knitr::opts_chunk$set(comment = NA)
getOption(x = 'DT.warn.size', default = FALSE)
```

#### Code:

<details><summary>Load libraries</summary>
```{r load_packages, echo = TRUE}
# load libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(plyr))
```
</details>

<details><summary>Directories</summary>
```{r}
# set data directory
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")

# set results directory
results_dir <- file.path(root_dir, "analyses", "molecular-subtyping-MB", "results")

# this file was created in the 01-filter-and-batch-correction.R
subset_clin_file <- file.path("input", "subset-mb-clinical.tsv")
```

</details>

<details><summary>Read input</summary>
```{r read_data, echo = TRUE}
# read in subset clinical mb file that was created in 01-filter-and-batch-correction.R
subset_clin <- read.delim(subset_clin_file)

exp_class <- readRDS(params$expected_input)
obs_class <- readRDS(params$observed_input)

# format ambiguous expected classes 
exp_class <- exp_class %>%
  mutate(pathology_subtype = recode(pathology_subtype, 
                                    "non-WNT" = "SHH, Group3, Group4", 
                                    "Group 3 or 4" = "Group3, Group4")) %>%
  mutate(pathology_subtype = gsub(" ", "", pathology_subtype))

# format subset clinical file
# RNA MB samples
clin_rna <- subset_clin  %>%
  filter(experimental_strategy == "RNA-Seq")  %>%
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = Kids_First_Biospecimen_ID) %>%
  dplyr::select(Kids_First_Participant_ID, sample_id, tumor_descriptor, Kids_First_Biospecimen_ID_RNA)

# DNA MB samples
clin_wgs <- subset_clin  %>%
  filter(experimental_strategy == "WGS")  %>%
  dplyr::rename(Kids_First_Biospecimen_ID_DNA = Kids_First_Biospecimen_ID) %>%
  dplyr::select(Kids_First_Participant_ID, sample_id, tumor_descriptor, Kids_First_Biospecimen_ID_DNA)

# combine both
clin_merged <- clin_wgs %>%
  right_join(clin_rna, by = c('sample_id', 'Kids_First_Participant_ID','tumor_descriptor'))
```
</details>

<details><summary>Data Table</summary>
```{r data_table, echo = TRUE}
# custom datatable function
viewDataTable <- function(dat){
  DT::datatable(dat,
                rownames = FALSE,
                filter = "top",
                class = 'cell-border stripe',
                options = list(pageLength = 5,
                               searchHighlight = TRUE,
                               scrollX = TRUE,
                               dom = 'tpi',
                               initComplete = JS("function(settings, json) {",
                                                 "$(this.api().table().header()).css({'background-color':
                                            '#004467', 'color': '#fff'});","}"))
  )
}
```
</details>

<details><summary>Merge expected and observed class + compute accuracy </summary>
```{r compute_accuracy, echo = TRUE}
# function for merging expected and observed subtypes and returning accuracy
compute_accuracy <- function(expected_input, observed_input){
  # merge expected and observed subtypes
  merged_table <- expected_input %>%
    left_join(observed_input, by = c('Kids_First_Biospecimen_ID' = 'sample')) %>%
    mutate(match = str_detect(pathology_subtype, best.fit))
  merged_table$sample_id <- as.factor(merged_table$sample_id)
  
  # compute accuracy
  accuracy <- merged_table %>%
    filter(!is.na(pathology_subtype)) %>%
    group_by(match) %>%
    dplyr::summarise(n = n()) %>%
    mutate(Accuracy = paste0(round(n/sum(n)*100, 2), '%')) %>%
    filter(match) %>%
    .$Accuracy
  
  # return both as list
  output <- list(merged_table = merged_table, accuracy = accuracy)
  return(output)
}
```
</details>

<details><summary>Create final output </summary>
```{r final_table, echo = TRUE}
# function for creating output for html and tsv files
create_final_output <- function(classifier_output, clin){
  # use molecular_subtype instead of best.fit (i.e. medulloPackage prediction column)
  classifier_output <- classifier_output %>%
    dplyr::rename(molecular_subtype = best.fit,
                  Kids_First_Biospecimen_ID_RNA = Kids_First_Biospecimen_ID) %>%
    dplyr::select(Kids_First_Biospecimen_ID_RNA, sample_id,
                  pathology_subtype, molecular_subtype, match)
  
  # merge with clinical data
  classifier_output <- clin %>%
    inner_join(classifier_output, by = c('sample_id', 'Kids_First_Biospecimen_ID_RNA'))
  
  # required for datatable filter
  classifier_output$sample_id <- as.factor(classifier_output$sample_id)
  
  # subset columns for html output
  classifier_output_html <- classifier_output %>%
    mutate(Kids_First_Biospecimen_ID = Kids_First_Biospecimen_ID_RNA) %>%
    dplyr::select(Kids_First_Biospecimen_ID, sample_id, pathology_subtype, molecular_subtype, match) %>%
    arrange(Kids_First_Biospecimen_ID)
  
  # subset columns for tsv output
  classifier_output_tsv <- classifier_output %>%
    dplyr::select(Kids_First_Participant_ID, sample_id, Kids_First_Biospecimen_ID_DNA,
                  Kids_First_Biospecimen_ID_RNA, molecular_subtype)
  
  # return as list
  output <- list(classifier_output_html = classifier_output_html,
                 classifier_output_tsv = classifier_output_tsv)
  return(output)
}
```
</details>

#### Details:

* % Accuracy is currently being calculated by matching observed and expected subtypes where expected subtype info is available. In case of ambiguous subtypes, we treat it as a match if the observed subtype matches with any one of the expected subtypes.

* Pathology report has subtype information on 43/122 (35.2%) samples. Following is the breakdown of pathology identified subtypes: 

```{r, echo=FALSE}
exp_count <- exp_class %>% 
  group_by(pathology_subtype) %>% 
  tally()
viewDataTable(exp_count)
```

#### Individual classifier and Consensus outputs:

## {.tabset .tabset-fade}

### MM2S

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- compute_accuracy(expected_input = exp_class, 
                        observed_input = obs_class$MM2S)

# accuracy
mm2s_accuracy <- out$accuracy
print(paste0("Accuracy: ", mm2s_accuracy))

# output table
mm2s_output <- out$merged_table
viewDataTable(mm2s_output)
```

### medulloPackage

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- compute_accuracy(expected_input = exp_class, 
                        observed_input = obs_class$medulloPackage)

# accuracy
medulloPackage_accuracy <- out$accuracy
print(paste0("Accuracy: ", medulloPackage_accuracy))

# output table
medulloPackage_output <- out$merged_table
viewDataTable(medulloPackage_output)
```

### Final output (medulloPackage)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- create_final_output(classifier_output = medulloPackage_output,
                           clin = clin_merged)


# output table
classifier_output_html <- out$classifier_output_html
viewDataTable(classifier_output_html)

# merged output with clinical ids
classifier_output_tsv <- out$classifier_output_tsv
classifier_output_tsv <- classifier_output_tsv %>%
  arrange(Kids_First_Biospecimen_ID_RNA)

classifier_output_tsv <- classifier_output_tsv %>%
  # adding "MB" to be consistent with other subtyping values
  dplyr::mutate(molecular_subtype = if_else(!is.na(molecular_subtype),
                                            paste("MB,",molecular_subtype),
                                            "MB, To be classified")) %>%
  write.table(file = file.path(results_dir, 'MB_molecular_subtype.tsv'),
            row.names = F, quote = F, sep = "\t")
```
