---
title: "Comparison of Expected and Observed MB Subtype Classification"
output:
  html_document:
    df_print: paged
params:
  expected_input:
    value: input/openPBTA-mb-pathology-subtypes.rds
  observed_input:
    value: results/mb-classified.rds
---
  
```{r include = FALSE}
knitr::opts_chunk$set(comment = NA)
getOption(x = 'DT.warn.size', default = FALSE)
```

#### Code:

<details><summary>Load libraries</summary>
```{r load_packages, echo = TRUE}
# load libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(plyr))
```
</details>

<details><summary>Directories</summary>
```{r}
# set data directory
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")

# set results directory
results_dir <- file.path(root_dir, "analyses", "molecular-subtyping-MB", "results")

# this file was created in the 01-filter-and-batch-correction.R
subset_clin_file <- file.path("input", "subset-mb-clinical.tsv")
```

</details>

<details><summary>Read input</summary>
```{r read_data, echo = TRUE}
# read in subset clinical mb file that was created in 01-filter-and-batch-correction.R
subset.clin <- read.delim(subset_clin_file)

exp.class <- readRDS(params$expected_input)
obs.class <- readRDS(params$observed_input)

# format ambiguous expected classes 
exp.class <- exp.class %>%
  mutate(pathology_subtype = recode(pathology_subtype, 
                                    "non-WNT" = "SHH, Group3, Group4", 
                                    "Group 3 or 4" = "Group3, Group4")) %>%
  mutate(pathology_subtype = gsub(" ", "", pathology_subtype))

# format subset clinical file
# RNA MB samples
clin.rna <- subset.clin  %>%
  filter(experimental_strategy == "RNA-Seq")  %>%
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = Kids_First_Biospecimen_ID) %>%
  dplyr::select(Kids_First_Participant_ID, sample_id, tumor_descriptor, Kids_First_Biospecimen_ID_RNA)

# DNA MB samples
clin.wgs <- subset.clin  %>%
  filter(experimental_strategy == "WGS")  %>%
  dplyr::rename(Kids_First_Biospecimen_ID_DNA = Kids_First_Biospecimen_ID) %>%
  dplyr::select(Kids_First_Participant_ID, sample_id, tumor_descriptor, Kids_First_Biospecimen_ID_DNA)

# combine both
clin.merged <- clin.wgs %>%
  right_join(clin.rna, by = c('sample_id','Kids_First_Participant_ID','tumor_descriptor'))
```
</details>

<details><summary>Data Table</summary>
```{r data_table, echo = TRUE}
# custom datatable function
viewDataTable <- function(dat){
  DT::datatable(dat,
                rownames = FALSE,
                filter = "top",
                class = 'cell-border stripe',
                options = list(pageLength = 5,
                               searchHighlight = TRUE,
                               scrollX = TRUE,
                               dom = 'tpi',
                               initComplete = JS("function(settings, json) {",
                                                 "$(this.api().table().header()).css({'background-color':
                                            '#004467', 'color': '#fff'});","}"))
  )
}
```
</details>

<details><summary>Merge expected and observed class + compute accuracy </summary>
```{r compute_accuracy, echo = TRUE}
# function for merging expected and observed subtypes and returning accuracy
compute.accuracy <- function(expected.input, observed.input){
  # merge expected and observed subtypes
  merged.table <- expected.input %>%
    left_join(observed.input, by = c('Kids_First_Biospecimen_ID' = 'sample')) %>%
    mutate(match = str_detect(pathology_subtype, best.fit))
  merged.table$sample_id <- as.factor(merged.table$sample_id)
  
  # compute accuracy
  accuracy <- merged.table %>%
    filter(!is.na(pathology_subtype)) %>%
    group_by(match) %>%
    dplyr::summarise(n = n()) %>%
    mutate(Accuracy = paste0(round(n/sum(n)*100, 2), '%')) %>%
    filter(match) %>%
    .$Accuracy

  # return both as list
  output <- list(merged.table = merged.table, accuracy = accuracy)
  return(output)
}
```
</details>

<details><summary>MB subtype propagation</summary>
```{r subtype_prop, echo = TRUE}
# function to propagate molecular subtype
prop.mb.subtype <- function(mat){
  # do this only for sample ids with multiple RNA ids
  if(nrow(mat) > 1){
    # A (rna id with predicted subtype)
    rna.id.with.molsubtype <- mat %>%
      filter(!is.na(molecular_subtype))
    
    # B (rna id without predicted subtype)
    rna.id.without.molsubtype <- mat %>%
      filter(is.na(molecular_subtype))
    
    # if A has a consensus molecular_subtype match and B has molecular_subtype as NA
    if(nrow(rna.id.with.molsubtype) == 1 & nrow(rna.id.without.molsubtype) == 1){
      # propagate consensus subtype of A if it matches to either prediction in B
      if(grepl(rna.id.with.molsubtype$molecular_subtype, rna.id.without.molsubtype$MM2S_best_fit) | 
         grepl(rna.id.with.molsubtype$molecular_subtype, rna.id.without.molsubtype$medulloPackage_best_fit)){
        rna.id.without.molsubtype <- rna.id.without.molsubtype %>%
          mutate(molecular_subtype = rna.id.with.molsubtype$molecular_subtype)
      }
    }
    
    # combine both rows
    mat <- rbind(rna.id.with.molsubtype, rna.id.without.molsubtype)
  } 
  return(mat)
}
```
</details>

<details><summary>Create consensus output + compute accuracy </summary>
```{r consensus_table, echo = TRUE}
# function for merging outputs from both classifiers
calc.consensus <- function(mm2s, medullo.classifier, clin){
  # create df with common cols
  mm2s <- mm2s %>%
    dplyr::rename(MM2S_best_fit = best.fit) %>%
    dplyr::select(Kids_First_Biospecimen_ID, sample_id, 
                  pathology_subtype, MM2S_best_fit)
  medullo.classifier <- medullo.classifier %>%
    dplyr::rename(medulloPackage_best_fit = best.fit) %>%
    dplyr::select(Kids_First_Biospecimen_ID, sample_id, 
                  pathology_subtype, medulloPackage_best_fit)
  
  # join both
  consensus.mat <- mm2s %>%
    inner_join(medullo.classifier, 
               by = c('Kids_First_Biospecimen_ID', 'sample_id',
                      'pathology_subtype'))
  
  # create consensus molecular subtype if both classifiers agree 
  # match consensus molecular subtype to the pathology report
  consensus.mat  <- consensus.mat %>%
    mutate(molecular_subtype = ifelse(MM2S_best_fit == medulloPackage_best_fit, MM2S_best_fit, NA)) %>%
    mutate(match = str_detect(pathology_subtype, molecular_subtype))
  
  # compute % accuracy
  consensus.accuracy <- consensus.mat %>%
    filter(!is.na(pathology_subtype)) %>%
    group_by(match) %>%
    dplyr::summarise(n = n()) %>%
    mutate(Accuracy = paste0(round(n/sum(n)*100, 2), '%')) %>%
    filter(match) %>%
    .$Accuracy
  
  # merge with clinical data
  consensus.mat.out <- clin.merged %>%
    inner_join(consensus.mat %>% 
                 mutate(Kids_First_Biospecimen_ID_RNA = Kids_First_Biospecimen_ID), 
               by = c('sample_id', 'Kids_First_Biospecimen_ID_RNA'))
  
  # resolve discrepancy with multiple sample ids
  # apply prop.mb.subtype function on same event (`sample_id`) with the same `tumor_descriptor`
  consensus.mat.out <- plyr::ddply(consensus.mat.out, 
                                   .variables = c("sample_id", "tumor_descriptor"), 
                                   .fun = function(x) prop.mb.subtype(mat = x))
  consensus.mat.out$sample_id <- as.factor(consensus.mat.out$sample_id)
  
  # add non-ambiguous molecular subtype reported from pathology report where there is no consensus 
  consensus.mat.out = consensus.mat.out %>%
    mutate(molecular_subtype = ifelse(MM2S_best_fit == medulloPackage_best_fit, MM2S_best_fit,
                                      ifelse(pathology_subtype %in% c("SHH", "Group3", "Group4", "WNT"), pathology_subtype, NA)))
  
  # subset columns for html output
  consensus.mat.html <- consensus.mat.out %>%
    mutate(Kids_First_Biospecimen_ID = Kids_First_Biospecimen_ID_RNA) %>%
    dplyr::select(Kids_First_Biospecimen_ID, sample_id, pathology_subtype, 
                  MM2S_best_fit, medulloPackage_best_fit, molecular_subtype, match) %>%
    arrange(Kids_First_Biospecimen_ID)
  
  # subset columns for tsv output
  consensus.mat.tsv <- consensus.mat.out %>%
    dplyr::select(Kids_First_Participant_ID, sample_id, Kids_First_Biospecimen_ID_DNA,
                  Kids_First_Biospecimen_ID_RNA, molecular_subtype)
  
  # return as list
  output <- list(consensus.mat = consensus.mat.html, 
                 consensus.mat.out = consensus.mat.tsv, 
                 consensus.accuracy = consensus.accuracy)
  return(output)
}
```
</details>

#### Details:

* % Accuracy is currently being calculated by matching observed and expected subtypes where expected subtype info is available. In case of ambiguous subtypes, we treat it as a match if the observed subtype matches with any one of the expected subtypes.

* Consensus tabs: For cases in which both classifiers agree, molecular subtype reflects the consensus prediction. For cases in which there is no consensus prediction, but we have a non-ambiguous pathology subtype, the molecular subtype reflects the pathology subtype.


* Propagation of molecular subtypes: 

  * IF there are two RNA specimens from the same event (`sample_id`) with the same `tumor_descriptor`, and 
  * IF one of the two RNA specimens has a consensus match from the MB classifiers, and 
  * IF the RNA specimen with mismatched classifications includes a match to the subtype of the consensus of the other RNA specimen, 
  * THEN propagate the subtype to the second RNA specimen.

* Pathology report has subtype information on 43/122 (35.2%) samples. Following is the breakdown of pathology identified subtypes: 

```{r, echo=FALSE}
exp.count <- exp.class %>% 
  group_by(pathology_subtype) %>% 
  tally()
viewDataTable(exp.count)
```

#### Individual classifier and Consensus outputs:

## {.tabset .tabset-fade}

### MM2S (batch-corrected)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- compute.accuracy(expected.input = exp.class, 
                        observed.input = obs.class$corrected_mat_MM2S)

# accuracy
mm2s.corrected.acc <- out$accuracy
print(paste0("Accuracy: ", mm2s.corrected.acc))

# output table
mm2s.corrected <- out$merged.table
viewDataTable(mm2s.corrected)
```

### MM2S (uncorrected)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- compute.accuracy(expected.input = exp.class, 
                        observed.input = obs.class$uncorrected_mat_MM2S)

# accuracy
mm2s.uncorrected.acc <- out$accuracy
print(paste0("Accuracy: ", mm2s.uncorrected.acc))

# output table
mm2s.uncorrected <- out$merged.table
viewDataTable(mm2s.uncorrected)
```

### medulloPackage (batch-corrected)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- compute.accuracy(expected.input = exp.class, 
                        observed.input = obs.class$`corrected_mat_medullo-classifier`)

# accuracy
medullo.classifier.corrected.acc <- out$accuracy
print(paste0("Accuracy: ", medullo.classifier.corrected.acc))

# output table
medullo.classifier.corrected <- out$merged.table
viewDataTable(medullo.classifier.corrected)
```

### medulloPackage (uncorrected)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- compute.accuracy(expected.input = exp.class, 
                        observed.input = obs.class$`uncorrected_mat_medullo-classifier`)

# accuracy
medullo.classifier.uncorrected.acc <- out$accuracy
print(paste0("Accuracy: ", medullo.classifier.uncorrected.acc))

# output table
medullo.classifier.uncorrected <- out$merged.table
viewDataTable(medullo.classifier.uncorrected)
```

### Consensus (batch-corrected)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- calc.consensus(mm2s = mm2s.corrected, 
                      medullo.classifier = medullo.classifier.corrected,
                      clin = clin.merged)

# accuracy
consensus.corrected.acc <- out$consensus.accuracy
print(paste0("Accuracy: ", consensus.corrected.acc))

# output table
consensus.corrected <- out$consensus.mat
viewDataTable(consensus.corrected)

# merged output with clinical ids 
consensus.corrected.out <- out$consensus.mat.out
consensus.corrected.out <- consensus.corrected.out %>%
  arrange(Kids_First_Biospecimen_ID_RNA)

consensus.corrected.out <- consensus.corrected.out %>%
  # adding "MB" to be consistent with other subtyping values
  dplyr::mutate(molecular_subtype = if_else(!is.na(molecular_subtype),
                                            paste("MB,",molecular_subtype),
                                            "MB, To be classified")) %>%
  write.table(
            file = file.path(results_dir, 'MB_batchcorrected_molecular_subtype.tsv'), 
            row.names = F, quote = F, sep = "\t")
```

### Consensus (uncorrected)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
out <- calc.consensus(mm2s = mm2s.uncorrected, 
                      medullo.classifier = medullo.classifier.uncorrected,
                      clin = clin.merged)

# accuracy
consensus.uncorrected.acc <- out$consensus.accuracy
print(paste0("Accuracy: ", consensus.uncorrected.acc))

# output table
consensus.uncorrected <- out$consensus.mat
viewDataTable(consensus.uncorrected)

# merged output with clinical ids 
consensus.uncorrected.out <- out$consensus.mat.out
consensus.uncorrected.out <- consensus.uncorrected.out %>%
  arrange(Kids_First_Biospecimen_ID_RNA)

consensus.uncorrected.out <- consensus.uncorrected.out %>%
  # adding "MB" to be consistent with other subtyping values
  dplyr::mutate(molecular_subtype = if_else(!is.na(molecular_subtype),
                                            paste("MB,",molecular_subtype),
                                            "MB, To be classified")) %>%
  write.table(file = file.path(results_dir, 'MB_molecular_subtype.tsv'), 
            row.names = F, quote = F, sep = "\t")
```

### Consensus comparison

```{r, echo = TRUE, warning = FALSE, message = FALSE}
# count of matches with pathology (corrected consensus)
consensus.corrected.match <- consensus.corrected %>%
  group_by(match) %>% 
  tally()
print(consensus.corrected.match)

# count of matches with pathology (uncorrected consensus)
consensus.uncorrected.match <- consensus.uncorrected %>%
  group_by(match) %>% 
  tally()
print(consensus.uncorrected.match)

# matches between the two consensus outputs
matches <- intersect(consensus.corrected %>%
                       filter(match == TRUE) %>%
                       .$Kids_First_Biospecimen_ID, 
                     consensus.uncorrected %>%
                       filter(match == TRUE) %>%
                       .$Kids_First_Biospecimen_ID)
print(length(matches))

# what's the difference?
consensus.corrected.matches <- consensus.corrected %>%
  filter(match) %>%
  .$Kids_First_Biospecimen_ID 
consensus.uncorrected.matches <- consensus.uncorrected %>%
  filter(match) %>%
  .$Kids_First_Biospecimen_ID 

# match in consensus corrected but not consensus uncorrected
print(setdiff(consensus.corrected.matches, consensus.uncorrected.matches))

# match in consensus uncorrected but not consensus corrected
print(setdiff(consensus.uncorrected.matches, consensus.corrected.matches))
```
