---
title: "Subtyping chordoma"
output: html_notebook
author: Mateusz Koptyra 
date: 20191121
---

This notebook prepares _SMARCB1_ copy number and expression data for chordoma samples for the purpose of identifying poorly-differentiated chordoma samples, which are characterized by loss of _SMARCB1_.

## Set up

```{r}
library(dplyr)
library(readr)
library(ggplot2)
```

### Read in data

```{r}
data_dir <- file.path("..", "..", "data")
histologies_df <- read_tsv(file.path(data_dir,
                                     "pbta-histologies.tsv"),
                           col_types = cols(
                             molecular_subtype = col_character()
                           ))
```

```{r}
# TODO: update to use consensus file with annotations included in the data 
# download?
focal_cn_df <- 
  read_tsv(file.path("..", "focal-cn-file-preparation", "results",
                     "consensus_seg_annotated_cn_autosomes.tsv.gz"))
```

```{r}
# we need to include the sample_id field from pbta-histologies.tsv in the final 
# table (field will allow #us to map between RNA-seq (e.g., SMARCB1 expression 
# values) and WGS data (e.g., SMARCB1 focal copy number status) from the same 
# event for a given individual). To get the SMARCB1 jitter plot in the photo 
# here #250 (comment), you will first need to read in the collapsed expression 
# data
expression_data <- 
  read_rds(file.path(data_dir, 
                     "pbta-gene-expression-rsem-fpkm-collapsed.stranded.rds"))
```

### Output

```{r}
# scatterplot
plot_dir <- "plots"
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir)
}
plot_file <- file.path(plot_dir, "smarcb1_expression_copy_status.png")

# tabular data
results_dir <- "results"
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
output_file <- file.path(results_dir, "chordoma_smarcb1_status.tsv")
```

## Prepare the data

Extracting the biospecimen IDs that correspond to chordoma samples

```{r}
chordoma_samples <- histologies_df %>%
  filter(short_histology == "Chordoma") %>% 
  pull(Kids_First_Biospecimen_ID)
```

Extracting the chordoma samples that have a loss of SMARCB1 from `focal_cn_df`

```{r}
chordoma_loss <- focal_cn_df %>% 
  filter(biospecimen_id %in% chordoma_samples, 
         gene_symbol == "SMARCB1",
         status == "loss")
chordoma_loss 
```

```{r}
# remove large copy number data frame
rm(focal_cn_df)
```

Modifying the output file to filter only chorodoma diagnoses and removing columns with information not relevant to the study
```{r}
chordoma_id_df <- histologies_df %>% 
  # only rows with chordoma samples
  filter(short_histology == "Chordoma") %>%
  # select only these columns that we'll need later
  select(Kids_First_Biospecimen_ID, sample_id, Kids_First_Participant_ID,
         experimental_strategy)
chordoma_id_df
```

Distinguishing the chordoma samples with no copy number change chromosome 22
```{r}
copy_neutral_df <- chordoma_id_df %>% 
  # the copy events can only be taken from WGS data not RNA-seq data
  # we also only want biospecimens where a loss was not recorded to avoid duplicates
  filter(experimental_strategy == "WGS",
         !(Kids_First_Biospecimen_ID %in% chordoma_loss$biospecimen_id)) %>%
  # if there's no loss, let's assume status is copy neutral
  mutate(status = "neutral") %>%
  # let's get the columns to match chordoma_loss
  dplyr::rename(biospecimen_id = Kids_First_Biospecimen_ID) %>%
  select(biospecimen_id, status)
copy_neutral_df
```

Joining chordoma samples (both with chr 22 loss and no copy number change)
```{r}
chordoma_copy <- chordoma_loss %>% 
  #join the losses with the neutrals to get a new data frame
  select(biospecimen_id, status) %>%
  bind_rows(copy_neutral_df)
chordoma_copy
```

Need to get the sample_id that corresponds to biospecimen_id into chordoma_copy so we can match WGS and RNA-seq biospecimens from the same event/sample:
```{r}
chordoma_copy <- chordoma_copy %>%
  # get only the Kids_First_Biospecimen_ID, sample_id columns from our identifier data.frame
  # then use biospecimen IDs to add the sample_id info
  inner_join(select(chordoma_id_df,
                    Kids_First_Biospecimen_ID,
                    sample_id),
             by = c("biospecimen_id" = "Kids_First_Biospecimen_ID"))
chordoma_copy
```

Look at SMARCB1 expression values only in chordoma

```{r}
# get the row that contains the SMARCB1 values
# gene symbols are rownames
smarcb1_expression <- expression_data[which(rownames(expression_data) == "SMARCB1"), ]

# now only the columns correspond to chordoma samples
smarcb1_expression <- smarcb1_expression[, which(colnames(expression_data) %in% chordoma_samples) ]
smarcb1_expression
```

```{r}
# remove large expression matrix that's no longer needed
rm(expression_data)
```

The `smarcb1_expression` is a not a friendly form ^^; Transposing needed: 

```{r}
# transpose such that samples are rows
smarcb1_expression <- t(smarcb1_expression) %>%
  # make a data.frame
  as.data.frame() %>%
  # we want the rownames that are biospecimen identifers as their own column called Kids_First_Biospecimen_ID
  tibble::rownames_to_column("Kids_First_Biospecimen_ID") %>%
  # give SMARCB1 column a slightly better column name
  dplyr::rename(SMARCB1_expression = SMARCB1)
smarcb1_expression
```

This also needs sample_id to add it in
```{r}
smarcb1_expression <- smarcb1_expression %>%
  inner_join(select(chordoma_id_df,
                    Kids_First_Biospecimen_ID,
                    sample_id),
             by = "Kids_First_Biospecimen_ID")
smarcb1_expression
```

Joining the copy number data with the expression data in this step
```{r}
chordoma_smarcb1_df <- smarcb1_expression %>%
  # any missing samples will get filled with NA when using a full join
  full_join(chordoma_copy, by = "sample_id") %>%
  dplyr::rename(Kids_First_Biospecimen_ID_DNA = Kids_First_Biospecimen_ID,
         Kids_First_Biospecimen_ID_RNA = biospecimen_id)

# this step adds in the participant identifier (sample_id to match between the two data.frame)
chordoma_smarcb1_df <- chordoma_id_df %>%
  select(sample_id, Kids_First_Participant_ID) %>%
  distinct() %>%
  inner_join(chordoma_smarcb1_df,
             by = "sample_id")

chordoma_smarcb1_df
```

```{r}
chordoma_smarcb1_df <- chordoma_smarcb1_df %>%
  select(Kids_First_Participant_ID, 
         Kids_First_Biospecimen_ID_DNA,
         Kids_First_Biospecimen_ID_RNA,
         sample_id,
         status,
         SMARCB1_expression) %>%
  # 'status' is replaced a more descriptive name
  dplyr::rename(focal_SMARCB1_status = status)
chordoma_smarcb1_df
```

### Plot _SMARCB1_ expression

copy loss vs. copy neutral 

```{r}
# this specifies that this is the data we want to plot
chordoma_smarcb1_df %>%
  # drop the sample that doesn't have WGS data
  tidyr::drop_na() %>%
  # this step specifies what should go on the x- and y-axes
  ggplot(aes(x = focal_SMARCB1_status,
             y = SMARCB1_expression)) +
  # we want a jitter plot where the points aren't too far 
  # apart that's what width does
  geom_jitter(width = 0.1) +
  # this is plotting the median as a blue diamond
  stat_summary(fun.y = "median", 
               geom = "point", 
               size = 3, 
               color = "blue", 
               shape = 18) +
               xlab("SMARCB1 status") +
   xlab("SMARCB1 status") +
   ylab("SMARCB1 expression")
```

```{r}
ggsave(filename = plot_file)
```

Write the table to file.

```{r}
chordoma_smarcb1_df %>%
  write_tsv(output_file)
```
