---
title: "Evaluate concordance between Mutect2 and Strelka2"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

Candace Savonen - CCDL for ALSF

This notebook analyzes the overlap and distinctions between between MuTect2 and 
Strelka2 results after having set the data up in 01-set-up.Rmd.
It addresses [issue \# 30 in OpenPBTA](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/30).

### Summary of Findings:

Due to some of the findings here with Mutect2 data, I suggest we should move 
forward with only Strelka2 data *OR* move forward with the variants that are 
detected by both algorithms. 

*Notable Findings:*  

- MuTect2 and Strelka2 detect [55,808 of the same variants](#venn-diagrams). 
This as defined by having the same Hugo gene symbol, base change, chromosomal 
start site, and sample of origin (see notebook [01-set-up.Rmd](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/mutect2-vs-strelka2/01-set-up.Rmd) 
for more details).
If moving forward we want only the most reliably called variants, this set of 
55,808 variants would give us plenty to work with. 

- MuTect2 and Strelka2 highly agree in their [Variant allele frequency (VAF) calculations](#variant-allele-frequency-scatterplot). 
This is good regardless of our choices moving forward. 

- Variants detected only by MuTect2 have a particularly low VAF compared to
variants detected only by Strelka2. (See [VAF Violin plots](#vaf-violin-plots)). 
These density plots suggest some of these MuTect2 calls may be noise. 
Although these low-VAF of MuTect2 could be identifying true variants, our further
analyses would probably benefit from a more robust, higher confidence set of 
variants. 

- MuTect2 also registers dinucleotide and larger variants where Strelka2 seems 
to break these variants into their single nucleotide changes.
In these analyses, these base changes have been grouped together and collectively
called [`long_changes`](#characterizing-the-variants-identified). 
The higher base resolution of Strelka2, and its ability to parse apart the SNVs
from each other, is more useful to us for this particular analyses, as the larger 
structural variants are better detected in the Manta or LUMPY analyses.

### Outline of analyses completed:

- [Venn diagram of variants detected](#venn-diagrams)  
- [VAF correlation hex plot](#variant-allele-frequency-scatterplot)   
- [Violin plot of VAF from MuTect2 or Strelka2 when they were detected by one or both algorithms](#vaf-violin-plots)  
- [Characterization of variants detected by one or both algorithms](#characterizing-the-variants-identified)  
  - [Number of base changes detected](#number-of-variants-categorized-by-their-base-change)  
  - [Percent of base changes detected](#percent-of-variants-by-their-base-change)  
  - [Percents of variants detected in coding versus non-coding regions](#percent-of-variants-that-are-in-coding-or-non-coding-regions)  
  - [Percent of variant type detected](#percent-of-variant-type)  

## Usage

To run this from the command line, use:
```
Rscript -e "rmarkdown::render('analyses/mutect2-vs-strelka2/02-analyze-concordance.Rmd', 
                              clean = TRUE)"
```
_This assumes you are in the top directory of the repository._
 
## Set Up

```{r}
# Will need hexbin for the hex plot
if (!("hexbin" %in% installed.packages())) {
  install.packages("hexbin")
}

# Will use colorblindr to make the plots colorblind-friendly
if (!("colorblindr" %in% installed.packages())) {
  devtools::install_github("clauswilke/colorblindr")
}

# Get magrittr pipe
`%>%` <- dplyr::`%>%`
```

Barplot wrapper function.

```{r}
barplot_maf_var <- function(df = vaf_df, variable_name = NULL, filter_cutoff = 0,
                            as_percent = TRUE, omit.na = FALSE) {
  # Creates a ggplot barplot for the numbers of variants that fall into each
  # category of the variable selected.
  #
  # Args:
  #   df: combined data.frame with both datasets (vaf_df)
  #   variable_name: a character string that indicates the base variable name
  #                  from the original MAF file
  #   filter_cutoff: variable categories with less than this cutoff will be removed
  #   as_percent: If true, will report numbers as percentages of the group they
  #               are from
  #   omit.na: If true, will remove NA's as a category
  #
  # Returns:
  #   ggplot2 grouped barplot with mutect_only, strelka2_only, and both as groups

  # Reconfigure dataset based on chosen variable, get numbers of variants for
  # each dataset and variable category
  count_df <- df %>%
    dplyr::select(
      paste0(variable_name, c(".strelka2", ".mutect2")),
      dataset
    ) %>%
    tidyr::gather(
      key = "data", value = "variable",
      paste0(variable_name, c(".strelka2", ".mutect2"))
    ) %>%
    dplyr::mutate("data" = gsub(paste0(variable_name, "."), "", data)) %>%
    dplyr::group_by(dataset, data, variable) %>%
    dplyr::summarise(count = dplyr::n())

  # Calculate sum totals for each data group
  totals <- count_df %>%
    dplyr::group_by(dataset) %>%
    dplyr::summarise(total = sum(count, na.rm = TRUE))

  # Calculate percentages based on totals and apply filter_cutoff
  count_df <- count_df %>%
    as.data.frame() %>%
    dplyr::mutate("percent" = count_df$count / totals$total[match(
      count_df$dataset,
      totals$dataset
    )]) %>%
    dplyr::mutate(report = ifelse(rep(as_percent, nrow(count_df)), percent, count)) %>%
    dplyr::filter(report > filter_cutoff)

  # Omit NAs if that was specified
  if (omit.na) {
    count_df <- dplyr::filter(count_df, !is.na(variable))
  }
  # Plot this!
  ggplot2::ggplot(count_df, ggplot2::aes(
    x = reorder(variable, -report),
    y = report, fill = dataset
  )) +
    ggplot2::geom_bar(position = "dodge", stat = "identity") +
    ggplot2::theme_classic() +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1)) +
    colorblindr::scale_fill_OkabeIto() +
    ggplot2::xlab("") +
    ggplot2::ylab(ifelse(as_percent, "Percent of Data Group", "Count"))
}
```

Read in the combined dataset that was set up in `01-set-up.Rmd`, 
'combined_results.tsv'.
This file has been zipped so it can be shared on GitHub more easily. 
If no unzipped version exists, we will unzip this.

```{r}
if (!file.exists(file.path("results", "combined_results.tsv"))) {
  unzip(file.path("results", "combined_results.tsv.zip"))
}
```

Read in the combined dataset, 'combined_results.tsv.'

```{r}
vaf_df <- readr::read_tsv(file.path("results", "combined_results.tsv"))
```

## Venn Diagrams

How many variants do both MuTect2 and Strelka2 agree on? Here we define a variant
as being identical in both datasets if its Hugo gene symbol, base change, 
chromosomal start site, and sample of origin. 
The variable `mutation_id` contains this information for each sample in a 
concatenated form, e.g. "ZMYND19_T>G_137590214_BS_5ZRZC3ZM".
It was previously set up in [01-set-up.Rmd](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/mutect2-vs-strelka2/01-set-up.Rmd).

```{r}
# Get the total count for variants found by each algorithm or by both.
count <- summary(as.factor(vaf_df$dataset))
```

Use the count object from above to make a Venn Diagram

```{r}
# Make the Venn diagram
grid::grid.newpage()
venn.plot <- VennDiagram::draw.pairwise.venn(
  area1 = count[3] + count[1],
  area2 = count[2] + count[1],
  cross.area = count[1],
  category = c("Strelka2", "MuTect2"),
  fill = c("blue", "yellow"),
  cex = 2,
  cat.cex = 1.5,
  cat.dist = c(-0.04, -0.031),
  ext.pos = 0,
  ext.dist = -0.01,
  ext.length = .8,
  ext.line.lwd = 2,
  ext.line.lty = "dashed"
)
grid::grid.draw(venn.plot) # Draw plot
```

Save the Venn diagram plot to a PDF. 

```{r}
# Make filename to save plot as
venn.plot.file <- file.path(
  "plots",
  "strelka2_mutect2_venn_diagram.pdf"
)
pdf(venn.plot.file)
grid::grid.draw(venn.plot)
dev.off()
```

## Variant allele frequency scatterplot

For variants identified in both datasets, are the VAF's similar? 
The VAFs for each mutation in each dataset were calculated by 
`t_alt_count) / (t_ref_count + t_alt_count) `, as following the [code used in 
`maftools`](https://github.com/PoisonAlien/maftools/blob/master/R/plot_vaf.R)
(See [01-set-up.Rmd](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/mutect2-vs-strelka2/01-set-up.Rmd)).

```{r}
# Calculate Pearson's correlation
pearson_r <- cor.test(vaf_df$vaf.strelka2, vaf_df$vaf.mutect2)

# Plot these data as a hex plot
vaf_df %>%
  ggplot2::ggplot(ggplot2::aes(x = vaf.strelka2, y = vaf.mutect2)) +
  ggplot2::geom_hex() +
  ggplot2::theme_classic() +
  ggplot2::xlab("VAF for each mutation for Strelka2") +
  ggplot2::ylab("VAF for each mutation for MuTect2") +
  ggplot2::ggtitle(
    label = "",
    subtitle = paste("R = ", round(pearson_r$estimate, 3))
  )
```

Overall, the VAF are highly correlated between the two algorithms.

## VAF Violin plots

How do the VAF distributions of MuTect2, Strelka2 look for variants found by
both algorithms, versus variants found only by one algorithm? 

```{r}
# Set up the VAF data for the density plot
vaf_df %>%
  tidyr::gather(key = "data", value = "vaf", vaf.strelka2, vaf.mutect2) %>%
  dplyr::mutate(data = gsub("^vaf.", "", data)) %>%
  dplyr::mutate(data.group = paste0(dataset, ":", data, " VAF")) %>%
  dplyr::filter(!is.na(vaf)) %>%
  # Plot it
  ggplot2::ggplot(ggplot2::aes(data.group, vaf)) +
  ggplot2::geom_violin(fill = "light blue") +
  ggplot2::theme_classic() +
  ggplot2::ylab("Density of VAF") +
  ggplot2::xlab(" ")
```

Variants that are only detected by MuTect2 appear to be particularly low VAF, 
even compared to variants only detected by Strelka2.
This may indicate it's picking up noise. 

## Characterizing the variants identified

What types of variants are are the most discrepant between the algorithms as 
compared to variants that are detected by both consistently?
We will make a series of grouped bar plots to try to get an overall picture of 
these variants.
I made the 

### Number of variants categorized by their base change

Note that `long_change` category is a combination of all base changes that are
more than a single nucleotide (See [01-set-up.Rmd](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/mutect2-vs-strelka2/01-set-up.Rmd)
for the code on how this was set up). 
Note that only MuTect2 reports these data as multiple base changes as opposed to 
parsing these changes a part as single base changes at a time.
Due to the lower base resolution lack of discernment of MuTect2 in this regard, 
this is another reason avoid using MuTect2 calls alone. 

```{r}
barplot_maf_var(
  df = vaf_df, variable_name = "change",
  as_percent = FALSE, omit.na = TRUE, filter_cutoff = 0
)
```

### Percent of variants by their base change

This is the same plot as above except normalized to be the percent of each data
group (`mutect2_only`, `strelka2_only`, or `both`).

MuTect2 shows a disproportional amount of deletions. 
It appears C>T and G>A changes are easiest for both algorithms to detect. 

```{r}
barplot_maf_var(
  df = vaf_df, variable_name = "change", filter_cutoff = 0,
  as_percent = TRUE, omit.na = TRUE
)
```

### Percent of variants that are in coding or non-coding regions

It's not surprising that these are mostly coding variants detected, however, 
there doesn't appear to be any strong biases between the groups. 

```{r}
barplot_maf_var(
  df = vaf_df, variable_name = "coding", filter_cutoff = 0,
  as_percent = TRUE, omit.na = TRUE
)
```

### Percent of variant type 

Only MuTect2 appears to record changes as DNP's or TNP's. 
Again we see MuTect2 shows more deletions. 

```{r}
barplot_maf_var(
  df = vaf_df, variable_name = "Variant_Type", filter_cutoff = 0,
  as_percent = TRUE, omit.na = TRUE
)
```

Session Info: 

```{r}
sessionInfo()
```
