---
title: "Survival analysis by TP53 and telomerase activity"
authors: Run Jin (D3B), Jo Lynne Rokita (D3B), and Stephanie Spielman (CCDL)
output: 
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---

Note that for models that consider the `cancer_group` predictor or are performed separately for each cancer group *only consider cancer groups where `DECEASED >= 3`.* There are seven such groups.

## Setup

#### Packages and functions


```{r libraries}
library(survival)
library(ggpubr)
library(tidyverse)
```

#### Set up files and directories

Set up input and output directories:

```{r Set up directories}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses")

tp53_dir <- file.path(analysis_dir, "tp53_nf1_score")
telomerase_dir <- file.path(analysis_dir, "telomerase-activity-prediction")

results_dir <- "results"
plots_dir <- "plots"
util_dir <- "util"

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

Make output directories and source the util function file.

```{r source_functions}
# Source functions used to build and export models
source(file.path(util_dir,"survival_models.R"))
```

Set up file paths.

```{r file_paths}

# Define input files:
metadata_file <- file.path(data_dir, "pbta-histologies.tsv")

# These analyses consider only stranded libraries:
tp53_file <- file.path(tp53_dir, 
                        "results", 
                        "tp53_altered_status.tsv")

tel_stranded_file <- file.path(telomerase_dir,
                               "results",
                               "TelomeraseScores_PTBAStranded_FPKM.txt")
```

## Prepare data for modeling

#### Import and process the metadata

```{r Read in metadata}
# get primary tumor and remove cell lines
histologies_rna <- read_tsv(metadata_file, guess_max = 10000) %>%
  filter(composition=="Solid Tissue",
                  tumor_descriptor == "Initial CNS Tumor",
                  experimental_strategy == "RNA-Seq", 
                  RNA_library == "stranded") %>%
  rename(Kids_First_Biospecimen_ID_RNA = Kids_First_Biospecimen_ID) %>% 
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID_RNA, cancer_group, OS_status, OS_days, PFS_days) %>%
  distinct() 

# read in tp53 scores
tp53_scores <- read_tsv(tp53_file) %>%
  select(Kids_First_Biospecimen_ID_RNA, sample_id, tp53_score) %>%
  distinct()

# read in telomerase scores
tel_scores <- read_tsv(tel_stranded_file) %>%
  rename(Kids_First_Biospecimen_ID_RNA = SampleID,
                telomerase_score = NormEXTENDScores) %>%
  arrange(Kids_First_Biospecimen_ID_RNA)


# combine tp53 scores and telomerase score to the dataframe
# note that we do not need to recode categories since the survival_models.R functions will take care of it
meta_indep <- histologies_rna %>%
  # add tp53 and telomerase scores by matching Kids_First_Biospecimen_ID_RNA
  left_join(tp53_scores, by = "Kids_First_Biospecimen_ID_RNA") %>%
  left_join(tel_scores, by = "Kids_First_Biospecimen_ID_RNA") %>% 
  arrange(Kids_First_Biospecimen_ID_RNA) %>%
  distinct(Kids_First_Participant_ID, .keep_all=TRUE) %>%
  # remove samples without survival data
  drop_na(OS_days) %>%
  # convert to years
  mutate(OS_years = OS_days / 365.25)

# Show the data we will use for analysis
meta_indep
```
 
#### Create subsets of data for different models

We create two new tibbles from `meta_indep`:

+ `meta_indep_hgg` which contains a new column `hgg_group` that contains levels (in order) `non-HGAT` and `HGAT` indicating whether it's an HGG sample
+ `meta_indep_cancers` which is subsetted to only certain cancer groups of interest that have suitable sample sizes which we find here.

```{r hgg_split}
# Define HGG groups
hgg_cancer_groups <- c("High-grade glioma astrocytoma", "Diffuse midline glioma", "Diffuse intrinsic pontine glioma") 
hgg_levels <- c("non-HGAT", "HGAT")

# separate into HGG or non-HGG and make the new column a factor
meta_indep_hgg <- meta_indep %>% 
   mutate(hgg_group = ifelse(cancer_group %in% hgg_cancer_groups,
                             hgg_levels[2], # second is HGAT 
                             hgg_levels[1]), # first is non-HGAT
          hgg_group = factor(hgg_group, levels = hgg_levels) # ensures non-HGAT is the reference level
   )
```

```{r find_cancer_groups_to_use}
# Define cancer groups of interest with higher Ns
# We want to only include groups that have at least THREE in the DECEASED category. 
# Find them here:
cancer_groups_of_interest_df <- meta_indep %>%
  drop_na() %>%
  count(cancer_group,OS_status) %>%
  spread(OS_status, n,
         # any missing values (indicating 0 deceased or 0 living) should be 0, not NA
         fill = 0) %>%
  filter(DECEASED >= 3) 
# show result: there are 7 groups to keep
cancer_groups_of_interest_df

# Subset to only the cancer groups with potentially suitable sample size
meta_indep_cancer_groups <- meta_indep %>%
  filter(cancer_group %in% cancer_groups_of_interest_df$cancer_group) %>%
  # set reference level to Medulloblastoma
  mutate(cancer_group = fct_relevel(cancer_group, "Medulloblastoma"))

# How many patients got removed?
nrow(meta_indep) - nrow(meta_indep_cancer_groups)
```

## Univariate Models

 
Here, we run a survival analyses for several _independent_ predictors, one at a time. We use cox regression for all analyses (including categorical), as demonstrated in `survival-analysis_template.Rmd`.

+ `tp53_score` (continuous)
+ `telomerase_score` (continuous)
+ `hgg_group` (categorical)
+ `cancer_group` (categorical)



### TP53 


```{r cox_tp53}
fit_save_model(meta_indep, 
              "tp53_score", 
             file.path(results_dir, "cox_terms_tp53.RDS"))

```


### Telomerase score

```{r cox_telo}
fit_save_model(meta_indep, 
              "telomerase_score", 
             file.path(results_dir, "cox_terms_telomerase.RDS"))
```


###  HGG group

> non-HGAT is the reference level.

```{r cox_hgg}
fit_save_model(meta_indep_hgg, 
              "hgg_group", 
             file.path(results_dir, "cox_terms_hgg_group.RDS"))
```

### Cancer group (cox)

> MB is the reference level

```{r cox_cancer_group}
fit_save_model(meta_indep_cancer_groups, 
               "cancer_group", 
               file.path(results_dir, "cox_terms_cancer_group.RDS"))
```



### Cancer group (log rank)

Here, we perform a log-rank analysis specifically for the `cancer_group` predictor, using data that only considers the 7 cancer groups determined to have suitable sample sizes (`DECEASED >= 3`). The output below shows:

+ Table of model coefficients
+ Matrix of `Pairwise comparisons using Log-Rank test`, which shows which cancer groups are significantly different from one another

Note that confidence intervals are so large that survival plots are not clearly legible, so we don't make that visualization here.

```{r cancer_group_log_rank}

# Print model coefficients
fit_save_model(meta_indep_cancer_groups,
               "cancer_group",
               file.path(results_dir,
                         "log_rank_survival_per_cancer_group.RDS"),
               test = "log.rank")


# Compare pairwise with Bonferroni correction for multiple testing
survminer::pairwise_survdiff(survival::Surv(time = OS_years, event = OS_status) ~ cancer_group,
                             meta_indep_cancer_groups %>% mutate(OS_status = ifelse(OS_status == "LIVING", 0, 1)), 
                             p.adjust.method = "bonferroni")

```

## Multivariate models 


Background on hazard ratios and output, from [ref](http://www.sthda.com/english/wiki/cox-proportional-hazards-model):

>"Hazard ratio above 1 indicates a covariate that is positively associated with the event probability, and thus negatively associated with the length of survival.
>
>In summary,
>
>HR = 1: No effect
>HR < 1: Reduction in the hazard
>HR > 1: Increase in Hazard" 
>
>The Cox regression results can be interpreted as follow:
_Statistical significance._ The column marked "z" gives the Wald statistic value. It corresponds to the ratio of each regression coefficient to its standard error (z = coef/se(coef)). The wald statistic evaluates, whether the beta (β) coefficient of a given variable is statistically significantly different from 0. From the output above, we can conclude that the variable sex have highly statistically significant coefficients.
>
>_The regression coefficients._ The second feature to note in the Cox model results is the the sign of the regression coefficients (coef). A positive sign means that the hazard (risk of death) is higher, and thus the prognosis worse, for subjects with higher values of that variable. The variable sex is encoded as a numeric vector. 1: male, 2: female. The R summary for the Cox model gives the hazard ratio (HR) for the second group relative to the first group, that is, female versus male. The beta coefficient for sex = -0.53 indicates that females have lower risk of death (lower survival rates) than males, in these data.
>
>_Hazard ratios._ The exponentiated coefficients (exp(coef) = exp(-0.53) = 0.59), also known as hazard ratios, give the effect size of covariates. For example, being female (sex=2) reduces the hazard by a factor of 0.59, or 41%. Being female is associated with good prognostic.
>
>_Confidence intervals of the hazard ratios._ The summary output also gives upper and lower 95% confidence intervals for the hazard ratio (exp(coef)), lower 95% bound = 0.4237, upper 95% bound = 0.816.
>
>_Global statistical significance of the model._ Finally, the output gives p-values for three alternative tests for overall significance of the model: The likelihood-ratio test, Wald test, and score logrank statistics. These three methods are asymptotically equivalent. For large enough N, they will give similar results. For small N, they may differ somewhat. The Likelihood ratio test has better behavior for small sample sizes, so it is generally preferred.
>
> Below, HR = exp(coeff) 


### Interaction model: `tp53_score*telomerase_score*hgg_group` 

Interaction terms are generally not signifcant, except `tp53_score:telomerase_score` which is only marginally significant at `P=0.0358`. This result suggests an additive model is more appropriate.

```{r multi_interaction_3terms}
fit_save_model(meta_indep_hgg,    
             "tp53_score*telomerase_score*hgg_group",
             file.path(results_dir, "cox_interaction_terms_tp53_telomerase_hgg.RDS"),
             "multivariate"
            )

  
```



### Additive model: `tp53_score+telomerase_score+hgg_group` 



```{r multi_additive_3terms}
fit_save_model(meta_indep_hgg,    
             "tp53_score+telomerase_score+hgg_group",
             file.path(results_dir, "cox_additive_terms_tp53_telomerase_hgg.RDS"),
             "multivariate")
```



### Additive models: ``tp53_score+telomerase_score` _for each cancer group_

> Note the model for DIPG does not converge!
> ONLY the EPN model shows significant effects for both `tp53_score` and `telomerase_score`, although CIs are fairly wide. The DPG model is marginally significant for `tp53_score` only.

```{r multivar-fit-2terms}
base_filename <- "cox_per_cg_terms_tp53_telomerase.RDS" # all file names here will follow this convention

for (cg in cancer_groups_of_interest_df$cancer_group) {
  cat("==============", cg, "==================")
  # filter to cancer group of interest
  fit_data <- meta_indep %>% 
    filter(cancer_group == cg)
  
  # Define filename where model will be saved
  filename <- file.path(
    results_dir,
    str_replace(base_filename, "cg", cg)
  )
  
  # Fit model
  fit_save_model(fit_data, 
                 "tp53_score + telomerase_score", 
                 filename, 
                 model_type = "multivariate") %>% print()
}


```


## Session Info

```{r Print session info}
sessionInfo()
```