---
title: "Survival Analysis for molecular subtypes of HGG"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: C. Savonen for ALSF CCDL, Krutika Gaonkar for D3b, Jo Lynne Rokita for D3b
date: 2019,, 2021
params:
  plot_ci: TRUE
---

**Purpose:** 

Runs survival analysis models for subtypes of HGG tumors. 

## Usage 

Uses a wrapper function (`survival_analysis`) from utils folder. 

## Setup

#### Packages and functions

Read in set up script.

```{r Set up library}
library(survival)
library(ggpubr)
library(tidyverse)

# This script contains a wrapper function that can conduct the survival analyses
source(file.path("util", "survival_models.R"))

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

#### Set up files and directories

Set up output directories. 

```{r Set up directories}
data_dir <- file.path("..", "..", "data")
results_dir <- "results"
plots_dir <- "plots"
```

Make output directories.

```{r Make output directories}
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

Declare input file paths and useful variables

```{r Set input file paths}
metadata_file <- file.path(data_dir, "pbta-histologies.tsv")
independent_primary_file <- file.path(
  data_dir,
  "independent-specimens.wgswxs.primary.tsv")

# Number of participants required to consider a subtype in survival analysis
n_participants <- 3
```

Declare output file paths for HGG/DMG all subtypes. 

```{r Set output file paths}
kap_meier_plot_file <- file.path(plots_dir, "KM_hgg_subtypes.pdf")
kap_meier_model_file <- file.path(results_dir, "logrank_hgg_subtypes.tsv")
```

## Import the metadata

```{r Read in metadata}
# choose HGG samples
# remove RNA-Seq because subtyping in HGG is from DNA abberations and samples where only RNA-Seq is available cannot be subtyped
# remove cell-lines from this analysis as well
metadata <- readr::read_tsv(
  metadata_file,
  col_types = readr::cols(
    molecular_subtype = readr::col_character())) %>%
  dplyr::filter(
    experimental_strategy != "RNA-Seq" &
      broad_histology == "Diffuse astrocytic and oligodendroglial tumor" &
      composition=="Solid Tissue" &
      # remove oligo
      cancer_group != "Oligodendroglioma" &
      # remove samples with no OS data
      !is.na(OS_days)) %>%
  mutate(
    OS_years = OS_days / 365.25
  )
```

#### Using independent_primary if multiple subtypes per patient are found 
This way we subset the metadata to 1 single primary sample per patient ID

```{r Read in independent}

# Patients with multiple different molecular subtype might affect downstream analysis for survival so using independent_primary sample set which has 1-1 sample-patient matches
independent_primary <- readr::read_tsv(independent_primary_file)

# select only independent primary bs_ids to get overall survival
metadata <- metadata %>%
  dplyr::filter(Kids_First_Biospecimen_ID %in%
                  independent_primary$Kids_First_Biospecimen_ID)

# find molecular subtypes which have more than n_participants (3) participants
filter_molecular_subtype <- metadata %>% group_by(molecular_subtype) %>% dplyr::tally() %>% dplyr::filter(n>n_participants)

# select only  that have more than n_participants (3) participants
metadata <- metadata %>%
  dplyr::filter(molecular_subtype %in%
                  filter_molecular_subtype$molecular_subtype)

```
## Kaplan-Meier for all HGG/DMG subtypes

The [Kaplan-Meier](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3059453/) 
survival curve shows the probability of survival over time.

```{r Kaplan Meier}
kap_fit <- survival_analysis(metadata,
  ind_var = "molecular_subtype",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID"
)
# median and 0.95CI survival days
kap_fit$model
```
Note that upper 95% bounds (0.95UCL) which are denoted "NA" are the consequence of underlying skew in the data. The upper confidence limit is therefore not defined (or may be interpreted as infinity).

`survminer` package comes with a convenient plotting function which uses `ggplot2`
arguments. 
It requires the original Kaplain-Meier `survfit` object, so we will extract that with `$model`.

```{r Make survival plot}
surv_plot <- survminer::ggsurvplot(kap_fit$model,
  pval = TRUE,
  data = kap_fit$original_data,
  risk.table = TRUE,
  xlim = c(0, 3000),
  break.time.by = 500,
  ggtheme = theme_minimal(),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE
)

surv_plot$plot <- surv_plot$plot +
  ggtitle(paste0("Kaplan Meier HGG tumors by molecular subtype")) +
  theme(legend.position = "right")

# Make this plot a combined plot
surv_plot_all_subtype <-
  cowplot::plot_grid(surv_plot[[1]], surv_plot[[2]], nrow = 2,
                     rel_heights = c(3, 2))

# Print it out here
surv_plot_all_subtype
```

Save the plot to a file. 

```{r Save survival plot}
# We can save the plot like a normal ggplot
cowplot::save_plot(filename = kap_meier_plot_file, plot = surv_plot_all_subtype, base_height = 4,  base_width = 8)
# Let's save the model for CI and median survial as well
readr::write_tsv(surv_median(kap_fit$model), kap_meier_model_file)
```

## Analysis: cox regression for HGG molecular subtypes
We observe a significant poor prognostic effect of TP53 loss in tumors with H3 K28 mutations.

```{r cox hgg subtype}
# Check Ns
as.data.frame(table(metadata$OS_status, metadata$molecular_subtype))

# Run model
fit_save_model(metadata, 
               "molecular_subtype",
               file.path(results_dir, "cox_hgg_subtype.RDS"))
```

Why are we not seeing H3 K28 subtypes?
```{r cox manual check}
# Double check the function isn't missing them - get the same result below
metadata_recode <- metadata %>%
      mutate(OS_status = ifelse(OS_status == "LIVING", 0, 1))

coxph(Surv(OS_years, OS_status) ~ molecular_subtype, metadata_recode) 
```






