---
title: "Survival Analysis"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: C. Savonen for ALSF CCDL
date: 2019
---

**Purpose:** 

### Summary of Findings:

#### Usage

To run this from the command line, use:
```
Rscript -e "rmarkdown::render('analyses/survival-analysis/survival-analysis.Rmd', 
                              clean = TRUE)"
```

_This assumes you are in the top directory of the repository._

## Setup

#### Packages and functions

Read in set up script.

```{r}
if (!("survminer" %in% installed.packages())) {
  install.packages("survminer")
}
library(survminer)

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

#### Set up files and directories

Set up output directories. 

```{r}
data_dir <- file.path("..", "..", "data")
results_dir <- "results"
plots_dir <- "plots"
```

Make output directories.

```{r}
if(!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if(!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

Declare input file paths. 

```{r}
metadata_file <- file.path(data_dir, "pbta-histologies.tsv")
polya <- file.path(data_dir, "pbta-gene-expression-rsem-fpkm-collapsed_table.polya.rds")
```

## Import the data

```{r}
metadata <- readr::read_tsv(metadata_file)
```

Reformat the `OS_days` variable. 
The `survival` R packages want overall survival status to be in the form of a 
numeric variable where `0  = Living` and `1 = Deceased`. 

```{r}
metadata$OS_status <- factor(metadata$OS_status, levels = c("LIVING", "DECEASED"))
metadata$OS_status_num <- as.numeric(metadata$OS_status)
```

This is the template model that you can follow for all survival analyses here. 
What will change is the independent variables `<INDEPENDENT_VARIABLES>` you use 
to try to predict survival. 
Depending on what type of variable you are using for your independent variables, 
this will influence what statistical model is appropriate `<STAT_MODEL_FUNCTION>`.

```
<STAT_MODEL_FUNCTION>(survival::Surv(OS_days, OS_status_num) ~ <INDEPENDENT_VARIABLES>,
                 data = metadata)
```

## Kaplan-Meier

[Kaplan-Meier](https://en.wikipedia.org/wiki/Kaplan%E2%80%93Meier_estimator)


```{r}
kap_fit <- survival::survfit(
  survival::Surv(metadata$OS_days, metadata$OS_status_num) ~ reported_gender,
  data = metadata)

kap_fit
```
`survminer` package comes with a convenient plotting function which uses `ggplot2`
arguments. 

```{r}
survminer::ggsurvplot(kap_fit, data = metadata, risk.table = TRUE, 
                      pval = TRUE, conf.int = TRUE,
                      xlim = c(0,2000),
                      break.time.by = 500,
                      ggtheme = theme_minimal(),
                      risk.table.y.text.col = T, 
                      risk.table.y.text = FALSE 
                      )

# We can save the plot like a normal ggplot
ggplot2::ggsave(file.path(plots_dir, "survival_curve_gender.pdf"))
```

## Log-Rank 

For testing whether two survival curves are different, we can use Log-Rank which
can be implemented using `survival::survdiff` function. 
In this example, we will test for survival curve differences between for 
`reported_gender` categories. 

```{r}
# First let's filter out data that does not have reported_gender available
metadata_filt <- metadata %>% 
  dplyr::filter(reported_gender != 'Not Available')
```

Use the `survdiff` function to test the differences between the `Male` and `Female` 
curves.
```{r}
logrank_fit <- survival::survdiff(
  survival::Surv(metadata_filt$OS_days, metadata_filt $OS_status_num) ~ reported_gender,
                                  data = metadata_filt)
```

Make a pretty table of the results.

```{r}
data.frame(group = gsub("reported_gender=", "", names(logrank_fit$n)), 
           Freq = logrank_fit$n, 
           Observed = logrank_fit$obs, 
           Expected = logrank_fit$exp, 
           Chi.Sq = logrank_fit$chisq) %>%
  dplyr::select(-Freq.groups) %>% 
  knitr::kable()
```

## Cox Regression 

Cox Regression is a model suited for survival analysis where your independent 
variabels are continuous. 
For this example, we'll use 

```{r}
polya_fpkm <- readr::read_rds(polya)
```

```{r}
cox_fit <- survival::coxph(
  survival::Surv(metadata$OS_days, metadata$OS_status_num)~ me,
                           data = metadata)
```

```{r}
summary(cox_fit)$coefficients %>%
  tibble::rownames_to_column("group") %>% 
  knitr::kable()
```

## Session Info

```{r}
sessionInfo()
```
