---
title: "Survival analysis by TP53 and telomerase activity"
output: html_notebook
---

This notebook will do the following multivariate survival analyses:
1. Multivariate analysis using immune cell types (minus uncharacterized), PDL1 expression, and cancer group as covariates
2. Multivariate analysis for each cancer group, using immune cell types (minus uncharacterized), PDL1 expression, and molecular subtype as covariates

## Setup

#### Packages

```{r Set up library}
library(survival)
library(dplyr)
library(ggplot2)
library(ggpubr)
```

#### Establish directories
```{r}
# define directories
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
input_dir <- file.path(root_dir, "analyses", "immune-deconv")
analysis_dir <- file.path(root_dir, "analyses", "survival-analysis")

results_dir <- file.path(analysis_dir, "results")
if(!dir.exists(results_dir)){
  dir.create(results_dir)
}
```

#### Read in required files
```{r}
# Read in analysis data
quantiseq <- readRDS(file.path(input_dir, "results", "quantiseq_deconv-output.rds")) %>%
  # remove uncharacterized
  dplyr::filter(cell_type != "uncharacterized cell")
polya <- readRDS(file.path(data_dir, "pbta-gene-expression-rsem-fpkm-collapsed.polya.rds"))
stranded <- readRDS(file.path(data_dir, "pbta-gene-expression-rsem-fpkm-collapsed.stranded.rds"))

# Read in histology file
histology_df <- readr::read_tsv(file.path(data_dir, "pbta-histologies.tsv"))

```

#### filter to cancer group of interest 
```{r}
# define cancer group of interest
cancer_groups_of_interest <- c("High-grade glioma astrocytoma", 
                              "Diffuse midline glioma",
                              "Low-grade glioma astrocytoma", 
                              "Ependymoma", 
                              "Medulloblastoma")

# filter to cancer group of interest 
histologies_df_filtered <- histology_df %>%
  # RNA-Seq samples only
  dplyr::filter(experimental_strategy == "RNA-Seq",
                composition=="Solid Tissue",
                tumor_descriptor == "Initial CNS Tumor",
                cancer_group %in% cancer_groups_of_interest) %>%
  dplyr::select(Kids_First_Biospecimen_ID, Kids_First_Participant_ID, cancer_group, OS_status, OS_days, molecular_subtype) %>%
  distinct() %>%
  dplyr::arrange(Kids_First_Biospecimen_ID)

```

#### Add immune information
```{r}
# filter to samples of interest
quantiseq_subset <- quantiseq %>%
  dplyr::filter(sample %in% histologies_df_filtered$Kids_First_Biospecimen_ID) %>%
  dplyr::mutate(cell_type_modi = gsub(" ", "_", cell_type),
                cell_type_modi = gsub("\\+", "", cell_type_modi),
                cell_type_modi = gsub("\\(", "", cell_type_modi),
                cell_type_modi = gsub("\\)", "", cell_type_modi),
                cell_type_modi = gsub("\\-", "_", cell_type_modi)) %>%
  dplyr::mutate() %>%
  dplyr::select(-cell_type)

# spread the results
quantiseq_subset_spread <- quantiseq_subset %>%
  tidyr::spread(cell_type_modi, score) %>%
  dplyr::rename(Kids_First_Biospecimen_ID = sample) %>%
  dplyr::left_join(histologies_df_filtered) %>%
  dplyr::select(-method)
```

#### Add RNA-Seq expression
```{r}
# get poly-A samples
polya_samples <- quantiseq_subset_spread %>%
  dplyr::filter(library == "polya") %>%
  pull(Kids_First_Biospecimen_ID)

# get stranded samples
stranded_samples <- quantiseq_subset_spread %>%
  dplyr::filter(library == "stranded") %>%
  pull(Kids_First_Biospecimen_ID)

# get CD274 expression for polya
polya_cd274 <- polya %>%
  dplyr::select(all_of(polya_samples)) %>%
  tibble::rownames_to_column("gene_symbol") %>%
  dplyr::filter(gene_symbol == "CD274") %>%
  tibble::column_to_rownames("gene_symbol") %>% 
  t() %>% as.data.frame() %>%
  tibble::rownames_to_column("Kids_First_Biospecimen_ID")

# get CD274 expression for stranded
stranded_cd274 <- stranded %>%
  dplyr::select(all_of(stranded_samples)) %>%
  tibble::rownames_to_column("gene_symbol") %>%
  dplyr::filter(gene_symbol == "CD274") %>%
  tibble::column_to_rownames("gene_symbol") %>% 
  t() %>% as.data.frame() %>%
  tibble::rownames_to_column("Kids_First_Biospecimen_ID")
  
# combine stranded and polya
combined_cd274 <- bind_rows(polya_cd274, stranded_cd274)

# add expression to quantiseq results
combined_df <- quantiseq_subset_spread %>%
  dplyr::left_join(combined_cd274)
```

#### get distinct sample
```{r}
# get distinct results 
combined_df_distinct <- combined_df %>%
  dplyr::distinct(Kids_First_Participant_ID, .keep_all=TRUE) %>%
  # recode the categories - 
  # DECEASED maps to a survival event status of 1, LIVING maps to a censored observation with value 0
  dplyr::mutate(OS_status_fixed = case_when(
    OS_status == "LIVING" ~ 0,
    OS_status == "DECEASED" ~1
 )) %>% 
  dplyr::mutate(OS_years = OS_days / 365.25) 

```

#### Run multivariate analysis for all cancer groups
```{r}
# find all immune cell types 
immune_cells <- quantiseq_subset %>% pull(cell_type_modi) %>% unique() %>%
  paste0(collapse = "+")

# define model
model <- paste0("survival::Surv(time = OS_years, event = OS_status_fixed) ~ ",
                immune_cells, 
                "+cancer_group+CD274")

fit <- survival::coxph(formula(model),
                       data = combined_df_distinct)
# generate output
table <- broom::tidy(fit) 

# Save the table data in a TSV
readr::write_tsv(table, file.path(results_dir, "cox_reg_results_5cgs.tsv"))

print(table)

# printout the plot
forest_coxph <- survminer::ggforest(fit, data = combined_df_distinct)
print(forest_coxph)
```

#### Run multivariate analysis for individual cancer group
```{r}
for(cg_each in cancer_groups_of_interest){
  # filter to the cancer group of interset 
  combined_df_distinct_each <- combined_df_distinct %>%
    dplyr::filter(cancer_group == cg_each)
  
  # define model
  model <- paste0("survival::Surv(time = OS_years, event = OS_status_fixed) ~ ",
                  immune_cells, 
                  "+molecular_subtype+CD274")
  
  fit <- survival::coxph(formula(model),
                         data = combined_df_distinct_each)
  # generate output
  table <- broom::tidy(fit) 
  
  # Save the table data in a TSV
  readr::write_tsv(table, file.path(results_dir, paste0("cox_reg_results_in_", cg_each, ".tsv")))
  
  print(table)
  
  # printout the plot
  forest_coxph <- survminer::ggforest(fit, data = combined_df_distinct_each)
  print(forest_coxph)
}

```

