---
title: "VAF Cutoffs and TMB Calculation"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: C. Savonen for ALSF CCDL
date: 2019
---

**Purpose:** How do the TMB calculations change with adjustments to the VAF cutoff
filter? More specifically, if we increase VAF filter cutoff, do the callers' TMB stats 
relate to each other more? 

### Summary of Findings:

VarDict is the least related to the other callers, but increasing the VAF filter cutoff does somewhat
recover how its TMB stats relate to the other callers' TMB stats. 
Regardless, VarDict's data should still be dropped as it is too aberrant from the other callers as 
well as overly sensitive. 

#### Usage

To run this from the command line, use:
```
Rscript -e "rmarkdown::render('analyses/survival-analysis/survival-analysis.Rmd', 
                              clean = TRUE)"
```

_This assumes you are in the top directory of the repository._

## Setup

#### Packages and functions

Read in set up script.

```{r}
if (!("survminer" %in% installed.packages())) {
  install.packages("survminer")
}

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

#### Set up files and directories

Set up output directories. 

```{r}
data_dir <- file.path("..", "..", "data")
results_dir <- "results"
plots_dir <- "plots"
```

Make output directories.

```{r}
if(!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if(!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

Declare input file paths. 

```{r}
metadata_file <- file.path(data_dir, "pbta-histologies.tsv")
```

## Import the data

```{r}
metadata <- readr::read_tsv(metadata_file)
```

```{r}
fit <- survminer::survfit(survminer::Surv(OS_days, OS_status) ~ short_histology,
               data = metadata)
```

```{r}
# Visualize with survminer
ggsurvplot(fit, data = metadata, risk.table = TRUE)
```

## Session Info

```{r}
sessionInfo()
```
