---
title: "Survival Analysis for TP53 and telomerase phenotypes"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Jo Lynne Rokita, Run Jin for D3B
date: 2021
params:
  plot_ci: TRUE
---
  
  **Purpose:** 
  
  Runs survival analysis for TP53 and telomerase phenotypes. This module was most recently run on release-v21-20210820/pbta-histologies.tsv 

## Usage 

Uses a wrapper function (`survival_analysis`) from utils folder. 

## Setup

#### Packages and functions

Read in set up script.

```{r Set up library}
# This script contains a wrapper function that can conduct the survival analyses
library(tidyverse)
library(survminer)
source(file.path("util", "survival_models.R"))
source(file.path("util", "quantile_calc.R"))

```

#### Set up files and directories

Set up output directories. 

```{r Set up directories}
data_dir <- file.path("..", "..", "data")
analysis_dir <- file.path("..", "..", "analyses")
figure_dir <- file.path("..", "..", "figures")
results_dir <- "results/model"
stats_dir <- "results/stats"
plots_dir <- "plots"
```

Make output directories.

```{r Make output directories}
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
if (!dir.exists(stats_dir)) {
  dir.create(stats_dir)
}
```

Declare input file paths and useful variables

```{r Set input file paths}
hist_file <- file.path(data_dir, "pbta-histologies.tsv")
indep_file <- file.path(data_dir,
                        "independent-specimens.rnaseq.primary-plus-stranded.tsv")
tp53_file <- file.path(analysis_dir, 
                       "tp53_nf1_score/results/tp53_altered_status.tsv")

# only use stranded since tp53 scores for polyA did not have good accuracy
tel_stranded_file <- file.path(analysis_dir,
                        "telomerase-activity-prediction/results/TelomeraseScores_PTBAStranded_FPKM.txt")

# color palette
color_file <- file.path(figure_dir, "palettes/broad_histology_cancer_group_palette.tsv")

```

## Import the metadata

```{r Read in metadata}
# read in histologies file
indep_rna <- readr::read_tsv(indep_file)

# remove cell lines by selecting only solid tissue
histologies <- readr::read_tsv(hist_file, guess_max = 10000) %>%
  dplyr::filter(composition=="Solid Tissue" & tumor_descriptor == "Initial CNS Tumor") %>%
  select(sample_id, broad_histology, short_histology, cancer_group, molecular_subtype, OS_status, OS_days, PFS_days) %>%
  distinct()

# read in tp53 scores
tp53_scores <- readr::read_tsv(tp53_file) %>%
  select(Kids_First_Biospecimen_ID_DNA, Kids_First_Biospecimen_ID_RNA, sample_id, tp53_score)

# read in telomerase scores
tel_scores <- readr::read_tsv(tel_stranded_file) %>%
  mutate(Kids_First_Biospecimen_ID_RNA = SampleID,
         telomerase_score = NormEXTENDScores) %>%
  select(Kids_First_Biospecimen_ID_RNA, telomerase_score)

# remove samples without RNA-Seq since both classifiers' results use RNA-Seq
# independent use primary plus rna-seq samples
meta_indep <- tp53_scores %>%
  left_join(tel_scores) %>%
  left_join(histologies) %>%
  dplyr::filter(Kids_First_Biospecimen_ID_RNA %in% indep_rna$Kids_First_Biospecimen_ID)

```

Caclulte the quantiles for both tp53 scores and telomarase scores. 
These scores are used later on for classifying samples into different categories for survival analysis.
```{r}
meta_indep_no_hgat <- meta_indep %>%
  filter(short_histology != "HGAT")

tp53_quantiles <- quantile(meta_indep$tp53_score, probs = seq(0, 1, 0.25))
telomerase_quantiles<- quantile(meta_indep$telomerase_score, probs = seq(0, 1, 0.25))

tp53_quantiles_no_hgat <- quantile(meta_indep_no_hgat$tp53_score, probs = seq(0, 1, 0.25))
telomerase_quantiles_no_hgat <- quantile(meta_indep_no_hgat$telomerase_score, probs = seq(0, 1, 0.25))
```

Create groupings for scores - basically, based on the TP53 and telomarase activity scores, additional columns are added indicating which group a particular sample fits in:
- `tp53_0.5`: this columns indicates whether a sample is `tp53_high` or `tp53_low` based on whehter its TP53 score is over or below 0.5
- `tel_0.5`: this columns indicates whether a sample is `telomerase_high` or `telomerase_low` based on whehter its telomerase score is over or below 0.5
- `tp53_strict`: this columns indicates whether a sample is `tp53_high`, `tp53_mid`, or `tp53_low` based on whether they are above 75 quantile (`tp53_high`), below 25 quantile (`tp53_low`), or in between (`tp53_mid`)
- `tel_strict`: this columns indicates whether a sample is `telomerase_high`, `telomerase_mid`, or `telomerase_low` based on whether they are above 75 quantile (`telomerase_high`), below 25 quantile (`telomerase_low`), or in between (`telomerase_mid`)
- `pheno_0.5`: this column combines info from `tp53_0.5` and `tel_0.5`
- `pheno_strict`: this column combines info from `tp53_strict` and `tel_strict`
- `pheno_extremes`: this column only retains categorizations from `pheno_strict` column when the sample does not fit into `mid` group (between 25 and 75 percentile) for neither TP53 nor telomase scores

```{r}
meta_indep <- calc_groups(meta_indep, tp53_quantiles, telomerase_quantiles)
meta_indep_no_hgat <- calc_groups(meta_indep_no_hgat, tp53_quantiles_no_hgat, telomerase_quantiles_no_hgat)
```

Set phenotypes to loop through. These are columns created by the `calc_groups` function. These columns, as described above, contain information about whether the sample has `high`, `mid` or `low` TP53/telomerase scores.
```{r}
# phenotypes <- c("tp53_0.5", "tel_0.5", "tp53_strict", "tel_strict", "pheno_0.5", "pheno_strict", "pheno_extremes")
phenotypes <- c("tp53_0.5", "tel_0.5", "tp53_strict", "tel_strict", "pheno_0.5", "pheno_strict")
```

## Kaplan-Meier
Run KM survival analysis on all PBTA, HGAT only, and PBTA samples that are not HGAT.
The [Kaplan-Meier](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3059453/) survival curve shows the probability of survival over time.

```{r Kaplan Meier}

# let's look at HGAT
meta_all <- meta_indep
meta_hgat <- meta_indep %>%
  filter(short_histology == "HGAT")
meta_no_hgat <- meta_indep_no_hgat

as.data.frame(table(meta_indep$short_histology))

prefix_list <- c("all-pbta","hgat","no-hgat")
# define data frame list in the same order as prefix list 
df_list <-list(meta_all, meta_hgat, meta_no_hgat)
  
for (pheno in phenotypes) {
  for(i in 1:length(prefix_list)){
    
    prefix <- prefix_list[i]
    df_of_interest <- df_list[[i]]
    
    kap_fit <- survival_analysis(df_of_interest,
                                 ind_var = pheno,
                                 test = "kap.meier",
                                 metadata_sample_col = "Kids_First_Biospecimen_ID_RNA"
    )
    
    # median and 0.95CI survival days
    kap_fit$model
    
    #`survminer` package comes with a convenient plotting function which uses `ggplot2` arguments. 
    #It requires the original Kaplan-Meier `survfit` object, so we will extract that with `$model`.
    surv_plot <- survminer::ggsurvplot(kap_fit$model,
                                       surv.plot.height = 0.7,
                                       tables.height = 0.3,
                                       pval = TRUE,
                                       data = kap_fit$original_data,
                                       risk.table = TRUE,
                                       xlim = c(0, 8000),
                                       break.time.by = 1000,
                                       ggtheme = theme_pubr(),
                                       risk.table.y.text.col = TRUE,
                                       risk.table.y.text = FALSE,
                                       palette = "Paired",
                                       color = pheno)
    
    #Save the plot to a file. 
    pdf(file.path(plots_dir, paste(prefix, pheno, "KM.pdf", sep = "-")), height = 7, width = 10)
    print(surv_plot, newpage = FALSE)
    dev.off()
    # Let's save the model for CI and median survival as well. Note: per this notebook, https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/32a81d2b5b307c5473c5dbb4408c9a64de40e7d5/analyses/survival-analysis/survival-analysis_HGG_DMG.Rmd#L129, NAs in the output are derived from skew in the underlying data.
    readr::write_tsv(surv_median(kap_fit$model), file.path(results_dir, paste(prefix, pheno, "model.tsv", sep = "-")))
    }
  }

```

#### output statistics for paper writing 

```{r}
# first format the metadata to have appropriate OS_status column
df_list <- lapply(df_list, format_meta)

# iterate through all data frames with all the models
for(i in 1:length(df_list)){
  df_of_interest <- df_list[[i]]
  prefix <- prefix_list[i]
  
  for(j in 1:length(phenotypes)){
    
    pheno_of_interest <- phenotypes[j]
    # define the formula
    model <- as.formula(paste0("survival::Surv(OS_days, OS_status) ~ ", pheno_of_interest))
    
    # use the formula to output the stat files
    output_stats(metadata=df_of_interest, 
                 model=model,
                 prefix=prefix,
                 pheno=pheno_of_interest)
  }
}

```


Print out the session info
```{r}
sessionInfo()
```
