---
title: "Survival Analysis for subtypes of HGG/DMG"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: C. Savonen for ALSF CCDL, Krutika Gaonkar for D3B
date: 2019
params:
  plot_ci: TRUE
---

**Purpose:** 

Runs survival analysis models for subtypes of Diffuse astrocytic and oligodendroglial tumors

## Usage 

Uses a wrapper function (`survival_analysis`) from utils folder. 

## Setup

#### Packages and functions

Read in set up script.

```{r Set up library}
# This script contains a wrapper function that can conduct the survival analyses
source(file.path("util", "survival_models.R"))

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

#### Set up files and directories

Set up output directories. 

```{r Set up directories}
data_dir <- file.path("..", "..", "data")
results_dir <- "results"
plots_dir <- "plots"
```

Make output directories.

```{r Make output directories}
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

Declare input file paths. 

```{r Set input file paths}
metadata_file <- file.path(data_dir, "pbta-histologies.tsv")
independent_primary_file <- file.path(
  data_dir,
  "independent-specimens.wgswxs.primary.tsv")
```

Declare output file paths for HGG/DMG all subtypes. 

```{r Set output file paths}
kap_meier_plot_file <- file.path(plots_dir, "survival_curve_HGG_DMG.pdf")
kap_meier_model_file <- file.path(results_dir, "survival_model_HGG_DMG.tsv")
```

## Import the metadata

```{r Read in metadata}
# choose Diffuse astrocytic and oligodendroglial tumor samples
# remove RNA-Seq because subtyping in HGG is from DNA abberations and samples where only RNA-Seq is available cannot be subtyped
# remove cell-lines from this analysis as well
metadata <- readr::read_tsv(
  metadata_file,
  col_types = readr::cols(
    molecular_subtype = readr::col_character())) %>%
  dplyr::filter(
    experimental_strategy != "RNA-Seq" &
      broad_histology == "Diffuse astrocytic and oligodendroglial tumor" &
      broad_composition == "tumor") %>%
  as.data.frame()

```

#### Using independent_primary if multiple subtypes per patient are found 
This way we subset the metadata to 1 single primary sample per patient ID

```{r Read in independent}

# Patients with multiple different molecular subtype might affect downstream analysis for survival so using independent_primary sample set which has 1-1 sample-patient matches
independent_primary <- readr::read_tsv(independent_primary_file)
# select only independent primary bs_ids to get overall survival
metadata <- metadata %>%
  dplyr::filter(Kids_First_Biospecimen_ID %in%
                  independent_primary$Kids_First_Biospecimen_ID)



```
## Kaplan-Meier for all HGG/DMG subtypes

The [Kaplan-Meier](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3059453/) 
survival curve shows the probability of survival over time.

```{r Kaplan Meier}
kap_fit <- survival_analysis(metadata,
  ind_var = "molecular_subtype",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID"
)
# median and 0.95CI survival days
kap_fit$model
```

`survminer` package comes with a convenient plotting function which uses `ggplot2`
arguments. 
It requires the original Kaplain-Meier `survfit` object, so we will extract that with `$model`.

```{r Make survival plot}
surv_plot <- survminer::ggsurvplot(kap_fit$model,
  pval = TRUE,
  data = kap_fit$original_data,
  risk.table = TRUE,
  xlim = c(0, 3000),
  break.time.by = 500,
  ggtheme = theme_minimal(),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE
)

surv_plot$plot <- surv_plot$plot +
  ggtitle(paste0("Kap_meier_Diffuse_astrocytic_and_oligodendroglial_tumor")) +
  theme(legend.position = "right")

# Make this plot a combined plot
surv_plot_all_subtype <-
  cowplot::plot_grid(surv_plot[[1]], surv_plot[[2]], nrow = 2,
                     rel_heights = c(3, 2))

# Print it out here
surv_plot_all_subtype
```

Save the plot to a file. 

```{r Save survival plot}
# We can save the plot like a normal ggplot
cowplot::save_plot(filename = kap_meier_plot_file, plot = surv_plot_all_subtype)
# Let's save the model for CI and median survial as well
readr::write_tsv(surv_median(kap_fit$model), kap_meier_model_file)
```

### Subtypes comapred to wildtype
Lets look at Kap Meier plots for each mutant subtype to H3 wildtype

```{r Kaplan Meier per subtype compared to wildtype}

list_subtypes <- c("DMG, H3 K28",
                   "BRAF V600E",
                   "HGG, H3 G35",
                   "HGG, IDH",
                   "DMG, H3 K28,BRAF V600E")
for (i in 1:length(list_subtypes)){
  kap_meier_plot_file <-
    file.path(plots_dir,
              paste0("Kap_meier_Diffuse_astrocytic_and_oligodendroglial_tumor_",
                     gsub(" ", "_", list_subtypes[i]), ".pdf"))
  kap_meier_model_file <-
    file.path(results_dir,
              paste0("Kap_meier_Diffuse_astrocytic_and_oligodendroglial_tumor_",
                     gsub(" ", "_", list_subtypes[i]), ".tsv"))
  metadata_test <- metadata %>%
    dplyr::filter(
      molecular_subtype %in% c(list_subtypes[i], "HGG, H3 wildtype") )
  kap_fit <- survival_analysis(metadata_test,
                               ind_var = "molecular_subtype",
                               test = "kap.meier",
                               metadata_sample_col = "Kids_First_Biospecimen_ID"
  )

  #plot curve
  surv_plot <- survminer::ggsurvplot(kap_fit$model,
                                     pval = TRUE,
                                     data = kap_fit$original_data,
                                     risk.table = TRUE,
                                     xlim = c(0, 3000),
                                     break.time.by = 500,
                                     ggtheme = theme_minimal(),
                                     risk.table.y.text.col = TRUE,
                                     risk.table.y.text = FALSE
  )

  surv_plot$plot <- surv_plot$plot + ggtitle(
    paste0("Kap_meier_Diffuse_astrocytic_and_oligodendroglial_tumor")) +
    theme(legend.position = "right")

  # Make this plot a combined plot
  surv_plot <- cowplot::plot_grid(
    surv_plot[[1]],
    surv_plot[[2]],
    nrow = 2,
    rel_heights = c(2.5, 1))

  # Print to file
  cowplot::save_plot(
    filename = kap_meier_plot_file,
    plot = surv_plot,
    base_height = 10)
  # Save to file
  readr::write_tsv(surv_median(kap_fit$model), kap_meier_model_file)
}
```




