---
title: "Survival analysis by TP53 and telomerase activity"
output: html_notebook
---

This notebook will do the following survival analysis

1. Univariate analysis 
- TP53 classifier score (as a continuous variable)
- EXTEND score (as a continuous variable)
- Broad histology specified in the color palette file
- HGG vs. non-HGG

2. Multivariate analysis
- TP53 classifier score, EXTEND score and HGG group
- TP53 classifier score, EXTEND score and broad histology specified in the color palette file

3. Plots
- Bivariate distribution of TP53 classifier score and EXTEND score in density plot stratified by HGAT status

## Setup

#### Packages and functions

Read in set up script.

```{r Set up library}
library(survival)
library(dplyr)
library(ggplot2)
library(ggpubr)
```

#### Set up files and directories

Set up output directories. 

```{r Set up directories}
data_dir <- file.path("..", "..", "data")
analysis_dir <- file.path("..", "..", "analyses")
figure_dir <- file.path("..", "..", "figures")

results_dir <- "results"
plots_dir <- "plots"
```

Make output directories.

```{r Make output directories}
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

## Declare the direcotry for all the data to be used 

```{r define data path}
metadata_file <- file.path(data_dir, "pbta-histologies.tsv")
tp53_file <- file.path(analysis_dir, 
                        "tp53_nf1_score/results/tp53_altered_status.tsv")
# only use stranded since tp53 scores for polyA did not have good accuracy
tel_stranded_file <- file.path(analysis_dir,
                         "telomerase-activity-prediction/results/TelomeraseScores_PTBAStranded_FPKM.txt")
```

## Import the metadata

```{r Read in metadata}
# get primary tumor and remove cell lines
histologies_rna <- readr::read_tsv(metadata_file, guess_max = 10000) %>%
  dplyr::filter(composition=="Solid Tissue" & 
                  tumor_descriptor == "Initial CNS Tumor" &
                  experimental_strategy == "RNA-Seq" & 
                  RNA_library == "stranded") %>%
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = Kids_First_Biospecimen_ID) %>% 
  dplyr::select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID_RNA, cancer_group, OS_status, OS_days, PFS_days) %>%
  distinct() %>%
  dplyr::arrange(Kids_First_Biospecimen_ID_RNA)

# read in tp53 scores
tp53_scores <- readr::read_tsv(tp53_file) %>%
  select(Kids_First_Biospecimen_ID_DNA, Kids_First_Biospecimen_ID_RNA, sample_id, tp53_score) %>%
  dplyr::arrange(Kids_First_Biospecimen_ID_RNA)

# read in telomerase scores
tel_scores <- readr::read_tsv(tel_stranded_file) %>%
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = SampleID,
                telomerase_score = NormEXTENDScores) %>%
  dplyr::arrange(Kids_First_Biospecimen_ID_RNA)

# combine tp53 scores and telomerase score to the dataframe
meta_indep <- histologies_rna %>%
  # add tp53 and telomerase scores by matching Kids_First_Biospecimen_ID_RNA
  left_join(tp53_scores, by = "Kids_First_Biospecimen_ID_RNA") %>%
  left_join(tel_scores, by = "Kids_First_Biospecimen_ID_RNA") %>% 
  dplyr::arrange(Kids_First_Biospecimen_ID_RNA) %>%
  dplyr::distinct(Kids_First_Participant_ID, .keep_all=TRUE) %>%
  # recode the categories - 
  # DECEASED maps to a survival event status of 1, LIVING maps to a censored observation with value 0
  dplyr::mutate(OS_status_fixed = case_when(
    OS_status == "LIVING" ~ 0,
    OS_status == "DECEASED" ~1
 )) %>% 
  dplyr::mutate(OS_years = OS_days / 365.25)

nrow(meta_indep)
```
 
### Categorical variable cancer group and HGG or not 
```{r}
# separate into HGG or non-HGG
meta_indep <- meta_indep %>% 
  dplyr::mutate(hgg_group = case_when(
    cancer_group %in% c("High-grade glioma astrocytoma", "Diffuse midline glioma", "Diffuse intrinsic pontine glioma") ~ "HGAT",
    !cancer_group %in% c("High-grade glioma astrocytoma", "Diffuse midline glioma", "Diffuse intrinsic pontine glioma") ~ "non-HGAT"
  ))

# make non-HGAT the reference level
meta_indep$hgg_group <- factor(meta_indep$hgg_group, levels = c("non-HGAT", "HGAT"))

```

## Cox Regression 
### TP53 scores & Telomerase score - results
```{r Run Cox regression model for tp53 and telomerase for all covariates}
# run the model with different covariates
for(ind_var in c("tp53_score", "telomerase_score", "hgg_group")){
  model <- paste0("survival::Surv(time = OS_years, event = OS_status_fixed) ~ ", ind_var)
  
  fit <- survival::coxph(
        formula(model),
        data = meta_indep
      )
  # generate output
  table <- broom::tidy(fit) 
  
  # Save the table data in a TSV
  readr::write_tsv(table, file.path(results_dir, paste0("cox_reg_results_per_", ind_var, ".tsv")))
  
  print(table)
  
  # output the plot
  pdf(file.path(plots_dir, paste0("coxph_survival_per_", ind_var, ".pdf")))
  plot_coxph <- survminer::ggsurvplot(survfit(fit),
                                      data=meta_indep,
                                      xlim = c(0, 14),
                                      break.time.by = 1,
                                      pval = TRUE, 
                                      conf.int = TRUE,
                                      risk.table = TRUE, # Add risk table
                                      linetype = "strata", # Change line type by groups
                                      surv.median.line = "hv", # Specify median survival
                                      ggtheme = theme_bw())
  print(ggpar(plot_coxph[[1]], font.legend=6, legend="right"))
  print(ggpar(plot_coxph[[2]], font.legend=6))

  dev.off()

  print(ggpar(plot_coxph[[1]], font.legend=6, legend="right"))
  print(ggpar(plot_coxph[[2]], font.legend=6))
}

```

## Log Rank analysis 
### Generate output for categorical files - only hgg group is used  
```{r}
# define the independent variable that we are using for analyzing survival
ind_var <- "hgg_group"

# define model
model <- paste0("survival::Surv(time = OS_years, event = OS_status_fixed) ~ ", ind_var)

# run survival analysis
fit <- survival::survdiff(formula(model),
                          data = meta_indep)
# Obtain p value for Chi-Squared stat
fit$p.value <- pchisq(fit$chisq, df = length(fit$n) - 1, lower = FALSE)

print(fit)
# save the output
saveRDS(fit, file.path(results_dir, paste0("log_rank_survival_per_", ind_var, ".RDS")))

# generate plots fit
fit_plot <- survfit(formula(model), data = meta_indep)

# output the plot
pdf(file.path(plots_dir, paste0("logrank_survival_per_", ind_var, ".pdf")))
plot_logrank <- survminer::ggsurvplot(fit_plot,
                                      data=meta_indep,
                                      xlim = c(0, 14),
                                      break.time.by = 1,
                                      pval = TRUE, 
                                      conf.int = TRUE,
                                      risk.table = TRUE, # Add risk table
                                      linetype = "strata", # Change line type by groups
                                      surv.median.line = "hv", # Specify median survival
                                      ggtheme = theme_bw())
print(ggpar(plot_logrank[[1]], font.legend=6, legend="right"))
print(ggpar(plot_logrank[[2]], font.legend=6))

dev.off()

print(ggpar(plot_logrank[[1]], font.legend=6, legend="right"))
print(ggpar(plot_logrank[[2]], font.legend=6))


```

## Multivariate analysis 
### Comparison made is `tp53_score + telomerase_score + hgg_group` 

```{r}
# define multi-variates that we are using for analyzing survival
ind_var <- "tp53_score+telomerase_score+hgg_group"

# define model
model <- paste0("survival::Surv(time = OS_years, event = OS_status_fixed) ~ ", ind_var)

fit <- survival::coxph(
      formula(model),
      data = meta_indep
    )
# generate output
table <- broom::tidy(fit) 

# Save the table data in a TSV
readr::write_tsv(table, file.path(results_dir, paste0("cox_reg_results_per_", ind_var, ".tsv")))

print(table)

# printout the plot
forest_coxph <- survminer::ggforest(fit, data = meta_indep)
print(forest_coxph)


```

### Further we want to separate to HGAT and non-HGAT and see how tp53 and telomerase affect survival in each group
```{r}
# get two separate groups
meta_indep_hgg <- meta_indep %>% dplyr::filter(hgg_group == "HGAT")
meta_indep_non_hgg <- meta_indep %>% dplyr::filter(hgg_group == "non-HGAT")

# define a list to iterate through
meta_list <- list(meta_indep_hgg, meta_indep_non_hgg)
names(meta_list) <- c("hgg", "non_hgg")

# define multi-variates that we are using for analyzing survival
ind_var <- "tp53_score+telomerase_score"
# define model
model <- paste0("survival::Surv(time = OS_years, event = OS_status_fixed) ~ ", ind_var)

for(i in 1:length(meta_list)){
  # find the meta file
  meta_df <- meta_list[[i]]
  
  fit <- survival::coxph(
      formula(model),
      data = meta_df
    )
  # generate output
  table <- broom::tidy(fit) 
  
  #names of subgroup
  subgroup <- names(meta_list[i])
  
  # Save the table data in a TSV
  readr::write_tsv(table, file.path(results_dir, paste0("cox_reg_results_per_", ind_var, "_in_", subgroup, ".tsv")))
  
  print(table)
  
  # printout the plot
  forest_coxph <- survminer::ggforest(fit, data = meta_df)
  print(forest_coxph)
}


```




## Bivariate density plot of tp53 score and telomerase activity score stratified by HGG group
```{r}

tp53_vs_telomerase <- ggplot(meta_indep, aes(x = tp53_score, y = telomerase_score))
tp53_vs_telomerase + 
  geom_density_2d(aes(color = hgg_group)) +
  labs(subtitle = "Density stratified by HGG group") +
  xlim(-0.1,1.2) +
  theme_bw()

```


## Session Info

```{r Print session info}
sessionInfo()
```