---
title: "Survival Analysis Example"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: C. Savonen for ALSF CCDL
date: 2019
params:
  plot_ci: TRUE
---

**Purpose:** 

An example for running basic survival analysis models which can be applied to 
various other data. 

## Usage 

This notebook is a template model that you can follow for all survival analyses. 
This includes using a wrapper function (`survival_analysis`) that can test given 
independent variables in the survival model and return back a table. 

**Example use**

```
survival_analysis(metadata, 
                  ind_var = <INDEPENDENT_VARIABLE(S)>, 
                  test = <TEST_ABBR>, 
                  ind_data = <DATA_FRAME>, 
                  metadata_sample_col = "Kids_First_Biospecimen_ID",
                  ind_data_sample_col = <COLUMN THAT MATCHES METADATA's Kids_First_Biospecimen_ID>)
```

#### Options explanation:

- `metadata` : A data.frame that contains columns `OS_status` and `OS_days` to use 
             for the survival model. This also assumes `LIVING` and `DECEASED`
             are the two statuses. This will be converted to a numeric variable 
             for use with `survival` R functions. Samples with `NA`s are dropped.
- `ind_var`: A character string noting the name of the independent variable to 
             test as a predictor for survival. 
- `test`  : A character string noting which test to use. Options: `"kap.meier"`, 
             `"log.rank"`, or `"cox.reg"` (See below).
- `ind_data`: If the data for the independent variable needed for the test is not
              in `metadata`, add it here. This assumes it will be a data.frame with 
              either a `Tumor_Sample_Barcode`, `Kids_First_Biospecimen_ID`, or 
              `sample` column which to `dplyr::inner_join` with `metadata` by. 
- `ind_data_sample_col` : A character string that states the name of the column in 
              `ind_data` that contains the sample IDs that match to the metadata. 
- `metadata_sample_col` : A character string that states the name of the column in 
              `metadata` that contains the sample IDs that match to the ind_data. 
- `os_days_col` : A character string that states the name of the column in 
                `metadata` that contains the overall survival days information.
                Default is `OS_days`.
- `os_status_col` : A character string that states the name of the column in 
                `metadata` that contains the overall survival status information
                 The data in this column will be converted to numeric  where 
                 `LIVING` = 1 and `DECEASED` = 2. (This is what `survminer` functions
                 look for.) Default is `OS_status`.
  
#### `test` options:  
- ["kap.meier"](#kaplan-meier-example) : For basic survival analysis.  
- ["log.rank"](#log-rank-example) : For use with categorical variables.  
- ["cox.reg"](#cox-regression-example) : For use with continuous variables.  
  
  
#### Output: 

A **list** with:   
  1) The original `$model` fit object.  
  2) The tidied results `$table`.  
  3) The `$original_data` that was used in the model.  
  

## Setup

#### Packages and functions

Read in set up script.

```{r Set up library}
# This script contains a wrapper function that can conduct the survival analyses
source(file.path("util", "survival_models.R"))

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

#### Set up files and directories

Set up output directories. 

```{r Set up directories}
data_dir <- file.path("..", "..", "data")
results_dir <- "results"
plots_dir <- "plots"
```

Make output directories.

```{r Make output directories}
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

Declare input file paths. 

```{r Set input file paths}
metadata_file <- file.path(data_dir, "pbta-histologies.tsv")
tmb_file <- file.path(data_dir, "pbta-snv-consensus-mutation-tmb-coding.tsv")
```

Declare output file paths. 

```{r Set output file paths}
kap_meier_plot_file <- file.path(plots_dir, "survival_curve_gender.pdf")
logrank_table_file <- file.path(results_dir, "logrank_gender.tsv")
cox_table_file <- file.path(results_dir, "cox_regression_tmb.tsv")
```

## Import the metadata

```{r Read in metadata}
metadata <- readr::read_tsv(metadata_file)
```

## Kaplan-Meier Example

The [Kaplan-Meier](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3059453/) 
survival curve shows the probability of survival over time.

```{r Kaplan Meier}
kap_fit <- survival_analysis(metadata,
  ind_var = "germline_sex_estimate",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID",
  ind_data_sample_col = "sample"
)
```

`survminer` package comes with a convenient plotting function which uses `ggplot2`
arguments. 
It requires the original Kaplain-Meier `survfit` object, so we will extract that with `$model`.

```{r Make survival plot}
surv_plot <- survminer::ggsurvplot(kap_fit$model,
  pval = TRUE,
  data = kap_fit$original_data,
  risk.table = TRUE,
  xlim = c(0, 2000),
  break.time.by = 500,
  ggtheme = theme_minimal(),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE
)

# Make this plot a combined plot
surv_plot <- cowplot::plot_grid(surv_plot[[1]], surv_plot[[2]], nrow = 2, 
                                rel_heights = c(2.5, 1))

# Print it out here
surv_plot
```

Save the plot to a file. 

```{r Save survival plot}
# We can save the plot like a normal ggplot
cowplot::save_plot(filename = kap_meier_plot_file, plot = surv_plot)
```

## Log-Rank Example

For testing whether two survival curves are different, we can use Log-Rank which
can be implemented using `survival::survdiff` function. 
In this example, we will test for survival curve differences between for 
`germline_sex_estimate` categories. 
Use the `survdiff` function to test the differences between the `Male` and `Female` 
curves.

```{r Do log rank test}
logrank_fit <- survival_analysis(metadata,
  ind_var = "germline_sex_estimate",
  test = "log.rank",
  metadata_sample_col = "Kids_First_Biospecimen_ID",
  ind_data_sample_col = "sample"
)
```

Print table of the results.

```{r Make log rank table}
# Print out the table here
logrank_fit$table %>%
  dplyr::mutate(p.value = logrank_fit$model$p.value) %>%
  knitr::kable(digits = 2)
```

```{r Write log rank to table}
# Save the table data in a TSV
readr::write_tsv(logrank_fit$table, logrank_table_file)
```

## Cox Regression Example

Cox Regression is a model suited for survival analysis where your independent 
variables are continuous. 
For this example, we'll use Tumor Mutation Burden as our continuous variable. 

```{r Read in TMB data}
# Read in the data
tmb <- readr::read_tsv(tmb_file)
```

Plug this into a Cox Regression model using the function `survival::coxph`. 

```{r Run Cox regression model}
cox_fit <- survival_analysis(metadata,
  ind_var = "tmb",
  test = "cox.reg",
  ind_data = tmb,
  metadata_sample_col = "Kids_First_Biospecimen_ID",
  ind_data_sample_col = "Tumor_Sample_Barcode"
)
```

Take the output from this and make a table. 

```{r Make Cox regression table}
# Print out the table here
cox_fit$table %>%
  knitr::kable(digits = 3)
```

```{r Write Cox regression to tsv}
# Save the table data in a TSV
readr::write_tsv(cox_fit$table, cox_table_file)
```

## Session Info

```{r Print session info}
sessionInfo()
```
