---
title: "Survival analysis by immune scores and molecular subtypes"
authors: Run Jin (D3B), Stephanie Spielman (CCDL), Jo Lynne Rokita (D3b)
output: 
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---

## Setup

#### Packages

```{r Set up library}
library(survival)
library(ggpubr)
library(tidyverse)
```


#### Set up files and directories

Set up input and output directories:

```{r directories}
# define directories
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
immune_dir <- file.path(root_dir, "analyses", "immune-deconv")
analysis_dir <- file.path(root_dir, "analyses", "survival-analysis")

results_dir <- file.path(analysis_dir, "results/immune")
if(!dir.exists(results_dir)){
  dir.create(results_dir)
}

plots_dir <- file.path(analysis_dir, "plots")
if(!dir.exists(plots_dir)){
  dir.create(plots_dir)
}

util_dir <- file.path(analysis_dir, "util")
```


Source the util function file.

```{r source_functions}
# Source functions used to build and export models
source(file.path(util_dir,"survival_models.R"))
```

Read in and process relevant data files:

```{r cd274_function}
# First, we define a function for use here only to process expression files and obtain CD274 expression across libraries
extract_cd274_expression <- function(df, library_type){
  
  # Get the library samples from `quantiseq` df
  samples <- quantiseq %>%
    filter(library == library_type) %>%
    pull(Kids_First_Biospecimen_ID)
  
  # Return the expression levels for those samples from `df`
  df %>% 
    select(samples) %>%
    rownames_to_column("gene_symbol") %>%
    filter(gene_symbol == "CD274") %>%
    column_to_rownames("gene_symbol") %>% 
    t() %>% 
    as.data.frame() %>%
    rownames_to_column("Kids_First_Biospecimen_ID") %>%
    mutate(library = library_type) %>%
    as_tibble()
  
}
```


```{r read_data}
# Read in and clean up the quantiseq immune deconvoluted scores file 
quantiseq <- read_rds(
  file.path(immune_dir, "results", "quantiseq_deconv-output.rds")
  ) %>%
  # remove uncharacterized cell types
  filter(cell_type != "uncharacterized cell") %>%
  # remove unneeded `method` column
  select(-method) %>%
  # rename sample column
  rename(Kids_First_Biospecimen_ID = sample) %>%
  # clean up the cell_type column with underscores for downstream use 
  mutate(cell_type_underscores = gsub(" ", "_", cell_type),
         cell_type_underscores = gsub("\\+", "", cell_type_underscores),
         cell_type_underscores = gsub("\\(", "", cell_type_underscores),
         cell_type_underscores = gsub("\\)", "", cell_type_underscores),
         cell_type_underscores = gsub("\\-", "_", cell_type_underscores)) %>%
  # remove original column
  select(-cell_type)

# Read in histology file and filter to RNAseq samples with relevant columns
# The following patients have discrepant extent of tumor resection
# 7316-116 has Biopsy only;Gross/Near total resection
# 7316-182 has Gross/Near total resection;Partial resection
# Ticket was created [here](https://github.com/PediatricOpenTargets/ticket-tracker/issues/298) to track this database update, and I will recode below

histology_df <- read_tsv(file.path(data_dir, "pbta-histologies.tsv"), guess_max = 10000) %>%
  filter(composition=="Solid Tissue",
          tumor_descriptor == "Initial CNS Tumor",
          experimental_strategy == "RNA-Seq") %>%
  mutate(extent_of_tumor_resection = case_when(
    sample_id == "7316-116" ~ "Biopsy only",
    sample_id == "7316-182" ~ "Partial resection",
    sample_id == "7316-356" ~ "Partial resection",
    TRUE ~ as.character(extent_of_tumor_resection))) %>%
  select(Kids_First_Participant_ID, 
         Kids_First_Biospecimen_ID, 
         broad_histology,
         cancer_group,
         molecular_subtype,
         extent_of_tumor_resection,
         OS_status, 
         OS_days, 
         PFS_days) %>%
  distinct() 

# create lgg grouping from broad_histology: https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/1416
lgg_levels <- c("non-LGG", "LGG")
histology_df <- histology_df %>%
  mutate(lgg_group = ifelse(broad_histology == "Low-grade astrocytic tumor",
                             lgg_levels[2], # second is LGG 
                             lgg_levels[1]), # first is non-LGG
          lgg_group = factor(lgg_group, levels = lgg_levels)) # ensures non-LGG is the reference 

# Read in gene expression files - both poly-A and stranded samples
polya <- read_rds(file.path(data_dir, "pbta-gene-expression-rsem-fpkm-collapsed.polya.rds"))
stranded <- read_rds(file.path(data_dir, "pbta-gene-expression-rsem-fpkm-collapsed.stranded.rds"))

# Process gene expression files to obtain PDL1 (aka CD274)
cd274_expression <- bind_rows(
  extract_cd274_expression(polya, "polya"),
  extract_cd274_expression(stranded, "stranded")
)
```

Now, we can combine data for modeling:

```{r combine_data}
# merge data frames and convert days ---> years
combined_df <- histology_df %>%
  inner_join(cd274_expression) %>%
  inner_join(quantiseq) %>%
  mutate(
    OS_years = OS_days / 365.25
  ) %>%
  # get a column for each cell type
  spread(cell_type_underscores, score)
```

## Analysis: PDL-1 expression, controlling for library, across entire cohort

We use a cox regression to explore influence of PDL-1 expression on survival, while controling for library (polyA vs stranded).

```{r cox_pdl1_library}
fit_save_model(combined_df, 
               "CD274 + library",
               file.path(results_dir, "cox_terms_pdl1_library.RDS"), 
               model_type = "multivariate") %>%
  print()

```

We observe a _highly significant_ library effect, suggesting strong batch effects in the analysis. **Therefore, we will model using only stranded data for the rest of this notebook to ameliorate this influence.**

### Subset to stranded only

```{r stranded}
combined_df_stranded <- filter(combined_df, library == "stranded")

# how many samples do we now have?
nrow(combined_df_stranded)
```

## Re-analyze with stranded only: PDL-1 expression across entire cohort

There is a marginally signficant effect of PDL-1 expression (P=0.033). 
Higher expression is associated with lowered survival.


```{r cox_pdl1}
fit_save_model(combined_df_stranded, 
               "CD274",
               file.path(results_dir, "cox_terms_pdl1_stranded.RDS"))

```

## Analysis: quantiseq cell type fractions + extent of tumor resection + lgg group across entire cohort

Three effects are significant: 

+ extent of tumor resection "partial resection" is signficantly different from biopsy
+ extent of tumor resection "gross/near total resection" is signficantly different from biopsy
+ LGG and non-LGG



```{r cox_cell_type}

# Define the terms for a quantiseq model. These are used several times in this notebook
quantiseq_terms <- "B_cell + Macrophage_M1 + Macrophage_M2 + Monocyte + Myeloid_dendritic_cell + Neutrophil + NK_cell + T_cell_CD4_non_regulatory + T_cell_CD8 + T_cell_regulatory_Tregs" 

# For this model, we need to remove unknown resections from the data:
combined_df_stranded_resection <- combined_df_stranded %>%
  filter(!(extent_of_tumor_resection %in% c("Not Reported", "Unavailable")))

# Note the order of resection - the reference is "Biopsy only"
unique(combined_df_stranded_resection$extent_of_tumor_resection)

# saving so P-values can be programmatically grabbed
fitted <- fit_save_model(combined_df_stranded_resection,
                         paste0(quantiseq_terms,"+extent_of_tumor_resection +lgg_group"),
                         file.path(results_dir, "cox_terms_quantiseq.RDS"),
                         model_type = "multivariate")
fitted
```


## Analysis: quantiseq cell type fractions + extent of tumor resection + lgg group + PDL-1  across entire cohort

Three effects are significant: 

+ extent of tumor resection "partial resection" is signficantly different from biopsy
+ extent of tumor resection "gross/near total resection" is signficantly different from biopsy
+ LGG and non-LGG

PDL-1 expression is not signficant.  

```{r cox_cell_type_cd274}
# again use data with unknown resections removed
fit_save_model(combined_df_stranded_resection,
               paste0(quantiseq_terms, "+extent_of_tumor_resection+lgg_group+CD274"),
               file.path(results_dir, "cox_terms_quantiseq.RDS"),
               model_type = "multivariate")
```


## Determine suitable cancer groups for analysis

For cancer group analyses, we only want to consider groups with at least THREE DECEASED, again considering stranded only. We find those groups here:
```{r cancer_groups}
suitable_cancer_groups <- combined_df_stranded %>%
  drop_na() %>%
  count(cancer_group,OS_status) %>%
  spread(OS_status, n,
         # When there are no events, record the event count as 0 instead of the default NA
         fill = 0) %>%
  filter(DECEASED >= 3) 

# show result: there are 5 groups to keep
suitable_cancer_groups
```

For molecular signatures analysis, we want to only consider cancer groups, of those identified above, that have more than 1 molecular subtype.

```{r suitable_subtypes}
combined_df_stranded %>%
  filter(cancer_group %in% suitable_cancer_groups$cancer_group) %>%
  select(cancer_group, molecular_subtype) %>%
  distinct() %>%
  # count number of subtypes
  count(cancer_group) %>%
  filter(n > 1) -> suitable_subtypes


# n is how many subtypes the group has. There needs to be more than 1.
suitable_subtypes
```



## Analysis: quantiseq cell type fractions and PDL-1 + extent of tumor resection for each cancer group of interest (no LGG here)

> Models here are generally not robust. They either do not converge, or converge with infinite coefficients.


```{r cell_type_cg}

base_filename <- "cox_per_cg_terms_quantiseq.RDS" # all file names here will follow this convention

for (cg in suitable_cancer_groups$cancer_group) {
  cat("==============", cg, "==================")
  # filter to cancer group of interest, using data with unknown resections removed
  fit_data <- combined_df_stranded_resection %>% 
    filter(cancer_group == cg)
  
  # Define filename where model will be saved
  filename <- file.path(
    results_dir,
    str_replace(base_filename, "cg", cg)
  )
  
  # Fit model
  fit_save_model(fit_data, 
                 paste0(quantiseq_terms, "+CD274+extent_of_tumor_resection"), 
                 filename, 
                 model_type = "multivariate") %>% print()
}
```


#### Session Info

```{r}
sessionInfo()
```
