---
title: "Survival analysis by TP53 and telomerase activity"
authors: Run Jin (D3B), Jo Lynne Rokita (D3B), and Stephanie Spielman (CCDL)
output: 
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---

Note that for models that consider the `cancer_group` predictor or are performed separately for each cancer group *only consider cancer groups where `DECEASED >= 3`.* There are seven such groups.

## Setup

#### Packages and functions


```{r libraries}
library(survival)
library(ggpubr)
library(tidyverse)
```

#### Set up files and directories

Set up input and output directories:

```{r Set up directories}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses")

tp53_dir <- file.path(analysis_dir, "tp53_nf1_score")
telomerase_dir <- file.path(analysis_dir, "telomerase-activity-prediction")

results_dir <- "results/tp53_telomerase/"
plots_dir <- "plots"
util_dir <- "util"

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

Source the util function file.

```{r source_functions}
# Source functions used to build and export models
source(file.path(util_dir,"survival_models.R"))
```

Set up file paths.

```{r file_paths}

# Define input files:
metadata_file <- file.path(data_dir, "pbta-histologies.tsv")

# These analyses consider only stranded libraries:
tp53_file <- file.path(tp53_dir, 
                        "results", 
                        "tp53_altered_status.tsv")

tel_stranded_file <- file.path(telomerase_dir,
                               "results",
                               "TelomeraseScores_PTBAStranded_FPKM.txt")
```

## Prepare data for modeling

#### Import and process the metadata

```{r Read in metadata}
# The following patients have discrepant extent of tumor resection
# 7316-116 has Biopsy only;Gross/Near total resection
# 7316-182 has Gross/Near total resection;Partial resection
# Ticket was created [here](https://github.com/PediatricOpenTargets/ticket-tracker/issues/298) to track this database update, and I will recode below

# get primary tumor and remove cell lines
histologies_rna <- read_tsv(metadata_file, guess_max = 10000) %>%
  filter(composition=="Solid Tissue",
                  tumor_descriptor == "Initial CNS Tumor",
                  experimental_strategy == "RNA-Seq", 
                  RNA_library == "stranded") %>%
  rename(Kids_First_Biospecimen_ID_RNA = Kids_First_Biospecimen_ID) %>% 
  # clean up extent of tumor resection
  mutate(extent_of_tumor_resection = case_when(
    sample_id == "7316-116" ~ "Biopsy only",
    sample_id == "7316-182" ~ "Partial resection",
    sample_id == "7316-356" ~ "Partial resection", # https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/1430#pullrequestreview-1014340267
    TRUE ~ as.character(extent_of_tumor_resection))) %>%
  select(Kids_First_Participant_ID, 
         Kids_First_Biospecimen_ID_RNA, 
         sample_id, 
         broad_histology,
         cancer_group, 
         OS_status, 
         OS_days, 
         PFS_days, 
         extent_of_tumor_resection) %>%
  distinct() 
  
# read in tp53 scores
tp53_scores <- read_tsv(tp53_file) %>%
  select(Kids_First_Biospecimen_ID_RNA, sample_id, tp53_score) %>%
  distinct()

# read in telomerase scores
tel_scores <- read_tsv(tel_stranded_file) %>%
  rename(Kids_First_Biospecimen_ID_RNA = SampleID,
                telomerase_score = NormEXTENDScores) %>%
  arrange(Kids_First_Biospecimen_ID_RNA)


# combine tp53 scores and telomerase score to the dataframe
# note that we do not need to recode categories since the survival_models.R functions will take care of it
meta_indep <- histologies_rna %>%
  # add tp53 and telomerase scores by matching Kids_First_Biospecimen_ID_RNA, with only one of them joining sample_id
  left_join(tp53_scores %>% select(-sample_id), by = "Kids_First_Biospecimen_ID_RNA") %>%
  left_join(tel_scores, by = "Kids_First_Biospecimen_ID_RNA") %>% 
  rename(Kids_First_Biospecimen_ID = Kids_First_Biospecimen_ID_RNA) %>%
  distinct(Kids_First_Participant_ID, .keep_all=TRUE) %>%
  # remove samples without survival data
  drop_na(OS_days) %>%
  # convert to years
  mutate(OS_years = OS_days / 365.25)
```
 
#### Create subsets of data for different models

We create two new tibbles from `meta_indep`:

+ `meta_indep_grouped` which contains new columns `hgg_group` that contains levels (in order) `non-HGG` and `HGG` indicating whether it's a HGG sample and `lgg_group` that contains levels (in order) `non-LGG` and `LGG` indicating whether it's a LGG sample
+ `meta_indep_cancers` which is subsetted to only certain cancer groups of interest that have suitable sample sizes which we find here.

```{r lgg_hgg_split}
# Define LGG and HGG groups
hgg_cancer_groups <- c("High-grade glioma astrocytoma", "Diffuse midline glioma", "Diffuse intrinsic pontine glioma") 
hgg_levels <- c("non-HGG", "HGG")
lgg_levels <- c("non-LGG", "LGG")

# separate into HGG or non-HGG and make the new column a factor
meta_indep_grouped <- meta_indep %>% 
   mutate(hgg_group = ifelse(cancer_group %in% hgg_cancer_groups,
                             hgg_levels[2], # second is HGG 
                             hgg_levels[1]), # first is non-HGG
          hgg_group = factor(hgg_group, levels = hgg_levels), # ensures non-HGG is the reference level
          lgg_group = ifelse(broad_histology == "Low-grade astrocytic tumor",
                             lgg_levels[2], # second is LGG 
                             lgg_levels[1]), # first is non-LGG
          lgg_group = factor(lgg_group, levels = lgg_levels) # ensures non-HGG is the reference level
   )
# Show the data we will use for analysis
meta_indep_grouped
```

```{r find_cancer_groups_to_use}
# Define cancer groups of interest with higher Ns
# We want to only include groups that have at least THREE in the DECEASED category. 
# Find them here:
cancer_groups_of_interest_df <- meta_indep %>%
  drop_na() %>%
  count(cancer_group,OS_status) %>%
  spread(OS_status, n,
         # When there are no events, record the event count as 0 instead of the default NA
         fill = 0) %>%
  filter(DECEASED >= 3) 
# show result: there are 7 groups to keep
cancer_groups_of_interest_df

# Subset to only the cancer groups with potentially suitable sample size
meta_indep_cancer_groups <- meta_indep %>%
  filter(cancer_group %in% cancer_groups_of_interest_df$cancer_group) %>%
  # set reference level to Medulloblastoma
  mutate(cancer_group = fct_relevel(cancer_group, "Medulloblastoma"))

# How many patients got removed?
nrow(meta_indep) - nrow(meta_indep_cancer_groups)
```

## Univariate Models

 
Here, we run a survival analyses for several _independent_ predictors, one at a time. We use cox regression for all analyses (including categorical), as demonstrated in `survival-analysis_template.Rmd`.

+ `extent_of_tumor_resection` (categorical)
+ `lgg_group` (categorical)
+ `hgg_group` (categorical)
+ `cancer_group` (categorical)

### Extent of tumor resection 
Confirming gross/near total tumor resection has a very significant prognostic effect. The greater the resection, the better the OS.

```{r cox_resect}
# We want to remove all unavailable/not reported from this model
meta_indep_grouped_resection <- meta_indep_grouped %>%
  # REMOVE unreported resections
  filter(!(extent_of_tumor_resection %in% c("Not Reported", "Unavailable"))) 
  
  
# Note the order of extent of resection - Biopsy only is the reference level
unique(meta_indep_grouped_resection$extent_of_tumor_resection)

# full dataset
fit_save_model(meta_indep_grouped_resection, 
              "extent_of_tumor_resection", 
             file.path(results_dir, "cox_terms_resect_all.RDS"))

# Are total resected tumors mainly LGG? About 30% of all fully resected are LGG, so LGG and extent_of_tumor_resection == "Gross/Near Total" do not make up the same cohort
table(meta_indep_grouped_resection$lgg_group, meta_indep_grouped_resection$extent_of_tumor_resection)

table(meta_indep_grouped_resection$hgg_group, meta_indep_grouped_resection$extent_of_tumor_resection)

```

###  LGG or HGG group

> non-LGG/non-HGG are the reference levels. No surprises here: LGG group has better overall survival and HGG has worse overall survival.

```{r cox_hgg}
fit_save_model(meta_indep_grouped, 
              "hgg_group", 
             file.path(results_dir, "cox_terms_hgg_group.RDS"))

fit_save_model(meta_indep_grouped, 
              "lgg_group", 
             file.path(results_dir, "cox_terms_lgg_group.RDS"))
```

### Cancer group (cox)

> MB is the reference level

```{r cox_cancer_group}
fit_save_model(meta_indep_cancer_groups, 
               "cancer_group", 
               file.path(results_dir, "cox_terms_cancer_group.RDS"))
```



### Cancer group (log rank)

Here, we perform a log-rank analysis specifically for the `cancer_group` predictor, using data that only considers the 7 cancer groups determined to have suitable sample sizes (`DECEASED >= 3`). The output below shows:

+ Table of model coefficients
+ Matrix of `Pairwise comparisons using Log-Rank test`, which shows which cancer groups are significantly different from one another

Note that confidence intervals are so large that survival plots are not clearly legible, so we don't make that visualization here.

```{r cancer_group_log_rank}

# Print model coefficients
fit_save_model(meta_indep_cancer_groups,
               "cancer_group",
               file.path(results_dir,
                         "log_rank_survival_per_cancer_group.RDS"),
               test = "log.rank")


# Compare pairwise with Bonferroni correction for multiple testing
survminer::pairwise_survdiff(survival::Surv(time = OS_years, event = OS_status) ~ cancer_group,
                             meta_indep_cancer_groups %>% mutate(OS_status = ifelse(OS_status == "LIVING", 0, 1)), 
                             p.adjust.method = "bonferroni")

```

## Multivariate models 


Background on hazard ratios and output, from [ref](http://www.sthda.com/english/wiki/cox-proportional-hazards-model):

>"Hazard ratio above 1 indicates a covariate that is positively associated with the event probability, and thus negatively associated with the length of survival.
>
>In summary,
>
>HR = 1: No effect
>HR < 1: Reduction in the hazard
>HR > 1: Increase in Hazard" 
>
>The Cox regression results can be interpreted as follow:
_Statistical significance._ The column marked "z" gives the Wald statistic value. It corresponds to the ratio of each regression coefficient to its standard error (z = coef/se(coef)). The wald statistic evaluates, whether the beta (β) coefficient of a given variable is statistically significantly different from 0. From the output above, we can conclude that the variable sex have highly statistically significant coefficients.
>
>_The regression coefficients._ The second feature to note in the Cox model results is the the sign of the regression coefficients (coef). A positive sign means that the hazard (risk of death) is higher, and thus the prognosis worse, for subjects with higher values of that variable. The variable sex is encoded as a numeric vector. 1: male, 2: female. The R summary for the Cox model gives the hazard ratio (HR) for the second group relative to the first group, that is, female versus male. The beta coefficient for sex = -0.53 indicates that females have lower risk of death (lower survival rates) than males, in these data.
>
>_Hazard ratios._ The exponentiated coefficients (exp(coef) = exp(-0.53) = 0.59), also known as hazard ratios, give the effect size of covariates. For example, being female (sex=2) reduces the hazard by a factor of 0.59, or 41%. Being female is associated with good prognostic.
>
>_Confidence intervals of the hazard ratios._ The summary output also gives upper and lower 95% confidence intervals for the hazard ratio (exp(coef)), lower 95% bound = 0.4237, upper 95% bound = 0.816.
>
>_Global statistical significance of the model._ Finally, the output gives p-values for three alternative tests for overall significance of the model: The likelihood-ratio test, Wald test, and score logrank statistics. These three methods are asymptotically equivalent. For large enough N, they will give similar results. For small N, they may differ somewhat. The Likelihood ratio test has better behavior for small sample sizes, so it is generally preferred.
>
> Below, HR = exp(coeff) 

### Interaction model: `tp53_score*telomerase_score*extent_of_tumor_resection*lgg_group` 

```{r multi_interaction_4terms extent_tumor_resect glioma_group}

# we save this for the purpose of printing out a p-value below
int_model <- fit_save_model(meta_indep_grouped_resection,     # use the data without unknown resection data for this model
             "tp53_score*telomerase_score*extent_of_tumor_resection*lgg_group*hgg_group",
             file.path(results_dir, "cox_interaction_terms_tp53_telomerase_resect_glioma_group.RDS"),
             "multivariate"
            )
int_model
```

Interaction terms are not significant, except `tp53_score:extent_of_tumor_resectionUnavailable:hgg_groupHGG` and `extent_of_tumor_resectionUnavailable:hgg_groupHGG`, suggesting an additive model is more appropriate.


### Additive model: `tp53_score+telomerase_score+extent_of_tumor_resection+lgg_group+hgg_group` 

> The strongest predictor here is extent of tumor resection with gross/near total resection having improved survival compared to biopsy.
> Non-LGG is a significant risk factor for poorer overall survival, and similar HGG is a risk factor for poorer survival.
> We do not see a significant effect of TP53 scores, but we do see a highly significant effect of telomerase scores.

```{r multi_additive_4terms extent_tumor_resect glioma_group}
# we save this for the purpose of printing out a p-value below
add_model <- fit_save_model(meta_indep_grouped_resection,     # use the data unknown resection data for this model
             "tp53_score+telomerase_score+extent_of_tumor_resection+lgg_group+hgg_group",
             file.path(results_dir, "cox_additive_terms_tp53_telomerase_resect_glioma_group.RDS"),
             "multivariate"
            )
add_model
```


### Additive models: `tp53_score+telomerase_score+extent_of_tumor_resection` _for each cancer group_


```{r multivar-fit-3terms}
base_filename <- "cox_per_cg_terms_tp53_telomerase_resect.RDS" # all file names here will follow this convention
  
# remove unknown extent of resections from this data
meta_indep <- meta_indep %>%
  filter(!(extent_of_tumor_resection %in% c("Not Reported", "Unavailable")))

for (cg in cancer_groups_of_interest_df$cancer_group) {
  cat("==============", cg, "==================")
  # filter to cancer group of interest
  fit_data <- meta_indep %>% 
    filter(cancer_group == cg)

  # Define filename where model will be saved
  filename <- file.path(
    results_dir,
    str_replace(base_filename, "cg", cg)
  )
  
  # Fit model
  fit_save_model(fit_data, 
                 "tp53_score + telomerase_score + extent_of_tumor_resection", 
                 filename, 
                 model_type = "multivariate") %>% print()
}


```


## Session Info

```{r Print session info}
sessionInfo()
```