---
title: "00-prepare-extraction-table"
author: "Jo Lynne Rokita"
date: 2022
output: 
  html_notebook:
    toc: true
    toc_float: true
---

## This notebook wrangles manually curated files which have been collated to designate whether a tuomor DNA and RNA specimen came from the same extraction, same tissue chunk ("same tissue aliquot"), or different tissue chunk ("different tissue aliquot"). 


## Load library/input files
```{r}
library(tidyverse)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
input_dir <- file.path(root_dir,
                      "analyses",
                      "molecular-subtyping-integrate",
                       "input")

results_dir <- file.path(root_dir,
                         "analyses",
                         "molecular-subtyping-integrate",
                         "results")

# Read in PNOC and CBTN metadata files
pnoc_aliquots <- read_tsv(file.path(input_dir, "pnoc_aliquots.tsv"))
cbtn_aliquots <- read_tsv(file.path(input_dir, "cbtn_aliquots.tsv")) %>%
  # necessary for combining with PNOC later
  mutate(aliquot_id = as.character(aliquot_id))
```

# Get extraction status from PNOC cohort
```{r}
# Using the parental_aliquot_id, determine which DNA + RNA pair together (ie came from the same extraction).
pnoc_aliquots_status <- pnoc_aliquots %>%
  # split sample and aliquot id by experimental strategy to group by parental aliquot id later
  mutate(sample_id_rna = ifelse(experimental_strategy == "RNA-Seq", sample_id, ""),
         sample_id_wxs = ifelse(experimental_strategy == "WXS", sample_id, ""),
         sample_id_wgs = ifelse(experimental_strategy == "WGS", sample_id, ""),
         aliquot_id_rna = ifelse(experimental_strategy == "RNA-Seq", aliquot_id, ""),
         aliquot_id_wxs = ifelse(experimental_strategy == "WXS", aliquot_id, ""),
         aliquot_id_wgs = ifelse(experimental_strategy == "WGS", aliquot_id, "")) %>%
  distinct() %>%
  select(-c(experimental_strategy, sample_id, aliquot_id)) %>%
  group_by(parent_aliquot_id) %>%
  # collapse
  summarise_all(funs(trimws(paste(., collapse = '')))) %>%
  # if there is an RNA sample and DNA sample which have the same parental aliquot id, deem them as from the same extraction
  mutate(extraction_type = ifelse(aliquot_id_rna != "" & (aliquot_id_wxs != "" | aliquot_id_wgs != ""),
                                    "Same extraction", NA_character_)) %>%
  mutate_all(na_if,"") %>%
  # rename to be compatible with CBTN data
  dplyr::rename(parental_aliquot_id = parent_aliquot_id)
```

# Combine CBTN and PNOC RNA sample and aliquot ids with extraction status
```{r}
combined <- pnoc_aliquots_status %>%
  # only keep those with RNA-Seq (had extraction annotated)
  filter(!is.na(extraction_type)) %>%
  dplyr::rename(sample_id = sample_id_rna,
                aliquot_id = aliquot_id_rna) %>%
  select(sample_id, aliquot_id, parental_aliquot_id, extraction_type) %>%
  # combine with CBTN
  bind_rows(cbtn_aliquots) %>%
  # clean up parental_aliquot id
  mutate(parental_aliquot_id = gsub("^.*\\[|\\]", "", parental_aliquot_id)) %>%
  write_tsv(file.path(results_dir, "pbta_extraction_status.tsv"))
```

# How many samples in each group
```{r}
combined %>%
count(extraction_type) %>%
  arrange(desc(n))
```


```{r}
sessionInfo()
```

