---
title: "Integrate molecular subtyping results"
output: 
  html_notebook:
    toc: true
    toc_float: true
author: Krutika Gaonkar for D3b
date: 2020
params:
   is_ci: FALSE
---

The purpose of this notebook is to integrate molecular subtyping results from 
[molecular-subtyping-pathology](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/molecular-subtyping-pathology) with `pbta-histologies-base.tsv`.

Here we are using v18 `pbta-histologies-base.tsv`


## Set up

```{r}
library(tidyverse)
data_dir <- "../../data/release-v18-20201123/"
base_histology <- read_tsv(file.path(data_dir,"pbta-histologies-base.tsv"),
                           col_types = readr::cols(molecular_subtype = readr::col_character(),
                                                   short_histology = readr::col_character(),
                           broad_histology = readr::col_character(),
                           Notes = readr::col_character())) %>%
  unique()
```

### Read molecular-subtyping-pathology results

Reading molecular_subtype, integrated_diagnosis, short_histology, broad_histology and Notes from `compiled_molecular_subtypes_with_clinical_pathology_feedback.tsv` 

```{r}

compiled_subtyping<-read_tsv(file.path("..", "molecular-subtyping-pathology", "results", "compiled_molecular_subtypes_with_clinical_pathology_feedback.tsv"))

```

Update "Other" sample broad/short histology and harmonized_diagnosis and add to `compiled_subtyping`

```{r}

Other_subtypes <- read_tsv(file.path("input","pathology_dx and pathology_free_text_diagnosis to broad_histology for subtyping module - rules_without_subtype_WIP.tsv"))

compiled_subtyping_other <- base_histology %>%
  select(
  Kids_First_Participant_ID ,
  sample_id,
  Kids_First_Biospecimen_ID,
  molecular_subtype,
  integrated_diagnosis,
  pathology_diagnosis,
  pathology_free_text_diagnosis,
  tumor_descriptor
) %>%
  # remove Kids_First_Biospecimen_ID which are subtypes "ETMR/Embryonal"
  filter(!Kids_First_Biospecimen_ID %in% compiled_subtyping$Kids_First_Biospecimen_ID) %>%
  # get pathology_diagnosis=="Other"
  filter(pathology_diagnosis=="Other") %>%
  left_join(Other_subtypes,by=c("pathology_free_text_diagnosis")) %>%
  mutate(Notes = "Updated by manual review of WHO diagnosis") %>%
select(
  # gather only columns needed to format as `compiled_subtyping`
  Kids_First_Participant_ID ,
  sample_id,
  Kids_First_Biospecimen_ID,
  molecular_subtype,
  integrated_diagnosis,
  tumor_descriptor,
  broad_histology,
  short_histology,
  Notes,
  # adding harmonized_diagnosis from manual review 
  # for pathology_diagnosis=="Other"
  harmonized_diagnosis
) %>%
  unique()


# combined OpenPBTA subtypes and manual "Other" subtypes
compiled_subtyping <- compiled_subtyping_other %>%
  bind_rows(compiled_subtyping)

```



### Add molecular-subtyping-pathology results

We will add molecular_subtype, integrated_diagnosis and Notes from `compiled_subtyping`

short_histology and broad_histology will be added from base histology for samples that are not subtyped as part of `molecular-subtype-pathology`

```{r}

histology <- base_histology %>% 
  select(-Notes,-molecular_subtype,-integrated_diagnosis) %>%
  left_join(compiled_subtyping,by=c("Kids_First_Biospecimen_ID","sample_id","Kids_First_Participant_ID","tumor_descriptor"),suffix=c(".base",".subtyped")) %>%
  unique() %>%
  mutate(
    broad_histology = if_else(!is.na(broad_histology.subtyped),
                              broad_histology.subtyped,
                              broad_histology.base),
    short_histology = if_else(!is.na(short_histology.subtyped),
                              short_histology.subtyped,
                              short_histology.base),
    harmonized_diagnosis = 
      case_when(!is.na(integrated_diagnosis) ~ integrated_diagnosis,
                is.na(integrated_diagnosis) & 
                  !is.na(harmonized_diagnosis) ~ harmonized_diagnosis,
                is.na(integrated_diagnosis) & 
                  is.na(harmonized_diagnosis) &
                  !is.na(pathology_diagnosis) ~ pathology_diagnosis
    )) 
```



### Check if any duplicates

```{r}
dup_ids<-histology$Kids_First_Biospecimen_ID[duplicated(histology$Kids_First_Biospecimen_ID)]

histology[which(histology$Kids_First_Biospecimen_ID %in% dup_ids),]
```

No duplicates

### Check if broad_histology, short_histology or harmonized_diagnosis

Are there NA in broad_histology, short_histology or harmonized_diagnosis

```{r}
histology %>% 
  filter(sample_type=="Tumor",
         (is.na(broad_histology)| is.na(short_histology)| is.na(harmonized_diagnosis))) %>%
  tally()

```

No NAs in broad_histology, short_histology or harmonized_diagnosis

Just a note, integrated_diagnosis is expected to be `NA` for samples where subtyping is not performed or if molecular_subtype is "XYZ,To be classified". 
This means no evidence was provided/available for these samples so we are not able to add integrated_diagnosis.

#### Check differences in broad_histology 
Checking for differences in broad_histology to look for changes in molecular_subtype 


```{r}
diff_broad_histology<- histology %>%
  filter(toupper(broad_histology.base) != toupper(broad_histology.subtyped)) %>%
  select(Kids_First_Biospecimen_ID,starts_with("broad_histology"),starts_with("short_histology")) %>%
  unique()

diff_broad_histology
```

#### Check differences in short_histology
Here we want to check for short_histology changes not part of `Check differences in broad_histology` chunk.
This will help us check what string assignment path_dx to short_histology mapping has changed from `molecular-subtyping-pathology`

```{r}

histology %>% 
  filter(!Kids_First_Biospecimen_ID %in% diff_broad_histology$Kids_First_Biospecimen_ID) %>%
   filter(toupper(short_histology.base) != toupper(short_histology.subtyped) 
         ) %>%
  select(Kids_First_Biospecimen_ID,starts_with("broad_histology"),starts_with("short_histology")) %>%
  unique() 

```
The above 172 changes occurred because of changes in string value assignment
In `compiled_molecular_subtypes_with_clinical_pathology_feedback.tsv` a sample with broad_histology that is `Ependymal tumor` is `EPN`, but in `pbta-histologies-base.tsv` it is `Ependymoma`.
For samples where broad_histology is `Embryonal tumor`, short_histology is also `Embryonal tumor` but in base histology it was `ETMR`.

45 Benign ,Non-(CNS) tumor and other samples where pathology_diagnosis == "Other",have short_histology updated from manual review of WHO diagnosis terms.

### Save
Let's save the final file.

But first need to remove broad_histology.base, broad_histology.subtyped and short_histology.base
 and short_histology.subtyped

```{r}
histology %>% 
  select(-broad_histology.base,
         -broad_histology.subtyped,
         - short_histology.base,
         -short_histology.subtyped) %>%
  write_tsv("results/pbta-histologies.tsv")
```
