---
title: "Integrate molecular subtyping results"
output: 
  html_notebook:
    toc: true
    toc_float: true
author: Krutika Gaonkar for D3b
date: 2020
params:
   is_ci: FALSE
---

The purpose of this notebook is to integrate molecular subtyping results from 
[molecular-subtyping-pathology](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/molecular-subtyping-pathology) with `pbta-histologies-base.tsv`.

Here we are using v18 `pbta-histologies-base.tsv`


## Set up

```{r}
library(tidyverse)
data_dir <- "../../data/release-v18-20201123/"
base_histology <- read_tsv(file.path(data_dir,"pbta-histologies-base.tsv"),
                           col_types = readr::cols(molecular_subtype = readr::col_character(),
                                                   short_histology = readr::col_character(),
                           broad_histology = readr::col_character(),
                           Notes = readr::col_character())) %>%
  unique()
```

### Read molecular-subtyping-pathology results

Reading molecular_subtype,	integrated_diagnosis, short_histology,	broad_histology and	Notes from `compiled_molecular_subtypes_with_clinical_pathology_feedback.tsv` 

```{r}

compiled_subtyping<-read_tsv("../molecular-subtyping-pathology/results/compiled_molecular_subtypes_with_clinical_pathology_feedback.tsv")

```

### Add molecular-subtyping-pathology results

We will add molecular_subtype,	integrated_diagnosis and	Notes from `compiled_subtyping`

short_histology and	broad_histology will be added from base histology for samples that are not subtyped as part of `molecular-subtype-pathology`

```{r}

histology <- base_histology %>% 
  select(-Notes,-molecular_subtype,-integrated_diagnosis) %>%
  left_join(compiled_subtyping,by=c("Kids_First_Biospecimen_ID","sample_id","Kids_First_Participant_ID","tumor_descriptor"),suffix=c(".base",".subtyped")) %>%
  unique() %>%
  mutate(
    broad_histology = if_else(!is.na(broad_histology.subtyped),broad_histology.subtyped,broad_histology.base),
    short_histology = if_else(!is.na(short_histology.subtyped),short_histology.subtyped,short_histology.base),
    harmonized_diagnosis = if_else(!is.na(integrated_diagnosis),
                                        integrated_diagnosis,
                                        pathology_diagnosis)) 

```


### Check if any duplicates

```{r}
dup_ids<-histology$Kids_First_Biospecimen_ID[duplicated(histology$Kids_First_Biospecimen_ID)]

histology[which(histology$Kids_First_Biospecimen_ID %in% dup_ids),]
```

No duplicates

### Check if broad_histology, short_histology or harmonized_diagnosis

Are there NA in broad_histology, short_histology or harmonized_diagnosis

```{r}
histology %>% 
  filter(sample_type=="Tumor",
         (is.na(broad_histology)| is.na(short_histology)| is.na(harmonized_diagnosis))) %>%
  tally()

```

No NAs in broad_histology, short_histology or harmonized_diagnosis

#### Check differences in broad_histology 
Checking for differences in broad_histology to look for changes in molecular_subtype 


```{r}
diff_broad_histology<- histology %>%
  filter(toupper(broad_histology.base) != toupper(broad_histology.subtyped)) %>%
  select(Kids_First_Biospecimen_ID,starts_with("broad_histology"),starts_with("short_histology")) %>%
  unique()

diff_broad_histology
```

#### Check differences in short_histology
Here we want to check for short_histology changes not part of `Check differences in broad_histology` chunk.
This will help us check what string assignment path_dx to short_histology mapping has changed from `molecular-subtyping-pathology`

```{r}

histology %>% 
  filter(!Kids_First_Biospecimen_ID %in% diff_broad_histology$Kids_First_Biospecimen_ID) %>%
   filter(toupper(short_histology.base) != toupper(short_histology.subtyped) 
         ) %>%
  select(Kids_First_Biospecimen_ID,starts_with("broad_histology"),starts_with("short_histology")) %>%
  unique()

```
The above 172 changes occurred because of changes in string value assignment
In `compiled_molecular_subtypes_with_clinical_pathology_feedback.tsv` sample with broad_histology== `Ependymal tumor` is `EPN`, but in base histology it is `Ependymoma`.
For samples with broad_histology==`Embryonal tumor`, short_histology is `Embryonal tumor` but in base histology it was `ETMR`.

### Save
Let's save the final file.

But first need to remove broad_histology.base, broad_histology.subtyped and short_histology.base
 and short_histology.subtyped

```{r}
histology %>% 
  select(-broad_histology.base,
         -broad_histology.subtyped,
         - short_histology.base,
         -short_histology.subtyped) %>%
  write_tsv("results/pbta-histologies.tsv")
```


