---
title: "LGAT final table for BRAF V600E and _BRAF_ fusion"
output: html_notebook
---

This script uses subsetted LGAT mutation and fusion files from data release to identify samples with BRAF V600E mutations and _BRAF_ fusions per [#631](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/631).

```{r}
library("tidyverse")
```

### Directories

```{r}
# to get root directory
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")

results_dir <- "results"
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```


### Read files
    
```{r}
# clinical file
clinical_df <- readr::read_tsv(file.path(data_dir, 
                                      "pbta-histologies.tsv"),
                            guess_max = 10000)

# identifiers only for joining with other data
identifiers_df <- clinical_df %>%
  select(Kids_First_Participant_ID,
         Kids_First_Biospecimen_ID,
         sample_id)

# get all RNA-Seq biospecimen IDs where short_histology == LGAT
lgat_rnaseq_bsids <- clinical_df %>% 
  filter(short_histology == "LGAT",
         experimental_strategy == "RNA-Seq") %>%
  pull(Kids_First_Biospecimen_ID)

# get subsetted mutation file
subsetWgs <- readr::read_tsv(file.path("lgat-subset", 
                                       "LGAT_snv_subset.tsv")) %>% 
  select(Tumor_Sample_Barcode, 
         BRAF_V600E) %>%
  rename(Kids_First_Biospecimen_ID = Tumor_Sample_Barcode)
  
# get putative oncogene fusion list
brafFusion <- readr::read_tsv(file.path(data_dir,
                                        "pbta-fusion-putative-oncogenic.tsv")) %>%
  select(Sample, FusionName) %>% 
  distinct() %>%
  filter(grepl("BRAF", FusionName),
         Sample %in% lgat_rnaseq_bsids) %>%
  rename(Kids_First_Biospecimen_ID = Sample)
```

### Fusion Data

Some samples will have multiple _BRAF_ fusions, so we'll collapse multiple fusions into a single column.

```{r}
brafFusion <- brafFusion %>%
  group_by(Kids_First_Biospecimen_ID) %>%
  summarize(FusionName = paste(sort(unique(FusionName)),
                               collapse = ", "))
```


All RNA-seq samples are represented in at least one of the fusion methods files.
_BRAF_ fusions are in the list that are prioritized in the putative oncogenic fusion file (which can be retained even in instances where only one caller supports the call).
Therefore, a _BRAF_ fusion can be detected in any RNA-seq biospecimen and if a _BRAF_ fusion is not present in `brafFusion` we can assume no _BRAF_ is detected.

```{r}
fusion_df <- brafFusion %>%
  right_join(filter(identifiers_df,
                    Kids_First_Biospecimen_ID %in% lgat_rnaseq_bsids)) %>%
  replace_na(list(FusionName = "None"))
```

### Merge all subset data for _BRAF_ fusion and BRAF V600E mutations

```{r}
# LGAT WGS subset with kids_first_participant and sample_id
subsetWgs <- subsetWgs %>% 
  left_join(identifiers_df, 
            by = "Kids_First_Biospecimen_ID") 
```

```{r}
all_lgat_df <- subsetWgs %>%
  full_join(fusion_df, 
            by = c("sample_id", "Kids_First_Participant_ID"),
            suffix = c("_DNA", "_RNA"))

# write the data underlying the calls to file
all_lgat_df %>% 
  write_tsv(file.path(results_dir, "lgat_cleaned_all_table.tsv"))
```

### Logic for subtyping

```{r}
# TODO: if we update to tidyr 1.0.0 I think using unite would be better here
final_table <- all_lgat_df %>%
  mutate(molecular_subtype = case_when(
    # When we don't have RNA-seq or mutation data, this is to be classified
    is.na(FusionName) & is.na(BRAF_V600E) ~ "LGG, To be classified",
    # When there is no BRAF fusion detected and there is no DNA-seq data
    # to be classified
    FusionName == "None" & is.na(BRAF_V600E) ~ "LGG, To be classified",
    # If there is no fusion data or BRAF fusion detection, but there is a 
    # BRAF V600E mutation, call BRAF V600E
    (FusionName %in% c(NA, "None")) & 
      BRAF_V600E == "Yes" ~ "LGG, BRAF V600E",
    # If there's no fusion data or BRAF fusion detected or V600E, wildtype call 
    (FusionName %in% c(NA, "None"))  &
      BRAF_V600E == "No" ~ "LGG, BRAF wildtype",
    # If there is a BRAF fusion detected but no BRAF V600E / the absence of 
    # a DNA-seq sample, call is BRAF fusion
    !(FusionName %in% c(NA, "None")) & 
      (BRAF_V600E %in% c("No", NA)) ~ "LGG, BRAF fusion",
    # If there is a BRAF fusion detected and a BRAF V600E mutation, note both
    !(FusionName %in% c(NA, "None")) & 
      BRAF_V600E == "Yes" ~ "LGG, BRAF fusion/V600E"
  )) %>%
  select(Kids_First_Participant_ID,
         sample_id,
         Kids_First_Biospecimen_ID_DNA,
         Kids_First_Biospecimen_ID_RNA,
         molecular_subtype)

write_tsv(final_table, file.path(results_dir,"lgat_subtyping.tsv"))
```
