---
title: "Annotate SNV subtype status for LGAT biospecimens"
output: html_notebook
author: K S Gaonkar 
date: 2020
---

In this PR we will use identify LGAT biospecimens from pathology diagnosis and annotate subtype specific SNV status per biospecimen.

As per [issue](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/790) we will be subtyping LGAT based on SNV in the following genes:

- LGG, NF1
somatic loss of NF1 via either missense, nonsense mutation 

- LGG, BRAF V600E
contains BRAF V600E or V599 SNV or non-canonical BRAF alterations such as p.V600ins or p.D594N

- LGG, other MAPK
contains KRAS, NRAS, HRAS, MAP2K1, MAP2K2, MAP2K1, ARAF SNV or indel

- LGG, RTK
harbors a MET SNV 
harbors a KIT SNV or
harbors a PDGFRA SNV 

- LGG, FGFR
harbors FGFR1 p.N546K, p.K656E, p.N577, or p. K687 hotspot mutations or

- LGG, IDH
harbors an IDH R132 mutation

- LGG, H3.3
harbors an H3F3A or H3F3B K28M or G35R/V mutation

-  LGG, H3.1
harbors an HIST1H3B K28M
harbors and HIST1H3C  K28M


### Setup
```{r}
library(tidyverse)

# Look for git root folder
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

# get subset folder
subset_dir <- file.path(root_dir, 
                        "analyses",
                        "molecular-subtyping-LGAT",
                        "lgat-subset")

# create if doesn't exist
if (!dir.exists(subset_dir)) {
  dir.create(subset_dir)
}
```

### Input

```{r}
# File from 00-LGAT-select-pathology-dx that is used for the pathology diagnosis
# inclusion/exclusion criteria
path_dx_list <- jsonlite::fromJSON(
  file.path(subset_dir, 
            "lgat_subtyping_path_dx_strings.json")
)

# clinical file
clinical <- read_tsv(file.path(root_dir, 
                               "data",
                               "pbta-histologies-base.tsv"), 
                     guess_max = 10000)

# consensus mutation data
consensusMutation <- read_tsv(file.path(root_dir,
                                        "data",
                                        "pbta-snv-consensus-mutation.maf.tsv.gz"))

```

Filter to tumor samples that should be included on the basis of pathology diagnosis
```{r}
lgat_specimens_df <- clinical %>%
  dplyr::filter(str_detect(str_to_lower(pathology_diagnosis),  # Inclusion criteria
                           paste(path_dx_list$include_path_dx, collapse = "|")),
                # Exclusion criteria
                str_detect(str_to_lower(pathology_diagnosis),
                           paste(path_dx_list$exclude_path_dx, collapse = "|"),
                           negate = TRUE),
                # Tumors
                sample_type == "Tumor",
                composition == "Solid Tissue")

# Write this intermediate file to the subset directory as it allows for
# inspection
write_tsv(lgat_specimens_df, file.path(subset_dir, "lgat_metadata.tsv"))

# Filter to dna samples
lgat_dna_df <- lgat_specimens_df %>%
  dplyr::filter(experimental_strategy != "RNA-Seq") %>%
  # will keep Kids_First_Biospecimen_ID 
  # sample_id is kept to be able to match with RNA-Seq in the later step
  dplyr::select(Kids_First_Biospecimen_ID,sample_id)
```


Gather gene(s) that define LGAT subtypes
Additional information for genes with known hotspots/canonical mutation is also provided in the list. 
If the gene has multiple hotspots we will gather the protein hotspots site to use it in the below chunks to grep the HGVSp_Short column.

```{r}
# combined list for SNV of interest per subtype
#
snvOI <- jsonlite::fromJSON(file.path(root_dir,
                                      "analyses",
                                      "molecular-subtyping-LGAT",
                                      "input",
                                      "snvOI_list.json"))

# Collapse multiple hotspots in genes with "|" so easy grep calls
BRAF_hotspot <-paste(snvOI$BRAF_V600E$hotspot[!is.na( snvOI$BRAF_V600E$hotspot)],collapse = "|")
FGFR_hotspot <-paste(snvOI$FGFR$hotspot[!is.na( snvOI$FGFR$hotspot)],collapse = "|")
IDH1_hotspot<- paste(snvOI$IDH1$hotspot[!is.na( snvOI$IDH1$hotspot)],collapse = "|")
```


### Subset consensus maf and annotate per subtype based on SNV
We will gather SNV calls that satisfy the conditions per subtype and save the subset along with the HGVSp_Short, DOMAINS and Variant_Classification 

```{r}
# Filter consensus mutation files for LGAT subset
consensusMutationSubset <- consensusMutation %>%
  # find lgat samples
  dplyr::filter(Tumor_Sample_Barcode %in% lgat_dna_df$Kids_First_Biospecimen_ID) %>%
  # select tumor sample barcode, gene, short protein annotation, domains, and variant classification
  dplyr::select(Tumor_Sample_Barcode,
                Hugo_Symbol,
                HGVSp_Short,
                DOMAINS,
                Variant_Classification,
                IMPACT,
                SIFT,
                PolyPhen) %>%
  dplyr::filter(
    # get BRAF mutation status
    # canonical mutations V600E
    HGVSp_Short %in% snvOI$BRAF_V600E$canonical[!is.na(snvOI$BRAF_V600E$canonical)] &
      Hugo_Symbol=="BRAF" | # OR
      # hotspot mutations in p.600 and p.599
      grepl(BRAF_hotspot,HGVSp_Short) &
      Hugo_Symbol=="BRAF" | # OR
      # and kinase domain mutation for non-canonical mutation 
      # Family: PK_Tyr_Ser-Thr https://pfam.xfam.org/family/PF07714
      grepl("PF07714",DOMAINS) & 
      Hugo_Symbol=="BRAF" | # OR
      
      # get NF1 mutation status
      Hugo_Symbol %in% snvOI$NF1$gene & 
      Variant_Classification %in% c("Missense_Mutation","Nonsense_Mutation") |
      
      # get other MAPK mutation status
      # all mutations in MAPK genes
      Hugo_Symbol %in% snvOI$MAPK$gene | # OR
      
      # get RTK mutation status
      # all mutations in RTK genes
      Hugo_Symbol %in% snvOI$RTK$gene | # OR
      
      # get FGFR mutation status
      # canonical mutations
      HGVSp_Short %in% snvOI$FGFR$canonical[!is.na(snvOI$FGFR$canonical)] &
      Hugo_Symbol=="FGFR1" | # OR
      # hotspot mutations 
      grepl(FGFR_hotspot,HGVSp_Short) &
      Hugo_Symbol=="FGFR1" | # OR
      
      # get IDH mutation status
      # hostspot mutations
      grepl(IDH1_hotspot,HGVSp_Short) & 
      Hugo_Symbol %in% snvOI$IDH1$gene | # OR
      
      # get histone mutation status
      # H3F3A canonical mutations
      HGVSp_Short %in% snvOI$H3F3A$canonical & Hugo_Symbol %in% "H3F3A" | # OR
      # H3F3B canonical mutations
      HGVSp_Short %in% snvOI$H3F3B$canonical & Hugo_Symbol %in% "H3F3B" | # OR
      # HIST1H3B canonical mutations
      HGVSp_Short %in% snvOI$HIST1H3B$canonical & Hugo_Symbol %in% "HIST1H3B" | # OR
      # HIST1H3C canonical mutations
      HGVSp_Short %in% snvOI$HIST1H3C$canonical & Hugo_Symbol %in% "HIST1H3C" | # OR
      # HIST2H3C canonical mutations     
      HGVSp_Short %in% snvOI$HIST2H3C$canonical & Hugo_Symbol %in% "HIST2H3C" 
  ) 

consensusMutationSubset
```

What is the distribution of Variant_Classification of the SNVs captured per gene?
```{r}
consensusMutationSubset %>% 
  group_by(Hugo_Symbol,Variant_Classification) %>%
  tally() 

```
We have some synonymous variants (classified as Intron, 5`Flank and Silent) captured per gene of interest.    
 
 
We will now remove synonymous variant calls from the list above. The Variant_Classification terms for synonymous SNV are selected as per [interaction-plots/scripts/02-process_mutations.R](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/master/analyses/interaction-plots/scripts/02-process_mutations.R).
```{r}
# Variant Classification with Low/Modifier variant consequences 
#  from maftools http://asia.ensembl.org/Help/Glossary?id=535
synonymous <- c(
  "Silent",
  "Start_Codon_Ins",
  "Start_Codon_SNP",
  "Stop_Codon_Del",
  "De_novo_Start_InFrame",
  "De_novo_Start_OutOfFrame"
)

consensusMutationSubset <- consensusMutationSubset %>%
  dplyr::filter(!Variant_Classification %in% synonymous)
```

### Review predicted impact of SNVs

```{r}
consensusMutationSubset %>% 
  dplyr::select("Tumor_Sample_Barcode", "IMPACT","SIFT","PolyPhen","Hugo_Symbol","Variant_Classification")

```
 We have some Modifier mutations in the filtered snv subset.
 
 We are removing the LOW/MODIFIER IMPACT annotated by [VEP](https://docs.gdc.cancer.gov/Data/File_Formats/MAF_Format/#impact-categories)
 
 - LOW (L): Assumed to be mostly harmless or unlikely to change protein behavior
 - MODIFIER (MO): Usually non-coding variants or variants affecting non-coding genes, where predictions are difficult or there is no evidence of impact

```{r}
consensusMutationSubset <- consensusMutationSubset %>%
  dplyr::filter(!IMPACT %in% c("MODIFIER","LOW"))
```
 

Let's check whether all LGAT DNA biospecimens have a subtype defining SNV.
 
```{r}
all(lgat_dna_df$Kids_First_Biospecimen_ID %in% consensusMutationSubset$Tumor_Sample_Barcode)
```
 
### Gather status of all SNVs per biospecimen
To gather the biospecimens, we will use all LGAT DNA biospecimen IDs and assign "Yes" or "No" to each subtype column depending on the presence or absence of SNV that define the subtype.

columnname  | description 
 --- | --- 
NF1_mut | somatic loss of NF1 via either missense, nonsense mutation 
BRAF_V600E_mut | contains BRAF V600E or V599 SNV or non-canonical BRAF alterations such as p.V600ins or p.D594N 
MAPK_mut | contains mutation in KRAS, NRAS, HRAS, MAP2K1, MAP2K2, MAP2K1, ARAF SNV or indel 
RTK_mut | harbors a MET, KIT, or PDGFRA SNV 
FGFR_mut | harbors FGFR1 p.N546K, p.K656E, p.N577, or p. K687 hotspot mutations 
IDH_mut | harbors an IDH R132 mutation 
H3.1_mut | harbors an HIST1H3B K28M or HIST1H3C  K28M
H3.2_mut | harbors an HIST2H3C K28M
H3.3_mut | harbors an H3F3A K28M or G35R/V mutation 

```{r}
consensusMutationSubset <- lgat_dna_df %>%
  left_join(consensusMutationSubset,by=c("Kids_First_Biospecimen_ID"="Tumor_Sample_Barcode")) %>%
  reshape2::dcast(Kids_First_Biospecimen_ID + sample_id~ Hugo_Symbol) %>%
  dplyr::mutate(BRAF_V600E_mut = if_else(BRAF>=1,"Yes","No"),
                FGFR_mut = if_else(rowSums(
                  dplyr::select(.,snvOI$FGFR$gene)) >=1,"Yes","No"),
                IDH_mut  = if_else(any(colnames(.) %in% snvOI$IDH$gene) & 
                                     rowSums(dplyr::select(.,one_of(snvOI$IDH$gene))) >=1,
                                   "Yes","No"),
                H3.1_mut = if_else(rowSums(
                  dplyr::select(.,one_of(c("HIST1H3B","HIST1H3C")))) >=1 ,"Yes","No"), 
                # No HIST1H3C, HIST2H3C, or H3F3A mutations found
                H3.2_mut = if_else(any(colnames(.) %in% snvOI$HIST2H3C$gene) & 
                                     rowSums(dplyr::select(.,one_of(snvOI$HIST2H3C$gene))) >=1,
                                   "Yes","No"), 
                H3.3_mut = if_else(any(colnames(.) %in% c("H3F3A","H3F3B")) & 
                                     rowSums(dplyr::select(.,one_of(c("H3F3A","H3F3B")))) >= 1 ,"Yes","No"),
                MAPK_mut  = if_else(
                  rowSums(dplyr::select(.,one_of(snvOI$MAPK$gene)))>=1,"Yes","No"),
                RTK_mut  = if_else(
                  rowSums(dplyr::select(.,one_of(snvOI$RTK$gene)))>=1,"Yes","No"),
                NF1_mut = if_else(
                  rowSums(dplyr::select(.,one_of(snvOI$NF1$gene)))>=1,"Yes","No") ) %>%
  dplyr::select("Kids_First_Biospecimen_ID","sample_id",ends_with("mut")) %>%
  dplyr::arrange(Kids_First_Biospecimen_ID, sample_id)
consensusMutationSubset 
```

```{r}
# remove consensusMutation
rm(consensusMutation)

# save to subset folder
write_tsv(consensusMutationSubset,file.path(subset_dir, "LGAT_snv_subset.tsv"))
```


```{r}
sessionInfo()
```