---
title: "Gather CNV changes to subtype LGAT biospecimens"
author: "K S Gaonkar (D3B)"
output: html_notebook

---

In this notebook, we will look for the following CNV changes that define subtypes of LGAT
 - LGG, FGFR
 harbors FGFR1 TKD (tyrosine kinase domain tandem duplication)
 
 - LGG, CDKN2A/B
 harbors focal CDKN2A and/or CDKN2B deletion

### Setup
```{r}
library("tidyverse")
```

### Input
```{r}
# Look for git root folder
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

# get subset folder
subset_dir <- file.path(root_dir, 
                        "analyses",
                        "molecular-subtyping-LGAT",
                        "lgat-subset")

# create if doesn't exist
if (!dir.exists(subset_dir)) {
  dir.create(subset_dir)
}

# manta calls
manta <- read_tsv(file.path(root_dir,
                            "data",
                            "pbta-sv-manta.tsv.gz")) 

# consensus seg calls
cnv <- read_tsv(file.path(root_dir,
                          "data",
                          "pbta-cnv-consensus.seg.gz"))

# consensus seg annotated with gene
cnv_anno <- read_tsv(file.path(root_dir,
                          "data",
                          "consensus_seg_annotated_cn_autosomes.tsv.gz"))

# Gene location and domain overlap file 
bioMartDataPfamFGFR1 <- 
  readRDS(system.file("extdata", "pfamDataBioMart.RDS", package = "annoFuse")) %>%
  # to check if biospecimen harbors FGFR1 TKD (tyrosine kinase domain tandem duplication) 
  dplyr::filter(hgnc_symbol == "FGFR1",
                DESC == "Protein tyrosine kinase")

```

### Gather lgat subset

```{r}
# File from 00-LGAT-select-pathology-dx that is used for the pathology diagnosis
# was used to subset clinical file
lgat_specimens_df<-read_tsv(file.path(subset_dir, "lgat_metadata.tsv"))

# Filter to RNA-Seq samples
lgat_wgs_df <- lgat_specimens_df %>%
  dplyr::filter(experimental_strategy == "WGS") %>%
  dplyr::select(Kids_First_Biospecimen_ID)

# Filter CNV calls
manta_lgat <- manta %>%
  dplyr::filter(Kids.First.Biospecimen.ID.Tumor %in% lgat_wgs_df$Kids_First_Biospecimen_ID)

cnv_lgat <- cnv %>%
  dplyr::filter( ID %in% lgat_wgs_df$Kids_First_Biospecimen_ID)

```

### Overlap tandem duplication in FGFR1 with exons 10-18 which make up the kinase domain, per [comment](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/790#issuecomment-729229244).

First, create genomic ranges objects for CNV seg file and gene.

```{r}
# Genomic range for cnv seg file 
cnv_gr <- cnv %>%
  # format
  dplyr::rename(chr = chrom, start = loc.start, end = loc.end,
                copy_number = copy.num) %>%
  dplyr::select(-num.mark, -seg.mean) %>%
  # make genomic ranges
  GenomicRanges::makeGRangesFromDataFrame(keep.extra.columns = TRUE,
                                          starts.in.df.are.0based = FALSE)

# Genomic range for gene location and domain overlap file 
domain_gr <- bioMartDataPfamFGFR1 %>%
  # keep only domain info
  dplyr::filter(!is.na(domain_start),!is.na(domain_end)) %>%
  # format
  dplyr::mutate(strand = if_else(strand=="-1","-","+"),
                domain_chr = paste0("chr",domain_chr)) %>%
  dplyr::rename(chr = domain_chr, start = domain_start, end = domain_end) %>%
  # make genomic ranges
  GenomicRanges::makeGRangesFromDataFrame(keep.extra.columns = TRUE,
                                          starts.in.df.are.0based = FALSE,seqnames.field = "chr")

```



### Distribution of manta SV
SV should be annotated as tandem duplicate and overlap 
"Protein tyrosine kinase" domain of FGFR1 in biospecimens

```{r}

# Genomic range for manta file 
manta_gr <- manta_lgat %>%
  # format
  dplyr::rename(chr = SV.chrom, start = SV.start, end = SV.end,
                ALT = ALT) %>%
  # format with chr to match to cnv
  dplyr::mutate(chr=paste0("chr",chr)) %>%
  # make genomic ranges
  GenomicRanges::makeGRangesFromDataFrame(keep.extra.columns = TRUE,
                                          starts.in.df.are.0based = FALSE)

# overlap manta and domain
overlaps_manta <- IRanges::mergeByOverlaps(manta_gr, domain_gr)

# filter for fgfr only overlap
overlaps_manta_fgfr <- overlaps_manta[which(overlaps_manta$Gene.name=="FGFR1" &
                # filter for tandem duplication
                overlaps_manta$ALT=="<DUP:TANDEM>"), ]

overlaps_manta_fgfr_df <- data.frame(
  biospecimen_id = overlaps_manta_fgfr$Kids.First.Biospecimen.ID.Tumor,
  ALT = overlaps_manta_fgfr$ALT,
  NAME = overlaps_manta_fgfr$NAME,
  hgnc_symbol = overlaps_manta_fgfr$hgnc_symbol,
  # annnotation for dup tandem
  FGFR_DUP_TANDEM = "Yes",
  stringsAsFactors = FALSE
) %>%
  dplyr::distinct() 

overlaps_manta_fgfr_df 

```
4 biospecimens have tandem duplication that overlap tyrosine kinase domain in FGFR1


### Distribution of copy number
CNV that are within FGFR1 gene region only and overlap "Protein tyrosine kinase" domain in FGFR1 in biospecimens

```{r}

# gene region overlap
fgfr_gene = GenomicRanges::GRanges(seqnames=paste0("chr",domain_gr$chromosome_name),
          ranges=IRanges::IRanges(start = domain_gr$gene_start, end = domain_gr$gene_end))

# subset cnv+domain grange to be within FGFR1 gene region
cnv_fgfr_gr <- IRanges::subsetByOverlaps(cnv_gr, fgfr_gene,type="within")

# overlap consensus seg with exon
overlaps_fgfr_cnv <- IRanges::mergeByOverlaps(cnv_fgfr_gr,domain_gr) 

# filter for fgfr only overlap
overlaps_fgfr_cnv <- overlaps_fgfr_cnv[which(overlaps_fgfr_cnv$hgnc_symbol=="FGFR1" &
               # filter for duplication
               overlaps_fgfr_cnv$copy_number>2), ]

# dataframe
overlaps_cnv_df <- data.frame(
  biospecimen_id = overlaps_fgfr_cnv$ID,
  copy_number = overlaps_fgfr_cnv$copy_number,
  NAME = overlaps_fgfr_cnv$NAME,
  stringsAsFactors = FALSE
) %>%
  dplyr::distinct() %>%
  # keep copy_number > 2
  dplyr::mutate(FGFR_DUP = if_else(copy_number >2,"Yes","No")) 

overlaps_cnv_df 

```
5 samples have copy_number changes > 2 that overlap with tyrosine kinase domain

Are there overlap in SV and CNV calls?

```{r}
intersect(overlaps_manta_fgfr_df$biospecimen_id,overlaps_cnv_df$biospecimen_id)
```
Only BS_MW6YMRBJ shows tandem duplication in manta and duplication in consensus seg file that overlaps tyrosine kinase domain in FGFR1

### Checking which exons overlap the above SV and CNV

```{r}
# Define the annotations for the hg38 genome
txdb <- GenomicFeatures::makeTxDbFromGFF(
    file = file.path(root_dir,"data",
                     "gencode.v27.primary_assembly.annotation.gtf.gz"),
    format = "gtf"
  )
# get exons and exon_id
tx_exons<-GenomicFeatures::exons(txdb, 
                                 columns = list("exon_rank"),
                                 # genocode canonical transcript https://genome.ucsc.edu/cgi-bin/hgGene?db=hg38&hgg_gene=FGFR1
                                 filter=list(tx_name = "ENST00000447712.6"))
  
```

For manta SV calls that are within FGFR1 gene region

```{r}
# overlap filtered manta with exons
overlaps_manta_fgfr_exon <- IRanges::mergeByOverlaps(overlaps_manta_fgfr$manta_gr,tx_exons) 

# overlap filtered cnv with exons
overlaps_cnv_fgfr_exon <- IRanges::mergeByOverlaps(overlaps_fgfr_cnv$cnv_fgfr_gr,tx_exons) 

# plot exon rank for manta and cnv
ggplot(rbind(data.frame("exon_number"=unlist(overlaps_manta_fgfr_exon$exon_rank),
                  "type"="tandem_dup"),
             data.frame("exon_number"=unlist(overlaps_cnv_fgfr_exon$exon_rank),
                  "type"="cnv_dup"))
       ,aes(x=exon_number,color=type))+geom_density()

```
Majority SV and CNV overlap exons downstream of exon 5

any SV overlap which overlaps exons 10 through 18 
```{r}
overlaps_manta_fgfr_exon[all(overlaps_manta_fgfr_exon$exon_rank>10  &
                        overlaps_manta_fgfr_exon$exon_rank<18)==TRUE,"Kids.First.Biospecimen.ID.Tumor"] %>% unique()

```

any CNV overlap which overlaps exons 10 through 18 
```{r}

overlaps_cnv_fgfr_exon[all(overlaps_cnv_fgfr_exon$exon_rank>10 & overlaps_cnv_fgfr_exon$exon_rank<18)==TRUE,"ID"] %>% unique()
```

All 4 biospecimens with tandem duplication in FGFR1 and 5 biospecimens with duplication in FGFR1 have atleast 1 SV/CNV overlap where all exons are between 10 and 18. 


### CDKN deletion 

```{r}

cnv_anno_cdkn2 <- cnv_anno %>%
  dplyr::filter(gene_symbol %in% c("CDKN2A","CDKN2B")) %>%
  reshape::cast(biospecimen_id+ploidy~gene_symbol,value = "copy_number") %>%
  # keep only if copy_number <=1
  dplyr::mutate(CDKN2A_DEL= if_else(CDKN2A<=1,"Yes","No"),
                CDKN2B_DEL= if_else(CDKN2B<=1,"Yes","No")) %>%
  dplyr::distinct() 

cnv_anno_cdkn2
  

```

### Gather CNV subsets

```{r}

# merge CNV subtype annotation
merge_CNV_substype <-dplyr::select(overlaps_manta_fgfr_df,biospecimen_id,FGFR_DUP_TANDEM) %>%
  full_join(dplyr::select(overlaps_cnv_df ,biospecimen_id,FGFR_DUP)) %>%
  full_join(dplyr::select(cnv_anno_cdkn2, biospecimen_id,CDKN2A_DEL,CDKN2B_DEL)) 

# merge all lgat WGS samples
lgat_wgs_df %>%
  left_join(merge_CNV_substype,by=c("Kids_First_Biospecimen_ID"="biospecimen_id")) %>%
  # change NA to value==No
  replace_na(list(FGFR_DUP_TANDEM="No",
                         FGFR_DUP="No",
                         CDKN2A_DEL="No",
                         CDKN2B_DEL ="No"
                         ))%>%
  write_tsv(file.path(subset_dir,"LGAT_cnv_subset.tsv"))


```
