---
title: "R Notebook"
output: html_notebook
---

In this notebook, we will be compiling the subtype annotation for LGAT samples. Subtypes are described in [#790](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/790) which were gathered through the following scripts : 

`01-subset-files-for-LGAT.Rmd` gathers subtype associated SNV annotation 
`02-subset-fusion-files-LGAT.Rmd` gathers subttype associated Fusion annotation
`03-subset-cnv-files-LGAT.Rmd` gathers subtype associated CNV annotation

along with the above molecular data based annotation in this notebook we will include NF-1 cancer_predispositions in histology file as a criteria for LGG, NF1

### Setup
```{r}
library("tidyverse")
```

### Input
```{r}
# Look for git root folder
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

# get subset folder
subset_dir <- file.path(
  root_dir,
  "analyses",
  "molecular-subtyping-LGAT",
  "lgat-subset"
)

# clinical file
clinical <- read_tsv(file.path(root_dir, 
                               "data",
                               "pbta-histologies-base.tsv"), 
                     guess_max = 10000)
```



