---
title: "Compare Variant Callers"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: C. Savonen for ALSF CCDL
date: 2019
---

Purpose: After running the initial analyses of each variant caller, this notebook 
combines the output from those calculations and compares them. 

#### Usage

To run this from the command line, use:
```
Rscript -e "rmarkdown::render('analyses/tmb-calculations/compare_variant_callers.Rmd', 
                              clean = TRUE)"
```

_This assumes you are in the top directory of the repository._

## Setup

#### Packages and functions

Read in set up script.

```{r}
if (!("ggupset" %in% installed.packages())) {
  install.packages("ggupset", repos='http://cran.us.r-project.org')
}
if (!("UpSetR" %in% installed.packages())) {
  install.packages("UpSetR", repos='http://cran.us.r-project.org')
}
if (!("GGally" %in% installed.packages())) {
  install.packages("GGally", repos='http://cran.us.r-project.org')
}
# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

Set up output directories. 

```{r}
base_results_dir <- "results"
base_plots_dir <- "plots"
```

Make new directories for the comparison analysis.

```{r}
compare_results_dir <- file.path(base_results_dir, "comparisons")
compare_plots_dir <- file.path(base_plots_dir, "comparisons")

# Make caller specific plots folder
if (!dir.exists(compare_results_dir)) {
  dir.create(compare_results_dir)
}
# Make caller specific plots folder
if (!dir.exists(compare_plots_dir)) {
  dir.create(compare_plots_dir)
}
```

## Import data 

```{r}
dir(base_results_dir)
```

Get the lists of each type of file. 

```{r}
# Get the lists of each type of file
vaf_files <- list.files(base_results_dir, pattern = "_vaf.tsv$", recursive = TRUE, 
                        full.names = TRUE) 
tmb_files <- list.files(base_results_dir, pattern = "_tmb.tsv$", recursive = TRUE, 
                        full.names = TRUE) 
region_annot_files <- list.files(base_results_dir, pattern = "_region.tsv$", recursive = TRUE, 
                        full.names = TRUE) 
```


```{r}
# Get the caller names
caller_names <- stringr::word(vaf_files, sep = "/", 2)
```

Make a big VAF data.frame

```{r}
vaf_list <- lapply(vaf_files, function(file) {
  readr::read_tsv(file) %>%
    dplyr::select("mutation_id", "vaf", "base_change", "change")
})
names(vaf_list) <- caller_names

vaf_df <- dplyr::bind_rows(vaf_list, .id = "caller") %>% 
  dplyr::mutate(caller = factor(caller))
```

Take a look the density plot of VAFs for each caller. 

```{r}
ggplot2::ggplot(vaf_df, ggplot2::aes(x = vaf, fill = caller, color = caller)) +
  ggplot2::geom_density(alpha = 0.15) + 
  ggplot2::theme_classic() + 
  ggplot2::xlab("VAF") + 
  ggplot2::ylab("Density") +
  colorblindr::scale_fill_OkabeIto() +
  colorblindr::scale_color_OkabeIto()
```

Make a big tmb data.frame

```{r}
tmb_list <- lapply(tmb_files, readr::read_tsv)
names(tmb_list) <- caller_names
tmb_df <- dplyr::bind_rows(tmb_list, .id = "caller") %>% 
  dplyr::mutate(caller = factor(caller))
```

Make a density plot of the TMB data. 

```{r}
ggplot2::ggplot(tmb_df, ggplot2::aes(x = tmb, fill = caller, color = caller)) +
  ggplot2::geom_density(alpha = 0.15) + 
  ggplot2::theme_classic() + 
  ggplot2::xlab("TMB") + 
  ggplot2::ylab("Density") +
    colorblindr::scale_fill_OkabeIto() +
  colorblindr::scale_color_OkabeIto()
```

Make a big genomic region data.frame

```{r}
region_annot <- lapply(region_annot_files,function(file) {
  readr::read_tsv(file) %>%
    dplyr::select("vaf", "base_change", "mutation_id", "change", "type")
})
names(region_annot) <- caller_names
region_annot_df <- dplyr::bind_rows(region_annot, .id = "caller") %>% 
  dplyr::mutate(caller = factor(caller))
```

## Upsett graph

```{r}
mutation_id_list <- lapply(vaf_list, function(caller) caller$mutation_id)

UpSetR::upset(UpSetR::fromList(mutation_id_list), order.by = "freq")
```

Graph mutations by combinations of callers and plot the VAF density.

```{r}
callers_per_mutation <- vaf_df %>% 
  dplyr::select(mutation_id, caller) %>% 
  reshape2::melt() %>% 
  dplyr::group_by(mutation_id) %>% 
  dplyr::summarize(callers = paste0(caller, collapse = "-"),
                   med_vaf = median(vaf))

callers_per_mutation %>%
  ggplot2::ggplot(ggplot2::aes(x = callers, y = med_vaf)) +
  ggplot2::geom_violin() +
  ggupset::scale_x_mergelist(sep = "-") +
  ggupset::axis_combmatrix(sep = "-")
```

## Base Changes 

```{r}
region_annot_df %>% 
  dplyr::group_by(change, caller) %>% 
  dplyr::summarize(count = dplyr::n()) %>%
  ggplot2::ggplot(ggplot2::aes(x = change, y = count)) + 
  ggplot2::theme_classic() + 
  ggplot2::geom_bar(position = "dodge", stat = "identity",
        ggplot2::aes(fill = caller)) + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
  ggplot2::xlab("") +
  ggplot2::ylab("Count") +
  colorblindr::scale_fill_OkabeIto()
```

## VAF distributions

```{r}
vaf_df %>% 
  ggplot2::ggplot(ggplot2::aes(x = caller, y = vaf, color = caller)) + 
  ggforce::geom_sina(alpha = 0.05) + 
  ggplot2::theme_classic() + 
  colorblindr::scale_color_OkabeIto()
```

## TMB distributions

```{r}
tmb_df %>% 
  ggplot2::ggplot(ggplot2::aes(x = caller, y = tmb, color = short_histology)) + 
  ggforce::geom_sina() + 
  ggplot2::theme_classic() 
```

## VAF correlations

```{r}
vaf_long <- vaf_df %>% 
  dplyr::group_by(mutation_id, caller) %>% 
  dplyr::summarize(vaf = mean(vaf)) %>% 
  dplyr::distinct(mutation_id, caller, vaf, .keep_all = TRUE) %>% 
  tidyr::spread(caller, vaf) %>% 
  tibble::column_to_rownames("mutation_id")
```

Make the plot with ggpairs. 

```{r}
GGally::ggpairs(vaf_long, mapping = ggplot2::aes(alpha = 0.05)) +
  ggplot2::theme_classic()
``` 

## TMB correlations

```{r}
tmb_long <- tmb_df %>% 
  dplyr::distinct(Tumor_Sample_Barcode, caller, tmb) %>% 
  tidyr::spread(caller, tmb) %>% 
  tibble::column_to_rownames("Tumor_Sample_Barcode")
```

Make the TMB correlation plot.

```{r}
GGally::ggpairs(tmb_long, mapping = ggplot2::aes(alpha = 0.05)) +
  ggplot2::theme_classic()
```

## Genomic regions 

```{r}
region_annot_df %>% 
  dplyr::group_by(type, caller) %>% 
  dplyr::summarize(count = dplyr::n()) %>%
  ggplot2::ggplot(ggplot2::aes(x = type, y = count)) + 
  ggplot2::theme_classic() + 
  ggplot2::geom_bar(position = "dodge", stat = "identity",
        ggplot2::aes(fill = caller)) + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
  ggplot2::xlab("") +
  ggplot2::ylab("Count") +
  colorblindr::scale_fill_OkabeIto()
```


```{r}
region_annot_df %>% 
  ggplot2::ggplot(ggplot2::aes(x = caller, y = vaf)) + 
  ggplot2::theme_classic() + 
  ggforce::geom_sina(alpha = 0.1, ggplot2::aes(color = caller)) + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
  ggplot2::xlab("") +
  ggplot2::ylab("VAF") +
  colorblindr::scale_color_OkabeIto() + 
  ggplot2::facet_wrap(~type)
```

## Session Info

```{r}
sessionInfo()
```
