---
title: "High-grade Glioma Molecular Subtyping - Focal and Broad Copy Number Alterations"
author: "Chante Bethell, Stephanie J. Spielman, and Jaclyn Taroni for ALSF CCDL"
date: "2020"
output:
  html_document:
    df_print: paged
    toc: yes
  html_notebook:
    toc: yes
    toc_float: yes
---

This notebook prepares focal and broad copy number alteration data for the 
purposes of use for subtyping HGG samples 
([`AlexsLemonade/OpenPBTA-analysis#249`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/249)).

#### TODO: This analysis should be revisited when consensus copy data is made available.

## Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

`Rscript -e "rmarkdown::render('analyses/molecular-subtyping-HGG/03-HGG-molecular-subtyping-cnv.Rmd', clean = TRUE)"`

## Set Up

```{r}
library(tidyverse)
```

### Directories

```{r}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
# File path to results directory
input_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-HGG", "hgg-subset")
# File path to results directory
results_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-HGG", "results")

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

### Read in Files

```{r}
# Read in HGG subset focal CN data
cn_df <- read_tsv(file.path(input_dir, "hgg_focal_cn.tsv.gz"))
# Read in relevant broad chromosome arm loss for HGG samples
gistic_df <- read_tsv(file.path(input_dir, "hgg_gistic_broad_values.tsv"))
```

### Output File

```{r}
output_file <- file.path(results_dir, "HGG_cleaned_cnv.tsv")
```

## CNVkit Data

We are using CNVkit data that has been annotated via the [`focal-cn-file-preparation`](../focal-cn-file-preparation/) module.
Note that a gene may have evidence for two kinds of copy number alteration (e.g., both _loss_ and _gain_) in these data.
This may or may not be a caller-specific artifact that will be alleviated by generating consensus copy number calls 
([`AlexsLemonade/OpenPBTA-analysis#128`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/128)).
See also: https://jaclyn-taroni.github.io/openpbta-notebook-concept/both-gain-and-loss.nb.html.



```{r}
# Select only the ID information, for use in joining below
cn_df_ids <- cn_df %>%  
  select(sample_id, 
         Kids_First_Biospecimen_ID = biospecimen_id,
         Kids_First_Participant_ID)

# Filter focal CN data for PDGFRA, PTEN and MYCN status
cn_df_genes <- cn_df %>%
  rename(Kids_First_Biospecimen_ID = biospecimen_id) %>%
  select(-cytoband) %>%
  filter(gene_symbol %in% c("PDGFRA", "PTEN", "MYCN"))

# Summarize the data to have a single row per sample where _all_ kinds of copy number alterations are present.
cn_df_genes_status <- cn_df_genes %>% 
  group_by(gene_symbol, sample_id, Kids_First_Biospecimen_ID) %>% 
  summarize(full_status = paste(status, collapse=", ")) %>% 
  spread(gene_symbol, full_status) %>% 
  right_join(cn_df_ids) %>%
  distinct() %>%
  replace_na(list(PDGFRA = "neutral", 
                  PTEN = "neutral", 
                  MYCN = "neutral")) %>% 
  # Rename columns to indicate focal status explicitly
  rename(MYCN_focal_status   = MYCN, 
         PDGFRA_focal_status = PDGFRA,
         PTEN_focal_status   = PTEN) %>%
  # Arrange order of columns
  dplyr::select(sample_id, 
                Kids_First_Biospecimen_ID, 
                Kids_First_Participant_ID, 
                everything())

head(cn_df_genes_status, n = 10)
```


## GISTIC Data

The GISTIC `broad_values_by_arm.txt` file is a matrix with integers that indicate chromosome arm status. 
Much like the annotated CNVkit data above, we'll convert this to `"loss"`, `"neutral"`, and `"gain"`.

```{r}
head(gistic_df, n = 10) 
```

```{r}
gistic_status_df <- gistic_df %>%
  # for the numeric columns (e.g., not the identifiers!), go from the integers
  # to loss, gain, neutral
  mutate_if(is.numeric, 
            list(~ case_when(
              . < 0 ~ "loss",
              . > 0 ~ "gain",
              . == 0 ~ "neutral"
            ))) %>%
  # drop sample_id
  select(-sample_id)

head(gistic_status_df, n = 10)
```

## Join Together

```{r}
# There are some samples that had too many segments and were flagged by GISTIC
# These values will be filled with NA -- we'll change them to indicate that 
# they failed GISTIC QC
final_hgg_cn_df <- left_join(cn_df_genes_status,
                             gistic_status_df,
                             by = "Kids_First_Biospecimen_ID") %>%
  replace(is.na(.), "Failed GISTIC QC")
head(final_hgg_cn_df, n = 10)
```

Write this to file.

```{r}
write_tsv(final_hgg_cn_df, output_file)
```

## Session Info

```{r}
sessionInfo()
```

