---
title: "High-grade Glioma Molecular Subtyping - Focal and Broad Copy Number Alterations"
author: "Chante Bethell, Stephanie J. Spielman, and Jaclyn Taroni for ALSF CCDL"
date: "2020"
output:
  html_notebook:
    toc: yes
    toc_float: yes
---

This notebook prepares focal and broad copy number alteration data for the 
purpose of subtyping HGG samples 
([`AlexsLemonade/OpenPBTA-analysis#249`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/249)).

## Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

`Rscript -e "rmarkdown::render('analyses/molecular-subtyping-HGG/03-HGG-molecular-subtyping-cnv.Rmd', clean = TRUE)"`

## Set Up

```{r}
library(tidyverse)
```

### Directories

```{r}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
# File path to results directory
input_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-HGG", "hgg-subset")
# File path to results directory
results_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-HGG", "results")

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

### Read in Files

We will need the clinical/histologies file for included samples to distinguish between missing samples or samples that had only neutral copy number calls.

```{r message=FALSE}
hgg_metadata_df <- read_tsv(file.path(input_dir,
                                     "hgg_metadata.tsv"), 
                           guess_max = 10000)
```

#### Inclusion Criteria

Here, we'll use the metadata output from  `02-HGG-molecular-subtyping-subset-files.R` to identify the samples we will include.
Only WGS biospecimens have CNV and SV data, so we'll filter based on `experimental_strategy`.

```{r}
included_wgs_specimens <- hgg_metadata_df %>%
  dplyr::filter(experimental_strategy == "WGS") %>%
  dplyr::pull(Kids_First_Biospecimen_ID)
```

Are all the WGS samples that pass the inclusion criteria in the consensus SEG file committed to the repository?

```{r}
consensus_seg_ids <- read_tsv(file.path(root_dir, 
                                        "data",
                                        "pbta-cnv-consensus.seg.gz")) %>%
  pull(ID)
consensus_seg_ids <- unique(consensus_seg_ids)
```

```{r}
all(included_wgs_specimens %in% consensus_seg_ids)
```

No, they're not.
How many are missing?

```{r}
sum(!(included_wgs_specimens %in% consensus_seg_ids))
```

We'd then expect `r sum(included_wgs_specimens %in% consensus_seg_ids)` samples to have non-NA values in the cleaned CNV files.

#### Read in CN Data

```{r}
# Read in HGG subset focal CN data
cn_df <- read_tsv(file.path(input_dir, "hgg_focal_cn.tsv.gz"))
# Read in relevant broad chromosome arm loss for HGG samples
gistic_df <- read_tsv(file.path(input_dir, "hgg_gistic_broad_values.tsv"))
```

### Output File

```{r}
output_file <- file.path(results_dir, "HGG_cleaned_cnv.tsv")
```

## Annotated consensus CN data

We are using consensus copy number data data that has been annotated via the [`focal-cn-file-preparation`](../focal-cn-file-preparation/) module.
Note that a gene may have evidence for two kinds of copy number alteration (e.g., both _loss_ and _gain_) in these data, but no longer see instances of this when using the consensus data for these samples.

```{r}
# Select only the ID information, for use in joining below
cn_df_ids <- cn_df %>%  
  select(sample_id,
         Kids_First_Biospecimen_ID = biospecimen_id,
         Kids_First_Participant_ID)

# Filter focal CN data for PDGFRA, PTEN and MYCN status
cn_df_genes <- cn_df %>%
  rename(Kids_First_Biospecimen_ID = biospecimen_id) %>%
  select(-cytoband) %>%
  filter(gene_symbol %in% c("PDGFRA", "PTEN", "MYCN"))

# Summarize the data to have a single row per sample where _all_ kinds of copy number alterations are present.
cn_df_genes_status <- cn_df_genes %>% 
  group_by(gene_symbol, sample_id, Kids_First_Biospecimen_ID) %>% 
  summarize(full_status = paste(status, collapse=", ")) %>% 
  spread(gene_symbol, full_status) %>% 
  right_join(cn_df_ids) %>%
  distinct() %>%
  replace_na(list(PDGFRA = "neutral", 
                  PTEN = "neutral", 
                  MYCN = "neutral")) %>% 
  # Rename columns to indicate focal status explicitly
  rename(MYCN_focal_status   = MYCN, 
         PDGFRA_focal_status = PDGFRA,
         PTEN_focal_status   = PTEN) %>%
  # Arrange order of columns
  dplyr::select(sample_id,
                Kids_First_Biospecimen_ID, 
                Kids_First_Participant_ID, 
                everything())

head(cn_df_genes_status, n = 10)
```

## GISTIC Data

The GISTIC `broad_values_by_arm.txt` file is a matrix with integers that indicate chromosome arm status. 
Much like the annotated CNVkit data above, we'll convert this to `"loss"`, `"neutral"`, and `"gain"`.

```{r}
head(gistic_df, n = 10) 
```

```{r}
gistic_status_df <- gistic_df %>%
  # for the numeric columns (e.g., not the identifiers!), go from the integers
  # to loss, gain, neutral
  mutate_if(is.numeric, 
            list(~ case_when(
              . < 0 ~ "loss",
              . > 0 ~ "gain",
              . == 0 ~ "neutral"
            )))

head(gistic_status_df, n = 10)
```

Add in participant IDs.

```{r}
gistic_status_df <- hgg_metadata_df %>% 
  select(Kids_First_Biospecimen_ID,
         Kids_First_Participant_ID,
         sample_id) %>%
  right_join(gistic_status_df)
```

## Join Together

```{r}
final_hgg_cn_df <- full_join(cn_df_genes_status,
                             gistic_status_df)

nrow(final_hgg_cn_df)
```

`final_hgg_cn_df` includes all the samples that are in the consensus SEG file.
So any missing CNA values at this stage come from the fact that neutral values are _not_ included in the annotated copy number files.
We can fill any "missing values" with neutral calls as a result.

```{r}
final_hgg_cn_df <- final_hgg_cn_df %>%
  replace(is.na(.), "neutral") %>%
  dplyr::arrange(Kids_First_Participant_ID, sample_id)
```

Write this to file.

```{r}
write_tsv(final_hgg_cn_df, output_file)
```

## Session Info

```{r}
sessionInfo()
```

