---
title: "High-Grade Glioma Molecular Subtyping - Mutations"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell and Jaclyn Taroni for ALSF CCDL
date: 2020
---

This notebook prepares consensus mutation data for the purpose of subtyping HGG samples 
([`AlexsLemonade/OpenPBTA-analysis#249`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/249)).

## Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

```
Rscript -e "rmarkdown::render('analyses/molecular-subtyping-HGG/04-HGG-molecular-subtyping-mutation.Rmd', clean = TRUE)"
```

## Set Up

### Libraries and Functions

```{r}
`%>%` <- dplyr::`%>%`
```

We'll take a similar approach to the [`tp53_nf1_score`](../tp53_nf1_score/) module and use functions from the [`snv-callers`](../snv-callers/) module.

```{r}
# TODO: if a common util folder is established, use that instead
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
source(file.path(root_dir, "analyses", "snv-callers", "util",
                 "tmb_functions.R"))
```

### Directories

```{r}
# File path to results directory
input_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-HGG", "hgg-subset")
# File path to results directory
results_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-HGG", "results")
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

### Read in Files

```{r message=FALSE}
# CDS bed file, to be used to filter the consensus MAF file
bed_file <- file.path(root_dir, 
                      "scratch", 
                      "gencode.v27.primary_assembly.annotation.bed")
gencode_cds_bed <- readr::read_tsv(bed_file, col_names = FALSE)
# Read in HGG subset snv maf data 
snv_df <- readr::read_tsv(file.path(input_dir, "hgg_snv_maf.tsv.gz"))
# Read in defining lesions file
defining_lesions_df <- readr::read_tsv(file.path(results_dir,
                                                 "HGG_defining_lesions.tsv"))
```

## Output File

```{r}
output_file <- file.path(results_dir, "HGG_cleaned_mutation.tsv")
```

## What mutations are we interested in?

We'll summarize the contents of ([`AlexsLemonade/OpenPBTA-analysis#249`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/249)).

### Defining Lesions

From [`@jharenza`](https://github.com/jharenza):

> Any sample that harbors _H3F3A_ K28M, _H3F3A_ G35R/V, or _HIST1H3B_ K28M, if not classified as a high-grade glioma, should be, as this is a defining lesion. Eg: three former PNET tumors were re-classified as such.

Samples were examined for the presence or absence of these defining lesions in [`01-HGG-molecular-subtyping-defining-lesions`](./01-HGG-molecular-subtyping-defining-lesions.Rmd).
Samples that were classified as HGAT in the `short_histology` column of `pbta-histologies-base.tsv` and those with one of these defining lesions were included in the subset files (see: [`02-HGG-molecular-subtyping-subset-files.R`](./02-HGG-molecular-subtyping-subset-files.R)).
Thus, we are not concerned with cleaning data for these specific mutations here.
Instead, we can look at the defining lesions `data.frame` we read in above.

```{r}
head(defining_lesions_df, n = 10)
```

### H3 K28 mutant

Co-occurring lesions include _ACRV1_, _TP53_, _ATRX_ mutations.
Mutually exclusive lesions include _FGFR1_ and _IDH1_ mutations and _BRAF_ V600E (specific for low-grade gliomas).

```{r}
k28_genes_of_interest <- c("ACRV1", "TP53", "ATRX", "FGFR1", "IDH1", "BRAF")
```

### H3 G35 mutant

Co-occurring lesions include _ATRX/DAXX_, _TP53_, _SETD2_ mutations.
Mutually exclusive lesions include _IDH1_ mutations.

```{r}
g35_genes_of_interest <- c("ATRX", "DAXX", "TP53", "SETD2", "IDH1")
```

### IDH mutant

These tumors harbor _IDH1_ R132H mutations.
Co-occurring lesions include _TP53_ mutations.

```{r}
idh_genes_of_interest <- c("IDH1", "TP53")
```

### H3.3 and IDH wildtype

These tumors do not have _H3F3A_ or _IDH1_ mutations, but harbor _TP53_ and _TERT_ mutations.

```{r}
wildtype_genes_of_interest <- c("H3F3A", "HIST1H3B", "HIST1H3C", "HIST2H3C",
                                "IDH1", "TP53", "TERT")
```

### 1p/19q co-deleted oligodendrogliomas

> * Nearly all have a co-occurring _IDH1_ mutation (R132 or R172)
> * Other co-occurring lesions include: _TERT_ promoter, _CIC_, and _FUBP1_ mutations

```{r}
codeleted_genes_of_interest <- c("IDH1", "TERT", "CIC", "FUBP1")
```

## Clean SNV data

Get a vector of _all_ genes of interest, so we can use this to filter the HGG subset consensus MAF file.

```{r}
genes_of_interest <- unique(c(k28_genes_of_interest,
                              g35_genes_of_interest,
                              idh_genes_of_interest,
                              wildtype_genes_of_interest,
                              codeleted_genes_of_interest))
```

### Filter the SNV data

```{r}
filtered_snv_df <- snv_df %>%
  dplyr::filter(Hugo_Symbol %in% genes_of_interest)
```

#### Coding mutations in genes other than _TERT_

For all genes other than _TERT_, we'll restrict to the coding sequences and drop silent mutations.

```{r}
coding_snv_df <- snv_ranges_filter(maf_df = filtered_snv_df,
                                   keep_ranges = gencode_cds_bed) %>%
  dplyr::filter(Variant_Classification != "Silent",
                Hugo_Symbol != "TERT")

head(coding_snv_df, n = 10)
```

#### _TERT_ mutation

We likely want to be permissive with the _TERT_ mutations in terms of *regions.*

```{r}
tert_snv_df <- filtered_snv_df %>%
  dplyr::filter(Hugo_Symbol == "TERT",
                Variant_Classification == "5'Flank",
                Start_Position %in% c("1295113","1295135"),
                End_Position %in% c("1295113","1295135"))

tert_snv_df
```

#### Specific _IDH1_ mutations

_IDH1_ mutations are specific - we'll break them out here.

```{r}
idh_snv_df <- filtered_snv_df %>%
  dplyr::filter(Hugo_Symbol == "IDH1",
                grepl("R132|R172", HGVSp_Short))

head(idh_snv_df)
```

_Note that some samples that did not get included in the subset file (e.g., LGAT, Ganglioglioma, etc.) have R132H mutations.
Should they be included in the subset files?_

#### _BRAF_ V600E

```{r}
braf_snv_df <- filtered_snv_df %>%
  dplyr::filter(Hugo_Symbol == "BRAF",
                grepl("V600E", HGVSp_Short))

head(braf_snv_df, n = 10)
```

### Wrangle SNV Data

Now that we've filtered the data, we'll get it into a form that's more amenable to presentation.
First, let's include a column that contains all of the coding mutations of interest (excluding _TERT_).
This will contain duplicated information -- e.g., _H3F3A_ will be present for tumors with _H3F3A_ K28M mutations -- but serve as a "sanity check."

```{r}
coding_summary_df <- coding_snv_df %>% 
  dplyr::select(Tumor_Sample_Barcode,
                Hugo_Symbol) %>%
  # For each tumor biospecimen, get a comma separated
  dplyr::group_by(Tumor_Sample_Barcode) %>%
  dplyr::summarize(relevant_coding_mutations = paste(sort(unique(Hugo_Symbol)),
                                                     collapse = ", "))

head(coding_summary_df, n = 10)
```

Add tumors that had no mutations in the relevant genes into this `data.frame`.

```{r}
# This will be all biospecimen IDs that were included in the HGG subset
# SNV file
subset_bsid <- unique(snv_df$Tumor_Sample_Barcode)
# Add the 'missing' biospecimen IDs into the coding summary data.frame but
# mark relevant_coding_mutations as "None"
coding_summary_df <- coding_summary_df %>%
  dplyr::bind_rows(data.frame(
    Tumor_Sample_Barcode = setdiff(subset_bsid,
                                   coding_summary_df$Tumor_Sample_Barcode)
  )) %>%
  replace(is.na(.), "None")
```

We'll make a `data.frame` that indicates the `Variant_Classification` of any _TERT_ mutation in a tumor biospecimen.

```{r}
tert_summary_df <- tert_snv_df %>%
  dplyr::select(Tumor_Sample_Barcode,
                Variant_Classification) %>%
  dplyr::group_by(Tumor_Sample_Barcode) %>%
  dplyr::summarize(TERT_variant_classification = 
                     paste(sort(unique(Variant_Classification)),
                           collapse = ", "))
head(tert_summary_df)
```

Add in tumor biospecimens where no _TERT_ mutation is present.

```{r}
tert_summary_df <- tert_summary_df %>%
  dplyr::bind_rows(data.frame(
    Tumor_Sample_Barcode = setdiff(subset_bsid,
                                   tert_summary_df$Tumor_Sample_Barcode)
  )) %>%
  replace(is.na(.), "None")
```

Now a `data.frame` that contains the specific IDH1 mutations.

```{r}
idh_summary_df <- idh_snv_df %>%
  dplyr::select(Tumor_Sample_Barcode, HGVSp_Short) %>%
  dplyr::rename(IDH1_mutation = HGVSp_Short) %>%
  dplyr::bind_rows(data.frame(
    Tumor_Sample_Barcode = setdiff(subset_bsid,
                                   idh_snv_df$Tumor_Sample_Barcode)
  )) %>%
  replace(is.na(.), "No R132 or R172")
```

And a `data.frame` that contains the specific _BRAF_ mutation.

```{r}
braf_summary_df <- braf_snv_df %>%
  dplyr::select(Tumor_Sample_Barcode, HGVSp_Short) %>%
  dplyr::rename(BRAF_V600E = HGVSp_Short) %>%
  dplyr::bind_rows(data.frame(
    Tumor_Sample_Barcode = setdiff(subset_bsid,
                                   braf_snv_df$Tumor_Sample_Barcode)
  )) %>%
  replace(is.na(.), "No V600E")
```

## Final Mutations Table

We'll include specific information from the defining lesions table.

```{r}
defining_lesions_df <- defining_lesions_df %>%
  # We're not concerned with disease type classifications at the moment
  dplyr::select(-dplyr::starts_with("disease"),
                -dplyr::contains("histology")) %>%
  # Filter to tumor biospecimens that are in the subset file
  dplyr::filter(Kids_First_Biospecimen_ID %in% subset_bsid) %>%
  # All other data.frame use 'Tumor_Sample_Barcode' for the column that contains
  # the tumor biospecimen identifiers
  dplyr::rename(Tumor_Sample_Barcode = Kids_First_Biospecimen_ID)
```

Join all mutation `data.frame` together.

```{r}
final_mutations_df <- list(defining_lesions_df,
                           coding_summary_df,
                           idh_summary_df,
                           tert_summary_df,
                           braf_summary_df) %>%
  purrr::reduce(dplyr::inner_join, by = "Tumor_Sample_Barcode") %>%
  dplyr::select(-defining_lesion) %>%
  dplyr::arrange(Kids_First_Participant_ID, sample_id)

head(final_mutations_df, n = 10)
```

Write the final table to file.

```{r}
readr::write_tsv(final_mutations_df, output_file)
```

## Session Info

```{r}
sessionInfo()
```
