---
title: "Create a minimal palette for displaying multiple disease labels"
output:   
  html_notebook: 
    toc: true
    toc_float: true
author: Candace Savonen, Krutika Gaonkar, and Jaclyn Taroni
date: 2021
---

## Purpose

There are multiple "disease labels" in the `pbta-histologies.tsv` file, including (from most broad to most narrow) `broad_histology`, `cancer_group`, and `harmonized_diagnosis`.
For context, it is helpful to note that an individual `cancer_group` will be nested under a single `broad_histology` and that `cancer_group` is a shorter form of `harmonized_diagnosis` with the following edits:

- Removed Other, Benign tumor and Dysplasia/Gliosis, Dysplasia/Gliosis-Glial-neuronal tumor NOS removed from `cancer_group`
- Neurofibroma/Plexiform;Other updated to Neurofibroma/Plexiform
- Non-germinomatous germ cell tumor;Teratoma updated to Teratoma
- Anaplastic (malignant) meningioma, Meningothelial meningioma and Clear cell meningioma updated to Meningioma
- Embryonal Tumor with Multilayered Rosettes updated to Embryonal tumor with multilayer rosettes

It is often useful to use color to indicate disease label in a plot where multiple groups are visualized when we can not rely particularly heavily on labels (e.g., scatter plots).
Unfortunately, there are too many potential labels for us to generate an effective color palette (e.g., of sufficiently distinct colors).
In addition, some groupings will contain very few samples.

The purpose of this notebook is to create color palettes for the following:

* `broad_histology` values, where a `broad_histology` contains at least one `cancer_group` with n >= 10
* `cancer_group` values with n >= 10

**Note: This is tied to `release-v21-20210820`.**

### Background

You may find [#1174](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/1174) to be helpful context.

## Usage

This notebook can be run via the command line from the top directory of the repository as follows:

```
Rscript -e "rmarkdown::render('figures/mapping-histology-labels.Rmd', 
                              clean = TRUE)"
```

## Set Up

```{r}
library(tidyverse)
```

### Directories and Files

```{r}
# Path to input directory
input_dir <- file.path("..", "data", "release-v21-20210820")
output_dir <- "palettes"
```

## Read in metadata 

Let's read in the `pbta-histologies.tsv` file from `release-v21-20210820`.

```{r}
histologies_df <-
  readr::read_tsv(file.path(input_dir, "pbta-histologies.tsv"), guess_max = 10000)
```

## Identify values to include in palettes

We will use `cancer_group` with n >= 10 to guide what values to include in both our `cancer_group` and `broad_histology` palettes.

```{r}
included_labels_df <- histologies_df %>% 
  # Exclude normal samples
  filter(sample_type == "Tumor") %>%
  # Filter to unique sample--"disease label" pairs
  select(sample_id, 
         broad_histology, 
         cancer_group) %>% 
  distinct() %>%
  # Count samples (e.g., sample_id)
  group_by(broad_histology, cancer_group) %>% 
  tally() %>%
  # Add a column called included which is a logical that can be used as 
  # a sample size filter & also to drop the NA values
  filter(n >= 10, 
         !is.na(cancer_group))

included_labels_df
```

So the unique values for `broad_histology` and `cancer_group` above are what we need to take into account for our palette.

## Create palettes

Outside of this notebook, we've done quite a bit of work to identify suitable palettes using http://phrogz.net/css/distinct-colors.html as a reference/starting point.
Check out the discussion on [#1174](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/1174)!

### `broad_histology`

```{r}
broad_histology_df <- data.frame(
  broad_histology = c("Benign tumor",
                      "Diffuse astrocytic and oligodendroglial tumor",
                      "Embryonal tumor",
                      "Ependymal tumor",
                      "Germ cell tumor",
                      "Low-grade astrocytic tumor",
                      "Meningioma",
                      "Mesenchymal non-meningothelial tumor",
                      "Neuronal and mixed neuronal-glial tumor",
                      "Tumor of cranial and paraspinal nerves",
                      "Tumors of sellar region"), 
  broad_histology_hex = c("#590024",
                          "#ff80e5",
                          "#220040",
                          "#2200ff",
                          "#0074d9",
                          "#8f8fbf",
                          "#2db398",
                          "#7fbf00",
                          "#685815",
                          "#ffaa00",
                          "#b2502d"),
  stringsAsFactors = FALSE
)
```

Now to create a legend with `legend()` (h/t [this StackOverflow answer](https://stackoverflow.com/questions/48966645/how-can-i-create-a-legend-without-a-plot-in-r/48966924))

```{r}
plot(NULL, xaxt = "n", yaxt = "n", bty = "n", ylab = "", xlab = "", 
     xlim = 0:1, ylim = 0:1)
legend("topleft", 
       legend = broad_histology_df$broad_histology,
       col = broad_histology_df$broad_histology_hex,
       pch = 15, pt.cex = 2, cex = 1, bty = "n")
mtext("Broad Histology", at = 0.135, cex = 1.5)
```

### `cancer_group`

There are 17 `cancer_group` values that we need to account for.
These are best used in conjunction with _labels_ in figures, but are intended to allow readers to "track" labels _across figures_.

Where there's a 1:1 mapping between `broad_histology` and `cancer_group`, the hex codes will be the same.

```{r}
cancer_group_df <- data.frame(
  cancer_group = c("Choroid plexus papilloma",
                   "Diffuse intrinsic pontine glioma",
                   "Diffuse midline glioma",
                   "High-grade glioma astrocytoma",
                   "Atypical Teratoid Rhabdoid Tumor",
                   "CNS Embryonal tumor",
                   "Medulloblastoma",
                   "Ependymoma",
                   "Teratoma",
                   "Ganglioglioma",
                   "Low-grade glioma astrocytoma",
                   "Meningioma",
                   "Ewing sarcoma",
                   "Dysembryoplastic neuroepithelial tumor",
                   "Neurofibroma Plexiform",
                   "Schwannoma",
                   "Craniopharyngioma"),
  cancer_group_hex = c("#592d3e",
                       "#ffccf5",
                       "#ff40d9",
                       "#bf0099",
                       "#4d0d85",
                       "#7739ad",
                       "#9426fb",
                       "#2200ff",
                       "#058aff",
                       "#8c8cff",
                       "#000080",
                       "#2db398",
                       "#9fbf60",
                       "#614e01",
                       "#e6ac39",
                       "#ab7200",
                       "#b33000"),
  stringsAsFactors = FALSE
)
```

And again, we'll create a legend with `legend()`

```{r}
plot(NULL, xaxt = "n", yaxt = "n", bty = "n", ylab = "", xlab = "", 
     xlim = 0:1, ylim = 0:1)
legend("topleft", 
       legend = cancer_group_df$cancer_group,
       col = cancer_group_df$cancer_group_hex,
       pch = 15, pt.cex = 1.5, cex = 0.75, bty = "n")
mtext("Cancer Group", at = 0.0625, cex = 1)
```

### Output

We can create a data frame that contains both palettes with a series of left joins, where we will then fill the NA values with a single (gray) hex code per column (`#808080` for `broad_histology`, `#d3d3d3` for `cancer_group`.)

```{r}
palette_df <- histologies_df %>%
  # Exclude normal samples
  filter(sample_type == "Tumor") %>%
  # Filter to unique broad histology--cancer group pairs
  select(broad_histology, 
         cancer_group) %>% 
  distinct() %>%
  # Add broad histology palette
  left_join(broad_histology_df, by = "broad_histology") %>%
  # Add cancer group palette
  left_join(cancer_group_df, by = "cancer_group") %>%
  # Fill all other values with gray colors
  replace_na(list(broad_histology_hex = "#808080", 
                  cancer_group_hex = "#b5b5b5")) %>%
  # The exception being - if cancer_group == NA, so should cancer_group_hex!
  mutate(cancer_group_hex = if_else(is.na(cancer_group), 
                                    NA_character_, 
                                    cancer_group_hex)) %>%
  # Sort by broad_histology for easy browsing
  arrange(broad_histology)
```

And now let's take a look!

```{r}
palette_df
```

### Add display names for convenience

When multiple values are using the same color, it can be helpful to have a separate value for the legend, e.g., for all `#808080` broad histologies, we may want to display `Other`. 
We'll add a couple columns for legend-making convenience.

```{r}
palette_df <- palette_df %>%
  mutate(broad_histology_display = if_else(broad_histology_hex == "#808080",
                                           "Other",
                                           broad_histology),
         cancer_group_display = if_else(cancer_group_hex == "#b5b5b5",
                                        "Other",
                                        cancer_group))
```

### Add `broad_histology_order` 

Previously, we had a concept known as `display_order` where we ordered categories based on their number of samples (from large to small).
Now that we've dropped `display_group`, let's apply this same concept to `broad_histology`.

```{r}
broad_histology_order_df <- histologies_df %>%  
  # Exclude normal samples
  filter(sample_type == "Tumor",
         # Only count histologies that we'll have a hex code for
         broad_histology %in% included_labels_df$broad_histology) %>%
  # Filter to unique sample--broad_histology pairs
  select(sample_id, 
         broad_histology) %>%
  distinct() %>%
  # Count samples within a broad histology
  count(broad_histology) %>%
  # Add Other placeholder
  bind_rows(data.frame(broad_histology = "Other",
                       n = 0, 
                       stringsAsFactors = FALSE)) %>%
  # Reorder based on sample size except Benign tumor and Other should come last
  # And then add numeric column with the order
  mutate(broad_histology = forcats::fct_reorder(broad_histology,
                                                      n,
                                                      .desc = TRUE) %>%
           forcats::fct_relevel("Benign tumor",
                                "Other",
                                after = Inf),
         broad_histology_order = as.numeric(broad_histology)) %>%
  # No longer require the sample size
  select(-n)

broad_histology_order_df
```

And now we're ready to add this to the palette data frame.

```{r}
palette_df <- palette_df %>%
  left_join(broad_histology_order_df, 
            by = c("broad_histology_display" = "broad_histology"))
```

### Add `oncoprint_include` and `oncoprint_group`

For most plots that make use of the `cancer_group` palette, such as a box or violin plot, we will rely heavily on labels and therefore using the gray hex code for multiple groups will not be a problem.
In oncoprints, we'll want to drop `cancer_group` with small sample size ([#1177](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/1177)).
We'll add a logical column to help with that.

```{r}
palette_df <- palette_df %>%
  mutate(oncoprint_include = if_else(cancer_group_display == "Other",
                                     FALSE,
                                     TRUE)) %>%
  # If cancer_group is NA, then set to FALSE
  replace_na(list(oncoprint_include = FALSE))
```

We will have four panels of individual oncoprints, where many `broad_histology` values will get grouped together into the `Other CNS` panel which you can see [here](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/d31c927a27813ec0b8032fbe768002f31723636f/analyses/oncoprint-landscape/02-plot-oncoprint.R#L181).
We can move this information into our palette data frame.

```{r}
# Taken from the current plot oncoprint script as of the writing of this
# See permalink above
other_cns_broad_histologies <- c(
  "Ependymal tumor",
  "Tumors of sellar region",
  "Neuronal and mixed neuronal-glial tumor",
  "Tumor of cranial and paraspinal nerves",
  "Meningioma",
  "Mesenchymal non-meningothelial tumor",
  "Germ cell tumor",
  "Choroid plexus tumor",
  "Histiocytic tumor",
  "Tumor of pineal region",
  "Metastatic tumors",
  "Other astrocytic tumor",
  "Lymphoma",
  "Melanocytic tumor",
  "Other tumor"
)

palette_df <- palette_df %>%
  mutate(oncoprint_group = case_when(
    # If the broad histology is in the vector above and the cancer group will 
    # be included, set to Other CNS
    broad_histology %in% other_cns_broad_histologies &
      oncoprint_include != FALSE ~ "Other CNS",
    # If this cancer group doesn't get included in the oncoprint, set to NA
    # Same with Benign tumors -- those do not get included in that module
    oncoprint_include == FALSE | 
      broad_histology == "Benign tumor" ~ NA_character_,
    # Otherwise use broad histology
    TRUE ~ broad_histology
  ))
```

### Save to TSV 

```{r}
readr::write_tsv(palette_df, 
                 file.path(output_dir, 
                           "broad_histology_cancer_group_palette.tsv"))
```

## Session Info

```{r}
sessionInfo()
```
