---
title: "Mapping histology labels for plots"
output:   
  html_notebook: 
    toc: true
    toc_float: true
author: Candace Savonen for ALSF - CCDL
date: 2021
---

# Purpose: 

The histology label variables included in `pbta-histologies.tsv` from data releases are not always useful for visualizing the full set of biospecimens due to the large number of different values.
Having too many different possible values makes the colors harder to distinguish.
In addition, there are some groups that are represented by only a very few samples; giving such groups a distinct color may be counterproductive.

The goal of this notebook is to use the currently existing `broad_histology` groups from `pbta-histologies.tsv`, to form 10-15 "high level histology" group labels that can used for plotting purposes.

## The output table

The output of this notebook is a TSV file: `palettes/histology_label_color_table.tsv` that contains the following fields:

**Copied from `pbta-histologies.tsv`**:    
- `Kids_First_Biospecimen_ID` (from `pbta-histologies.tsv`)  
- All the original histology label variables (`broad_histology`, `short_histology`, etc.)  
  
**Created in this notebook**:  
- `display_group` - the high-level histology labels that should be used for plotting   
- `hex_codes` the direct colors that correspond to display_groups
- `cancer_group_hex_codes` the direct colors that correspond to cancer_groups

With this info, `histology_label_color_table.tsv` can be used by all plots and figures that summarize high level data  while displaying histology information. 

# How `display_group` is made:

Here's how `broad-histology` groups are [combined into the higher-level groupings of `display_group`](#declare-new-equivalent-groups).

1) "Lymphoma", "Melanocytic tumor", "Other tumor", "Metastatic tumors", "Non-CNS tumor" are combined into a `Other tumor` in `display_group`. 

2) `Benign tumor` and `Non-tumor` biospecimens are combined into a `Benign` group. 

3) `Other astrocytic tumor` biospecimens are combined into the existing `Low-grade astrocytic tumor`. These biospecimens  in `other astrocytic tumors` were low-grade SEGA tumors. 

4) Anything not in the above categories gets its `broad_histology` label carried over. 

# Usage

This notebook can be run via the command line from the top directory of the 
repository as follows:

```
Rscript -e "rmarkdown::render('figures/mapping-histology-labels.Rmd', 
                              clean = TRUE)"
```

## Set Up

```{r}
# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

### Directories and Files

```{r}
# Path to input directory
input_dir <- file.path("..", "data")
output_dir <- "palettes"
```

# Read in metadata 

Which variables are we keeping for this table? 

```{r}
histology_variables <- 
  c("integrated_diagnosis", 
    "Notes", 
    "harmonized_diagnosis",
    "broad_histology", 
    "short_histology",
    "cancer_group")
```

Let's read in the current release's `pbta-histologies.tsv` file. 

```{r}
metadata <-
  readr::read_tsv(file.path(input_dir, "pbta-histologies.tsv"), guess_max = 10000)
```

Now we'll select histology variables we mentioned above and so capitalization differences don't get in the way with this process, we will change everything to lower case for now. 

```{r}
working_metadata <- metadata %>% 
  dplyr::select(Kids_First_Biospecimen_ID, sample_type, histology_variables) %>% 
  dplyr::mutate(broad_histology_lower = tolower(broad_histology))
```

# Take a look at how many biospecimens per `broad_histology` group

Let's summarize `broad_histology`. 
Because the `Normal` samples don't have histologies, we'll look at just the `Tumor` samples at for this summary. 

```{r}
broad_summary <- working_metadata %>% 
  dplyr::filter(sample_type == "Tumor") %>%
  dplyr::count(broad_histology_lower) %>% 
  dplyr::arrange(n) 
```

Let's print out the summary. 

```{r}
broad_summary %>% 
  knitr::kable()
```

There's handful of very small groups (many are n = 2). 

## Declare new equivalent groups

These groups we'll combine into a non-CNS/other tumor group.

```{r}
other_tumor <- c("lymphoma", "melanocytic tumor", "other tumor", "metastatic tumors", "non-cns tumor")
```

These groups we'll combine as a benign.

```{r}
benign <- c("benign tumor", "non-tumor")
```

Add in the `Other astrocytic tumor` in with the LGAT group. 

```{r}
lgat <- c("other astrocytic tumor", "low-grade astrocytic tumor")
```

# Make new `display_group`

```{r}
histology_table <- working_metadata %>% 
  dplyr::mutate(
    # NAs are really Normals
    display_group = tidyr::replace_na(broad_histology_lower, "normal"),
    # Now do the group combining
    display_group = forcats::fct_collapse(display_group,
      "low-grade astrocytic tumor" = lgat,
      "other tumor" = other_tumor,
      "benign" = benign
    ),
    # Put this as a character for later handling
    display_group = as.character(display_group)
    )
```

Print out the number of `display_group` (including `normal`)!

```{r}
display_group_df <- histology_table %>% 
  dplyr::count(display_group) %>% 
  dplyr::arrange(n)

knitr::kable(display_group_df)
```

Make this notebook stop if there are more than 16 histology groups + `Normal`. 

```{r}
if (nrow(display_group_df) > 18) {
  stop("There are more than 18 categories in `display_group`. We may want to re-evaluate the high-level histology groupings")
}
```

# Make `display_order`

Get ranks in order of big to small and make them into a new column in `display_group_df`. 
We will always want the "normal", "benign", "other_tumor" groups to come last so we will push then to the end of the factor order. 

```{r}
display_order_df <- display_group_df %>% 
  dplyr::mutate(display_group = forcats::fct_reorder(display_group, n, .desc = TRUE) %>%
                  forcats::fct_relevel("benign", "other tumor", "normal", after = Inf),
                display_order = as.numeric(display_group)) # save the factor order for text table export
```

# Make `cancer_group_order`

`cancer_group` is a shorter form of `harmonized_diagnosis` with the following edits:
- Removed Other, Benign tumor and Dysplasia/Gliosis,Dysplasia/Gliosis-Glial-neuronal tumor NOS removed from `cancer_group`
- Neurofibroma/Plexiform;Other updated to Neurofibroma/Plexiform
- Non-germinomatous germ cell tumor;Teratoma updated to Teratoma
- Anaplastic (malignant) meningioma, Meningothelial meningioma and Clear cell meningioma updated to Meningioma
- Embryonal Tumor with Multilayered Rosettes updated to Embryonal tumor with multilayer rosettes

Get ranks in order of big to small and make them into a new dataframe `cancer_group_order_df`. 
   
```{r}
cancer_group_order_df <- histology_table %>% 
  dplyr::count(cancer_group,name = "cancer_group_n") %>% 
  dplyr::mutate(
    cancer_group = forcats::fct_reorder(cancer_group, cancer_group_n, .desc = TRUE),
                cancer_group_order = as.numeric(cancer_group)) # save the factor order for text table export
```


Add on the `display_order` column using `inner_join` 

```{r}
histology_table <- histology_table %>%
  # Join on the display orders
  dplyr::inner_join(display_order_df, by = "display_group") %>%
  # Join on the cancer_group orders
  dplyr::inner_join(cancer_group_order_df, by = "cancer_group") 
```

# Add hex codes for display_group and cancer_group

These hex codes were retrieved from http://phrogz.net/css/distinct-colors.html with the settings on default for 18 colors.

```{r}
color_palette_display <- 
  c("#ff0000", "#cc0000", "#995200", "#bfb300", "#fffbbf", 
    "#2e7300", "#00e65c", "#00ffee", "#103d40", "#0085a6", 
    "#003380", "#4073ff", "#737899", "#70008c", "#f2b6ee", 
    "#ff40bf", "#8c0038", "#330d12"
)

color_palette_cancer_group <-
  c("#ff0000", "#f20000", "#997373", "#403030", "#330700",
    "#ff9180", "#591800", "#b2502d", "#cca799", "#ff6600",
    "#ffb380", "#7f5940", "#cc6d00", "#331b00", "#ccb499",
    "#ffaa00", "#996600", "#594316", "#ffd580", "#ffee00",
    "#998f00", "#999673", "#303300", "#fbffbf", "#ccff00",
    "#494d39", "#b5d96c", "#6a8040", "#66ff00", "#42a600",
    "#bfffbf", "#003307", "#00661b", "#00ff88", "#86b39e",
    "#00b377", "#006652", "#00ffee", "#00a7b3", "#bffbff",
    "#567173", "#00ccff", "#003d4d", "#00aaff", "#267399",
    "#0088ff", "#0042a6", "#001a40", "#bfd9ff", "#0044ff",
    "#394973", "#000e66", "#bfbfff", "#9180ff", "#5800a6",
    "#754d99", "#aa00ff", "#3a3040", "#aa86b3", "#530059",
    "#ff00ee", "#a60085", "#330022", "#ff80d5", "#ff0088",
    "#804062", "#a60042", "#590024", "#ffbfd9", "#ff0044",
    "#990014", "#ff8091"
)

```

Declare how many colors we need. 

```{r}
n_colors_display <- nrow(display_group_df)
n_colors_cancer_group <- nrow(cancer_group_order_df)
```

Make a named list color key where histologies are the names. 

```{r}
# Set seed so the colors are consistent upon re-run
set.seed(2021)

# Sample from the 18 colors for display_group
subset_colors_display <- sample(color_palette_display, n_colors_display)
names(subset_colors_display) <- display_order_df$display_group

# Sample from the 62 colors for cancer_group
subset_colors_cancer_group <- sample(color_palette_cancer_group, n_colors_cancer_group)
names(subset_colors_cancer_group) <- cancer_group_order_df$cancer_group
```

Remove <NA> from subset_colors_cancer_group 
```{r}
# We will assign a gray color for NA below
subset_colors_cancer_group <- subset_colors_cancer_group[!is.na(names(subset_colors_cancer_group))]
```

We want `Other tumor` and the `Benign` in display_group to both always be gray. 

```{r}
subset_colors_display[names(subset_colors_display) == 'other tumor'] <- "#808080" 
subset_colors_display[names(subset_colors_display) == 'benign'] <-  "#D3D3D3"
```

Normal biospecimens should not get plotted in display_group, so we will put their hex code as black. 

```{r}
subset_colors_display[names(subset_colors_display) == 'normal'] <- "#000000"
```

Use `pie` function to preview what display_group these look like.

```{r}
pie(rep(1, n_colors_display), 
    col = subset_colors_display, 
    labels = names(subset_colors_display))
```

Use `pie` function to preview what cancer_group these look like.

```{r}
pie(rep(1, n_colors_cancer_group), 
    col = subset_colors_cancer_group, 
    labels = names(subset_colors_cancer_group))
```


Add the hex codes for display_group and cancer_group to the `histology_table`. 
Add gray color if cancer_group==NA
```{r}
histology_table <- histology_table %>%
  # We don't need this anymore
  dplyr::select(-broad_histology_lower) %>%
  # Add the hex_codes
  dplyr::mutate(hex_codes = dplyr::recode(display_group, !!!subset_colors_display),
		cancer_group_hex_codes = dplyr::recode(cancer_group, !!!subset_colors_cancer_group),
                # if cancer_group is NA for tumor sample add gray color
		cancer_group_hex_codes = dplyr::if_else(is.na(cancer_group_hex_codes) & sample_type=="Tumor"
                                                 ,"#808080",
                                                 cancer_group_hex_codes)
	       	) %>% 
  # Restore capitalization so its pretty for labeling
  dplyr::mutate(display_group = stringr::str_to_sentence(display_group),
                # Deal with CNS exception
                display_group = stringr::str_replace(display_group, "cns", "CNS")
                )
```

## Save to TSV 

```{r}
readr::write_tsv(histology_table, file.path(output_dir, "histology_label_color_table.tsv"))
```

# Session Info

```{r}
sessionInfo()
```
