---
title: "Mapping histology labels for plots"
output:   
  html_notebook: 
    toc: true
    toc_float: true
author: Candace Savonen for ALSF - CCDL
date: 2020
---
## Purpose: 

Our goal is to get 10-15 labels based on histologies we can use for plotting. 

### Usage

This notebook can be run via the command line from the top directory of the 
repository as follows:

```
Rscript -e "rmarkdown::render('figures/mapping-histology-labels.Rmd', 
                              clean = TRUE)"
```

### Set Up

```{r}
# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

### Directories and Files

```{r}
# Path to input directory
input_dir <- file.path("..", "data")
output_dir <- "palettes"
```

### Read in data 

Which variables are we keeping for this table? 

```{r}
histology_variables <- 
  c("Kids_First_Biospecimen_ID",
    "sample_type", 
    "integrated_diagnosis", 
    "Notes", 
    "harmonized_diagnosis",
    "broad_histology", 
    "short_histology")

```

Let's read in the current pbta-histologies.tsv file and select the variables we mentioned above. 

```{r}
metadata <-
  readr::read_tsv(file.path(input_dir, "pbta-histologies.tsv"), guess_max = 10000) %>% 
  dplyr::select(histology_variables)
```

Because the `Normal` samples don't have histologies, we'll look at just the Tumor samples at first. 

```{r}
tumor_only_metadata <- metadata %>% 
  dplyr::filter(sample_type == "Tumor") 
```

### Take a look at how many biospecimens per `broad_histology` group

Let's summarize `broad_histology`. 

```{r}
broad_summary <- tumor_only_metadata %>% 
  dplyr::count(broad_histology) 
```

Let's take a look at how big all these groups are with a barplot.

```{r}
ggplot2::ggplot(broad_summary, ggplot2::aes(x = reorder(broad_histology, -n), y = n)) +
  ggplot2::geom_bar(position = "dodge", stat = "identity") +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1)) + 
  ggplot2::xlab("broad_histology label")
```

```{r}
broad_summary %>% 
  dplyr::filter(n > 12) %>% 
  nrow()
```

There are 13 groups that are bigger than 12 samples and 8 groups that have less than 12 samples.

Now let's get a vector of those small `broad_histology` groups. 

```{r}
small_groups <- broad_summary %>% 
  dplyr::filter(n < 12) %>% 
  dplyr::pull(broad_histology)
```

### Make new `summary_group`

Here's the major recoding steps done in the next code chunk: 

1) All groups with less than 12 biospecimens assigned to them will be grouped together as `Other`. 

2) "Neuronal and mixed neuronal-glial tumor" and "Diffuse astrocytic and oligodendroglial tumor" were labeled as LGAT in previous releases so we will group these into `Low-grade astrocytic tumor` again. 

3) So its clear why they are `NA`s, we will label all `Normal` samples (non-tumor samples) as `Normal`. 

4) Anything not in the above categories gets its `broad_histology` label carried over. 

```{r}
histology_table <- metadata %>% 
  dplyr::mutate(
    summary_group = dplyr::case_when(
      broad_histology %in% small_groups ~ "Other",
      broad_histology %in% c("Neuronal and mixed neuronal-glial tumor", 
                             "Diffuse astrocytic and oligodendroglial tumor") ~ "Low-grade astrocytic tumor", 
      sample_type == "Normal" ~ "Normal",
      TRUE ~ broad_histology)
    )
```

Now there are 14 groups!

```{r}
summary_groups <- histology_table %>% 
  dplyr::count(summary_group)

summary_groups
```

### Add hex codes 

These hex codes were retrieved from http://phrogz.net/css/distinct-colors.html with the settings on default for 18 colors.

```{r}
color_palette <- 
  c("#ff0000", "#cc0000", "#995200", "#bfb300", "#fffbbf", 
    "#2e7300", "#00e65c", "#00ffee", "#103d40", "#0085a6", 
    "#003380", "#4073ff", "#737899", "#70008c", "#f2b6ee", 
    "#ff40bf", "#8c0038", "#330d12"
)
```

Declare how many colors we need. 

```{r}
n_colors <- nrow(summary_groups)
```

Make a named list color key where histologies are the names. 

```{r}
# Set seed so the colors are consistent upon re-run
set.seed(2021)

# Sample from the 18 colors
subset_colors <- sample(color_palette, n_colors)
names(subset_colors) <- summary_groups$summary_group
```

Use `pie` function to preview what these look like.

```{r}
pie(rep(1, n_colors), 
    col = subset_colors, 
    labels = summary_groups$summary_group)
```

Add the hex codes to the `histology_table`. 

```{r}
histology_table <- histology_table %>% 
  dplyr::mutate(hex_codes = dplyr::recode(summary_group, hex_codes = !!!subset_colors))
```

Save to the TSV. 

```{r}
readr::write_tsv(histology_table, file.path(output_dir, "histology_label_color_table.tsv"))
```

# Session Info

```{r}
sessionInfo()
```
