---
title: "Tables output for manuscript"
author: "Run Jin"
date: "10/27/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
install.packages("openxlsx")
library(tidyverse)
library(openxlsx)
```

## Output Tables for OpenPBTA Manuscript
This is a Rmd files that record scripts for generating tables 

```{r define directories}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
working_dir <- file.path(root_dir, "tables")
data_dir <- file.path(root_dir, "data", "release-v21-20210820")

results_dir <- file.path(working_dir, "results")
if(!dir.exists(results_dir)){
  dir.create(results_dir, recursive=TRUE)
}
```

## Table 1: Molecular subtypes determined for this project
```{r table 1}
histology_df <- readr::read_tsv(file.path(data_dir, "pbta-histologies.tsv"), guess_max =10000)

subtypes <- histology_df %>%
  dplyr::filter(!is.na(pathology_diagnosis) & !is.na(molecular_subtype) & !grepl("To be classified", molecular_subtype))
subtype_table <- subtypes %>%
  dplyr::group_by(broad_histology, molecular_subtype) %>%
  tally() %>%
  readr::write_tsv(file.path(results_dir, "Table1-molecular-subtypes.tsv"))

```

## Table S1: V21 histologies table

```{r v21 histology as excel}
# add back the new samples with `MB, To be classified"
histology_df <- histology_df %>% 
  dplyr::mutate(molecular_subtype = case_when(
    Kids_First_Biospecimen_ID %in% c("BS_3J4T2YYW", "BS_7F07M7JG", "BS_AG0BZN6F", "BS_HZNKSQ17", "BS_V6HKXX1Z", "BS_WP9J88EB", "BS_YYAPSA5P") ~ "MB, To be classified", 
    TRUE ~ molecular_subtype
  ))

histology_df %>%
  mutate_all(as.character) %>% 
  replace(is.na(.), "NA") %>%
  openxlsx::write.xlsx(file.path(results_dir, "TableS1-histologies.xlsx"),
                        overwrite=TRUE)

```

## Table S2: DNA results table
### TMB
```{r s2 DNA results part 1}

# read in tmb all file, select and rename columns
tmb_all <- readr::read_tsv(file.path(data_dir, "pbta-snv-mutation-tmb-all.tsv")) %>%
  dplyr::select(Tumor_Sample_Barcode, tmb) %>%
  dplyr::rename(Kids_First_Biospecimen_ID = Tumor_Sample_Barcode) %>% 
  dplyr::rename(Tmb_all = tmb) 

# read in tmb coding file, select and rename columns
tmb_coding <- readr::read_tsv(file.path(data_dir, "pbta-snv-mutation-tmb-coding.tsv")) %>%
  dplyr::select(Tumor_Sample_Barcode, tmb) %>%
  dplyr::rename(Kids_First_Biospecimen_ID = Tumor_Sample_Barcode) %>% 
  dplyr::rename(Tmb_coding = tmb)

# combine files
tmb_combined <- full_join(tmb_all, tmb_coding)  %>%
  mutate_all(as.character) %>% 
  replace(is.na(.), "NA")
```

### COSMIC mutational signatures
```{r s2 DNA results part 2}
# read in the file
cosmic_mut_df <- readr::read_tsv("../analyses/mutational-signatures/results/cosmic_signatures_results.tsv") %>%
  dplyr::select(Tumor_Sample_Barcode, signature, mut_per_mb) %>%
  dplyr::rename(Kids_First_Biospecimen_ID = Tumor_Sample_Barcode)

# get wide format
cosmic_mut_wide <- cosmic_mut_df %>%
  spread(signature, mut_per_mb ) 

# order the columns
unique_cosmic_sig <- cosmic_mut_df %>% 
  pull(signature) %>% unique()
cosmic_mut_wide <- cosmic_mut_wide %>%
  dplyr::select(c(Kids_First_Biospecimen_ID, all_of(unique_cosmic_sig)))

```

### Alexandrov mutational signatures
```{r s2 DNA results part 3}
# read in the file
alexandrov_mut_df <- readr::read_tsv("../analyses/mutational-signatures/results/nature_signatures_results.tsv") %>%
  dplyr::select(Tumor_Sample_Barcode, signature, mut_per_mb) %>%
  dplyr::rename(Kids_First_Biospecimen_ID = Tumor_Sample_Barcode)

# get wide format
alexandrov_mut_wide <- alexandrov_mut_df %>%
  spread(signature, mut_per_mb ) 

# order the columns
unique_alex_sig <- alexandrov_mut_df %>% 
  pull(signature) %>% unique()
alexandrov_mut_wide <- alexandrov_mut_wide %>%
  dplyr::select(c(Kids_First_Biospecimen_ID, all_of(unique_alex_sig))) 
```

### CNS mutational signatures
```{r s2 DNA results part 4}
cns_mut_list <- readRDS("../analyses/mutational-signatures/results/fitted_cns_signature_exposures.RDS")
cns_mean <- cns_mut_list[["mean"]] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Kids_First_Biospecimen_ID")  

```

### Chromothripsis regions per sample
```{r s2 DNA results part 5}
chromothripsis_region_df <- readr::read_tsv("../analyses/chromothripsis/results/chromothripsis_summary_per_sample.txt")  

```

### combine S2 table
```{r s2 DNA results combined}
list_s2_table <- list(tmb = tmb_combined,
                     cosmic_mut_sigs = cosmic_mut_wide,
                     alexandrov_mut_sigs = alexandrov_mut_wide,
                     cns_denovo_mut_sigs = cns_mean,
                     chromothripsis_events = chromothripsis_region_df
                     )
openxlsx::write.xlsx(list_s2_table, 
                     file.path(results_dir, "TableS2-DNA-results-table.xlsx"),
                     overwrite=TRUE)
```

## Table S3: RNA results table
### read in and process files
```{r s3 RNA results table}
# get tp53 scores 
tp53_scores <- readr::read_tsv("../analyses/tp53_nf1_score/results/tp53_altered_status.tsv") %>%
  mutate_all(as.character) %>% 
  replace(is.na(.), "NA")

# get extend scores file
telomerase_scores_polya_count <- readr::read_tsv("../analyses/telomerase-activity-prediction/results/TelomeraseScores_PTBAPolya_counts.txt") %>%
  dplyr::select(SampleID, NormEXTENDScores) %>% 
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = SampleID,
                NormEXTENDScores_counts = NormEXTENDScores)

telomerase_scores_polya_fpkm <-
  readr::read_tsv("../analyses/telomerase-activity-prediction/results/TelomeraseScores_PTBAPolya_FPKM.txt") %>%
  dplyr::select(SampleID, NormEXTENDScores) %>% 
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = SampleID,
                NormEXTENDScores_fpkm = NormEXTENDScores)

telomerase_scores_polya_combined <- full_join(telomerase_scores_polya_count,
                                              telomerase_scores_polya_fpkm)
  
telomerase_scores_stranded_count <- readr::read_tsv("../analyses/telomerase-activity-prediction/results/TelomeraseScores_PTBAStranded_counts.txt") %>%
  dplyr::select(SampleID, NormEXTENDScores) %>% 
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = SampleID,
                NormEXTENDScores_counts = NormEXTENDScores)

telomerase_scores_stranded_fpkm <- readr::read_tsv("../analyses/telomerase-activity-prediction/results/TelomeraseScores_PTBAStranded_FPKM.txt") %>%
  dplyr::select(SampleID, NormEXTENDScores) %>% 
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = SampleID,
                NormEXTENDScores_fpkm = NormEXTENDScores)

telomerase_scores_stranded_combined <- full_join(telomerase_scores_stranded_count,
                                                 telomerase_scores_stranded_fpkm)

telomerase_scores_combined <- bind_rows(telomerase_scores_polya_combined,
                                        telomerase_scores_stranded_combined)  %>%
  mutate_all(as.character) %>% 
  replace(is.na(.), "NA")

```

### combine and output file
```{r s3 table combine and output}
list_s3_table <- list(tp53_scores = tp53_scores,
                      telomerase_scores = telomerase_scores_combined
                      )
openxlsx::write.xlsx(list_s3_table, 
                     file.path(results_dir, "TableS3-RNA-results-table.xlsx"), 
                     overwrite=TRUE)

```

### print out session info
```{r}
sessionInfo()
```

