---
title: "Tables output for manuscript"
author: "Run Jin"
date: "10/27/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}

library(tidyverse)

```

## Output Tables for OpenPBTA Manuscript
This is a Rmd files that record scripts for generating tables 

```{r define directories}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
working_dir <- file.path(root_dir, "tables")
data_dir <- file.path(root_dir, "data")

results_dir <- file.path(working_dir, "results")
if(!dir.exists(results_dir)){
  dir.create(results_dir, recursive=TRUE)
}
```

## Table 1: Molecular subtypes determined for this project
```{r table 1}
histology_df <- readr::read_tsv(file.path(data_dir, "pbta-histologies.tsv"), guess_max =10000)

subtypes <- histology_df %>%
  dplyr::filter(!is.na(pathology_diagnosis) & !is.na(molecular_subtype) & !grepl("To be classified", molecular_subtype))
subtype_table <- subtypes %>%
  dplyr::group_by(broad_histology, molecular_subtype) %>%
  tally() %>%
  readr::write_tsv(file.path(results_dir, "subtypes-n.tsv"))

```

## Table S1: V21 histologies table

```{r v21 histology as excel}

openxlsx::write.xlsx(histology_df, 
                     file.path(results_dir, "OpenPBTA_histologies_v21.xlsx"),
                     overwrite=TRUE)

```

## Table S2: DNA results table
### TMB
```{r s2 DNA results part 1}

# read in tmb all file, select and rename columns
tmb_all <- readr::read_tsv(file.path(data_dir, "tcga-snv-mutation-tmb-all.tsv")) %>%
  dplyr::select(Tumor_Sample_Barcode, tmb) %>%
  dplyr::rename(Tmb_all = tmb)

# read in tmb coding file, select and rename columns
tmb_coding <- readr::read_tsv(file.path(data_dir, "tcga-snv-mutation-tmb-coding.tsv")) %>%
  dplyr::select(Tumor_Sample_Barcode, tmb) %>%
  dplyr::rename(Tmb_coding = tmb)

# combine files
tmb_combined <- full_join(tmb_all, tmb_coding)
```

### COSMIC mutational signatures
```{r s2 DNA results part 2}
# read in the file
cosmic_mut_df <- readr::read_tsv("../analyses/mutational-signatures/results/cosmic_signatures_results.tsv") %>%
  dplyr::select(Tumor_Sample_Barcode, signature, mut_per_mb)

# get wide format
cosmic_mut_wide <- cosmic_mut_df %>%
  spread(signature, mut_per_mb ) 

# order the columns
unique_cosmic_sig <- cosmic_mut_df %>% 
  pull(signature) %>% unique()
cosmic_mut_wide <- cosmic_mut_wide %>%
  dplyr::select(c(Tumor_Sample_Barcode, all_of(unique_cosmic_sig)))

```

### Alexandrov mutational signatures
```{r s2 DNA results part 3}
# read in the file
alexandrov_mut_df <- readr::read_tsv("../analyses/mutational-signatures/results/nature_signatures_results.tsv") %>%
  dplyr::select(Tumor_Sample_Barcode, signature, mut_per_mb)

# get wide format
alexandrov_mut_wide <- alexandrov_mut_df %>%
  spread(signature, mut_per_mb ) 

# order the columns
unique_alex_sig <- alexandrov_mut_df %>% 
  pull(signature) %>% unique()
alexandrov_mut_wide <- alexandrov_mut_wide %>%
  dplyr::select(c(Tumor_Sample_Barcode, all_of(unique_alex_sig)))
```

### CNS mutational signatures
```{r s2 DNA results part 4}
cns_mut_list <- readRDS("../analyses/mutational-signatures/results/fitted_cns_signature_exposures.RDS")
cns_mean <- cns_mut_list[["mean"]] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Tumor_Sample_Barcode")

```

### Chromothripsis regions per sample
```{r s2 DNA results part 5}
chromothripsis_region_df <- readr::read_tsv("../analyses/chromothripsis/results/chromothripsis_summary_per_sample.txt") %>%
  dplyr::rename(Tumor_Sample_Barcode = Kids_First_Biospecimen_ID)

```

### combine S2 table
```{r s2 DNA results combined}
list_s2_table <- list(tmb = tmb_combined,
                     cosmic_mut = cosmic_mut_wide,
                     alexandrov_mut = alexandrov_mut_wide,
                     cnv_mut = cns_mean,
                     chromothripsis=chromothripsis_region_df
                     )
openxlsx::write.xlsx(list_s2_table, 
                     file.path(results_dir, "tableS2_DNA_results_table.xlsx"),
                     overwrite=TRUE)
```

## Table S3: RNA results table
### read in and process files
```{r s3 RNA results table}
# get tp53 scores 
tp53_scores <- readr::read_tsv("../analyses/tp53_nf1_score/results/tp53_altered_status.tsv")

# get extend scores file
telomerase_scores_polya <- readr::read_tsv("../analyses/telomerase-activity-prediction/results/TelomeraseScores_PTBAPolya_counts.txt") %>%
  dplyr::select(SampleID, NormEXTENDScores) %>% 
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = SampleID)
  
telomerase_scores_stranded <- readr::read_tsv("../analyses/telomerase-activity-prediction/results/TelomeraseScores_PTBAStranded_FPKM.txt") %>%
  dplyr::select(SampleID, NormEXTENDScores) %>% 
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = SampleID)

telomerase_scores_combined <- bind_rows(telomerase_scores_polya, telomerase_scores_stranded)

```

### combine and output file
```{r s3 table combine and output}
list_s3_table <- list(tp53_scores = tp53_scores,
                      telomerase_scores = telomerase_scores_combined
                      )
openxlsx::write.xlsx(list_s3_table, 
                     file.path(results_dir, "tableS3_RNA_results_table.xlsx"), 
                     overwrite=TRUE)

```


