---
title: "Sample numbers for manuscript"
author: "Jo Lynne Rokita for D3b"
date: "2022"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
params:
  release: release-v22-20220505 
---

```{r setup, include=FALSE}
library(tidyverse)
```
This notebook tallies patient-level, sample-level, and biospecimen-level Ns used throughout the manuscript.

## Set directories 
```{r set directories}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "sample-distribution-analysis")
```

## Read in Files
```{r read in files}
hist_file <- file.path(data_dir, "pbta-histologies.tsv")
cell_lines_file <- file.path(root_dir, "analyses", "molecular-subtyping-HGG", "input", "cell-line-composition.tsv")

# Read in histologies file
hist <- read_tsv(hist_file, guess_max = 10000)

# Cell lines which have both suspension (S, serum-free media) or adherent (A, serum-containing media) compositions due to being grown in different media. 
cell_lines <- read_tsv(cell_lines_file)
```

## Calculate patient level Ns used for manuscript and by cohort


```{r sample size - patient-level}
# Total N patients
hist %>%
  pull(Kids_First_Participant_ID) %>%
  unique() %>%
  length()

# Total N patients by cohort
hist %>%
  select(Kids_First_Participant_ID, cohort) %>%
  unique() %>%
  count(cohort)
  
```

## Calculate sample (tumor/normal) level Ns used for manuscript

```{r sample size - tumor/normal-level}
# Total number of tumors
n_tumors <- hist %>%
  filter(sample_type == "Tumor" & composition != "Derived Cell Line") %>%
  pull(sample_id) %>%
  unique()
length(n_tumors)

# Total N assays by based on whether it is a cell line or tumor/normal tissue
hist <- hist %>%
  mutate(type = case_when(composition == "Derived Cell Line" ~ "Cell Line",
                                                sample_type == "Tumor" ~ "Tumor Sample",
                                                TRUE ~ "Normal Sample"),
         experimental_strategy = ifelse(experimental_strategy == "WXS" | experimental_strategy == "Targeted Sequencing", "WXS/Panel", experimental_strategy))
table(hist$type, hist$experimental_strategy)

```

## Calculate biospecimen level Ns used for manuscript

```{r sample size - biospecimen-level}
# Biospecimen counts by sample type
hist %>%
  count(sample_type)

# Biospecimen counts by sample type and composition (including duplicate cell lines)
hist %>%
  count(sample_type, composition)

# Biospecimen assay counts by sample type and composition
hist %>%
  count(sample_type, composition, experimental_strategy)
```
## Calculate cell line Ns used for manuscript

```{r sample size - cell lines}
# Total number of unique cell lines
cellline_bs <- hist %>%
  filter(sample_type == "Tumor" & composition == "Derived Cell Line") %>%
  left_join(cell_lines, by = c("Kids_First_Participant_ID", "Kids_First_Biospecimen_ID", "sample_id")) %>%
  select(Kids_First_Biospecimen_ID, sample_id, cell_line_composition, molecular_subtype) %>%
  arrange(sample_id)
 
cellline_bs_count <- cellline_bs %>%
    count(sample_id, cell_line_composition)
nrow(cellline_bs_count)


# Total number of cell lines, including duplicates
sum(cellline_bs_count$n)
```
## Calculate subtyped sample Ns used for manuscript

```{r samples subtyped}
# Total number of samples subtyped
all_samples_subtyped <- hist %>%
  filter(!is.na(pathology_diagnosis), 
          !is.na(molecular_subtype), 
          !grepl("To be classified", molecular_subtype)) %>% 
   # Count sample_id and subtype combinations 
  count(sample_id, molecular_subtype)
nrow(all_samples_subtyped)

# Total number of samples subtyped, not including cell lines
tumors_not_lines_subtyped <- hist %>%
  filter(composition != "Derived Cell Line",
         !is.na(pathology_diagnosis), 
         !is.na(molecular_subtype), 
         !grepl("To be classified", molecular_subtype)) %>% 
   # Count sample_id and subtype combinations 
  count(sample_id, molecular_subtype)
nrow(tumors_not_lines_subtyped)

# percent of tumor samples subtyped
(nrow(tumors_not_lines_subtyped)/length(n_tumors))*100

# which have non-identical tumor/cell line subtypes?
dup_samples <- all_samples_subtyped %>%
  filter(duplicated(.[["sample_id"]])) %>%
  pull(sample_id)

hist %>%
  filter(!is.na(molecular_subtype),
         sample_id %in% dup_samples) %>%
  count(sample_id, molecular_subtype, composition) %>%
  arrange(sample_id, composition)

# Do any cell line duplicates have different subtypes? No
cellline_bs %>%
  select(sample_id, molecular_subtype) %>%
  unique() %>%
  filter(duplicated(.[["sample_id"]])) %>%
  pull(sample_id)

# Total number of samples with WHO-aligned integrated diagnosis
who_dx <- hist %>%
  filter(!is.na(integrated_diagnosis)) %>%
  count(sample_id, composition, molecular_subtype) 
table(who_dx$composition)

```
## Calculate histology and cancer group Ns used for manuscript

```{r histology level Ns}
# Total number of annotated histologies
hist %>%
  filter(!is.na(pathology_diagnosis) & !is.na(cancer_group)) %>%
  pull(cancer_group) %>%
  unique() %>%
  length()

# Total number of broad histologies
hist %>%
  filter(!is.na(pathology_diagnosis) & !is.na(broad_histology) & broad_histology != "Non-tumor") %>%
  count(broad_histology) %>%
  nrow()
```

## Print session info
```{r print session info}
sessionInfo()
```
