---
title: "Visualization of CNS signature fitting with `deconstructSigs`"
author: "SJ Spielman"
date: "2022"
output:
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
params:
  is_ci: 0
---

<br><br><br><br>

#### Files and paths


```{r setup}
`%>%` <- dplyr::`%>%`

proj_root_path  <- file.path( rprojroot::find_root(rprojroot::has_dir(".git")) )
analysis_path   <- file.path(proj_root_path, "analyses", "mutational-signatures")
plot_path       <- file.path(analysis_path, "plots", "cns")
if (!dir.exists(plot_path)) {
  dir.create(plot_path, recursive = TRUE)
}

# Input files
fitted_file    <- file.path(analysis_path, "results", "fitted_exposures_signal-cns_deconstructSigs.rds")
meta_file      <- file.path(proj_root_path, "data", "pbta-histologies.tsv")


histology_palette_file   <- file.path(proj_root_path, "figures", "palettes", "broad_histology_cancer_group_palette.tsv")
binary_palette_file <- file.path(proj_root_path, "figures", "palettes", "binary_color_palette.tsv")
tumor_palette_file <-  file.path(proj_root_path, "figures", "palettes", "tumor_descriptor_palette.tsv")

# Output plots
exposures_sample_barplot_file <- file.path(analysis_path, "plots", "cns", "exposures_per_sample_barplot.pdf")
exposures_sina_IQR_file <- file.path(analysis_path, "plots", "cns", "exposures_sina_IQR.pdf")
exposures_presence_barplot_file <- file.path(analysis_path, "plots", "cns", "exposures_presence_barplot.pdf")
signature1_tumor_cancergroups_file <- file.path(analysis_path, "plots", "cns", "signature1_tumor-descriptor_cancer-groups.pdf")
signature1_tumor_molecularsubtypes_file <- file.path(analysis_path, "plots", "cns", "signature1_tumor-descriptor_molecular-subtypes.pdf")

# Output result file which contains merged/prepped data for visualization, also used in the `08` notebooks
results_df_file <- file.path("results", "deconstructsigs_exposures_merged.tsv")


# The real signature names: https://signal.mutationalsignatures.com/explore/studyTissueType/1-6
signature_names_df <- tibble::tibble(
  signature = factor(
    c("11", "1", "N6", "8", "MMR2", "18", "19", "3", "Other"), 
    levels = c("1", "3", "8", "11", "18", "19", "N6", "MMR2", "Other")),
  cns_names   = c(paste0("CNS_",LETTERS[1:8]), "Other")
)
```

First, we read in data and prepare exposures.

```{r read_prep_data, warning=FALSE}

# Read in metadata and palette files
meta <- readr::read_tsv(meta_file, guess_max = 10000)
palette_df <- readr::read_tsv(histology_palette_file) 
binary_df <- readr::read_tsv(binary_palette_file)
binary_colors <- binary_df$hex_codes[binary_df$color_names != "na_color"]
tumor_df <- readr::read_tsv(tumor_palette_file)


# Read in exposures from CNS signature fitting and tibble it up, including the "unknown" weight
raw_exposures <- readr::read_rds(fitted_file)
decon_weights <- lapply(raw_exposures, "[[", "weights")
exposures_wide <- do.call(rbind, decon_weights) %>%
  tibble::rownames_to_column("Kids_First_Biospecimen_ID") %>%
  tibble::as_tibble() 

unknown_weights <- lapply(raw_exposures, "[[", "unknown")
unknown_weights_array <- do.call(rbind, unknown_weights)
exposures <- exposures_wide %>%
  # use "Other" for unknown
  dplyr::mutate(Other = unknown_weights_array[,1]) %>%
  # key is column name in signature_names_df
  tidyr::gather(-Kids_First_Biospecimen_ID, key = "cns_names", value = "exposure") 

```


Now we prepare the data for visualization, including merging data frames and further setting up palettes.
Note that we are plotting here only the cancer groups used in the [`interaction-plots` module](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/master/analyses/interaction-plots/scripts/03-plot_interactions.R)

```{r merge_data}
# Cancer groups we are showing in figure 4
cancer_groups_to_plot <- c(
  "Diffuse midline glioma",
  "Other low-grade gliomas",
  "Craniopharyngioma",
  "Other high-grade gliomas",
  "Ganglioglioma",
  "Medulloblastoma",
  "Meningioma",
  "Ependymoma",
  "Pilocytic astrocytoma",
  "Pleomorphic xanthoastrocytoma"
)

# Join input dataframes together
results_df <- exposures %>% 
  # Merge with metadata
  dplyr::inner_join(
    meta %>%
      dplyr::select(Kids_First_Biospecimen_ID, broad_histology, cancer_group, tumor_descriptor)
  ) %>%
  # Grab the cancer group DISPLAY specifically (needed for joining later)
  dplyr::inner_join(
      dplyr::select(palette_df, cancer_group, cancer_group_display)
  ) %>%
  # no longer need cancer_group
  dplyr::select(-cancer_group) %>%
  # Retain only cancer groups for these visualizations
  dplyr::filter(cancer_group_display %in% cancer_groups_to_plot) %>%
  # Merge with correct signal names (already factored and ordered)
  dplyr::inner_join(signature_names_df)  %>%
  dplyr::select(-cns_names) # remove signature column 

# We need to create a new scheme for labeling that shows wrapped cancer groups with `(n=X)`
results_df <- results_df %>%
  dplyr::count(cancer_group_display, signature) %>%
  dplyr::select(-signature) %>%
  dplyr::distinct() %>%
  dplyr::inner_join(
      dplyr::select(palette_df, dplyr::contains("cancer_group"))
  ) %>%
  dplyr::select(cancer_group_display, n, cancer_group_hex) %>%
  # Create wrapped with (n=X) factor column for cancer groups
  dplyr::mutate(cancer_group_display_n = stringr::str_wrap(glue::glue("{cancer_group_display} (N={n})"), 30),
                cancer_group_display_n = forcats::fct_reorder(cancer_group_display_n, n, .desc=T)) %>%
  dplyr::inner_join(results_df)

# We also write `results_df` to a file since it will be used by the `08` notebooks.
readr::write_tsv(results_df, results_df_file)


```


<br><br>


### Sina/IQR of exposures across cancer groups 


_All_ samples are included in these sina plots, which show the distribution of exposures for each signature across samples in each cancer group.


```{r boxplot, fig.width = 14, fig.height = 4.5}

exposures_sina_boxplot <- results_df %>%
  # we DO visualize unknown also, but in case...
  #dplyr::filter(signature != "Other") %>%
  ggplot2::ggplot() + 
  ggplot2::aes(x = signature, 
               y = exposure,
               color = cancer_group_hex) + 
  ggforce::geom_sina(size = 0.7) + 
  ggplot2::geom_boxplot(outlier.size = 0, 
                        size = 0.3, 
                        color = "black", 
                        alpha = 0,
                        # remove whiskers
                        coef = 0) +
  ggplot2::facet_wrap(~cancer_group_display_n, nrow = 2) +
  ggplot2::scale_color_identity() + 
  # commenting out because 2 points are somehow removed..
  #ggplot2::scale_y_continuous(limits = c(0, 1.2), 
  #                            breaks = c(0, 0.25, 0.5, 0.75, 1.0)) +
  ggplot2::labs(
    x = "RefSig signature",
    y = "Signature weights across samples"
  ) +
  ggpubr::theme_pubr() + 
  cowplot::panel_border() +
  ggplot2::theme(
    # angle needed to fit "Other" category
    axis.text.x = ggplot2::element_text(size = 8.5, angle = 45, hjust = 1),
    axis.text.y = ggplot2::element_text(size = 10),
    axis.title = ggplot2::element_text(size = 12),
    strip.text = ggplot2::element_text(size = 8.5),
    legend.position = "none"
  )
exposures_sina_boxplot

ggplot2::ggsave(exposures_sina_IQR_file, exposures_sina_boxplot, width = 12, height = 5)

```


### Barplots of proportions of samples that are _unexposed_ 

Notably, the "rug" of 0's in each distribution represents samples which are unexposed to the given signature. 
It's useful to get a clear sense of the proportion of samples that are totally unexposed.
We do _not_ include "Other" in this plot. 

```{r proportion, fig.width = 5, fig.height = 10}
exposures_proportions <- results_df %>%
  # remove "other" - not useful to know if samples are unexposed to signatures that aren't even present.
  dplyr::filter(signature != "Other") %>%
  dplyr::group_by(cancer_group_display_n, cancer_group_hex, signature) %>%
  dplyr::summarize(percent_exposed = sum(exposure > 0)/dplyr::n()) %>%
  ggplot2::ggplot() + 
  ggplot2::aes(x = signature, y = percent_exposed, fill = cancer_group_hex) + 
  ggplot2::geom_col(color = "black") +
  ggplot2::facet_wrap(~cancer_group_display_n, nrow = 5) + 
  ggplot2::scale_fill_identity() +
  ggplot2::labs(x = "RefSig Signature", 
                y = "Percent of samples exposed to signature") +
  ggpubr::theme_pubr() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(size = 8))

exposures_proportions

ggplot2::ggsave(exposures_presence_barplot_file, exposures_proportions, width = 5, height = 10)

```


<br><br>


### Barplot of sample exposures faceted across cancer groups

This plot shows the signature exposures for each sample. 
Each bar represents one sample, with fills representing the different signatures.
We order plots by sample 1, and we include the unknown proportion so everything has the same height. 
In this case, "Other" specifically means mutations that were not assigned to the specific 8 CNS signatures examined, but are associated (if at all) with a different signature.

```{r barplot1, fig.width = 12, fig.height = 9}

# Make a column for exposure of interest to order samples by
samples_in_order <- results_df %>%
  dplyr::filter(signature == 1) %>%
  dplyr::select(exposure, Kids_First_Biospecimen_ID) %>%
  dplyr::arrange(-exposure) %>%
  dplyr::pull(Kids_First_Biospecimen_ID)


sample_exposures_barplot <- results_df %>%
  dplyr::mutate(Kids_First_Biospecimen_ID = factor(Kids_First_Biospecimen_ID, levels = samples_in_order)) %>%
  ggplot2::ggplot() +
  ggplot2::aes(x = Kids_First_Biospecimen_ID, 
               y = exposure, 
               fill = signature) + 
  ggplot2::geom_col(color = "black", size = 0.05) + 
  ggplot2::labs(x = "Sample", y = "Signature weight", fill = "RefSig Signature") + 
  # Only 1:8 palette. The "Other" gets filled as white by including `color="black"` in the geom
  # There will be a warning about palette size and the warning is OK because ^^
  colorblindr::scale_fill_OkabeIto() + 
  ggplot2::facet_wrap(~cancer_group_display_n, scales = "free_x", nrow = 5) + 
  ggpubr::theme_pubr() + 
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    axis.title = ggplot2::element_text(size = ggplot2::rel(1.125))
  )

sample_exposures_barplot
ggplot2::ggsave(exposures_sample_barplot_file, sample_exposures_barplot, width = 12, height = 9)
  
```

<br><br>

### Explore: Signature 1 presence


Is there a relationship between the presence of signature 1 in a given sample, and either i) its molecular subtype, or ii) its tumor descriptor?
Because signature 1 is a "normal" signature, we might expect that samples without signature 1 are more progressive/relapse and/or are associated with a specific subtype. 
For background discussion, see [here](https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/1227). 

We see somewhat of a relationship here, but the data sets are highly imbalanced across categories (include 0 and other very low counts), so the statistics are not extremely reliable.



```{r sig1_prop, fig.width = 8, fig.height =4}
# Grab signature1 exposures. 
sig1_exposures <- results_df %>%
  # Since exposures sum to 1 **including unknown**, we just need sig 1's exposure
  dplyr::filter(signature == 1) %>%
  dplyr::select(Kids_First_Biospecimen_ID, 
                sig1_proportion = exposure, 
                cancer_group_display_n) %>%
  # create indicator column of absence/presence
  dplyr::mutate(sig1_present = ifelse(sig1_proportion == 0, "Absent", "Present"),
                # Present should be first level
                sig1_present = forcats::fct_relevel(sig1_present, "Present")) %>%
  # merge in tumor descriptor and subtype information
  dplyr::inner_join(
    dplyr::select(meta, Kids_First_Biospecimen_ID, tumor_descriptor, molecular_subtype)
  )
```

Now, we can make contingency tables and compare signature 1 absence/presence with i) subtypes and ii) descriptors. We run some $\chi^2$-squared tests, **but** there are many low counts in these tables (and some 0!) _so we need to take these tests with a very very big boulder of salt._

```{r contingency_tables}

# Subtypes:
table(sig1_exposures$molecular_subtype, sig1_exposures$sig1_present)
chisq.test(table(sig1_exposures$molecular_subtype, sig1_exposures$sig1_present))

# Tumor descriptor:
table(sig1_exposures$tumor_descriptor, sig1_exposures$sig1_present)
chisq.test(table(sig1_exposures$tumor_descriptor, sig1_exposures$sig1_present))
```
 
We can also explore this using the signature 1 exposure directly instead of a binary absence/presence indicator, specifically across cancer groups of interest, and then specifically within `HGG, H3 wildtype` subtype samples because there is (mostly) enough data to do some light comparison across groups. There is somewhat of a trend here, but sample sizes are small that it's not very robust to perform robust statistics across groupings.




```{r sig1_jitter, fig.width = 12, fig.height = 6}

# This chunk should NOT run in CI. 
if (!(params$is_ci == 1)) {
   
  # Grab the tumor descriptor colors
  tumor_colors <- tumor_df$hex_codes
  names(tumor_colors) <- tumor_df$color_names
  
  # First we can display proportions and tumor desriptor, across cancer display groups
  sig1_tumor_cancergroups <- sig1_exposures %>%
    ggplot2::ggplot() + 
    ggplot2::aes(x = tumor_descriptor, 
                 y = sig1_proportion, 
                 color = tumor_descriptor) + 
    ggplot2::geom_jitter(width = 0.2, alpha = 0.5, size = 2) + 
    # light guiding line representing 0 exposure
    ggplot2::geom_hline(yintercept = 0, size = 0.15) + 
    ggplot2::scale_color_manual(values = tumor_colors) +
    # add in mean guiding point
    ggplot2::stat_summary(color = "black") + 
    ggplot2::facet_wrap(~cancer_group_display_n,nrow = 2) +
    ggplot2::labs(
      x = "Tumor descriptor",
      y = "Signature 1 Weight"
    ) +
    ggpubr::theme_pubr() + 
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 8),
      legend.position = "none"
    )
  
  # We can also viz with molecular subtypes, focusing only on subtypes with decent sample sizes. For now let's say N=15.
  subtypes <- sig1_exposures %>%
    dplyr::count(molecular_subtype) %>%
    dplyr::filter(n >= 15) %>%
    dplyr::pull(molecular_subtype)
  
  sig1_tumor_molsubs <- sig1_exposures %>%
    dplyr::filter(molecular_subtype %in% subtypes) %>%
    tidyr::drop_na(molecular_subtype) %>%
    ggplot2::ggplot() + 
    ggplot2::aes(x = tumor_descriptor, 
                 y = sig1_proportion, 
                 color = tumor_descriptor) + 
    ggplot2::geom_jitter(width = 0.2, alpha = 0.5, size = 1) + 
    # light guiding line representing 0 exposure
    ggplot2::geom_hline(yintercept = 0, size = 0.15) + 
    ggplot2::scale_color_manual(values = tumor_colors) +
    # add in mean guiding point
    ggplot2::stat_summary(color = "black", size = 0.25) + 
    ggplot2::facet_wrap(~molecular_subtype,nrow = 3) +
    ggplot2::labs(
      x = "Tumor descriptor",
      y = "Signature 1 Weight"
    ) +
    ggpubr::theme_pubr() + 
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 6),
      strip.text = ggplot2::element_text(size = 6),
      legend.position = "none"
    )
  
  # Plot the plots
  print(sig1_tumor_cancergroups)
  print(sig1_tumor_molsubs)
  
  # Save the plots
  ggplot2::ggsave(signature1_tumor_cancergroups_file, sig1_tumor_cancergroups, width = 12, height = 6)
  ggplot2::ggsave(signature1_tumor_molecularsubtypes_file, sig1_tumor_molsubs, width = 12, height = 6)
}

```




### Session info

```{r sessioninfo}
sessionInfo()
```