---
title: "Explore hypermutator samples and relationship to signatures"
author: "SJ Spielman (CCDL) and Jo Lynne Rokita (D3B)"
date: "2022"
output:
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---


#### Files and paths


```{r setup}
library(tidyverse)
library(pheatmap)
`%>%` <- dplyr::`%>%`

proj_root_path  <- file.path( rprojroot::find_root(rprojroot::has_dir(".git")) )
analysis_path   <- file.path(proj_root_path, "analyses")
plot_path       <- file.path(analysis_path, "plots", "cns")


# Input files
fitted_file    <- file.path(analysis_path, "mutational-signatures", "results", "deconstructsigs_exposures_merged.tsv")
meta_file      <- file.path(proj_root_path, "data", "pbta-histologies.tsv")
tmb_all_file    <- file.path(proj_root_path, "data", "pbta-snv-consensus-mutation-tmb-all.tsv")
tmb_coding_file <- file.path(proj_root_path, "data", "pbta-snv-consensus-mutation-tmb-coding.tsv")
```

First, we read in data and prepare TMB information.

```{r read_prep_data, warning=FALSE}

# Read in metadata
meta <- readr::read_tsv(meta_file, guess_max = 10000)

# Read in exposures
exposures <- readr::read_tsv(fitted_file)

# Read in and prepare tmb data
tmb_all <- readr::read_tsv(tmb_all_file)
tmb_coding <- readr::read_tsv(tmb_coding_file)

# Merge the all and coding tmb datasets
tmb_data <- tmb_coding %>% 
  dplyr::rename(coding_tmb = tmb) %>%
  dplyr::select(Tumor_Sample_Barcode, coding_tmb) %>%
  dplyr::inner_join(
    dplyr::select(tmb_all,
                  Tumor_Sample_Barcode, 
                  all_tmb = tmb) # rename `tmb` column  --> `all_tmb` to distinguish from coding_tmb
  ) %>%
  dplyr::rename(Kids_First_Biospecimen_ID = Tumor_Sample_Barcode)
```



<br><br>

### Explore: Signatures 3, 18, and MMR2 and mutation burden

GitHub references:

+ https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/1248
+ https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/1227

Which samples have high 3, 18, and/or MMR2, and are these hypermutators? 

Note that this visualization is strictly exploratory and not exported.  

We do **not** observe that samples with high MMR2 or signature 3 exposure have high mutational burdens. 
Moreover, the hypermutators (all are HGG) are unexposed to signature 3 entirely. 
Notably, three of the hypermutators have >=50% exposure to signature 18. 


```{r tmb_sigs, fig.width = 10, fig.height = 4}
sigs_of_interest <- c(3, 18, "MMR2")
tmb_sigs_df <- exposures %>%
  dplyr::select(signature, exposure, Kids_First_Biospecimen_ID, cancer_group_display) %>%
  dplyr::distinct() %>%
  dplyr::inner_join(tmb_data) %>%
  tidyr::gather(dplyr::contains("tmb"), key = "tmb_type", value = "tmb")

# plot TMB against exposure to those three signatures
tmb_sigs_df %>%
  dplyr::filter(signature %in% sigs_of_interest) %>%
  ggplot2::ggplot() + 
  ggplot2::aes(x = exposure, y = tmb, fill = cancer_group_display) +
  ggplot2::geom_point(shape = 21, color ="black") + 
  # guiding line for majority of signature exposure, >=0.5
  ggplot2::geom_vline(xintercept = 0.5, color = "red") +
  # gives a warning because insufficient colors - but OK since points of interest are colored!
  colorblindr::scale_fill_OkabeIto() +
  ggplot2::labs(x = "Exposure to given signature") +
  ggplot2::facet_grid(tmb_type ~ signature, scales = "free")
```


Do the hypermutators also show dysregulated TP53?

We define hypermutant samples as > 10 Mut/Mb, and ultrahypermutant samples as >100 Mut/Mb, from reference [Campbell et al., 2017](https://doi.org/10.1016/j.cell.2017.09.048).

The output below shows hypermutator samples and...

+ their molecular subtypes (which implicitly gives the cancer group)
  + this also gives the **TP53 status**
+ their tumor descriptor 
+ their _coding_ TMB
+ their exposures to all signatures (with those of interest appearing first: 3, 18, MMR2)




```{r tp53_sigs}
hypermutator_df <- tmb_sigs_df %>%
  # filter to the hypermutators
  dplyr::filter(tmb_type == "coding_tmb", 
                tmb >= 10, 
                signature != "Other") %>% 
  dplyr::select(-tmb_type) %>%
  # add in molecular subtype information, **which contains tp53 status** AND cancer group
  dplyr::inner_join(
    dplyr::select(meta, Kids_First_Biospecimen_ID, molecular_subtype, tumor_descriptor)
  ) %>%
  tidyr::spread(signature, exposure) %>%
  # it's all coding so can remove type column, but a rename is helpful
  dplyr::rename(coding_tmb = tmb) %>%
  # order columns so sigs of interest come first
  dplyr::select(Kids_First_Biospecimen_ID, molecular_subtype, tumor_descriptor, coding_tmb, sigs_of_interest, dplyr::everything()) %>%
  # we don't need cancer_group_display, the info is in molecular_subtype
  dplyr::select(-cancer_group_display) %>%
  # arrange in order of burden
  dplyr::arrange(-coding_tmb)

hypermutator_df
```

A couple interesting observations in the table above:

+ _No samples_ are exposed to signature 3, 8, or N6

+ The DMG hypermutator sample (TMB=35.89) is mostly exposed to [signature 11](https://signal.mutationalsignatures.com/explore/referenceCancerSignature/40). This signature was thought to be associated with temozolomide, except that original evidence has not been reproduced. 

+ The HGG hypermutator sample with lowest TMB (TMB=15.48) is primarily exposed to signature 1, which is a normal signature.

### Make heatmap for mutational signatures for patients with hypermutant tumors
```{r}
# Create signature matrix for only patients withh hypermutant samples
# Gather all BS_IDs (soi = "samples of interest"), whether hypermutant or not, for pts with hypermutant samples

soi_df <- meta %>%
  filter(Kids_First_Biospecimen_ID %in% hypermutator_df$Kids_First_Biospecimen_ID) %>%
  select(Kids_First_Participant_ID) %>%
  left_join(meta) %>%
  #keep only WGS samples; remove normals and cell lines
  filter(experimental_strategy == "WGS" & !is.na(pathology_diagnosis) & composition != "Derived Cell Line") %>%
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, tumor_descriptor) %>%
  unique() %>%
  arrange(Kids_First_Participant_ID)
         
sigs <- exposures %>%
  filter(signature != "Other") %>%
  tidyr::spread(signature, exposure) %>%
  select(-broad_histology, -cancer_group_display_n, -n, -cancer_group_display, -cancer_group_hex, -tumor_descriptor) %>%
  filter(Kids_First_Biospecimen_ID %in% soi_df$Kids_First_Biospecimen_ID) %>%
  as.data.frame()
rownames(sigs) <- sigs$Kids_First_Biospecimen_ID

sigs$Kids_First_Biospecimen_ID <- NULL
sigs <- as.matrix(sigs)
sapply(sigs, class)

# Create annotation dataframe
anno <- tmb_coding %>%
  mutate(Kids_First_Biospecimen_ID = Tumor_Sample_Barcode) %>%
  select(Kids_First_Biospecimen_ID, tmb) %>%
  # add a column for mutator
  mutate(
    Mutator = case_when(
      tmb < 10 ~ "Normal",
      tmb >= 10 & tmb < 100 ~ "Hypermutant",
      tmb >= 100 ~ "Ultra-hypermutant")
  ) %>%
  right_join(soi_df, by = "Kids_First_Biospecimen_ID") %>%
  unique() %>%
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, tumor_descriptor, Mutator) %>%
  as.data.frame()

# set rownames to match matrix, remove BS_ID
rownames(anno) <- anno$Kids_First_Biospecimen_ID
anno$Kids_First_Biospecimen_ID <- NULL

# Specify colors
ann_colors = list(
  # mutator color from [this PR](https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/1280/files#diff-bc8134008706f38e7795b59696928058d123e44b4172df68105a26ad8aa0a398R298-R301) 
  Mutator = c(Normal = "grey40",
              Hypermutant = "orange",
              `Ultra-hypermutant` = "red"),
  # tumor descriptor colors from [palette](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/master/figures/palettes/tumor_descriptor_palette.tsv) 
  tumor_descriptor = c(`Initial CNS Tumor` = "#709AE1", 
                       `Progressive` = "#38736D"))
# Make heatmap
pheatmap(sigs[, c("1", "3", "8", "11", "18", "19", "MMR2", "N6")], 
         cluster_rows = F, 
         cluster_cols = F,
         border_color = "white",
         color=colorRampPalette(c("#f1f1f1", "#67001f"))(50),
         gaps_row = c(2,3,5,6),
         annotation_row = anno,
         annotation_colors = ann_colors)

```


### Session info

```{r sessioninfo}
sessionInfo()
```