---
title: "Explore hypermutator samples and relationship to signatures"
author: "SJ Spielman (CCDL) and Jo Lynne Rokita (D3B)"
date: "2022"
output:
  html_notebook:
  toc: true
editor_options: 
  chunk_output_type: inline
params:
  is_ci: 0
---
  
  
#### Files and paths
  
  
```{r setup}
library(tidyverse)
library(ComplexHeatmap)
library(colorblindr)

`%>%` <- dplyr::`%>%`

proj_root_path  <- file.path( rprojroot::find_root(rprojroot::has_dir(".git")) )
analysis_path   <- file.path(proj_root_path, "analyses")
plot_path       <- file.path(analysis_path, "mutational-signatures", "plots", "cns")
results_path       <- file.path(analysis_path, "mutational-signatures", "results")

# Input files
fitted_file    <- file.path(analysis_path, "mutational-signatures", "results", "deconstructsigs_exposures_merged.tsv")
meta_file      <- file.path(proj_root_path, "data", "pbta-histologies.tsv")
tmb_all_file    <- file.path(proj_root_path, "data", "pbta-snv-consensus-mutation-tmb-all.tsv")
tmb_coding_file <- file.path(proj_root_path, "data", "pbta-snv-consensus-mutation-tmb-coding.tsv")
td_palette_file <- file.path(proj_root_path, "figures", "palettes", "tumor_descriptor_palette.tsv")
bin_palette_file <- file.path(proj_root_path, "figures", "palettes", "binary_color_palette.tsv")

# Output files
heatmap_file <- file.path(plot_path, "hypermutator_sigs_heatmap.pdf")
heatmap_legends_file <- file.path(plot_path, "hypermutator_sigs_heatmap_legends.pdf")
sig_matrix_file <- file.path(results_path, "hypermutator_sig_matrix.tsv")
```

First, we read in data and prepare TMB information.

```{r read_prep_data, warning=FALSE}
# Read in tumor descriptor and palettes
td_palette <- readr::read_tsv(td_palette_file)
bin_palette <- readr::read_tsv(bin_palette_file) %>%
  mutate(color_names = case_when(
        color_names == "binary_1" ~ "Solid Tissue",
        color_names == "binary_2" ~ "Derived Cell Line")) %>%
  filter(!is.na(color_names))

# Read in metadata
meta <- readr::read_tsv(meta_file, guess_max = 10000)

# Read in exposures
exposures <- readr::read_tsv(fitted_file)

# Read in and prepare tmb data
tmb_all <- readr::read_tsv(tmb_all_file)
tmb_coding <- readr::read_tsv(tmb_coding_file)

# Merge the all and coding tmb datasets
tmb_data <- tmb_coding %>% 
  dplyr::rename(coding_tmb = tmb) %>%
  dplyr::select(Tumor_Sample_Barcode, coding_tmb) %>%
  dplyr::inner_join(
    dplyr::select(tmb_all,
                  Tumor_Sample_Barcode, 
                  all_tmb = tmb) # rename `tmb` column  --> `all_tmb` to distinguish from coding_tmb
  ) %>%
  dplyr::rename(Kids_First_Biospecimen_ID = Tumor_Sample_Barcode)
```



<br><br>
  
  ### Explore: Signatures 3, 18, and MMR2 and mutation burden
  
  GitHub references:
  
  + https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/1248
+ https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/1227

Which samples have high 3, 18, and/or MMR2, and are these hypermutators? 
  
  Note that this visualization is strictly exploratory and not exported.  

We do **not** observe that samples with high MMR2 or signature 3 exposure have high mutational burdens. 
Moreover, the hypermutators (all are HGG) are unexposed to signature 3 entirely. 
Notably, three of the hypermutators have >=50% exposure to signature 18. 


```{r tmb_sigs, fig.width = 10, fig.height = 4}
sigs_of_interest <- c(3, 18, "MMR2")
tmb_sigs_df <- exposures %>%
  dplyr::select(signature, exposure, Kids_First_Biospecimen_ID, cancer_group_display) %>%
  dplyr::distinct() %>%
  dplyr::inner_join(tmb_data) %>%
  tidyr::gather(dplyr::contains("tmb"), key = "tmb_type", value = "tmb")

# plot TMB against exposure to those three signatures
tmb_sigs_df %>%
  dplyr::filter(signature %in% sigs_of_interest) %>%
  ggplot2::ggplot() + 
  ggplot2::aes(x = exposure, y = tmb, fill = cancer_group_display) +
  ggplot2::geom_point(shape = 21, color ="black") + 
  # guiding line for majority of signature exposure, >=0.5
  ggplot2::geom_vline(xintercept = 0.5, color = "red") +
  # gives a warning because insufficient colors - but OK since points of interest are colored!
  colorblindr::scale_fill_OkabeIto() +
  ggplot2::labs(x = "Exposure to given signature") +
  ggplot2::facet_grid(tmb_type ~ signature, scales = "free")
```


Do the hypermutators also show dysregulated TP53?
  
  We define hypermutant samples as > 10 Mut/Mb, and ultrahypermutant samples as >100 Mut/Mb, from reference [Campbell et al., 2017](https://doi.org/10.1016/j.cell.2017.09.048).

The output below shows hypermutator samples and...

+ their molecular subtypes (which implicitly gives the cancer group)
+ this also gives the **TP53 status**
  + their tumor descriptor 
+ their _coding_ TMB
+ their exposures to all signatures (with those of interest appearing first: 3, 18, MMR2)




```{r tp53_sigs}
hypermutator_df <- tmb_sigs_df %>%
  # filter to the hypermutators
  dplyr::filter(tmb_type == "coding_tmb", 
                tmb >= 10, 
                signature != "Other") %>% 
  dplyr::select(-tmb_type) %>%
  # add in molecular subtype information, **which contains tp53 status** AND cancer group
  dplyr::inner_join(
    dplyr::select(meta, Kids_First_Biospecimen_ID, molecular_subtype, tumor_descriptor, composition)
  ) %>%
  tidyr::spread(signature, exposure) %>%
  # it's all coding so can remove type column, but a rename is helpful
  dplyr::rename(coding_tmb = tmb) %>%
  # order columns so sigs of interest come first
  dplyr::select(Kids_First_Biospecimen_ID, molecular_subtype, tumor_descriptor, composition, coding_tmb, sigs_of_interest, dplyr::everything()) %>%
  # we don't need cancer_group_display, the info is in molecular_subtype
  dplyr::select(-cancer_group_display) %>%
  # arrange in order of burden
  dplyr::arrange(-coding_tmb)

hypermutator_df
```

A couple interesting observations in the table above:
  
  + _No samples_ are exposed to signature 3, 8, or N6

+ The DMG hypermutator sample (TMB=35.89) is mostly exposed to [signature 11](https://signal.mutationalsignatures.com/explore/referenceCancerSignature/40). This signature was thought to be associated with temozolomide, except that original evidence has not been reproduced. 

+ The HGG hypermutator sample with lowest TMB (TMB=15.48) is primarily exposed to signature 1, which is a normal signature.

### Make heatmap for mutational signatures for patients with hypermutant tumors
```{r}
# This chunk should NOT run in CI. 
if (!(params$is_ci == 1)) {
  
  
  # Create signature matrix for only patients withh hypermutant samples
  # Gather all BS_IDs (soi = "samples of interest"), whether hypermutant or not, for pts with hypermutant samples
  
  soi_df <- meta %>%
    filter(Kids_First_Biospecimen_ID %in% hypermutator_df$Kids_First_Biospecimen_ID) %>%
    select(Kids_First_Participant_ID) %>%
    left_join(meta) %>%
    #keep only WGS samples; remove normals; keep cell lines
    filter(experimental_strategy == "WGS" & !is.na(pathology_diagnosis)) %>%
    select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, tumor_descriptor, composition) %>%
    unique() %>%
    arrange(Kids_First_Participant_ID)
  
  sigs <- exposures %>%
    filter(signature != "Other") %>%
    tidyr::spread(signature, exposure) %>%
    select(-broad_histology, -cancer_group_display_n, -n, -cancer_group_display, -cancer_group_hex, -tumor_descriptor) %>%
    filter(Kids_First_Biospecimen_ID %in% soi_df$Kids_First_Biospecimen_ID) %>%
    as.data.frame() %>%
    arrange(Kids_First_Biospecimen_ID) %>%
    write_tsv(sig_matrix_file)
  
  # add spaces at the end of BS_ids to make a space between the heatmap and legend
  rownames(sigs) <- paste0(sigs$Kids_First_Biospecimen_ID, "     ")
  
  sigs$Kids_First_Biospecimen_ID <- NULL
  sigs <- as.matrix(sigs)
  
  # Create annotation dataframe
  anno <- tmb_coding %>%
    mutate(Kids_First_Biospecimen_ID = Tumor_Sample_Barcode) %>%
    select(Kids_First_Biospecimen_ID, tmb) %>%
    # add a column for mutation status
    mutate(
      `Mutation status` = case_when(
        tmb < 10 ~ "Normal",
        tmb >= 10 & tmb < 100 ~ "Hypermutant",
        tmb >= 100 ~ "Ultra-hypermutant")
    ) %>%
    right_join(soi_df, by = "Kids_First_Biospecimen_ID") %>%
    unique() %>%
    # mutate names for annotation
    mutate(Patient = Kids_First_Participant_ID,
           `Phase of Therapy` = tumor_descriptor,
           Composition = composition) %>%
    select(Patient, Kids_First_Biospecimen_ID, `Phase of Therapy`, Composition, `Mutation status`) %>%
    as.data.frame() %>%
    arrange(Kids_First_Biospecimen_ID)

  # set rownames to match matrix, remove BS_ID
  rownames(anno) <- paste0(anno$Kids_First_Biospecimen_ID, "     ")
  anno$Kids_First_Biospecimen_ID <- NULL
  
  # Add tumor descriptor palette from dataframe
  td_palette_filtered <- td_palette %>%
    filter(color_names %in% anno$`Phase of Therapy`)
  td_palette_col <- as.character(td_palette_filtered$hex_codes)
  names(td_palette_col) <- td_palette_filtered$color_names
  
  pt_palette <- colorRampPalette(palette_OkabeIto)(length(unique(anno$Patient)))
  names(pt_palette) <- unique(anno$Patient)
  
  comp_palette <- as.character(bin_palette$hex_codes)
  names(comp_palette) <- unique(anno$Composition)
  
  # Specify colors
  ann_colors = list(
    # mutator color from [this PR](https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/1280/files#diff-bc8134008706f38e7795b59696928058d123e44b4172df68105a26ad8aa0a398R298-R301) 
    `Mutation status` = c(Normal = "grey40",
                          Hypermutant = "orange",
                          `Ultra-hypermutant` = "red"),
    Composition = comp_palette,
    `Phase of Therapy` = td_palette_col,
    Patient = pt_palette)
  
  # Heatmap annotation
  row_anno = rowAnnotation(df = anno,
                           col = ann_colors, show_legend=FALSE)
  
  # Make heatmap without legends
  heat_plot <- Heatmap(sigs[, c("1", "3", "8", "11", "18", "19", "MMR2", "N6")], 
                       name = "Signature weights",
                       col = colorRampPalette(c("#f1f1f1", "#2166ac"))(50),
                       cluster_rows = FALSE,
                       show_row_names = TRUE,
                       show_heatmap_legend=FALSE,
                       cluster_columns = FALSE,
                       right_annotation = row_anno,
                       rect_gp = gpar(col = "white"),
                       row_split = anno$Patient,
                       row_title = NULL, 
                       column_title = "RefSig Mutational Signature Weights", 
                       column_title_side = "bottom")
  
  heat_plot
  
  # Make separate legends to compile in Illustrator
  patient_legend <- Legend(
    labels = names(ann_colors$Patient),
    legend_gp = gpar(fill = ann_colors$Patient),
    title = "Patient"
  )
  
  mutator_legend <- Legend(
    labels = names(ann_colors$`Mutation status`),
    legend_gp = gpar(fill = ann_colors$`Mutation status`),
    title = "Mutation status"
  )
  
  phase_legend <- Legend(
    labels = names(ann_colors$`Phase of Therapy`),
    legend_gp = gpar(fill = ann_colors$`Phase of Therapy`),
    title = "Phase of Therapy"
  )
  
  comp_legend <- Legend(
    labels = names(ann_colors$Composition),
    legend_gp = gpar(fill = ann_colors$Composition),
    title = "Composition"
  )
  
  weights_legend <- color_mapping_legend(heat_plot@matrix_color_mapping, plot = FALSE, 
                                         legend_direction = "horizontal")
  
  heat_legends <- packLegend(patient_legend,
             mutator_legend,
             phase_legend,
             comp_legend,
             weights_legend,
             direction = "horizontal",
             column_gap = unit(0.75, "cm")
  ) 

  
  # save heatmap 
  pdf(heatmap_file, width = 8, height = 4)
  print(heat_plot)
  dev.off()

  # save heatmap legends
  pdf(heatmap_legends_file, width = 8, height = 2)
  draw(heat_legends)
  dev.off()  
 }
```


### Session info

```{r sessioninfo}
sessionInfo()
```