---
title: "Mutational Signatures"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: C. Savonen for ALSF CCDL
date: 2019
---

**Purpose:**

Calculate and plot mutational signatures for all samples using [COSMIC signatures](https://cancer.sanger.ac.uk/cosmic) and 
[Alexandrov et al, 2013](https://www.ncbi.nlm.nih.gov/pubmed/23945592) mutational signatures. 

### Summary of Findings:

Coming soon. 

#### Usage

To run this from the command line, use:
```
Rscript -e "rmarkdown::render('analyses/mutational-signatures/mutation_signatures.Rmd', 
                              clean = TRUE)"
```

_This assumes you are in the top directory of the repository._

## Setup

#### Packages and functions

Import necessary functions.

```{r}
if (!("deconstructSigs" %in% installed.packages())) {
  BiocManager::install("BSgenome.Hsapiens.UCSC.hg38")
  install.packages("deconstructSigs")
}
# Magrittr pipe
`%>%` <- dplyr::`%>%`

# Import specialized functions
source(file.path("util", "mut_sig_functions.R"))

# Load this library
library(deconstructSigs)
```

Set up directory paths. 

```{r}
data_dir <- file.path("..", "..", "data")
results_dir <- "results"
plots_dir <- "plots"
cosmic_plots <- file.path(plots_dir, "cosmic")
nature_plots <- file.path(plots_dir, "nature")
```

Make new directories for the results. 

```{r}
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if (!dir.exists(cosmic_plots)) {
  dir.create(cosmic_plots, recursive = TRUE)
}
if (!dir.exists(nature_plots)) {
  dir.create(nature_plots, recursive = TRUE)
}
```

## Read in data

Download and read in the consensus MAF file. 

```{r}
# Declare file path for consensus file
consensus_file <- file.path(data_dir, "snv-consensus_11122019", "consensus_mutation.maf.tsv")
```

Read in the consensus MAF file. 

```{r}
# Read in the file
maf <- data.table::fread(consensus_file)
```

Read in the metadata. 

```{r}
metadata_df <- readr::read_tsv(file.path(data_dir, "pbta-histologies.tsv")) %>% 
  dplyr::rename(Tumor_Sample_Barcode = "Kids_First_Biospecimen_ID") %>%
  dplyr::select("Tumor_Sample_Barcode", "experimental_strategy", "short_histology")
```

Read in this list so we can make sure we keep only primary tumors for the 
grouped bar plots.

```{r}
ind_samples <- readr::read_tsv(file.path(
  data_dir,
  "independent-specimens.wgswxs.primary.tsv"
))
```

Read in the WGS and WXS regions so they can be used for the Mb denominator. 

```{r}
# Set up BED region files for TMB calculations
wgs_bed <- readr::read_tsv(file.path(data_dir, "WGS.hg38.strelka2.unpadded.bed"),
  col_names = FALSE
)
wxs_bed <- readr::read_tsv(file.path(data_dir, "WXS.hg38.100bp_padded.bed"),
  col_names = FALSE
)

# Calculate size of genome surveyed
# These files are BED files where the third column is the End position and
# the second column is the Start position.
# So End - Start gives the size of each range. Sum the gives the total size in bp.
wgs_size <- sum(wgs_bed[, 3] - wgs_bed[, 2])
wxs_size <- sum(wxs_bed[, 3] - wxs_bed[, 2])
```

## Set up data

Determine how many mutations we have per sample.

```{r}
mut_per_sample <- maf %>%
  dplyr::group_by(Tumor_Sample_Barcode) %>%
  dplyr::tally() %>%
  dplyr::arrange(n)

summary(mut_per_sample$n)
```

Graph this.

```{r}
ggplot2::ggplot(mut_per_sample, ggplot2::aes(x = n, geom = "density")) +
  ggplot2::geom_density() +
  ggplot2::theme_classic()
```

Make mutation data into `deconstructSigs` input format.

```{r}
# Convert to deconstructSigs input
sigs_input <- mut.to.sigs.input(
  mut.ref = maf,
  sample.id = "Tumor_Sample_Barcode",
  chr = "Chromosome",
  pos = "Start_Position",
  ref = "Reference_Allele",
  alt = "Allele",
  bsg = BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38
)
```

Add total mutations per sample. 

```{r}
# Count the total number of signature mutations for each sample
total_muts <- apply(sigs_input, 1, sum)
```

## Determine Signatures for COSMIC and Alexandrov et al, 2013

Get list of tumor sample ids. 

```{r}
tumor_sample_ids <- maf %>%
  dplyr::filter(Tumor_Sample_Barcode %in% rownames(sigs_input)) %>%
  dplyr::distinct(Tumor_Sample_Barcode) %>%
  dplyr::pull(Tumor_Sample_Barcode)
```

Get [COSMIC signatures](https://cancer.sanger.ac.uk/cosmic) for each sample. 
This step will take some time. 

```{r}
sample_sigs_cosmic <- lapply(tumor_sample_ids, function(sample_id) {
  # Determine the signatures contributing to the sample
  whichSignatures(
    tumor.ref = sigs_input,
    signatures.ref = signatures.cosmic,
    sample.id = sample_id,
    contexts.needed = TRUE
  )
})
# Bring along the names
names(sample_sigs_cosmic) <- tumor_sample_ids
```

Get [Alexandrov et al, 2013](https://www.ncbi.nlm.nih.gov/pubmed/23945592) signatures for each sample. 

```{r}
sample_sigs_nature <- lapply(tumor_sample_ids, function(sample_id) {
  # Determine the signatures contributing to the sample
  whichSignatures(
    tumor.ref = sigs_input,
    signatures.ref = signatures.nature2013,
    sample.id = sample_id,
    contexts.needed = TRUE
  )
})
# Bring along the names
names(sample_sigs_nature) <- tumor_sample_ids
```

### COSMIC signature plots

```{r}
sample_mut_sig_plot(sample_sigs_cosmic,
  label = "cosmic",
  output_dir = file.path(cosmic_plots, "individual_mutation_sig")
)
```

### Alexandrov et al, 2013 signature plots

```{r}
sample_mut_sig_plot(sample_sigs_nature,
  label = "nature",
  output_dir = file.path(nature_plots, "individual_mutation_sig")
)
```

### Calculate the mutations per Mb for each signature

Do this for COSMIC mutation signatures.

```{r}
# Calculate mutations per signature
cosmic_sigs_df <- calc_mut_per_sig(sample_sigs_cosmic,
  muts_per_sample = total_muts,
  wgs_genome_size = wgs_size,
  wxs_exome_size = wxs_size,
  metadata = metadata_df
) 

# Write this to a file  
readr::write_tsv(cosmic_sigs_df, 
                 file.path(results_dir, "cosmic_signatures_results.tsv"))

# Print out a preview
cosmic_sigs_df
```

Do this for Alexandrov et al, 2013 mutation signatures.

```{r}
# Calculate mutations per signature
nature_sigs_df <- calc_mut_per_sig(sample_sigs_nature,
  muts_per_sample = total_muts,
  wgs_genome_size = wgs_size,
  wxs_exome_size = wxs_size,
  metadata = metadata_df
) 

# Write this to a file  
readr::write_tsv(nature_sigs_df, 
                 file.path(results_dir, "nature_signatures_results.tsv"))

# Print out a preview
nature_sigs_df
```

## Mutation signature bubble matrix by histology groups

```{r}
bubble_matrix_plot(cosmic_sigs_df, label = "COSMIC Signatures")
```

```{r}
ggplot2::ggsave(file.path(cosmic_plots, paste0("bubble_matrix_cosmic_mutation_sig.png")), width = 30, height = 20, units = "cm")
```

```{r}
bubble_matrix_plot(nature_sigs_df, label = "Alexandrov et al, 2013 signatures")
```

```{r}
ggplot2::ggsave(file.path(nature_plots, paste0("bubble_matrix_nature_mutation_sig.png")), width = 30, height = 20, units = "cm")
```

## Mutation signature grouped bar plots for each histology group

We will make these plots for primary tumor samples only. 
Let's make these for COSMIC mutation signatures first. 

```{r}
# Keep only primary tumors
cosmic_sigs_primary <- cosmic_sigs_df %>%
  dplyr::filter(Tumor_Sample_Barcode %in% ind_samples$Kids_First_Biospecimen_ID)

# Make grouped bar plots
lapply(unique(cosmic_sigs_primary$short_histology),
  grouped_sig_barplot,
  sig_num_df = cosmic_sigs_primary,
  output_dir = file.path(cosmic_plots, "signature_grouped_barplots"), 
  label = "cosmic"
)
```

Make these plots for Alexandrov et al, 2013 signatures. 

```{r}
# Keep only primary tumors
nature_sigs_primary <- nature_sigs_df %>%
  dplyr::filter(Tumor_Sample_Barcode %in% ind_samples$Kids_First_Biospecimen_ID)

# Make grouped bar plots
lapply(unique(nature_sigs_primary$short_histology),
  grouped_sig_barplot,
  sig_num_df = nature_sigs_primary,
  output_dir = file.path(nature_plots, "signature_grouped_barplots"),
  label = "nature"
)
```

## Session Info

```{r}
sessionInfo()
```
