---
title: "Subgtyping-chordoma"
output: html_notebook
author: Mateusz Koptyra 
date: 20191121
---

```{r}
library(dplyr)
library(readr)
```
```{r}
histologies_df <- read_tsv("../../data/pbta-histologies.tsv")
```
```{r}
chordoma_samples <- histologies_df %>%
  filter(short_histology=="Chordoma") %>% 
  pull(Kids_First_Biospecimen_ID)
```

```{r}
focal_cn_df <- read_tsv("../focal-cn-file-preparation/results/controlfreec_annotated_cn_autosomes.tsv.gz")
```

```{r}
chordoma_loss <- focal_cn_df %>% 
  filter(biospecimen_id %in% chordoma_samples, gene_symbol=="SMARCB1")

chordoma_loss 
```

```{r}

```


```{r}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

