---
title: "Select pathology diagnoses for inclusion"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2020
---

Code for this notebook was adapted from the `analyses/molecular-subtyping-HGG/00-HGG-select-pathology-dx.Rmd`(https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/f65c803706d9c1eea7f2892cb8b638f3d47d15c6/analyses/molecular-subtyping-HGG/00-HGG-select-pathology-dx.Rmd) notebook.

## Background

Originally, we subtyped tumors in this module if the specimen satisfied one of the following criteria:

1. An RNA-seq biospecimen sample includes a _TTYH1_ fusion (5' partner) [per this comment](https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/401#issuecomment-573669727).
2. Any sample with "Embryonal tumor" in the `broad_histology` column of the metadata `pbta-histologies.tsv` that is not labeled "Medulloblastoma" or "Atypical Teratoid Rhabdoid Tumor (ATRT)" in `pathology_diagnosis` or `integrated_diagnosis` columns [per this comment](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/251#issuecomment-568220913).

In an upcoming release, `integrated_diagnosis`, which can be updated as the result of subtyping, will be used to populate the `broad_histology` column (see [#748](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/748)).
Thus, molecular subtyping modules need to be upstream of `broad_histology` and use the `pathology_diagnosis` and `pathology_free_text_diagnosis` fields.

Per a [related github comment](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/752#issuecomment-697000066), the following criteria should be used to subset samples based on `pathology_diagnosis` and `pathology_free_text_diagnosis` for this module:

1. Any sample with "Supratentorial or Spinal Cord PNET" in the `pathology_diagnosis` column of the metadata `pbta-histologies.tsv`.
2. Any sample with "Neuroblastoma" in the `pathology_diagnosis` column, where `primary_site` does not contain "Other locations NOS", `pathology_free_text_diagnosis` does not contain "peripheral" or "metastatic".
3. Any sample with "Other" in the `pathology_diagnosis` column of the metadata, and with "embryonal tumor with multilayer rosettes, ros (who grade iv)", "embryonal tumor, nos, congenital type", "ependymoblastoma" or "medulloepithelioma" in the `pathology_free_text_diagnosis` column.

## Set up

```{r}
library(tidyverse)
```

### Directories and files

We're going to tie this to a specific release.

```{r}
data_dir <- file.path("..", "..", "data", "release-v17-20200908")
histologies_file <- file.path(data_dir, "pbta-histologies.tsv")
```

## Read in data

```{r}
histologies_df <- read_tsv(histologies_file)
```

## Explore the pathology diagnoses

### `pathology_diagnosis == Supratentorial or Spinal Cord PNET`

We want to filter rows on the basis of the presence of the values "Supratentorial or Spinal Cord PNET" and "Neuroblastoma" in the `pathology_diagnosis` column, [per this comment](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/752#issuecomment-697000066).  

```{r}
histologies_df %>%
  filter(
      pathology_diagnosis %in% c("Supratentorial or Spinal Cord PNET",
                                 "Neuroblastoma")
  ) %>%
  count(pathology_diagnosis) %>%
  arrange(desc(n))
```

When filtering, we will want to keep CNS neuroblastoma samples and not peripheral neuroblastoma samples for this module.

### Can we use `pathology_diagnosis` alone to filter out metastatic neuroblastoma tumors?

There are instances of `Metastatic secondary tumors;Neuroblastoma` in `pathology_diagnosis`, which should be treated as metastatic tumors in this module ([GitHub comment](https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/788#issuecomment-705141520)). 

If we filter to `pathology_diagnosis === "Neuroblastoma"`, does that remove all specimens marked with metastatic?

```{r}
histologies_df %>%
  filter(
    pathology_diagnosis == "Neuroblastoma"
  ) %>%
  count(pathology_diagnosis, pathology_free_text_diagnosis) %>%
  arrange(desc(n))
```

No, it looks like we need to account for "metastatic" in the `pathology_free_text_diagnosis` column as we still capture "metastatic neuroblastoma" samples when we filter on `pathology_diagnosis` only.

We will also need to select free text terms for inclusions because above we can see values such as `peripheral neuroblastoma` that we do not want to include in this module.

## Pathology diagnosis strings for inclusion

These are the terms that we'll use to look for exact matches in `pathology_diagnosis`.
 
```{r}
path_dx_terms <- c(
  "Supratentorial or Spinal Cord PNET"
) # "neuroblastoma" will also need to be detected in the `pathology_diagnosis` column, but the logic around it is a bit more complex so we will handle this term separately
```

## Pathology free text diagnosis strings for inclusion

These are the terms that we'll use to look for exact matches in the `pathology_free_text_diagnosis` column.

```{r}
free_text_dx_terms <- c(
  "embryonal tumor with multilayer rosettes, ros (who grade iv)",
  "embryonal tumor, nos, congenital type",
  "ependymoblastoma",
  "medulloepithelioma",
  "medullooepithelioma" # It was noted that this misspelling was included in the `pathology_free_text_diagnosis` column of the histologies file so we will add it to the JSON file [per this comment](https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/788#issuecomment-700717514)
)
```

## Filter and join data

Let's take a look at a first attempt using these terms as described above.

First, we will create an object filtered to include only samples with exact matches for the `path_dx_terms` (which is currently "Supratentorial or Spinal Cord PNET") in the `pathology_diagnosis` column.

```{r}
pnet_df <- histologies_df %>%
  # Filter for exact match of the defined pathology diagnosis term in `path_dx_terms`
  filter(pathology_diagnosis %in% path_dx_terms)
```

Now we will create an objected filtered to include any sample with "Other" in the `pathology_diagnosis` column of the metadata, and with the `free_text_dx_terms` in the `pathology_free_text_diagnosis` column.

```{r}
pathology_free_text_df <- histologies_df %>%
  # Filter for exact matches of `pathology_diagnosis == "Other"` and defined pathology free text diagnosis terms in `free_text_dx_terms`
  filter(pathology_diagnosis == "Other",
         pathology_free_text_diagnosis %in% free_text_dx_terms)
```

Any sample with "Neuroblastoma" in the `pathology_diagnosis` column, where `primary_site` does not contain "Other locations NOS", `pathology_free_text_diagnosis` does not contain "peripheral", and `pathology_diagnosis` or `pathology_free_text_diagnosis` does not contain "metastatic".

```{r}
neuroblastoma_df <- histologies_df %>%
  filter(
    pathology_diagnosis == "Neuroblastoma",
    str_detect(str_to_lower(primary_site), "other locations nos", negate = TRUE),
    str_detect(
      str_to_lower(pathology_free_text_diagnosis),
      "peripheral|metastatic",
      negate = TRUE
    )
  )
neuroblastoma_df
```

Now let's bind all of the information in our filtered data frames together into one object.

```{r}
# Binding all of the rows
filtered_on_dx_df <- bind_rows(pnet_df,
                               pathology_free_text_df,
                               neuroblastoma_df) %>%
  select(
    Kids_First_Biospecimen_ID,
    sample_id,
    Kids_First_Participant_ID,
    pathology_diagnosis,
    pathology_free_text_diagnosis,
    primary_site,
    broad_histology
  )

filtered_on_dx_df
```

## Session Info

```{r}
sessionInfo()
```
