---
title: "Molecularly Subtyping Embryonal Tumors - Which samples to include?"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Stephanie J. Spielman and Jaclyn Taroni for ALSF CCDL
date: 2019
---

This notebook identifies samples to include in subset files for the purpose of molecularly subtyping embryonal tumors ([`AlexsLemonade/OpenPBTA-analysis#251`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/251)).

This closed pull request is also relevant to the task at hand: [`AlexsLemonade/OpenPBTA-analysis#401`](https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/401).

To summarize - we will identify biospecimen IDs that should be included for downstream analysis using the following criteria:

1. An RNA-seq biospecimen includes a _TTYH1_ fusion (5' partner).
   We can consume the files from the [`fusion-summary` module](../fusion-summary/) for this step.
   (See also: [this comment on `AlexsLemonade/OpenPBTA-analysis#401`](https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/401#issuecomment-573669727).)
2. Any sample with `"Embryonal tumor"` in the `broad_histology` column of `pbta-histologies.tsv` that is not labeled `"Medulloblastoma"` or `"Atypical Teratoid Rhabdoid Tumor (ATRT)"` in `disease_type_old` per [this comment on `AlexsLemonade/OpenPBTA-analysis#251`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/251#issuecomment-568220913).
   We'll look at the `disease_type_new` column as well.

## Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

```
Rscript -e "rmarkdown::render('analyses/molecular-subtyping-embryonal/01-samples-to-subset.Rmd', clean = TRUE)"
```

## Set Up

### Libraries and functions

```{r}
library(tidyverse)
```

### Directories

```{r}
data_dir <- file.path("..", "..", "data")
# the directory that contains the fusion summary files of interest
fusion_summary_dir <- file.path("..", "fusion-summary", "results")

results_dir <- "results"
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

### Read in relevant files

```{r message=FALSE}
histologies_df <- read_tsv(file.path(data_dir, "pbta-histologies.tsv"))
fusion_summary_df <- read_tsv(file.path(fusion_summary_dir,
                                        "fusion_summary_embryonal_foi.tsv"))
```

### Output file

```{r}
output_file <- file.path(results_dir, "biospecimen_ids_embryonal_subtyping.tsv")
```

## Identify relevant samples

### _TTYH1_ fusions

We'll use the fusion summary file to find samples that contain _TTYH1_.
We're only interested in fusions where _TTYH1_ is the 5' fusion partner.

```{r}
ttyh1_fusions <- fusion_summary_df %>%
  select(Kids_First_Biospecimen_ID, starts_with("TTYH1"))
```

Which samples have these fusions?

```{r}
fusion_biospecimens <- ttyh1_fusions %>%
  # add a column that counts how many TTYH1 fusions are present
  mutate(fusion_count = rowSums(select(., starts_with("TTYH1")))) %>%
  # if there's any TTYH1 fusion - pull out the biospecimen ID
  filter(fusion_count > 0) %>%
  pull(Kids_First_Biospecimen_ID)
```

### Disease labels

First subset to embryonal **tumors**, excluding any derived cell lines.

```{r}
embryonal_df <- histologies_df %>%
  filter(str_detect(broad_histology, "Embryonal tumor"),
         sample_type == "Tumor",
         composition == "Solid Tissue")
```

Drop medulloblastoma and ATRT samples using the `disease_type_old` labels.

```{r}
embryonal_df <- embryonal_df %>%
  filter(!(disease_type_old %in% c("Medulloblastoma",
                                   "Atypical Teratoid Rhabdoid Tumor (ATRT)")))
```

### Combine biospecimen IDs

Using the fusion information and the clinical file.

```{r}
all_biospecimen_ids <- 
  unique(c(embryonal_df %>% pull(Kids_First_Biospecimen_ID),
           fusion_biospecimens))
```

Write to file.

```{r}
data.frame(Kids_First_Biospecimen_ID = all_biospecimen_ids) %>%
  write_tsv(output_file)
```

