---
title: "Molecularly Subtyping Embryonal Tumors - Final Table"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Stephanie J. Spielman and Jaclyn Taroni for ALSF CCDL
date: 2020
---

This notebook identifies samples to include in subset files for the purpose of molecularly subtyping embryonal tumors ([`AlexsLemonade/OpenPBTA-analysis#251`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/251)).

This closed pull request is also relevant to the task at hand: [`AlexsLemonade/OpenPBTA-analysis#401`](https://github.com/AlexsLemonade/OpenPBTA-analysis/pull/401

We'll use the subset files generated in [`02-generate-subset-files.R`](./02-generate-subset-files.R) to construct tables that summarize the data on the subtyping issue.

## Usage


This notebook is intended to be run via the command line from the top directory
of the repository as follows:

```
Rscript -e "rmarkdown::render('analyses/molecular-subtyping-embryonal/04-table-prep.Rmd', clean = TRUE)"
```

## Set up

### Libraries and functions

```{r}
library(tidyverse)
```

```{r}
# TODO: consider moving this out of this notebook to use with ependymoma 
# subtyping?
wrangle_fusions <- function(fusions_df, string_to_match, column_name) {
  # This function takes the fusion summary binary matrix and returns a 
  # data frame that contains two columns: Kids_First_Biospecimen_ID and
  # a column containing a comma-separated 'list' of relevant fusions.
  # A "relevant fusion" is determined by passing the string_to_match argument
  # to dplyr::matches. To specify that a gene symbol should be a 5' partner,
  # you can pass "<gene symbol>--". The 2nd column name is determined by the
  # column_name argument.
  #
  # Args:
  #   fusions_df: a fusion_summary data.frame from `fusion-summary`
  #   string_to_match: character; passed to dplyr::matches to select relevant
  #                    columns
  #   column_name: the name of the relevant fusion column in the data.frame 
  #                that is returned
  # 
  # Returns: a data.frame with Kids_First_Biospecimen_ID and relevant fusions
  #          columns; biospecimens without any relevant fusions will have "None"
  #          in the second column
  
  specific_fusions <- fusions_df %>%
    # only include relevant fusions using string_to_match
    select(Kids_First_Biospecimen_ID, matches(string_to_match)) %>%
    reshape2::melt() %>%
    # a value of 0 means that fusion is not present in that biospecimen
    filter(value > 0) %>%
    select(-value) %>%
    group_by(Kids_First_Biospecimen_ID) %>%
    # get the comma separated 'list' of relevant fusions
    summarize(!!column_name := paste(sort(unique(variable)), collapse = ", "))
  
  # if a sample had none of the relevant fusions, it's missing from 
  # specific_fusions -- let's add them back in with "None" in the second
  # column
  missing_ids <- setdiff(fusions_df$Kids_First_Biospecimen_ID,
                         specific_fusions$Kids_First_Biospecimen_ID)
  specific_fusions <- specific_fusions %>%
    bind_rows(data.frame(Kids_First_Biospecimen_ID = missing_ids)) %>%
    replace(is.na(.), "None")
  
  return(specific_fusions)
}
```

### Directories

```{r}
subset_dir <- "subset-files"
results_dir <- "results"
data_dir <- file.path("..", "..", "data")
```

### Read in files

Set up filenames.

```{r}
# full clinical file
histologies_file <- file.path(data_dir, "pbta-histologies.tsv")
# TODO: update when the new files with union of sample IDs get included in the
# in the data download (v14)
fusion_file <- file.path("..", "fusion-summary", "results",
                         "fusion_summary_embryonal_foi.tsv")
# copy number consensus file (analysis file)
# TODO: update when included in data download (v14)
cn_consensus_seg <- file.path("..", "copy_number_consensus_call", "results",
                              "pbta-cnv-consensus.seg.gz")
manta_file <- file.path(subset_dir, "embryonal_manta_sv.tsv")
polya_file <- file.path(subset_dir, "embryonal_zscored_exp.polya.rds")
stranded_file <- file.path(subset_dir, "embryonal_zscored_exp.stranded.rds")
# file that contains the biospecimen IDs that met criteria for subtyping
biospecimen_file <- file.path(results_dir, 
                              "biospecimen_ids_embryonal_subtyping.tsv")
# annotated sex chromosomes copy number alterations
sex_chr_consensus <- file.path("..", "focal-cn-file-preparation", "results", 
                               "consensus_seg_annotated_cn_x_and_y.tsv.gz")
sex_chr_cnvkit <- file.path("..", "focal-cn-file-preparation", "results", 
                            "cnvkit_annotated_cn_x_and_y.tsv.gz")
sex_chr_controlfreec <- file.path("..", "focal-cn-file-preparation", "results", 
                                  "controlfreec_annotated_cn_x_and_y.tsv.gz")
```

Read in files.

```{r message=FALSE}
histologies_df <- read_tsv(histologies_file, 
                           col_types = cols(
                             molecular_subtype = col_character()
                          ))

# manta data for BCOR tandem duplication
manta_sv_df <- read_tsv(manta_file)

# all of the RNA data
polya_exp <- read_rds(polya_file)
stranded_exp <- read_rds(stranded_file)
fusions_df <- read_tsv(fusion_file)

# get biospecimen ids as a vector rather than data frame
biospecimen_ids <- read_tsv(biospecimen_file) %>%
  pull(Kids_First_Biospecimen_ID)

# we'll filter to only the relevant biospecimens for the copy number alteration
# data that is committed to the repository
cn_consensus <- read_tsv(cn_consensus_seg) %>%
  filter(ID %in% biospecimen_ids)
consensus_sex_chr <- read_tsv(sex_chr_consensus) %>%
  filter(biospecimen_id %in% biospecimen_ids)
cnvkit_sex_chr <- read_tsv(sex_chr_cnvkit) %>%
  filter(biospecimen_id %in% biospecimen_ids)
controlfreec_sex_chr <- read_tsv(sex_chr_controlfreec) %>%
  filter(biospecimen_id %in% biospecimen_ids)
```

### Output file

```{r}
output_all <- file.path(results_dir, 
                        "embryonal_tumor_subtyping_relevant_data.tsv")
```

## Wrangle data

### Fusion data

Summarizing the salient points from [`AlexsLemonade/OpenPBTA-analysis#251`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/251):

> **ETMR, C19MC-altered**: These tumors have focal amplification of the miRNA cluster on chr19 (denoted C19MC) and often have gene fusions involving _TTYH1_ and chr19 miRNA cluster genes

> **CNS HGNET-MN1**: Contain gene fusions involving 5' _MN1_. 3' fusion partners can include _BEND2_ and _CXXC5_.

> **CNS NB-FOXR2**: Over-expression and/or gene fusions in _FOXR2_

> **CNS EFT-CIC**: Alterations in _CIC_, commonly fused with _NUTM1_

```{r}
colnames(fusions_df)
```

```{r message=FALSE}
ttyh1_fusions <- wrangle_fusions(fusions_df, "TTYH1--", "TTYH1_fusions")
mn1_fusions <- wrangle_fusions(fusions_df, "MN1--", "MN1_fusions") 
foxr2_fusions <- wrangle_fusions(fusions_df, "FOXR2", "FOXR2_fusions")
cic_fusions <- wrangle_fusions(fusions_df, "CIC--", "CIC_fusions")
fusions_summary_df <- list(ttyh1_fusions,
                           mn1_fusions,
                           foxr2_fusions,
                           cic_fusions) %>%
  purrr::reduce(dplyr::inner_join, by = "Kids_First_Biospecimen_ID")
# remove data.frame we no longer need
rm(ttyh1_fusions, mn1_fusions, foxr2_fusions, cic_fusions, fusions_df)
```

```{r}
head(fusions_summary_df %>% arrange(desc(TTYH1_fusions)), n = 10)
```

### Expression data

Summarizing salient points from [`AlexsLemonade/OpenPBTA-analysis#251`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/251):

> **ETMR, C19MC-altered**: These tumors have high expression of _LIN28A_ and serves as a biomarker for ETMRs

> **CNS NB-FOXR2**: Over-expression and/or gene fusions in _FOXR2_

> **ETMR, NOS**: These tumors have high expression of _LIN28A_ and serves as a biomarker for ETMRs, but do not show focal amplification of C19MC.

```{r}
exp_genes_of_interest <- c("LIN28A", "FOXR2")
# set up poly-A data
polya_exp <- polya_exp[, exp_genes_of_interest] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Kids_First_Biospecimen_ID")
# set up stranded data
stranded_exp <- stranded_exp[, exp_genes_of_interest] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Kids_First_Biospecimen_ID")

# bind together both datasets, but keep track of what selection strategy was
# used
exp_df <- bind_rows("polya" = polya_exp, 
                    "stranded" = stranded_exp, 
                    .id = "exp_dataset") %>%
  select(-exp_dataset, everything())

# remove data.frame we no longer need
rm(polya_exp, stranded_exp, exp_genes_of_interest)
```

### Join RNA data

```{r}
rna_df <- exp_df %>%
  left_join(fusions_summary_df) %>%
  select(-exp_dataset, everything())

# remove data.frame we no longer need
rm(exp_df)
```

```{r}
head(rna_df, n = 10)
```

### Structural variant data

> **CNS HGNET-BCOR**

> * CNS high-grade neuroepithelial tumor with BCOR alteration

> * Tumors have internal tandem duplication of BCOR

The definition of internal tandem duplication from [Rustagi et al. _BMC Bioinformatics._ 2016.](https://doi.org/10.1186/s12859-016-1031-8):

> Detection of tandem duplication within coding exons, referred to as internal tandem duplication (ITD)

The annotated files from the [`focal-cn-file-preparation`](../focal-cn-file-preparation/) module use exons to check for overlaps.
So, we can try to supplement the Manta calls with the annotated sex chromosome files.

```{r}
# only include rows from Manta that make reference to BCOR, pass all filters
# https://github.com/Illumina/manta/blob/75b5c38d4fcd2f6961197b28a41eb61856f2d976/docs/userGuide/README.md#vcf-format-fields
# and are recorded as tandem duplication events
bcor_sv_df <- manta_sv_df %>%
  filter(str_detect(Gene.name, "BCOR"),
         FILTER == "PASS",
         ALT == "<DUP:TANDEM>")

# take a look at the genes 
bcor_sv_df %>% 
  head(n = 2) %>%
  pull(Gene.name)
```

Not necessarily specific to _BCOR_ alterations.

```{r}
consensus_sex_chr %>% filter(gene_symbol == "BCOR") 
```

```{r}
cnvkit_sex_chr %>% filter(gene_symbol == "BCOR") 
```

```{r}
controlfreec_sex_chr %>% filter(gene_symbol == "BCOR")
```

This is not conclusive.

## Final table

### Add `Kids_First_Participant_ID` and `sample_id`

```{r}
identifiers_df <- histologies_df %>%
  select(Kids_First_Participant_ID, 
         sample_id, 
         Kids_First_Biospecimen_ID)
# add IDs to chromosome 19 data
chr19_cn_df <- identifiers_df %>%
  inner_join(chr19_cn_df, by = c("Kids_First_Biospecimen_ID" = "ID"))
# add IDs to RNA data 
rna_df <- identifiers_df %>%
  inner_join(rna_df)
```

### Join molecular data

```{r}
molecular_data_df <- full_join(chr19_cn_df, rna_df,
                               by = c("Kids_First_Participant_ID",
                                      "sample_id")) %>%
  rename(Kids_First_Biospecimen_ID_DNA = Kids_First_Biospecimen_ID.x,
         Kids_First_Biospecimen_ID_RNA = Kids_First_Biospecimen_ID.y)
```

### Add clinical data

The following were mentioned in [`AlexsLemonade/OpenPBTA-analysis#251`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/251):

> **CNS HGNET-BCOR**: Median age of diagnosis less than 10 years

> **CNS HGNET-MN1**: Predominantly female patients.

> **CNS Embryonal, NOS**: Tumors previously called PNET that do not fit into other groups above.

Granted, we don't currently have enough information to look specifically at CNS HGNET-BCOR and therefore perhaps can not classify any tumor as CNS Embryonal, NOS. 

```{r}
relevant_clinical_data <- histologies_df %>% 
  filter(Kids_First_Biospecimen_ID %in% biospecimen_ids) %>%
  select(Kids_First_Participant_ID, 
         sample_id,
         age_at_diagnosis_days,
         germline_sex_estimate,
         disease_type_old,
         disease_type_new) %>%
  distinct() %>%
  # convert age to years
  mutate(age_at_diagnosis_yrs = 
           floor(as.numeric(age_at_diagnosis_days) / 365)) %>%
  select(-age_at_diagnosis_days)

all_data_df <- inner_join(relevant_clinical_data,
                          molecular_data_df)
```

Write to file.

```{r}
all_data_df %>%
  select(Kids_First_Participant_ID, 
         sample_id,
         Kids_First_Biospecimen_ID_DNA,
         Kids_First_Biospecimen_ID_RNA,
         age_at_diagnosis_yrs,
         germline_sex_estimate,
         disease_type_old,
         disease_type_new,
         LIN28A,
         TTYH1_fusions,
         chr19_amplification,
         FOXR2,
         FOXR2_fusions,
         MN1_fusions,
         CIC_fusions) %>%
  write_tsv(output_all)
```

## Subtyping

### ETMR, C19MC-altered and ETMR, NOS

The differences between ETMR, C19MC-altered and ETMR, NOS is a difference in C19MC amplification.
We'll filter on _LIN28A_ overexpression, as that is a feature of both.

*Recall that we z-scored these expression values before subsetting the matrix so they are in the context of **all** samples for a given selection strategy.*

```{r}
all_data_df %>%
  # 3 standard deviations above the mean
  filter(LIN28A > 3) %>%
  select(ends_with("ID"), 
         starts_with("disease"),
         LIN28A,
         TTYH1_fusions,
         chr19_amplification)
```

For some samples that have a _TTYH1_ fusion, we do not have DNA data to check for C19MC amplification.

### CNS HGNET-MN1

> * CNS high-grade neuroepithelial tumor with _MN1_ alteration
> * Likely previously diagnosed as PNET.
> * Contain gene fusions involving 5' _MN1_. 3' fusion partners can include _BEND2_ and _CXXC5_.
> * Predominantly female patients.

```{r}
all_data_df %>%
  # 3 standard deviations above the mean
  filter(MN1_fusions != "None") %>%
  select(ends_with("ID"), 
         starts_with("disease"),
         germline_sex_estimate,
         MN1_fusions)
```

No samples fit this criteria.
There were other samples that **did** contain _MN1_ fusions.

```{r}
mn1_fusions <- fusions_summary_df %>%
  filter(MN1_fusions != "None") 

mn1_fusions
```

```{r}
histologies_df %>%
  filter(Kids_First_Biospecimen_ID %in% pull(mn1_fusions, 
                                             Kids_First_Biospecimen_ID)) %>%
  select(Kids_First_Participant_ID, 
         sample_id, 
         Kids_First_Biospecimen_ID,
         disease_type_old,
         disease_type_new,
         germline_sex_estimate)
```

They are classified as `Ependymoma` and `Low-grade astrocytic tumor`.

### CNS HGNET-BCOR

> * CNS high-grade neuroepithelial tumor with BCOR alteration
> * Tumors have internal tandem duplication of BCOR
> * Median age of diagnosis less than 10 years

We have not included SV data, yet.

### CNS NB-FOXR2

> * Central nervous system (CNS) neuroblastoma with _FOXR2_ activation
> * Over-expression and/or gene fusions in _FOXR2_

```{r}
all_data_df %>%
  filter(FOXR2_fusions != "None" | FOXR2 > 3) %>%
  select(ends_with("ID"), 
         starts_with("disease"),
         FOXR2,
         FOXR2_fusions)
```

### CNS EFT-CIC

> * CNS Ewing sarcoma family tumor with _CIC_ alteration
> * Alterations in _CIC_, commonly fused with _NUTM1_

```{r}
all_data_df %>%
  filter(CIC_fusions != "None")
```

### CNS Embryonal, NOS

> * CNS Embryonal tumor, not otherwise specified
> * Tumors previously called PNET that do not fit into other groups above.

We do not have CNS HGNET-BCOR information currently.

## Session Info

```{r}
sessionInfo()
```

