---
title: "CN Status Heatmap"
output:   
  html_notebook: 
    toc: true
    toc_float: true
author: Candace Savonen for ALSF - CCDL
date: 2020
---

## Purpose: 
Create a summary heatmap of copy number status from the consensus CNV call data. 
This is done by binning the genome and calculating the segment's coverage of the
CNV consensus segments. 
A bin is declared a particular copy number status if that status's base pair 
coverage fraction is above a certain threshold (`frac_threshold_val`).

### Usage

This notebook can be run via the command line from the top directory of the 
repository as follows:

```
Rscript -e "rmarkdown::render('analyses/cnv-chrom-plot/cn_status_heatmap.Rmd', 
                              clean = TRUE)"
```

### Cutoffs: 

```{r}
# The max length of a segment to use the data.
# segments that are too long may dominate the heatmap and/or be indicators of
# broader structural changes
length_max <- 1e7

# Set minimum percentage of a bin that should be callable to report data.
frac_uncallable <- 0.75

# Absolute fraction needed for a bin to be called a particular status
frac_threshold <- 0.75
```

### Set Up

```{r}
# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

### Directories and Files

```{r}
# Path to input directory
input_dir <- file.path("..", "..", "data")
figure_dir <- file.path("..", "..", "figures")
scratch_dir <- file.path("..", "..", "scratch")

# Path to output directory
plots_dir <- "plots"

# Create the plots_dir if it does not exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}
```

Read in custom functions.

```{r}
source(file.path("util", "bin-coverage.R"))
```

Import color palettes. 

```{r}
# Import standard color palettes for project
histology_col_palette <- readr::read_tsv(
  file.path(figure_dir, "palettes", "histology_color_palette.tsv")
) %>%
  # We'll use deframe so we can use it as a recoding list
  tibble::deframe()
```

Read in the divergent color palette and set it up with three colors. 
In this instance, we only need three values for `gain`, `neutral`, and `loss`.

```{r}
divergent_col_palette <- readr::read_tsv(
  file.path(figure_dir, "palettes", "divergent_color_palette.tsv")
) %>%
  # Only keep only these three colors
  dplyr::filter(
    color_names %in% c("divergent_low_4", "divergent_neutral", "divergent_high_4")
  ) %>%
  dplyr::pull("hex_codes")
```

Binary color palette for chromosomes.

```{r}
# Read in color palette
binary_col_palette <- readr::read_tsv(
  file.path(figure_dir, "palettes", "binary_color_palette.tsv")
) %>%
  # Drop na color from this color palette
  dplyr::filter(color_names != "na_color")
```

### Read in data 

```{r}
# Read in metadata
metadata <-
  readr::read_tsv(file.path(input_dir, "pbta-histologies.tsv")) %>%
  # Easier to deal with NA short histologies if they are labeled something different
  dplyr::mutate(short_histology = as.character(tidyr::replace_na(short_histology, "none"))) %>%
  # Tack on the sample color using the short_histology column and a recode
  dplyr::mutate(sample_color = dplyr::recode(
    short_histology,
    !!!histology_col_palette
  ))
```

### Set up consensus copy number data

```{r}
# Read in the segment copy number data
seg_data <- data.table::fread(file.path(
  input_dir,
  "pbta-cnv-consensus.seg.gz"
),
data.table = FALSE
)
```

Set up the status for each consensus segment. 

```{r}
seg_data <- seg_data %>%
  # Join the histology column to this data
  dplyr::inner_join(dplyr::select(
    metadata,
    "Kids_First_Biospecimen_ID",
    "short_histology",
    "tumor_ploidy"
  ),
  by = c("ID" = "Kids_First_Biospecimen_ID")
  ) %>%
  # Reformat the chromosome variable to drop the "chr"
  dplyr::mutate(chrom = factor(gsub("chr", "", chrom),
    levels = c(1:22, "X", "Y")
  )) %>%
  # Recode the copy number status based on ploidy
  dplyr::mutate(status = dplyr::case_when(
    # when the copy number is less than inferred ploidy, mark this as a loss
    copy.num < tumor_ploidy ~ "loss",
    # if copy number is higher than ploidy, mark as a gain
    copy.num > tumor_ploidy ~ "gain",
    copy.num == tumor_ploidy ~ "neutral"
  )) %>%
  # Remove sex chromosomes
  dplyr::filter(
    !(chrom %in% c("X", "Y", "M")),
    !is.na(status)
  )
```

Set up seg data as GenomicRanges. 

```{r}
seg_ranges <- GenomicRanges::GRanges(
  seqnames = seg_data$chrom,
  ranges = IRanges::IRanges(
    start = seg_data$loc.start,
    end = seg_data$loc.end
  ),
  status = seg_data$status,
  histology = seg_data$short_histology,
  biospecimen = seg_data$ID
)
```

Explore the distribution of segment lengths. 

```{r}
ggplot2::qplot(seg_ranges@ranges@width, geom = "density") +
  ggplot2::theme_classic() +
  ggplot2::ylab("density") +
  ggplot2::xlab("Segment length in bp") +
  # Let's put a vertical line where we will make a filter cutoff
  ggplot2::geom_vline(xintercept = length_max, color = "red")
```

Filter out segments that are longer than our cutoff. 

```{r}
filtered_seg_ranges <- seg_ranges[which(seg_ranges@ranges@width < length_max)]
```

### Set up chromosomal sizes for making bins. 

(This has nothing to do with Strelka, but it just so happens this is a file 
with the sizes of the chromosomes in this genome build, hg38).

```{r}
chr_sizes <- readr::read_tsv(file.path(input_dir, "WGS.hg38.strelka2.unpadded.bed"),
  col_names = c("chrom", "start", "end")
) %>%
  # Reformat the chromosome variable to drop the "chr"
  dplyr::mutate(chrom = factor(gsub("chr", "", chrom),
    levels = c(1:22, "X", "Y", "M")
  )) %>%
  # Remove sex chromosomes
  dplyr::filter(!(chrom %in% c("X", "Y", "M")))


# Make chromosome size named vector for Heatmap annotation
chr_sizes_vector <- chr_sizes$end
names(chr_sizes_vector) <- chr_sizes$chrom
```

### Set up uncallable regions data 

Regions that were not able to be accurately called will need to be color coded gray later. 
Here, we are setting up the uncallable regions like we did with the callable regions.

```{r}
uncallable_bed <- readr::read_tsv(
  file.path(
    "..",
    "copy_number_consensus_call",
    "ref",
    "cnv_excluded_regions.bed"
  ),
  col_names = c("chrom", "start", "end")
) %>%
  # Reformat the chromosome variable to drop the "chr"
  dplyr::mutate(chrom = factor(gsub("chr", "", chrom),
    levels = c(1:22, "X", "Y")
  )) %>%
  dplyr::filter(
    # Drop CNVs that don't have chromosome labels
    !is.na(chrom),
    # Drop sex chromosomes
    !(chrom %in% c("X", "Y", "M"))
  )
```

Set up uncallable regions as GenomicRanges. 

```{r}
uncallable_ranges <- GenomicRanges::GRanges(
  seqnames = uncallable_bed$chrom,
  ranges = IRanges::IRanges(
    start = uncallable_bed$start,
    end = uncallable_bed$end
  )
)
```

## Call bin CN statuses for each sample

Set up binned genome ranges. 

```{r}
# Set up bins of ~1Mb size
bins <- GenomicRanges::tileGenome(
  chr_sizes_vector,
  tilewidth = 1e6
)
# Uncompress these ranges
bins <- unlist(bins)
```

Run the bin status calling on each sample. 

```{r echo=FALSE}
# Get a vector of the biospecimen IDs
sample_ids <- unique(seg_data$ID)

# Run call_bin_status for each biospecimen's segments.
bin_calls_list <- lapply(sample_ids,
  call_bin_status,
  bin_ranges = bins,
  seg_ranges = filtered_seg_ranges,
  uncallable_ranges = uncallable_ranges,
  frac_threshold_val = frac_threshold,
  frac_uncallable_val = frac_uncallable
)

# Bring along sample IDs
names(bin_calls_list) <- sample_ids

# Format into data.frame
bin_calls_df <- dplyr::bind_rows(bin_calls_list,
  .id = "biospecimen_id"
)
```

## Set up heatmap annotation objects

Make color key. 

```{r}
color_key <- structure(c(divergent_col_palette, "#9932CC", "#cccccc"),
  names = c("loss", "neutral", "gain", "unstable", "uncallable")
)
```

### Make column annotation object

Extract chromosome labels and make an alternating color key for them. 
This annotation object strategy was originally from [chromosomal-instability](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/b5a33838d1e9bd7e7913a89201ec26125c16c94c/analyses/chromosomal-instability/02a-plot-chr-instability-heatmaps.Rmd#L73).

```{r}
# Set up chromosome labels from bins as a factor vector
chrs <- paste0("chr", S4Vectors::decode(bins@seqnames))
chrs <- factor(chrs, levels = paste0("chr", 1:22))

# Make a key for assigning alternating colors to the chromosomes
chr_colors <- rep(binary_col_palette$hex_codes,
  length.out = length(unique(chrs))
)
names(chr_colors) <- unique(chrs)

# Get coordinate start positions
chr_pos <- match(unique(chrs), chrs)

# Get mid points of chromosome labels
mid_points <- ceiling((c(chr_pos[2:length(chr_pos)], length(chrs)) + chr_pos[1:length(chr_pos)]) / 2)
```

Make chromosomal labeling `HeatmapAnnotation` object.

```{r}
# Make text labels for chromosome text
chr_text <- ComplexHeatmap::anno_mark(
  at = mid_points,
  labels = levels(chrs),
  which = "column",
  side = "bottom",
  labels_rot = 45,
  labels_gp = grid::gpar(cex = 0.65)
)

# Create the Heatmap annotation object
chr_annot <- ComplexHeatmap::HeatmapAnnotation(
  df = data.frame(chrs),
  col = list(chrs = chr_colors),
  name = "",
  show_legend = FALSE,
  show_annotation_name = FALSE,
  mark = chr_text # Put the text in
)
```

### Make row annotation object

Make histology labeling `HeatmapAnnotation` object.
This annotation object strategy was originally from [chromosomal-instability](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/b5a33838d1e9bd7e7913a89201ec26125c16c94c/analyses/chromosomal-instability/02a-plot-chr-instability-heatmaps.Rmd#L73).

```{r}
# Get the histologies for the samples in this set and order them by histology
histologies <-
  data.frame(Kids_First_Biospecimen_ID = bin_calls_df$biospecimen_id) %>%
  dplyr::inner_join(metadata %>%
    dplyr::select(Kids_First_Biospecimen_ID, short_histology, sample_color)) %>%
  dplyr::mutate(short_histology = factor(short_histology)) %>%
  dplyr::arrange(short_histology) %>%
  tibble::column_to_rownames("Kids_First_Biospecimen_ID")

# Make color key specific to these samples
histologies_color_key_filtered <- unique(histologies$sample_color)
names(histologies_color_key_filtered) <- unique(histologies$short_histology)

# Drop this column so ComplexHeatmap isn't tempted to plot it
histologies <- dplyr::select(histologies, -sample_color)

# Get coordinate start positions
hist_pos <- match(names(histologies_color_key_filtered), histologies$short_histology)

# Get mid points of chromosome labels
mid_points <- ceiling((c(hist_pos[2:length(hist_pos)], length(histologies$short_histology)) + hist_pos[1:length(hist_pos)]) / 2)
```

```{r}
# Make text labels for chromosome text
hist_text <- ComplexHeatmap::anno_mark(
  at = mid_points,
  labels = levels(histologies$short_histology),
  which = "row",
  side = "right",
  labels_gp = grid::gpar(cex = 0.65),
  link_width = grid::unit(15, "mm")
)

# Create the Heatmap annotation object
hist_annot <- ComplexHeatmap::HeatmapAnnotation(
  df = data.frame(histologies),
  col = list(short_histology = histologies_color_key_filtered),
  which = "row",
  show_annotation_name = FALSE,
  show_legend = FALSE,
  mark = hist_text # Put the text in
)
```

Format `bin_calls_df` as a matrix with rownames for `ComplexHeatmap` to use. 

```{r}
bin_calls_mat <- bin_calls_df %>%
  tibble::column_to_rownames("biospecimen_id") %>%
  as.matrix()

# Ensure that this matrix is in the same order as the annotation
bin_calls_mat <- bin_calls_mat[rownames(histologies), ]

# Double check its in thte same order
all.equal(rownames(bin_calls_mat), rownames(histologies))
```

## Assemble CN status heatmap

```{r}
# Plot on a heatmap
heatmap <- ComplexHeatmap::Heatmap(
  bin_calls_mat,
  name = "CN status",
  col = color_key,
  cluster_columns = FALSE,
  cluster_rows = FALSE,
  show_column_names = FALSE,
  show_row_names = FALSE,
  bottom_annotation = chr_annot,
  right_annotation = hist_annot,
  heatmap_legend_param = list(nrow = 1)
)
```

Print out heatmap. 

```{r}
ComplexHeatmap::draw(heatmap, heatmap_legend_side = "bottom")
```

Save to PDF. 

```{r}
# Save plot as PDF
pdf(file.path(plots_dir, "cn_status_heatmap.pdf"))
ComplexHeatmap::draw(heatmap, heatmap_legend_side = "bottom")
dev.off()
```

# Session Info

```{r}
sessionInfo()
```


