---
title: "CN Status Heatmap"
output:   
  html_notebook: 
    toc: true
    toc_float: true
author: Candace Savonen for ALSF - CCDL
date: 2020
---

## Purpose: 
Create a summary heatmap of copy number status from the consensus CNV call data. 
This is done by binning the genome and calculating the segment's coverage of the
CNV consensus segments. 
A bin is declared a particular copy number status if that status's base pair 
coverage is a certain threshold percentage larger than the other status' 
coverage (`perc_delta_threshold`).

### Usage

This notebook can be run via the command line from the top directory of the 
repository as follows:

```
Rscript -e "rmarkdown::render('analyses/cnv-chrom-plot/cn_status_heatmap.Rmd', 
                              clean = TRUE)"
```

### Cutoffs: 

```{r}
# The max length of a segment to use the data. 
# segments that are too long may dominate the heatmap and/or be indicators of 
# broader structural changes
length_max <- 1e7

# Set minimum percentage of a bin that should be callable to report data. 
perc_uncallable <- 0.75

# Percent coverage difference for a bin to be called a particular status 
# (0.2 means anything where a status is 20% bigger than the others will 
# be called).
perc_delta_threshold <- 0.1
```

### Set Up

```{r}
# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

### Directories and Files

```{r}
# Path to input directory
input_dir <- file.path("..", "..", "data")
figure_dir <- file.path("..", "..", "figures")
scratch_dir <- file.path("..", "..", "scratch")

# Path to output directory
plots_dir <- "plots"

# Create the plots_dir if it does not exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}
```

Read in custom functions.

```{r}
source(file.path("util", "bin-coverage.R"))
```

Import color palettes. 

```{r}
# Import standard color palettes for project
histology_col_palette <- readr::read_tsv(
  file.path(figure_dir, "palettes", "histology_color_palette.tsv")
  ) %>% 
  # We'll use deframe so we can use it as a recoding list
  tibble::deframe()
```

Read in the divergent color palette and set it up with three colors. 

```{r}
divergent_col_palette <- readr::read_tsv(
  file.path(figure_dir, "palettes", "divergent_color_palette.tsv")) %>% 
  # Only keep only these three colors
  dplyr::filter(
    color_names %in% c("divergent_low_4", "divergent_neutral", "divergent_high_4")) %>% 
  dplyr::pull("hex_codes")
```

Binary color palette for chromosomes.

```{r}
# Read in color palette
binary_col_palette <- readr::read_tsv(
  file.path(figure_dir, "palettes", "binary_color_palette.tsv")
  ) %>%
  # Drop na color from this color palette
  dplyr::filter(color_names != "na_color")
```

### Read in data 

```{r}
# Read in metadata
metadata <-
  readr::read_tsv(file.path(input_dir, "pbta-histologies.tsv")) %>%
  # Easier to deal with NA short histologies if they are labeled something different
  dplyr::mutate(short_histology = as.character(tidyr::replace_na(short_histology, "none"))) %>% 
  # Tack on the sample color using the short_histology column and a recode
  dplyr::mutate(sample_color = dplyr::recode(short_histology, 
                                             !!!histology_col_palette))
```

### Set up consensus copy number data

```{r}
# Read in the segment copy number data
seg_data <- data.table::fread(file.path(
  input_dir,
  "pbta-cnv-consensus.seg.gz"
),
data.table = FALSE
)
```

Set up the status for each consensus segment. 

```{r}
seg_data <- seg_data %>%
  # Join the histology column to this data
  dplyr::inner_join(dplyr::select(
    metadata,
    "Kids_First_Biospecimen_ID",
    "short_histology",
    "tumor_ploidy"
  ),
  by = c("ID" = "Kids_First_Biospecimen_ID")
  ) %>%
  # Reformat the chromosome variable to drop the "chr"
  dplyr::mutate(chrom = factor(gsub("chr", "", chrom),
    levels = c(1:22, "X", "Y")
  )) %>%
  # Recode the copy number status based on ploidy
  dplyr::mutate(status = dplyr::case_when(
    # when the copy number is less than inferred ploidy, mark this as a loss
    copy.num < tumor_ploidy ~ "loss",
    # if copy number is higher than ploidy, mark as a gain
    copy.num > tumor_ploidy ~ "gain",
    copy.num == tumor_ploidy ~ "neutral"
  )) %>%
  # Remove sex chromosomes
  dplyr::filter(
    !(chrom %in% c("X", "Y", "M")),
    !is.na(status)
  )
```

Set up seg data as GenomicRanges. 

```{r}
seg_ranges <- GenomicRanges::GRanges(
  seqnames = seg_data$chrom,
  ranges = IRanges::IRanges(
    start = seg_data$loc.start,
    end = seg_data$loc.end
  ),
  status = seg_data$status,
  histology = seg_data$short_histology,
  biospecimen = seg_data$ID
)
```

Explore the distribution of segment lengths. 

```{r}
ggplot2::qplot(seg_ranges@ranges@width, geom= "density") + 
  ggplot2::theme_classic() + 
  ggplot2::ylab("density") + 
  ggplot2::xlab("Segment length in bp") + 
  # Let's put a vertical line where we will make a filter cutoff
  ggplot2::geom_vline(xintercept = length_max, color = "red")
```

Filter out segments that are longer than our cutoff. 

```{r}
filtered_seg_ranges <- seg_ranges[which(seg_ranges@ranges@width < length_max)]
```

### Set up chromosomal sizes for making bins. 

(This has nothing to do with Strelka, but it just so happens this is a file 
with the sizes of the chromosomes in this genome build, hg38).

```{r}
chr_sizes <- readr::read_tsv(file.path(input_dir, "WGS.hg38.strelka2.unpadded.bed"),
  col_names = c("chrom", "start", "end")
) %>%
  # Reformat the chromosome variable to drop the "chr"
  dplyr::mutate(chrom = factor(gsub("chr", "", chrom),
    levels = c(1:22, "X", "Y", "M")
  )) %>%
  # Remove sex chromosomes
  dplyr::filter(!(chrom %in% c("X", "Y", "M")))


# Make chromosome size named vector for Heatmap annotation
chr_sizes_vector <- chr_sizes$end
names(chr_sizes_vector) <- chr_sizes$chrom
```

### Set up uncallable regions data 
Regions that were not able to be accurately called will need to be color coded gray. 

```{r}
uncallable_bed <- readr::read_tsv(
  file.path(
    "..",
    "copy_number_consensus_call",
    "ref",
    "cnv_excluded_regions.bed"
  ),
  col_names = c("chrom", "start", "end")
) %>%
  # Reformat the chromosome variable to drop the "chr"
  dplyr::mutate(chrom = factor(gsub("chr", "", chrom),
    levels = c(1:22, "X", "Y")
  )) %>%
  dplyr::filter(
    # Drop CNVs that don't have chromosome labels
    !is.na(chrom),
    # Drop sex chromosomes
    !(chrom %in% c("X", "Y", "M"))
  )
```

Set up uncallable regions as GenomicRanges. 

```{r}
uncallable_ranges <- GenomicRanges::GRanges(
  seqnames = uncallable_bed$chrom,
  ranges = IRanges::IRanges(
    start = uncallable_bed$start,
    end = uncallable_bed$end
  )
)
```

## Call bin CN statuses for each sample

Set up binned genome ranges. 

```{r}
# Find which regions of the chromosome are uncallable
bins <- GenomicRanges::tileGenome(
  chr_sizes_vector,
  tilewidth = 1e6
)
# Uncompress these ranges
bins <- unlist(bins)
```

Run the bin status calling on each sample. 

```{r echo=FALSE}
# Get a vector of the biospecimen IDs
sample_ids <- unique(seg_data$ID)

# Run call_bin_status for each biospecimen's segments. 
bin_calls_list <- lapply(sample_ids,
  call_bin_status,
  bin_ranges = bins,
  seg_ranges = filtered_seg_ranges,
  # Percent coverage difference for a bin to be called a particular status
  perc_delta_threshold = perc_delta_threshold
)

# Bring along sample IDs
names(bin_calls_list) <- sample_ids

# Format into data.frame
bin_calls_df <- dplyr::bind_rows(bin_calls_list,
  .id = "biospecimen_id"
)
```

## Determine uncallable bins of the genome

We'll determine NA regions based on what percent of that window of that region is uncallable. 
These methods are borrowed from [chromosomal-instability methods](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/b5a33838d1e9bd7e7913a89201ec26125c16c94c/analyses/chromosomal-instability/util/chr-break-calculate.R#L216). 

```{r}
# Find the NA region(s) for each bin without combining close regions
na_regions <- GenomicRanges::pintersect(
  IRanges::findOverlapPairs(bins, uncallable_ranges)
)

# Find overlap between na_regions and bin
na_overlaps <- GenomicRanges::findOverlaps(bins, na_regions)

# Get the sum of the length of all excluded regions for each bin.
excluded_length_per_bin <- tapply(
  na_regions@ranges@width, # Get length of each sequence
  na_overlaps@from, # Index of which bin it overlaps
  sum
) # Add up per bin

# Get the total bin length for each bin that has excluded regions
bin_length <- bins[unique(na_overlaps@from)]@ranges@width

# Calculate the percent overlap, adding up the na regions within a bin
pct_overlap <- excluded_length_per_bin / bin_length

# Store bins as names for sanity checking
# Note that unique and tapply put the bins in the same order
names(pct_overlap) <- unique(na_overlaps@from)

# Get the bin indices that correspond to less than the cutoff
bin_indices <- names(pct_overlap)[which(pct_overlap > perc_uncallable)]
```

Replace these bins with NAs.

```{r}
bin_calls_df[, bin_indices] <- NA
```

Format a matrix with rownames for `ComplexHeatmap` to use. 

```{r}
bin_calls_mat <- bin_calls_df %>%
  tibble::column_to_rownames("biospecimen_id") %>%
  as.matrix()
```

## Set up heatmap annotation objects

Make color key. 

```{r}
color_key <- structure(divergent_col_palette,
  names = c("loss", "neutral", "gain")
)
```

### Make column annotation object

Extract chromosome labels and make an alternating color key for them. 
This annotation object strategy was originally from [chromosomal-instability](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/b5a33838d1e9bd7e7913a89201ec26125c16c94c/analyses/chromosomal-instability/02a-plot-chr-instability-heatmaps.Rmd#L73).

```{r}
# Set up chromosome labels from bins as a factor vector
chrs <- paste0("chr", S4Vectors::decode(bins@seqnames))
chrs <- as.factor(chrs)

# Make a key for assigning alternating colors to the chromosomes
chr_colors <- rep(binary_col_palette$hex_codes, length.out = length(unique(chrs)))
names(chr_colors) <- unique(chrs)
```

Make chromosomal labeling `HeatmapAnnotation` object.

```{r}
# Create the Heatmap annotation object
chr_annot <- ComplexHeatmap::HeatmapAnnotation(
  df = data.frame(chrs),
  col = list(chrs = chr_colors),
  name = "",
  show_legend = FALSE,
  show_annotation_name = FALSE
)
```

### Make row annotation object

Make histology labeling `HeatmapAnnotation` object.
This annotation object strategy was originally from [chromosomal-nstability](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/b5a33838d1e9bd7e7913a89201ec26125c16c94c/analyses/chromosomal-instability/02a-plot-chr-instability-heatmaps.Rmd#L73).

```{r}
# Get the histologies for the samples in this set and order them by histology
histologies <-
  data.frame(Kids_First_Biospecimen_ID = rownames(bin_calls_mat)) %>% 
  dplyr::inner_join(metadata %>% 
                      dplyr::select(Kids_First_Biospecimen_ID, short_histology, sample_color)) %>%
  dplyr::arrange(short_histology) %>% 
  tibble::column_to_rownames("Kids_First_Biospecimen_ID")

# Make color key specific to these samples
histologies_color_key_filtered <- unique(histologies$sample_color)
names(histologies_color_key_filtered) <- unique(histologies$short_histology)

# Drop this column so ComplexHeatmap isn't tempted to plot it
histologies <- dplyr::select(histologies, -sample_color)
```

```{r}
# Create the Heatmap annotation object
hist_annot <- ComplexHeatmap::HeatmapAnnotation(
  df = data.frame(histologies),
  col = list(short_histology = histologies_color_key_filtered),
  which = "row",
  show_annotation_name = FALSE
)
```

## Assemble CN status heatmap

```{r}
# Plot on a heatmap
heatmap <- ComplexHeatmap::Heatmap(bin_calls_mat,
  heatmap_legend_param = list(title = "CN status"),
  col = color_key,
  cluster_columns = FALSE,
  cluster_rows = FALSE,
  show_column_names = FALSE,
  show_row_names = FALSE,
  bottom_annotation = chr_annot,
  left_annotation = hist_annot,
  na_col = "#cccccc"
)
```

Print out heatmap. 

```{r}
heatmap
```

Save to PDF. 

```{r}
# Save plot as PDF
pdf(file.path(plots_dir, "cn_status_heatmap.pdf"))
heatmap
dev.off()
```

# Session Info

```{r}
sessionInfo()
```


