---
title: "CNV GISTIC Plots"
output:   
  html_notebook: 
    toc: true
    toc_float: true
author: Candace Savonen for ALSF - CCDL
date: 2020
---

### Usage

This notebook can be run via the command line from the top directory of the 
repository as follows:

```
Rscript -e "rmarkdown::render('analyses/cnv-chrom-plot/gistic_plot.Rmd', 
                              clean = TRUE)"
```

### Set Up

```{r}
if (!("ggbio" %in% installed.packages())) {
  install.packages("BiocManager")
  BiocManager::install("ggbio")
}

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

### Directories and Files

```{r}
# Path to input directory
input_dir <- file.path("..", "..", "data")

# Path to gistic zip file results
gistic_zip <- file.path(input_dir, "pbta-cnv-cnvkit-gistic.zip")

# Zip to here
gistic_results <- file.path(input_dir, "gistic")

# Path to output directory
plots_dir <- "plots"

# Create the plots_dir if it does not exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}
```

Unzip and set up GISTIC results folder. 

```{r}
# Get the name of the folder first since this changes with updates to the date
folder_name <- unzip(zipfile = gistic_zip, list = TRUE) %>%
  dplyr::filter(Length == 0) %>%
  dplyr::pull("Name")

# Extract files if it hasn't been done yet
if (!dir.exists(gistic_results)) {
  # Unzip but only extract the files not the folder
  unzip(
    zipfile = gistic_zip,
    exdir = gistic_results
  )
}
# Designate GISTIC results folder to extract to
gistic_results <- file.path(gistic_results, folder_name)
```

### Read in data 

```{r}
# Read in metadata
metadata <-
  readr::read_tsv(file.path(input_dir, "pbta-histologies.tsv"))
```

Read in the GISTIC scores file and format the data.

```{r}
gistic_scores <- readr::read_tsv(file.path(gistic_results, "scores.gistic")) %>%
  dplyr::rename("gscore" = "G-score") %>%
  # Recode 23 and 24 as X and Y.
  dplyr::mutate(
    Chromosome = as.character(Chromosome),
    Chromosome = dplyr::recode(Chromosome,
      "23" = "X",
      "24" = "Y"
    ),
    # Turn `Del` scores into negative `G-scores`
    # This is how GISTIC shows the scores.
    gscore = dplyr::case_when(
      Type == "Del" ~ -gscore,
      TRUE ~ gscore
    )
  )
```
Read in the `seg` file so we can plot data by histology. 

```{r}
# Read in the segment copy number data
seg_data <- data.table::fread(file.path(input_dir, "pbta-cnv-cnvkit.seg.gz"),
  data.table = FALSE
) %>%
  # Join the histology column to this data
  dplyr::inner_join(dplyr::select(
    metadata,
    "Kids_First_Biospecimen_ID",
    "short_histology"
  ),
  by = c("ID" = "Kids_First_Biospecimen_ID")
  ) %>% 
  # Reformat the chromosome variable to drop the "chr"
  dplyr::mutate(chrom = factor(gsub("chr", "", chrom), 
                               levels = c(1:22, "X", "Y")))
```

### Make GISTIC data into GRanges object

```{r}
gistic_ranges <- GenomicRanges::GRanges(
  seqnames = gistic_scores$Chromosome,
  ranges = IRanges::IRanges(
    start = gistic_scores$Start,
    end = gistic_scores$End
  ),
  score = gistic_scores$gscore,
  mcols = gistic_scores
)
```

### Plot the GISTIC scores

```{r}
gistic_plot <- ggbio::autoplot(gistic_ranges, ggplot2::aes(y = score, fill = mcols.Type),
  geom = "bar", scales = "free_x", space = "free_x"
) +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(size = 3, angle = 45, hjust = 1)) +
  colorblindr::scale_fill_OkabeIto(name = "Type") +
  ggplot2::ylab("G-scores")

# Print out plot here
gistic_plot@ggplot
```

### Save the plot as PNG. 

```{r}
ggplot2::ggsave(
  plot = gistic_plot@ggplot,
  filename = file.path(plots_dir, "gistic_plot.png"),
  width = 10,
  height = 3,
  units = "in"
)
```

### Segment mean plots by histology 

```{r, results='hide', message=FALSE}
# Make a plot for each histology group
lapply(unique(seg_data$short_histology), function(histology_group) {

  # How many samples are of this histology label?
  sample_num <- seg_data %>%
    dplyr::filter(short_histology == histology_group) %>%
    dplyr::pull(ID) %>%
    unique() %>%
    length()

  # Isolate the data to the histology group
  hist_df <- seg_data %>%
    dplyr::filter(short_histology == histology_group) %>%
    dplyr::group_by(short_histology, chrom, loc.start, loc.end) %>%
    dplyr::summarize(hist.mean = mean(seg.mean)) %>%
    # Make Del/Amp variable
    dplyr::mutate(Type = dplyr::case_when(
      hist.mean < 0 ~ "Del",
      hist.mean > 0 ~ "Amp",
      hist.mean == 0 ~ "Neutral"
    ))

  # Turn into a GRanges for easier mapping
  hist_ranges <- GenomicRanges::GRanges(
    seqnames = hist_df$chrom,
    ranges = IRanges::IRanges(
      start = hist_df$loc.start,
      end = hist_df$loc.end
    ),
    score = hist_df$hist.mean,
    mcols = hist_df$Type
  )
  # Map this on a plot
  hist_plot <- ggbio::autoplot(hist_ranges,
    ggplot2::aes(y = score, fill = mcols),
    geom = "bar", scales = "free_x", space = "free_x"
  ) +
    ggplot2::theme_classic() +
    ggplot2::theme(axis.text.x = ggplot2::element_text(size = 3, angle = 45, hjust = 1)) +
    colorblindr::scale_fill_OkabeIto(name = "Type") +
    ggplot2::labs(
      title = histology_group,
      y = "Average segment mean",
      subtitle = paste("n =", sample_num)
    )

  # Save the plot to a png
  # Have to ouse @ggplot to extract the portion of the ggbio object with the plot
  ggplot2::ggsave(
    plot = hist_plot@ggplot,
    filename = file.path(plots_dir, paste0(histology_group, "_plot.png")),
    width = 10,
    height = 3,
    units = "in"
  )
  
  # Print the plot here 
  hist_plot
})
```

# Session Info

```{r}
sessionInfo()
```

