---
title: "Summary of Ensembl identifiers for RNA-seq matrices"
output: html_notebook
params: 
  polya.annot.table: 
    value: 'results/pbta-gene-expression-rsem-fpkm-collapsed_table.polya.rds'
  stranded.annot.table:
    value: 'results/pbta-gene-expression-rsem-fpkm-collapsed_table.stranded.rds'
---

```{r include = FALSE}
knitr::opts_chunk$set(comment = NA)
getOption(x = 'DT.warn.size', default = FALSE)
```

```{r load_packages, echo = FALSE}
# load libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(DT))
```

```{r dt_function, echo = FALSE}
# custom datatable function
viewDataTable <- function(dat){
  DT::datatable(dat,
                rownames = FALSE,
                filter = "bottom",
                class = 'cell-border stripe',
                options = list(pageLength = 5,
                               searchHighlight = TRUE,
                               scrollX = TRUE,
                               dom = 'tpi',
                               initComplete = JS("function(settings, json) {",
                                            "$(this.api().table().header()).css({'background-color':
                                            '#004467', 'color': '#fff'});","}"))
                )
}
```

```{r example_run, echo = FALSE}
# Example Run:
# Rscript -e "rmarkdown::render(input = 'analyses/collapse-rnaseq/02-analyze-drops.Rmd',
#                   params = list(polya.annot.table = 'pbta-gene-expression-rsem-fpkm-collapsed_table.polya.rds',
#                                 stranded.annot.table = 'pbta-gene-expression-rsem-fpkm-collapsed_table.stranded.rds'), clean = TRUE)"
```

## {.tabset .tabset-fade}

### PolyA RNA-seq
#### Annotation Table

```{r, echo = FALSE, warning = FALSE}
# view full gene annotation for polya data for multi-mapped gene symbols where
# the correlation is not NA
polya.annot.table <- readRDS(params$polya.annot.table)
viewDataTable(polya.annot.table %>% 
                dplyr::filter(ensembl_id == "Multiple",
                              !is.na(avg.cor)))
```

#### Summary of biotypes that are dropped due to zero expression:

```{r, echo = FALSE, warning = FALSE}
# view table of dropped identifiers because of zero expression
dat <- polya.annot.table %>% 
  filter(expressed == "No") %>%
  group_by(gene_type)  %>%
  summarise(count = n()) 
viewDataTable(dat)
```

#### Summary of biotypes that are dropped due to merging multiple Ensembl identifiers:

```{r, echo = FALSE, warning = FALSE}
# view table of dropped identifiers by removing duplicates
dat <- polya.annot.table %>% 
  filter(ensembl_id == "Multiple" & keep == "No") %>%
  group_by(gene_type)  %>%
  summarise(count = n()) 
viewDataTable(dat)
```

### Stranded RNA-seq 
#### Annotation Table

```{r, echo = FALSE, warning = FALSE}
# view full gene annotation for stranded data for multi-mapped gene symbols
# where the correlation is not NA
stranded.annot.table <- readRDS(params$stranded.annot.table)
viewDataTable(stranded.annot.table %>% 
                dplyr::filter(ensembl_id == "Multiple",
                              !is.na(avg.cor)))
```

#### Summary of biotypes that are dropped due to low expression:

```{r, echo = FALSE, warning = FALSE}
# view table of dropped identifiers because of zero expression
dat <- stranded.annot.table %>% 
  filter(expressed == "No") %>%
  group_by(gene_type)  %>%
  summarise(count = n()) 
viewDataTable(dat)
```

#### Summary of biotypes that are dropped due to merging multiple Ensembl identifiers:

```{r, echo = FALSE, warning = FALSE}
# view table of dropped identifiers by removing duplicates
dat <- stranded.annot.table %>% 
  filter(ensembl_id == "Multiple" & keep == "No") %>%
  group_by(gene_type)  %>%
  summarise(count = n()) 
viewDataTable(dat)
```

##

### Number of multi-mapped genes with `NA` correlation values

Many Ensembl identifiers will have total FPKM > 0, but still have many zeroes.
This results in `NA` values in the `avg.cor` column for many genes where 
multiple Ensembl IDs map to the same gene symbol.

_Note: Where `ensembl_id == "Unique"`, correlation values are not calculated._

```{r}
data.frame(
  strategy = c("poly-A", "stranded"),
  number_of_NA = c(nrow(polya.annot.table %>% 
                          filter(ensembl_id == "Multiple",
                                 is.na(avg.cor))),
                   nrow(stranded.annot.table %>% 
                          filter(ensembl_id == "Multiple",
                                 is.na(avg.cor))))
)
```

