---
title: "GISTIC comparison of gene level status"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2020
---

This notebook will compare the GISTIC gene level calls to our calls using the overlaps with exons.

The purpose is to identify, if any, disagreement between GISTIC's gene level calls and the calls in the annotated CN files we prepared in `analyses/focal-cn-file-preparation/results`. 

## Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

```
Rscript -e "rmarkdown::render('analyses/gistic-cohort-vs-histology-comparison/gistic-gene-level-call-comparison.Rmd', clean = TRUE)"
```

## Set up

```{r}
library(tidyverse)
```

### Files and Directories

```{r}
# Path to input directory
focal_cn_data_dir <- file.path("..", "focal-cn-file-preparation", "results")
gistic_data_dir <- file.path("..", "run-gistic", "results")
```

### Unzip GISTIC results and read in `all_lesions.conf_90.txt` files

```{r}
# Unzip and set up GISTIC results folders and files.
cohort_gistic_dir <- file.path(gistic_data_dir, "pbta-cnv-consensus-gistic")
cohort_gistic_zip <- file.path(gistic_data_dir, "pbta-cnv-consensus-gistic.zip")

if (!dir.exists(cohort_gistic_dir)) {
  unzip(cohort_gistic_zip, 
        exdir = data_dir)
}

lgat_gistic_dir <- file.path(gistic_data_dir, "pbta-cnv-consensus-lgat-gistic")
lgat_gistic_zip <- file.path(gistic_data_dir, "pbta-cnv-consensus-lgat-gistic.zip")

if (!dir.exists(lgat_gistic_dir)) {
  unzip(lgat_gistic_zip, 
        exdir = data_dir)
}
```

### Define file paths for GISTIC `all_lesions.conf_90.txt` files

```{r}
cohort_all_lesions_file <- file.path(cohort_gistic_dir, "all_lesions.conf_90.txt")

lgat_all_lesions_file <- file.path(lgat_gistic_dir, "all_lesions.conf_90.txt")
```

### Define file paths for `focal-cn-file-preparation` result files

```{r}
consensus_autosomes_file <- file.path(focal_cn_data_dir, "consensus_seg_annotated_cn_autosomes.tsv.gz")

cnvkit_autosomes_file <- file.path(focal_cn_data_dir, "cnvkit_annotated_cn_autosomes.tsv.gz")
```

## Function to Wrangle Data

```{r}
merge_calls <- function (annotated_cn_file,
                         gistic_file) {
  # Given the file path to one of our annotated cn files from 
  # `analyses/focal-cn-file-prepration/results` and one of the 
  # GISTIC `all_lesions.conf_90.txt` files from `analyses/run-gistic/results`,
  # produce a table that contains the non-neutral GISTIC gene level calls
  # that disagree with the calls we have.
  #
  # Args:
  #   annotated_cn_file: file path to our annotated CN file
  #   gistic_file: file path to the GISTIC all lesions file
  #
  # Return:
  #   final_table: final table with the GISTIC calls and our calls
  #                that disagree
  
  # Read in the annotated CN file
  annotated_cn_df <- readr::read_tsv(annotated_cn_file)
  
  # Read in the GISTIC calls file
  gistic_calls_df <- data.table::fread(gistic_file,
                                       data.table = FALSE)
  
  # Wrangle the GISTIC calls data to be in a comparable format with our CN calls
  gistic_calls_df <- gistic_calls_df %>%
    dplyr::select(
      -c(
        "Unique Name",
        "Wide Peak Limits",
        "Peak Limits",
        "Region Limits",
        "q values",
        "Residual q values after removing segments shared with higher peaks",
        "Broad or Focal",
        "Amplitude Threshold"
      )
    ) %>%
    as.data.frame() %>%
    tidyr::gather("biospecimen_id", "gistic_call", -Descriptor) %>%
    arrange(desc(Descriptor)) %>%
    dplyr::mutate(
      gistic_call = dplyr::case_when(
        gistic_call %in% c("-1", "-1.2929") ~ "loss",
        gistic_call %in% c("1", " 1", " 2", "2", " 1.0000") ~ "gain",
        TRUE ~ "neutral"
      )
    ) %>%
    dplyr::distinct()
  
  # Merge our CN calls with the GISTIC calls
  all_calls_df <- gistic_calls_df %>%
    dplyr::inner_join(annotated_cn_df,
                      by = c("biospecimen_id", "Descriptor" = "cytoband")) %>%
    dplyr::select(-c("ensembl", "copy_number", "ploidy", "gene_symbol")) %>%
    dplyr::distinct() %>%
    # Filter out the instances where the calls disagree
    dplyr::filter(gistic_call != status)
  
  # Keep just the chromosome arm info
  all_calls_df$Descriptor <-
    gsub("(p|q).*", "\\1", all_calls_df$Descriptor)
  
  # Tally how many times the calls disagree for each chromsome arm
  final_table <- all_calls_df %>%
    dplyr::group_by(Descriptor, gistic_call, status) %>%
    tally()
}
```

## Generate final tables

### Use GISTIC results for LGAT histology

```{r warning = FALSE, message = FALSE}
lgat_merged_calls_table <- merge_calls(cnvkit_autosomes_file,
                                       lgat_all_lesions_file)

lgat_merged_calls_table
```

### Use GISTIC results for entire cohort

```{r warning = FALSE, message = FALSE}
consensus_merged_calls_table <- merge_calls(cnvkit_autosomes_file,
                                            cohort_all_lesions_file)

consensus_merged_calls_table
```

