---
title: "UMAP at tumor purity threshold"
author: "S. Spielman for CCDL"
date: "2023"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
library(magrittr)
library(ggplot2)

# set overall theme
theme_set(
  theme_bw() + 
    theme(
      legend.position = "bottom", 
      legend.text = element_text(size = rel(0.6)),
      legend.title = element_text(size = rel(0.75)),
      legend.key.size = unit(0.35, "cm")
    )
)

# seed
set.seed(2023)
```
This notebook explores how the UMAP made from stranded RNA-Seq data might change if only samples that pass a tumor purity threshold are considered.

## Setup

First, define directories and file names: 
```{r}
# Directories
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
palette_dir <- file.path(root_dir, "figures", "palettes")
dim_red_dir <- file.path(
  root_dir,
  "analyses",
  "transcriptomic-dimension-reduction"
)

# metadata file
meta_file <- file.path(data_dir, "pbta-histologies.tsv")

# palette mapping file
pal_file <- file.path(palette_dir,  "broad_histology_cancer_group_palette.tsv")

# UMAP embedding files
umap_all_file <- file.path(
  "results",
  "rsem_stranded_log_umap_scores_aligned.tsv"
)
umap_threshold_file <- file.path(
  "results",
  "rsem_stranded_log_tumor-purity-threshold_umap_scores_aligned.tsv"
)
```

And read in and prepare the files:
```{r}
metadata_df <- readr::read_tsv(meta_file, guess_max = 10000)

palette_mapping_df <- readr::read_tsv(pal_file)

# Names to retain from UMAP files
retain_cols <- c("X1", "X2", "Kids_First_Biospecimen_ID", "dataset")

# Read in and combine UMAP and palette files into a single df
all_umaps <- readr::read_tsv(umap_all_file) %>% 
  dplyr::mutate(dataset = "full data") %>%
  dplyr::select(retain_cols)

umap_df <- readr::read_tsv(umap_threshold_file) %>% 
  dplyr::mutate(dataset = "filtered to tumor purity threshold") %>%
  dplyr::select(retain_cols) %>%
  # bind rows with full results
  dplyr::bind_rows(all_umaps) %>%
  # rename umaps
  dplyr::rename(UMAP1 = X1, UMAP2 = X2) %>%
  # combine with metadata
  dplyr::inner_join(
    dplyr::select(metadata_df, 
                  Kids_First_Biospecimen_ID, 
                  cancer_group, 
                  broad_histology), 
    by = "Kids_First_Biospecimen_ID"
  ) %>%
  # combine with display names
  dplyr::inner_join(
    dplyr::select(palette_mapping_df,
                  broad_histology, broad_histology_display, 
                  cancer_group, cancer_group_display),
    by = c("cancer_group", "broad_histology")
)


# Data frame with all UMAP results
umap_df
```


Now we will prepare palettes:

```{r}
# set up broad histology display palette
bh_df <- palette_mapping_df %>%
  dplyr::select(broad_histology_display, broad_histology_hex) %>%
  dplyr::distinct() 
pal_bh <- bh_df$broad_histology_hex
names(pal_bh) <- bh_df$broad_histology_display

# set up cancer group display palette
cg_df <- palette_mapping_df %>%
  dplyr::select(cancer_group_display, cancer_group_hex) %>%
  dplyr::distinct() %>%
  tidyr::drop_na()
pal_cg <- cg_df$cancer_group_hex
names(pal_cg) <- cg_df$cancer_group_display
```


## UMAPs

First, we'll display UMAPs colored by `broad_histology_display`. 
The panel labeled "full data" represents the UMAP presented in the manuscript.


```{r}
ggplot(umap_df) + 
  aes(x = UMAP1, y = UMAP2, 
      color = broad_histology_display) + 
  geom_point(size = 1, alpha = 0.7) + 
  # facet by full data vs. tumor purity threshold data
  facet_grid(~dataset) +
  scale_color_manual(values = pal_bh)
```


And now by cancer group display:

```{r}
ggplot(umap_df) + 
  aes(x = UMAP1, y = UMAP2, 
      color = cancer_group_display) + 
  geom_point(size = 1, alpha = 0.7) + 
  # facet by full data vs. tumor purity threshold data
  facet_grid(~dataset) +
  scale_color_manual(values = pal_cg) + 
  # tweak legend so it fit in rendered HTML
  guides(color = guide_legend(title.position = "top", ncol=4))
```

## Conclusions

UMAPs are broadly consistent before and after filtering samples on tumor purity.
It is worth noting that the filtered data may have fewer histology groups compared to the full data set because of how filtering was applied. 


## sessionInfo

```{r print session info}
sessionInfo()
```
