---
title: "CNV Comparison Plots"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

This notebook plots and compares detected CNV aberrations given CNVkit and 
Control-FREEC output.

## Output Files

- `analyses/cnv-comparison/plots/compare_cnv_output_proportion.pdf`
- `analyses/cnv-comparison/plots/compare_cnv_output_frequency.pdf`
- `analyses/cnv-comparison/plots/compare_cnv_output_violin_plot.pdf`
- `analyses/cnv-comparison/plots/compare_cnv_output_barplot_histology.pdf`
- `analyses/cnv-comparison/plots/compare_cnv_output_barplot_aberration.pdf`

## Usage

This script is intended to be run via the command line from the top directory
of the repository as follows:

```
Rscript -e "rmarkdown::render('analyses/cnv-comparison/01-cnv-comparison-plotting.Rmd', 
                              clean = TRUE)"
```

# Set Up

```{r}
# This will be needed to create the frequency and proportion aberration plots 
if (!("GenVisR" %in% installed.packages())) {
  install.packages("BiocManager")
  BiocManager::install("GenVisR")
}

# This will be need to combine plots 
if (!("cowplot" %in% installed.packages())) {
  install.packages("cowplot")
}

# Magrittr pipe
`%>%` <- dplyr::`%>%`

# Source custom functions script
source(file.path("util", "cnv-comparison-functions.R"))
```

# Directories and Files

```{r}
# Path to input directory
input_directory <- file.path("..", "..", "data")

# Path to output directory
output_directory <- "plots"

# Create the output directory if it does not exist
if (!dir.exists(output_directory)) {
  dir.create(output_directory, recursive = TRUE)
}

# List of file paths to the CNV data 
cnv_list <-
  list(
    cnvkit = file.path(input_directory, "pbta-cnv-cnvkit.seg.gz"),
    controlfreec = file.path(input_directory, "pbta-cnv-controlfreec.seg.gz")
  )
```

# Read in data 

```{r}
# Read in list of CNV data using custom `read_in_cnv` function
cnv_data <- lapply(cnv_list, read_in_cnv)

# Read in metadata
metadata <-
  readr::read_tsv(file.path(input_directory, "pbta-histologies.tsv"), guess_max = 10000)
```

# Filter data

```{r}
# Filter CNV data by cutoff segmean using custom `filter_segmean` function
cnv_filtered <-
  lapply(cnv_data, filter_segmean, segmean_cutoff = 0.5)

# Bind rows of dataframes in cnv_filtered for use with ggplots
combined_cnv_filtered <-
  dplyr::bind_rows(cnv_filtered, .id = "cnv_caller")
```

# GenVisR plots

```{r, fig.height = 25, fig.width = 40}
# Run `GenVisR::cnFreq` 
cnv_proportion_plot <-
  lapply(
    cnv_filtered,
    GenVisR::cnFreq,
    genome = "hg38",
    CN_low_cutoff = 0,
    CN_high_cutoff = .2,
    plotType = "proportion"
  )
cnv_frequency_plot <- lapply(
  cnv_filtered,
  GenVisR::cnFreq,
  genome = "hg38",
  CN_low_cutoff = 0,
  CN_high_cutoff = .2,
  plotType = "frequency"
)

# Plot cowplot of frequency plots and save
plot_cowplot(
  cnv_proportion_plot,
  output_directory,
  "compare_cnv_output_proportion.pdf"
)

# Plot cowplot of proportion plots and save
plot_cowplot(
  cnv_frequency_plot,
  output_directory,
  "compare_cnv_output_frequency.pdf"
)
```

# Violin plots

These plots represent the size of aberrations. In other words, there is no 
differention between gain and loss. 
```{r, fig.height = 25, fig.width = 40}
# Run `plot_violin` on CNV data
cnv_violin_plots <- plot_violin(combined_cnv_filtered)

# Save plot
pdf(
  file.path("plots", "compare_cnv_output_violin_plot.pdf"),
  height = 12,
  width = 30
)
cnv_violin_plots
dev.off()

cnv_violin_plots
```

# Barplots

```{r, fig.height = 25, fig.width = 40}
# Run `plot_histology_barplot`
cnv_histology_barplots <-
  plot_histology_barplot(combined_cnv_filtered, metadata)

# Save plot
pdf(
  file.path("plots", "compare_cnv_output_barplot_histology.pdf"),
  height = 12,
  width = 30
)
cnv_histology_barplots
dev.off()

# Run `plot_aberration_barplot`
cnv_aberration_barplots <- plot_aberration_barplot(combined_cnv_filtered)

# Save plot
pdf(
  file.path("plots", "compare_cnv_output_barplot_aberration.pdf"),
  height = 12,
  width = 30
)
cnv_aberration_barplots
dev.off()

cnv_histology_barplots
cnv_aberration_barplots
```

# Session Info

```{r}
sessionInfo()
```

